# Individual Covariates

If there is an individual covariate you wish to solve for you may specify it by the `iCov` dataset:

```{r individual-covariates}
library(rxode2)

library(units)
library(xgxr)

mod3 <- function() {
  ini({
    TKA   <- 2.94E-01
    ## Clearance with individuals
    TCL   <- 1.86E+01
    TV2   <-4.02E+01
    TQ    <-1.05E+01
    TV3   <-2.97E+02
    TKin  <- 1
    TKout <- 1
    TEC50 <-200
  })
  model({
    KA            <- TKA
    CL            <- TCL * (WT / 70) ^ 0.75
    V2            <- TV2
    Q             <- TQ
    V3            <- TV3
    Kin           <- TKin
    Kout          <- TKout
    EC50          <- TEC50
    Tz            <- 8
    amp           <- 0.1
    C2            <- central/V2
    C3            <- peri/V3
    d/dt(depot)   <- -KA*depot
    d/dt(central) <- KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)    <-                    Q*C2 - Q*C3
    d/dt(eff)     <-  Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0)        <- 1  ## This specifies that the effect compartment starts at 1.
  })
}

ev <- et(amount.units="mg", time.units="hours") |>
  et(amt=10000, cmt=1) |>
  et(0,48,length.out=100) |>
  et(id=1:4)

set.seed(10)
rxSetSeed(10)
## Now use iCov to simulate a 4-id sample
r1 <- solve(mod3, ev,
            # Create individual covariate data-frame
            iCov=data.frame(id=1:4, WT=rnorm(4, 70, 10)))
print(r1)

plot(r1, C2, log="y")
```

# Time Varying Covariates
```{r, echo=FALSE}
## options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width=80, cli.width=80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH=80)
```

Covariates are easy to specify in rxode2, you can specify them as a
variable. Time-varying covariates, like clock time in a circadian
rhythm model, can also be used.  Extending the indirect response model
already discussed, we have:


```{r}
library(rxode2)
library(units)

mod4 <- mod3 |>
  model(d/dt(eff) <-  Kin - Kout*(1-C2/(EC50+C2))*eff) |>
  model(-Kin) |>
  model(Kin <- TKin + amp *cos(2*pi*(ctime-Tz)/24), append=C2, cov="ctime")

ev <- et(amountUnits="mg", timeUnits="hours") |>
    et(amt=10000, cmt=1) |>
    et(0,48,length.out=100)


## Create data frame of 8 am dosing for the first dose This is done
## with base R but it can be done with dplyr or data.table
ev$ctime <- (ev$time+set_units(8,hr)) %% 24
ev$WT <- 70

```

Now there is a covariate present in the event dataset, the system can
be solved by combining the dataset and the model:

```{r}
r1 <- solve(mod4, ev, covsInterpolation="linear")
print(r1)
```

When solving ODE equations, the solver may sample times outside of the
data.  When this happens, this ODE solver can use linear interpolation
between the covariate values. It is equivalent to R's `approxfun` with
`method="linear"`.

```{r time-varying-linear-central}
plot(r1,C2, ylab="Central Concentration")
```

```{r time-varying-linear-effect}
plot(r1,eff) + ylab("Effect") + xlab("Time")
```

Note that the linear approximation in this case leads to some kinks in
the solved system at 24-hours where the covariate has a linear
interpolation between near 24 and near 0. While linear seems
reasonable, cases like clock time make other interpolation methods
more attractive.

In rxode2 the default covariate interpolation is be the last
observation carried forward (`locf`), or constant approximation.  This is
equivalent to R's `approxfun` with `method="constant"`.


```{r}
r1 <- solve(mod4, ev,covsInterpolation="locf")
print(r1)
```

which gives the following plots:

```{r time-varying-constant-central}
plot(r1,C2, ylab="Central Concentration", xlab="Time")
```

```{r time-varying-constant-effect}
plot(r1,eff, ylab="Effect", xlab="Time")
```

In this case, the plots seem to be smoother.

You can also use NONMEM's preferred interpolation style of next
observation carried backward (NOCB):

```{r}
r1 <- solve(mod4, ev,covsInterpolation="nocb")
print(r1)
```

which gives the following plots:

```{r time-varying-nocb-central}
plot(r1,C2, ylab="Central Concentration", xlab="Time")
```

```{r time-varying-nocb-effect}
plot(r1,eff, ylab="Effect", xlab="Time")
```

# Forced parameters

Sometimes a parameter's value is owned *outside* the usual `params`/data/`ini`
channel and must be applied on **every** solve of a model -- for example
weights fit by an external optimizer that should travel with the model so it
stays self-contained.  `rxForcedPars()` stores such values in a hidden slot on
the model (a named numeric vector).  On each solve they are injected into every
subject/simulation column at solve setup, overriding whatever `params` or the
data supplied:

```{r}
mod <- function() {
  ini({ tcl <- -2 })
  model({
    cl <- exp(tcl) * WT
    d/dt(depot) <- -cl * depot
    cp <- depot
  })
}
ui <- rxode2(mod)

ev <- et(amt = 100) |> et(seq(1, 8, by = 1))
ev$WT <- 1

## baseline: WT = 1 from the data
s0 <- rxSolve(ui, ev)

## force WT = 2 for every solve of this model
rxForcedPars(ui) <- c(WT = 2)
s1 <- rxSolve(ui, ev)

head(data.frame(time = s0$time, cp_WT1 = s0$cp, cp_forcedWT2 = s1$cp))
```

Because the value lives on the model, it is carried through model piping and
into any `nlmixr2` fit built from the model, so `predict()`/`simulate()` on that
fit reuse it with no external state.  The forced values are also reported by
`rxInjectedPars()` on the solved object:

```{r}
rxInjectedPars(s1)
```

Clear forcing by assigning `NULL`:

```{r}
rxForcedPars(ui) <- NULL
```

Names that are not model parameters are ignored, and forcing applies only to
population-constant parameters/covariates (a single value per parameter), not
to per-record time-varying covariates.
