# Individual Covariates

If there is an individual covariate you wish to solve for you may specify it by the `iCov` dataset:

```{r individual-covariates}
library(rxode2)

library(units)
library(xgxr)

mod3 <- function() {
  ini({
    TKA   <- 2.94E-01
    ## Clearance with individuals
    TCL   <- 1.86E+01 
    TV2   <-4.02E+01
    TQ    <-1.05E+01
    TV3   <-2.97E+02
    TKin  <- 1
    TKout <- 1
    TEC50 <-200
  })
  model({
    KA            <- TKA
    CL            <- TCL * (WT / 70) ^ 0.75
    V2            <- TV2
    Q             <- TQ
    V3            <- TV3
    Kin           <- TKin
    Kout          <- TKout
    EC50          <- TEC50
    Tz            <- 8
    amp           <- 0.1
    C2            <- central/V2
    C3            <- peri/V3
    d/dt(depot)   <- -KA*depot
    d/dt(central) <- KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)    <-                    Q*C2 - Q*C3
    d/dt(eff)     <-  Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0)        <- 1  ## This specifies that the effect compartment starts at 1.
  })
}

ev <- et(amount.units="mg", time.units="hours") %>%
  et(amt=10000, cmt=1) %>%
  et(0,48,length.out=100) %>%
  et(id=1:4)

set.seed(10)
rxSetSeed(10)
## Now use iCov to simulate a 4-id sample
r1 <- solve(mod3, ev,
            # Create individual covariate data-frame
            iCov=data.frame(id=1:4, WT=rnorm(4, 70, 10)))
print(r1)

plot(r1, C2, log="y")
```

# Time Varying Covariates
```{r, echo=FALSE}
## options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width=80, cli.width=80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH=80)
```

Covariates are easy to specify in rxode2, you can specify them as a
variable. Time-varying covariates, like clock time in a circadian
rhythm model, can also be used.  Extending the indirect response model
already discussed, we have:


```{r}
library(rxode2)
library(units)

mod4 <- mod3 %>%
  model(d/dt(eff) <-  Kin - Kout*(1-C2/(EC50+C2))*eff) %>%
  model(-Kin) %>%
  model(Kin <- TKin + amp *cos(2*pi*(ctime-Tz)/24), append=C2, cov="ctime")
 
ev <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=10000, cmt=1) %>%
    et(0,48,length.out=100)


## Create data frame of 8 am dosing for the first dose This is done
## with base R but it can be done with dplyr or data.table
ev$ctime <- (ev$time+set_units(8,hr)) %% 24
ev$WT <- 70

```

Now there is a covariate present in the event dataset, the system can
be solved by combining the dataset and the model:

```{r}
r1 <- solve(mod4, ev, covsInterpolation="linear")
print(r1)
```

When solving ODE equations, the solver may sample times outside of the
data.  When this happens, this ODE solver can use linear interpolation
between the covariate values. It is equivalent to R's `approxfun` with
`method="linear"`.

```{r time-varying-linear-central}
plot(r1,C2, ylab="Central Concentration")
```

```{r time-varying-linear-effect}
plot(r1,eff) + ylab("Effect") + xlab("Time")
```

Note that the linear approximation in this case leads to some kinks in
the solved system at 24-hours where the covariate has a linear
interpolation between near 24 and near 0. While linear seems
reasonable, cases like clock time make other interpolation methods
more attractive.

In rxode2 the default covariate interpolation is be the last
observation carried forward (`locf`), or constant approximation.  This is
equivalent to R's `approxfun` with `method="constant"`.


```{r}
r1 <- solve(mod4, ev,covsInterpolation="locf")
print(r1)
```

which gives the following plots:

```{r time-varying-constant-central}
plot(r1,C2, ylab="Central Concentration", xlab="Time")
```

```{r time-varying-constant-effect}
plot(r1,eff, ylab="Effect", xlab="Time")
```

In this case, the plots seem to be smoother.

You can also use NONMEM's preferred interpolation style of next
observation carried backward (NOCB):

```{r}
r1 <- solve(mod4, ev,covsInterpolation="nocb")
print(r1)
```

which gives the following plots:

```{r time-varying-nocb-central}
plot(r1,C2, ylab="Central Concentration", xlab="Time")
```

```{r time-varying-nocb-effect}
plot(r1,eff, ylab="Effect", xlab="Time")
```
