## Prediction only models

Prediction only models are simple to create.  You use the rxode2 syntax
without any ODE systems in them.  A very simple example is a one-compartment model.

```{r predOnlyModel}
library(rxode2)

mod <- function(){
  model({
    ipre <- 10 * exp(-ke * t)
  })
}
```

Solving the rxode2 models are the same as saving the simple ODE system, but faster of course.

```{r predOnlyModelSolve}
et  <- et(seq(0,24,length.out=50))
cmt1 <- rxSolve(mod,et,params=c(ke=0.5))
cmt1
```

## Solved compartment models

Solved models are also simple to create. You simply place the
`linCmt()` pseudo-function into your code.  The `linCmt()` function
figures out the type of model to use based on the parameter names
specified.

Most often, pharmacometric models are parameterized in terms of volume
and clearances. Clearances are specified by NONMEM-style names of
`CL`, `Q`, `Q1`, `Q2`, etc. or distributional clearances `CLD`,
`CLD2`. Volumes are specified by Central (`VC` or `V`),
Peripheral/Tissue (`VP`, `VT`).  While more translations are
available, some example translations are below:

```{r clLinCmt, echo=FALSE}
clLinCmt <- read.csv("../../vignettes/cl-lincmt.csv")
if (knitr::is_latex_output()) {
  library(kableExtra)
  kbl(clLinCmt, longtable=TRUE, booktabs=TRUE, caption="Clearance Based linCmt() parameterizations") %>%
    kable_styling(latex_options=c("repeat_header", "striped", "hold_position"))
} else {
  library(DT)
  datatable(clLinCmt, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
}
```

Another popular parameterization is in terms of micro-constants. rxode2
assumes compartment `1` is the central compartment.  The elimination
constant would be specified by `K`, `Ke` or `Kel`. Some example translations are below:

```{r kelLinCmt, echo=FALSE}
kelLinCmt <- read.csv("../../vignettes/kel-lincmt.csv");
if (knitr::is_latex_output()) {
  kbl(kelLinCmt, longtable=TRUE, booktabs=TRUE, caption="Kel Based linCmt() parameterizations") %>%
    kable_styling(latex_options=c("repeat_header", "striped", "hold_position"))
} else {
  datatable(kelLinCmt, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
}
```

The last parameterization possible is using `alpha` and `V` and/or
`A`/`B`/`C`. Some example translations are below:

```{r alphaLinCmt, echo=FALSE}
alphaLinCmt <- read.csv("../../vignettes/alpha-lincmt.csv");
if (knitr::is_latex_output()) {
  kbl(alphaLinCmt, longtable=TRUE, booktabs=TRUE, caption="alpha Based linCmt() parameterizations") %>%
    kable_styling(latex_options=c("repeat_header", "striped", "hold_position"))
} else {
  datatable(alphaLinCmt, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
}
```


Once the `linCmt()` sleuthing is complete, the `1`, `2` or `3`
compartment model solution is used as the value of `linCmt()`.  After the linear
`linCmt()` models the following variables/state values will become available
(depenting on the type of compartment model): `depot`, `central`, `peripheral1` and
`peripheral2`

The compartments where you can dose in a linear solved system are
`depot` and `central` when there is an linear absorption constant in
the model `ka`.  Without any additional ODEs, these compartments are
numbered `depot=1` and `central=2`.

When the absorption constant `ka` is missing, you may only dose to the
`central` compartment.  Without any additional ODEs the compartment
number is `central=1`.

The peripheral compartments may be dosed to as well, though their numbers depend 
on if they are mixed with the ODEs.  The `central` and `depot` compartment can
have infusions, and all of the departments can have non-infusion doses.

These compartments take the same sort of events that a ODE model can
take, and are discussed in the [rxode2 events
vignette](rxode2-events.html). 

```{r modLinCmtExample}
mod <- function() {
  ini({
    kel <- 0.5
    V <- 1
  })
  model({
    ipre <- linCmt(V, kel)
  })
}
```

This then acts as an ODE model; You specify a dose to the depot
compartment and then solve the system:

```{r modLinCmtEt}
et  <- et(amt=10,time=0,cmt=depot) %>%
  et(seq(0,24,length.out=50))
cmt1 <- rxSolve(mod,et)
cmt1
```



## Mixing Solved Systems and ODEs

In addition to pure ODEs, you may mix solved systems and ODEs.  The
prior 2-compartment indirect response model can be simplified with a
`linCmt()` function:

```{r modMixedLinOde}
library(rxode2)
## Setup example model
mod1 <-function() {
  model({
    C2 = centr/V2
    C3 = peri/V3
    d/dt(depot) =-KA*depot
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  =                    Q*C2 - Q*C3
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff
  })
}

## Seup parameters and initial conditions

theta <-
    c(KA=2.94E-01, CL=1.86E+01, V2=4.02E+01, # central
      Q=1.05E+01,  V3=2.97E+02,              # peripheral
      Kin=1, Kout=1, EC50=200)               # effects

inits <- c(eff=1)

## Setup dosing event information
ev <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=10000, addl=9, ii=12) %>%
    et(amt=20000, addl=4, time=120, ii=24) %>%
    add.sampling(0:240)

## Setup a mixed solved/ode system:
mod2 <- function() {
  model({
    ## the order of variables do not matter, the type of compartmental
    ## model is determined by the parameters specified.
    C2   = linCmt(KA, CL, V2, Q, V3);
    eff(0) = 1  ## This specifies that the effect compartment starts at 1.
    d/dt(eff) =  Kin - Kout*(1-C2/(EC50+C2))*eff;
  })
}
```


This allows the indirect response model above to assign the
2-compartment model to the `C2` variable and the used in the indirect
response model.

When mixing the solved systems and the ODEs, the solved system's
compartment is always the last compartment.  This is because the
solved system technically isn't a compartment to be solved.  Adding
the dosing compartment to the end will not interfere with the actual
ODE to be solved.

Therefore,in the two-compartment indirect response model, the effect
compartment is compartment #1 while the PK dosing compartment for the
depot is compartment #2.

This compartment model requires a new event table since the
compartment number changed:

```{r modMixedLinOdeET}
ev <- et(amountUnits='mg', timeUnits='hours') %>%
  et(amt=10000, addl=9, ii=12, cmt=2) %>%
  et(amt=20000, addl=4, time=120, ii=24, cmt=2) %>%
  et(0:240)
```

This can be solved with the following command:

```{r modMixedLinOdeSovle}
x <- mod2 %>%  solve(theta, ev)
print(x)
```

Note this solving did not require specifying the effect compartment
initial condition to be `1`.  Rather, this is already pre-specified by
`eff(0)=1`.

This can be solved for different initial conditions easily:

```{r modMixedLinOdeIni}
x <- mod2 %>%  solve(theta, ev,c(eff=2))
print(x)
```

The rxode2 detective also does not require you to specify the variables
in the `linCmt()` function if they are already defined in the block.
Therefore, the following function will also work to solve the same
system.


```{r}
mod3 <- function() {
  ini({
    KA <- 2.94E-01
    CL <- 1.86E+01
    V2 <- 4.02E+01
    Q <- 1.05E+01
    V3 <- 2.97E+02
    Kin <- 1
    Kout <- 1
    EC50 <- 200
  })
  model({
    # Since the parameters are in the ini block, put them in linCmt so
    # that the model is detected correctly
    C2   <- linCmt(KA, CL, V2, Q, V3)
    eff(0) <- 1  ## This specifies that the effect compartment starts at 1.
    d/dt(eff) <-  Kin - Kout*(1-C2/(EC50+C2))*eff;
  })
}

x <- mod3 %>%  solve(ev)

print(x)
```

Note that you do not specify the parameters when solving the system
since they are built into the model, but you can override the parameters:

```{r}
x <- mod3 %>%  solve(c(KA=10),ev)
print(x)
```
