# Population Simulations with rxode2

## Simulation of Variability with rxode2

In pharmacometrics the nonlinear-mixed effect modeling software (like
nlmixr) characterizes the between-subject variability.  With this
between subject variability you can simulate new subjects.

Assuming that you have a 2-compartment, indirect response model, you
can set create an rxode2 model describing this system below:

### Setting up the rxode2 model

```{r}
library(rxode2)

set.seed(32)
rxSetSeed(32)

mod <- function() {
  ini({
    KA <- 2.94E-01
    TCl <- 1.86E+01
    # between subject variability
    eta.Cl ~ 0.4^2
    V2 <- 4.02E+01
    Q <- 1.05E+01
    V3 <- 2.97E+02
    Kin <- 1
    Kout <- 1
    EC50 <- 200
  })
  model({
    C2 <- centr/V2
    C3 <- peri/V3
    CL <-  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) <- -KA*depot
    d/dt(centr) <- KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                    Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0) <- 1
  })
}
```


### Simulating 

The next step to simulate is to create the dosing regimen for overall simulation:

```{r}
ev <- et(amountUnits="mg", timeUnits="hours") %>%
  et(amt=10000, cmt="centr")
```
If you wish, you can also add sampling times (though rxode2 can fill these in for you):

```{r}
ev <- ev %>% et(0,48, length.out=100)
```

Note the `et` takes similar arguments as `seq` when adding sampling
times.  There are more methods to adding sampling times and events to
make complex dosing regimens (See [the event
vignette](rxode2-events.html)). This includes ways to add variability
to the [both the sampling and dosing
times](rxode2-events.html#add-doses-and-samples-within-a-sampling-window)).

Once this is complete you can simulate using the `rxSolve` routine:

```{r}
sim  <- rxSolve(mod, ev, nSub=100)
```

To quickly look and customize your simulation you use the default
`plot` routine.  Since this is an rxode2 object, it will create a
`ggplot2` object that you can modify as you wish.  The extra parameter
to the `plot` tells `rxode2`/`R` what piece of information you are
interested in plotting. In this case, we are interested in looking at
the derived parameter `C2`:

### Checking the simulation with `plot`

```{r}
library(ggplot2)
## The plots from rxode2 are ggplots so they can be modified with
## standard ggplot commands.
plot(sim, C2, log="y") +
    ylab("Central Compartment") 
```

Of course this additional parameter could also be a state value, like `eff`:

```{r}
## They also takes many of the standard plot arguments; See ?plot
plot(sim, eff, ylab="Effect")
```

Or you could even look at the two side-by-side:

```{r}
plot(sim, C2, eff)
```

Or stack them with `patchwork`
```{r}
library(patchwork)
plot(sim, C2, log="y") / plot(sim, eff)
```


### Processing the data to create summary plots

Usually in pharmacometric simulations it is not enough to simply
simulate the system.  We have to do something easier to digest, like
look at the central and extreme tendencies of the simulation.

Since the `rxode2` solve object is a type of [data
frame](rxode2-data-frame.html)

It is now straightforward to perform calculations and generate plots
with the simulated data. You can 

Below, the 5th, 50th, and 95th percentiles
of the simulated data are plotted. 

```{r}
confint(sim, "C2", level=0.95) %>%
    plot(ylab="Central Concentration", log="y")
```

```{r}
confint(sim, "eff", level=0.95) %>%
    plot(ylab="Effect")
```

Note that you can see the parameters that were simulated for the example

```{r}
head(sim$param)
```

### Simulation of unexplained variability (sigma)

In addition to conveniently simulating between subject variability,
you can also easily simulate unexplained variability.

One way to do that is to create a rxode2 model with the endpoints
defined. Model piping can do this while keeping the model intact:

```{r}
mod2 <- mod %>%
  model(eff ~ add(eff.sd), append=TRUE) %>%
  model(C2 ~ prop(prop.sd), append=TRUE) %>%
  ini(eff.sd=sqrt(0.1), prop.sd=sqrt(0.1))
```

You can see how the dataset should be defined with
`$multipleEndpoint`:

```{r}
mod2$multipleEndpoint
```

Here you see the endpoints should be defined for `eff` and `C2`:

```{r}
ev <- et(amountUnits="mg", timeUnits="hours") %>%
  et(amt=10000, cmt="centr") %>%
  et(seq(0,48, length.out=100), cmt="eff") %>%
  et(seq(0,48, length.out=100), cmt="C2")
```

Which allows you to solve the system:

```{r}
sim  <- rxSolve(mod2, ev, nSub=100)
```

Since this is simulated from a model with the residual specification
included and a multiple endpoint model, you can summarize for each
endpoint by simply taking a confidence interval of `"sim"`:

```{r}
s <- confint(sim, "sim")

plot(s)
```

If you want to subset to a specific endpoint, like `eff` you can
create the confidence interval for only that endpoint by using the
specification `sim.eff`, where the endpoint name is separated from
`sim` by a dot:


```{r}
seff <- confint(sim, "sim.eff")

plot(seff)
```

### Simulation of Individuals

Sometimes you may want to match the dosing and observations of
individuals in a clinical trial.  To do this you will have to create a
data.frame using the `rxode2` event specification as well as an `ID`
column to indicate an individual. The rxode2 event vignette talks more about
how these datasets should be created.

```{r}
library(dplyr)
ev1 <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=10000, cmt=2) %>%
    et(0,48,length.out=10)

ev2 <- et(amountUnits="mg", timeUnits="hours") %>%
    et(amt=5000, cmt=2) %>%
    et(0,48,length.out=8)

dat <- rbind(data.frame(ID=1, ev1$get.EventTable()),
             data.frame(ID=2, ev2$get.EventTable()))


## Note the number of subject is not needed since it is determined by the data
sim  <- rxSolve(mod, dat)

sim %>% select(id, time, eff, C2)
```

# Simulation of Clinical Trials

By either using a simple single event table, or data from a clinical
trial as described above, a complete clinical trial simulation can be
performed.

Typically in clinical trial simulations you want to account for the
uncertainty in the fixed parameter estimates, and even the uncertainty
in both your between subject variability as well as the unexplained
variability.

`rxode2` allows you to account for these uncertainties by simulating
multiple virtual "studies," specified by the parameter `nStud`.  Each
of these studies samples a realization of fixed effect parameters and
covariance matrices for the between subject variability(`omega`) and
unexplained variabilities (`sigma`). Depending on the information you
have from the models, there are a few strategies for simulating a
realization of the `omega` and `sigma` matrices. 

The first strategy occurs when either there is not any standard errors
for standard deviations (or related parameters), or there is a modeled
correlation in the model you are simulating from.  In that case the
suggested strategy is to use the inverse Wishart (parameterized to
scale to the conjugate prior)/[scaled inverse chi
distribution](https://en.wikipedia.org/wiki/Scaled_inverse_chi-squared_distribution).
this approach uses a single parameter to inform the variability of the
covariance matrix sampled (the degrees of freedom).

The second strategy occurs if you have standard errors on the
variance/standard deviation with no modeled correlations in the
covariance matrix.  In this approach you perform separate simulations
for the standard deviations and the correlation matrix.  First you
simulate the variance/standard deviation components in the `thetaMat`
multivariate normal simulation.  After simulation and transformation
to standard deviations, a correlation matrix is simulated using the
degrees of freedom of your covariance matrix.  Combining the simulated
standard deviation with the simulated correlation matrix will give a
simulated covariance matrix. For smaller dimension covariance matrices
(dimension < 10x10) it is recommended you use the `lkj` distribution
to simulate the correlation matrix.  For higher dimension covariance
matrices it is suggested you use the inverse wishart distribution
(transformed to a correlation matrix) for the simulations.

The covariance/variance prior is simulated from `rxode2`s `cvPost()`
function.

## Simulation from inverse Wishart correlations

An example of this simulation is below:
```{r}

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(NULL, names(mod2$theta)[1:8])

sim  <- rxSolve(mod2, ev, nSub=100, thetaMat=tMat, nStud=10,
                dfSub=10, dfObs=100)

s <-sim %>% confint("sim")

plot(s)
```

If you wish you can see what `omega` and `sigma` was used for each
virtual study by accessing them in the solved data object with
`$omega.list` and `$sigma.list`:

```{r}
head(sim$omegaList)
```

```{r}
head(sim$sigmaList)
```

You can also see the parameter realizations from the `$params` data frame.

## Simulate using variance/standard deviation standard errors

Lets assume we wish to simulate from [the nonmem run included in
xpose](https://github.com/UUPharmacometrics/xpose/blob/master/inst/extdata/run001.lst)

First we setup the model; Since we are taking this from nonmem and
would like to use the more free-form style from the classic `rxode2`
model we start from the classic model:

```{r}
rx1 <- rxode2({
  cl <- tcl*(1+crcl.cl*(CLCR-65)) * exp(eta.cl)
  v <- tv * WT * exp(eta.v)
  ka <- tka * exp(eta.ka)
  ipred <- linCmt()
  obs <- ipred * (1 + prop.sd) + add.sd 
})
```

Next we input the estimated parameters:

```{r}
theta <- c(tcl=2.63E+01, tv=1.35E+00, tka=4.20E+00, tlag=2.08E-01,
           prop.sd=2.05E-01, add.sd=1.06E-02, crcl.cl=7.17E-03,
           ## Note that since we are using the separation strategy the ETA variances are here too
           eta.cl=7.30E-02,  eta.v=3.80E-02, eta.ka=1.91E+00)
```

And also their covariances; To me, the easiest way to create a named
covariance matrix is to use `lotri()`:

```{r}
thetaMat <- lotri(
    tcl + tv + tka + tlag + prop.sd + add.sd + crcl.cl + eta.cl + eta.v + eta.ka ~
        c(7.95E-01,
          2.05E-02, 1.92E-03,
          7.22E-02, -8.30E-03, 6.55E-01,
          -3.45E-03, -6.42E-05, 3.22E-03, 2.47E-04,
          8.71E-04, 2.53E-04, -4.71E-03, -5.79E-05, 5.04E-04,
          6.30E-04, -3.17E-06, -6.52E-04, -1.53E-05, -3.14E-05, 1.34E-05,
          -3.30E-04, 5.46E-06, -3.15E-04, 2.46E-06, 3.15E-06, -1.58E-06, 2.88E-06,
          -1.29E-03, -7.97E-05, 1.68E-03, -2.75E-05, -8.26E-05, 1.13E-05, -1.66E-06, 1.58E-04,
          -1.23E-03, -1.27E-05, -1.33E-03, -1.47E-05, -1.03E-04, 1.02E-05, 1.67E-06, 6.68E-05, 1.56E-04,
          7.69E-02, -7.23E-03, 3.74E-01, 1.79E-03, -2.85E-03, 1.18E-05, -2.54E-04, 1.61E-03, -9.03E-04, 3.12E-01))

evw <- et(amount.units="mg", time.units="hours") %>%
    et(amt=100) %>%
    ## For this problem we will simulate with sampling windows
    et(list(c(0, 0.5),
       c(0.5, 1),
       c(1, 3),
       c(3, 6),
       c(6, 12))) %>%
    et(id=1:1000)

## From the run we know that:
##   total number of observations is: 476
##    Total number of individuals:     74
sim  <- rxSolve(rx1, theta, evw,  nSub=100, nStud=10,
                thetaMat=thetaMat,
                ## Match boundaries of problem
                thetaLower=0, 
                sigma=c("prop.sd", "add.sd"), ## Sigmas are standard deviations
                sigmaXform="identity", # default sigma xform="identity"
                omega=c("eta.cl", "eta.v", "eta.ka"), ## etas are variances
                omegaXform="variance", # default omega xform="variance"
                iCov=data.frame(WT=rnorm(1000, 70, 15), CLCR=rnorm(1000, 65, 25)),
                dfSub=74, dfObs=476);

print(sim)
## Notice that the simulation time-points change for the individual

## If you want the same sampling time-points you can do that as well:
evw <- et(amount.units="mg", time.units="hours") %>%
    et(amt=100) %>%
    et(0, 24, length.out=50) %>%
    et(id=1:100)

sim  <- rxSolve(rx1, theta, evw,  nSub=100, nStud=10,
                thetaMat=thetaMat,
                ## Match boundaries of problem
                thetaLower=0, 
                sigma=c("prop.sd", "add.sd"), ## Sigmas are standard deviations
                sigmaXform="identity", # default sigma xform="identity"
                omega=c("eta.cl", "eta.v", "eta.ka"), ## etas are variances
                omegaXform="variance", # default omega xform="variance"
                iCov=data.frame(WT=rnorm(100, 70, 15), CLCR=rnorm(100, 65, 25)),
                dfSub=74, dfObs=476,
                resample=TRUE)

s <-sim %>% confint(c("ipred"))
plot(s)
```

## Simulate without uncertainty in `omega` or `sigma` parameters

If you do not wish to sample from the prior distributions of either
the `omega` or `sigma` matrices, you can turn off this feature by
specifying the `simVariability = FALSE` option when solving:

```{r}

sim  <- rxSolve(mod2, ev, nSub=100, thetaMat=tMat, nStud=10,
                simVariability=FALSE)

s <-sim %>% confint(c("centr", "eff"))

plot(s)
```

Note since realizations of `omega` and `sigma` were not simulated,
`$omegaList` and `$sigmaList` both return `NULL`.

