---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rxode2

## CRAN updating

[![CRAN-status](https://img.shields.io/badge/CRAN-Updating-red)](https://github.com/nlmixr2/rxode2/actions/workflows/R-CMD-check.yaml)

Currently we are updating `PreciseSums` on CRAN; `rxode2` has a binary
linked to `PreciseSums`.  This means while both `PreciseSums` and
`rxode2` are being submitted the most stable version of `rxode2` comes
from the `r-universe`:

```r
install.packages(c("dparser", "nlmixr2data", "lotri", "rxode2ll",
                   "rxode2parse", "rxode2random", "rxode2et",
                   "rxode2"),
                 repos = c('https://nlmixr2.r-universe.dev',
                           'https://cloud.r-project.org'))
```

This is temporary and should resolve itself in a couple of weeks.

<!-- badges: start -->
[![R-CMD-check](https://github.com/nlmixr2/rxode2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nlmixr2/rxode2/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/nlmixr2/rxode2/branch/main/graph/badge.svg)](https://app.codecov.io/gh/nlmixr2/rxode2?branch=main)
[![CRAN version](http://www.r-pkg.org/badges/version/rxode2)](https://cran.r-project.org/package=rxode2)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/rxode2)](https://cran.r-project.org/package=rxode2)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/rxode2)](https://cran.r-project.org/package=rxode2)
[![CodeFactor](https://www.codefactor.io/repository/github/nlmixr2/rxode2/badge)](https://www.codefactor.io/repository/github/nlmixr2/rxode2)
![r-universe](https://nlmixr2.r-universe.dev/badges/rxode2)
<!-- badges: end -->

## Overview

**rxode2** is an R package for solving and simulating from ode-based
models. These models are convert the rxode2 mini-language to C and
create a compiled dll for fast solving. ODE solving using rxode2 has a
few key parts:

 - `rxode2()` which creates the C code for fast ODE solving based on a
   [simple syntax](https://nlmixr2.github.io/rxode2/articles/rxode2-syntax.html) related to Leibnitz notation.
 - The event data, which can be:
   - a `NONMEM` or `deSolve` [compatible data frame](https://nlmixr2.github.io/rxode2/articles/rxode2-event-types.html), or
   - created with `et()` or `eventTable()` for [easy simulation of events](https://nlmixr2.github.io/rxode2/articles/rxode2-event-table.html)
   - The data frame can be augmented by adding
     [time-varying](https://nlmixr2.github.io/rxode2/articles/rxode2-covariates.html#time-varying-covariates)
     or adding [individual covariates](https://nlmixr2.github.io/rxode2/articles/rxode2-covariates.html#individual-covariates) (`iCov=` as needed)
 - `rxSolve()` which solves the system of equations using initial
   conditions and parameters to make predictions
   - With multiple subject data, [this may be
     parallelized](https://nlmixr2.github.io/rxode2/articles/rxode2-speed.html).
   - With single subject the [output data frame is adaptive](https://nlmixr2.github.io/rxode2/articles/rxode2-data-frame.html)
   - Covariances and other metrics of uncertanty can be used to
     [simulate while solving](https://nlmixr2.github.io/rxode2/articles/rxode2-sim-var.html)

## Installation

```{r child='man-roxygen/rmdhunks/rxode2-install-hunk.Rmd'}
```

# Illustrated Example

```{r child='man/rmdhunks/rxode2-intro-chunk.Rmd'}
```

# Related R Packages
```{r child='man-roxygen/rmdhunks/related-R-chunk.Rmd'}
```
