<!--
---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---
-->
<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- https://blog.r-hub.io/2019/12/03/readmes/ -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RxODE2

<!-- badges: start -->
[![R build status](https://github.com/nlmixr2/rxode2/workflows/R-CMD-check/badge.svg)](https://github.com/nlmixr2/rxode2/actions)
[![codecov.io](https://codecov.io/github/nlmixr2/rxode2/coverage.svg)](https://codecov.io/github/nlmixr2/rxode2)
[![CRAN version](http://www.r-pkg.org/badges/version/rxode2)](https://cran.r-project.org/package=rxode2)
[![CRAN checks](https://cranchecks.info/badges/worst/rxode2)](https://cran.r-project.org/web/checks/check_results_rxode2.html)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/rxode2)](https://cran.r-project.org/package=rxode2)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/rxode2)](https://cran.r-project.org/package=rxode2)
[![CodeFactor](https://www.codefactor.io/repository/github/nlmixr2/rxode2/badge)](https://www.codefactor.io/repository/github/nlmixr2/rxode2)
<!-- badges: end -->

## Overview

**rxode2** is an R package for solving and simulating from ode-based
models. These models are convert the rxode2 mini-language to C and
create a compiled dll for fast solving. ODE solving using rxode2 has a
few key parts:

 - `rxode()` which creates the C code for fast ODE solving based on a
   [simple syntax](https://nlmixr2.github.io/rxode2/articles/rxode2-syntax.html) related to Leibnitz notation.
 - The event data, which can be:
   - a `NONMEM` or `deSolve` [compatible data frame](https://nlmixr2.github.io/rxode2/articles/rxode2-event-types.html), or
   - created with `et()` or `EventTable()` for [easy simulation of events](https://nlmixr2.github.io/rxode2/articles/rxode2-event-table.html)
   - The data frame can be augmented by adding
     [time-varying](https://nlmixr2.github.io/rxode2/articles/rxode2-covariates.html#time-varying-covariates)
     or adding [individual covariates](https://nlmixr2.github.io/rxode2/articles/rxode2-covariates.html#individual-covariates) (`iCov=` as needed)
 - `rxSolve()` which solves the system of equations using initial
   conditions and parameters to make predictions
   - With multiple subject data, [this may be
     parallelized](https://nlmixr2.github.io/rxode2/articles/rxode2-speed.html).
   - With single subject the [output data frame is adaptive](https://nlmixr2.github.io/rxode2/articles/rxode2-data-frame.html)
   - Covariances and other metrics of uncertanty can be used to
     [simulate while solving](https://nlmixr2.github.io/rxode2/articles/rxode2-sim-var.html)

## Installation

```{r child='man-roxygen/rmdhunks/rxode2-install-hunk.Rmd'}
```

# Illustrated Example

```{r child='man/rmdhunks/rxode2-intro-chunk.Rmd'}
```

# Related R Packages
```{r child='man-roxygen/rmdhunks/related-R-chunk.Rmd'}
```
