---
title: "rxode2 model functions"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Learn how to define rxode2 models using the modern model function (UI) syntax,
  including initial estimates and model equations.
---

```{r setup, include=FALSE}
library(learnr)
library(rxode2)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

## Review: Syntax and Events

Before we explore the `rxode2` UI syntax, let's refresh our memory on model syntax and event tables.

### Exercise: A Simple Simulation

Try to create an event table `ev` for a 100 mg dose at time 0, with sampling every hour for 24 hours. Then solve the provided model `m1_syntax`.

```{r review_ex, exercise=TRUE}
m1_syntax <- rxode2({
  ka <- 0.5; cl <- 1; v <- 10
  d/dt(depot) <- -ka * depot
  d/dt(center) <- ka * depot - cl/v * center
  cp <- center/v
})

# Create event table ev here

# Solve the model here

```

```{r review_ex-solution}
m1_syntax <- rxode2({
  ka <- 0.5; cl <- 1; v <- 10
  d/dt(depot) <- -ka * depot
  d/dt(center) <- ka * depot - cl/v * center
  cp <- center/v
})

ev <- et(amt = 100) |> et(0:24)
res <- rxSolve(m1_syntax, ev)
plot(res, cp)
```

## Introduction to Model Functions

In `rxode2`, you can define a model using a standard R function. This approach is powerful because it allows you to package the model equations together with their initial parameter estimates and residual error models.

A `rxode2` model function typically consists of two main blocks:
- `ini({})`: Where you define the initial estimates for your parameters.
- `model({})`: Where you define the model equations (ODEs, algebraic equations, and residual error models).

### A Simple Example

Here is a simple one-compartment model defined as a function:

```r
m1 <- function() {
  ini({
    tka <- log(0.5)
    tcl <- log(1.0)
    tv  <- log(10)
    eta.ka ~ 0.1
    eta.cl ~ 0.1
    eta.v ~ 0.1
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

## The `ini()` Block

The `ini()` block is used to specify the parameters of your model. This includes "fixed effects" (population parameters) and "random effects" (between-subject variability).

### Defining Fixed Effects

Fixed effects can be defined in several ways:
- Simple assignment: `tcl <- log(1.0)`
- With bounds: `tcl <- c(0, log(1.0), 5)` (lower, estimate, upper)
- Fixed values: `tcl <- fix(log(1.0))` or `tcl <- c(0, log(1.0), 5, fixed=TRUE)`

### Defining Random Effects

Random effects (variances) are defined using the `~` operator:
- `eta.cl ~ 0.1` specifies that `eta.cl` has a variance of 0.1.
- You can also specify correlations/covariances by defining a matrix.

### Exercise: Define Parameters

Try to complete the `ini()` block for a model with:
- `ka` (on log scale `tka`) with an initial estimate of 0.5.
- `cl` (on log scale `tcl`) with an initial estimate of 1.0, fixed to this value.
- A variance for `eta.ka` of 0.05.

```{r ini_ex, exercise=TRUE}
my_model <- function() {
  ini({
    # Add tka here
    
    # Add tcl here (fixed)
    
    # Add eta.ka variance here
    
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl)
    d/dt(depot) <- -ka * depot
    cp <- depot / 10
    cp ~ add(0.1)
  })
}
rxode2(my_model)
```

```{r ini_ex-solution}
my_model <- function() {
  ini({
    tka <- 0.5
    tcl <- fix(1.0)
    eta.ka ~ 0.05
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl)
    d/dt(depot) <- -ka * depot
    cp <- depot / 10
    cp ~ add(0.1)
  })
}
rxode2(my_model)
```

## The `model()` Block

The `model()` block contains the mathematical description of your system.

### ODEs and LHS Assignments

- **ODEs**: Use Leibniz notation `d/dt(variable) <- ...`.
- **LHS (Left-Hand Side)**: Ordinary assignments like `cp <- center / v`. These variables will be calculated at each time point.

### Residual Error Models

One of the modern features of `rxode2` is the ability to specify residual error models directly in the `model()` block using the `~` operator.

Common error models include:
- `add(sd)`: Additive error.
- `prop(sd)`: Proportional error.
- `add(add.sd) + prop(prop.sd)`: Combined additive and proportional error.

### Combined Error Models

When you combine additive and proportional error, there are two common ways to represent the variance. You can specify which one to use by adding `combined1()` or `combined2()` to your error model:

- **`combined1()`**: Represents the form $(SD_{add} + Pred \cdot SD_{prop})^2$.
- **`combined2()`**: Represents the "standard" form $SD_{add}^2 + (Pred \cdot SD_{prop})^2$.

If neither is specified, `rxode2` uses a default (usually `combined2()`).

Example:
```r
cp ~ add(add.sd) + prop(prop.sd) + combined2()
```

### Exercise: Complete the Model

Complete the `model()` block below. Define the concentration `cp` and an additive residual error model for it.

```{r model_ex, exercise=TRUE}
my_model <- function() {
  ini({
    tka <- log(0.5); tcl <- log(1.0); tv <- log(10)
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v  <- exp(tv)
    d/dt(center) <- - cl / v * center
    
    # Define cp here
    
    # Define residual error for cp here
    
  })
}
rxode2(my_model)
```

```{r model_ex-solution}
my_model <- function() {
  ini({
    tka <- log(0.5); tcl <- log(1.0); tv <- log(10)
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v  <- exp(tv)
    d/dt(center) <- - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
rxode2(my_model)
```

## The UI Object

When you pass your function to `rxode2()`, it creates a "UI object". This object contains all the information about your model in a structured format.

### Inspecting the UI

You can access different parts of the UI object using the `$` operator:
- `$iniDf`: A data frame of the initial estimates.
- `$model`: The model equations.
- `$predDf`: A data frame describing the residual error models (predictions).
- `$lstExpr`: The model expressed as a list of R expressions.

### Exercise: Inspect a UI Object

Run the following code to create a UI object and inspect its initial estimates data frame.

```{r inspect_ui, exercise=TRUE}
m1 <- function() {
  ini({
    tka <- log(0.5)
    tcl <- log(1.0)
    tv  <- log(10)
    eta.cl ~ 0.1
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv)
    d/dt(center) <- - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

ui <- rxode2(m1)

# Inspect the initial estimates
print(ui$iniDf)

# Inspect the prediction data frame
print(ui$predDf)
```

## Summary

In this tutorial, you learned:
- How to define a `rxode2` model as an R function with `ini()` and `model()` blocks.
- How to specify parameters, bounds, and random effects in `ini()`.
- How to write ODEs and residual error models in `model()`.
- How to create and inspect a `rxode2` UI object.

## What's Next?

Excellent! You now know how to define models using the `rxode2` function syntax. This is the foundation for building more complex and realistic models. In the next tutorial, **[rxode2 piping and modifying](rxode2-piping.Rmd)**, you'll learn how to easily adjust and extend these models without starting from scratch.
