---
title: "rxode2 ODE solving syntax"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rxode2 ODE solving syntax}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

## Introduction

This briefly describes the syntax used to define models 
that `rxode2` will translate into R-callable compiled code. It also
describes the communication of variables between `R` and the
`rxode2` modeling specification.


## Creating rxode2 models

```{r child='../man/rmdhunks/rxode2-create-models.Rmd'}
```

## Syntax

```{r child='../man/rmdhunks/rxode2-syntax-hunk-1.Rmd'}
```

$$
y=f+(a+b\times f^c)\times\varepsilon
$$

The second, **combined2**, assumes that the additive and proportional
differences are combined on a variance scale:

$$
y=f+\left(\sqrt{a^2+b^2\times f^{2c}}\right)\times\varepsilon
$$


```{r child='../man/rmdhunks/rxode2-syntax-hunk-2.Rmd'}
```

## Bugs and/or deficiencies

- The modulo operator `%%` is currently unsupported. 

## Note

The ODE specification mini-language is parsed with the help of the
open source tool \code{dparser}, Plevyak (2015).

## Example model

Below is a commented example to quickly show the capabilities of
`rxode2` syntax.

## Example

```{r}
f <- function() {
  ini({
    
  })
  model({
    # An rxode2 model specification (this line is a comment).

    if(comed==0) {  # concomitant medication (con-med)?
      F <- 1.0     # full bioavailability w.o. con-med
    } else {
      F <- 0.80    # 20% reduced bioavailability
    }

    C2 <- centr/V2  # concentration in the central compartment
    C3 <- peri/V3   # concentration in the peripheral compartment

    # ODE describing the PK and PD

    d/dt(depot) <- -KA*depot
    d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                      Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0)      <- 1
  })
}
```



## Interface and data handling between R and the generated C code

Users specify which variables are the dynamic system's state variables
via the `d/dt(identifier)` operator as part of the model specification,
and which are model parameters via the `params=` argument in `rxode2`
`solve()` method:

```
m1 <- rxode2(model = ode, modName = "m1")

# model parameters -- a named vector is required
theta <- 
   c(KA=0.29, CL=18.6, V2=40.2, Q=10.5, V3=297, Kin=1, Kout=1, EC50=200)

# state variables and their amounts at time 0 (the use of names is
# encouraged, but not required)
inits <- c(depot=0, centr=0, peri=0, eff=1)

# qd1 is an eventTable specification with a set of dosing and sampling 
# records (code not shown here)

solve(theta, event = qd1, inits = inits)
```


The values of these variables at pre-specified time points are saved
during model fitting/integration and returned as part of the fitted
values (see the function `et()`, to define a set of time points when
to capture the values of these variables) and returned as part of the
modeling output.

The ODE specification mini-language is parsed with the help of the
open source tool *DParser*, Plevyak (2015).
