---
title: "rxode2 ODE solving syntax"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rxode2 ODE solving syntax}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

## Introduction

This briefly describes the syntax used to define models 
that `rxode2` will translate into R-callable compiled code. It also
describes the communication of variables between `R` and the
`rxode2` modeling specification.


## Creating rxode2

```{r child='../man/rmdhunks/rxode2-create-models.Rmd'}
```

## Syntax

```{r child='../man/rmdhunks/rxode2-syntax-hunk.Rmd'}
```


## Supported functions

All the supported functions in rxode2 can be seen with the
`rxSupportedFuns()`.

A brief description of the built-in functions are
in the following table:

```{r, echo=FALSE, screenshot.force=FALSE}
library(rxode2)
## rxSyntaxFunctions <- read.csv("syntax-functions.csv");usethis::use_data(rxSyntaxFunctions)
library(DT)
datatable(rxSyntaxFunctions, rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=T) )
```

Note that `lag(cmt) = ` is equivalent to `alag(cmt) = ` and not the same as `= lag(wt)`

## Reserved keywords

There are a few reserved keywords in a rxode2 model.  They are  in the following table:

```{r, echo=FALSE, screenshot.force=FALSE}
## rxReservedKeywords <- read.csv("reserved-keywords.csv");names(rxReservedKeywords)[1] <- "Reserved Name";usethis::use_data(rxReservedKeywords)
datatable(rxReservedKeywords, rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=T) )
```

Note that `rxFlag` will always output `11` or `calc_lhs` since that is
where the final variables are calculated, though you can tweak or test
certain parts of `rxode2` by using this flag.

## Residual functions when using rxode2 functions

In addition to `~` hiding output for certain types of output, it also
is used to specify a residual output or endpoint when the input is an
`rxode2` model function (that includes the residual in the `model({})`
block).

These specifications are of the form:

``` r
var ~ add(add.sd)
```

Indicating the variable `var` is the variable that represents the
individual central tendencies of the model and it also represents the
compartment specification in the data-set.

You can also change the compartment name using the `|` syntax, that is:

``` r
var ~ add(add.sd) | cmt
```

In the above case `var` represents the central tendency and `cmt`
represents the compartment or `dvid` specification. 

### Normal and t-related distributions

For the normal and t-related distributions, we wanted to keep the
ability to use skewed distributions additive and proportional in the
t/cauchy-space, so these distributions are specified differently in
comparison to the other supported distributions within `nlmixr2`:

+-------------------------+----------------+----------------------------+
| Distribution            | How to Add     | Example                    |
+=========================+================+============================+
| Normal (log-likelihood) | `+dnorm()`     | `cp~add(add.sd)+dnorm()`   |
+-------------------------+----------------+----------------------------+
| T-distribuion           | `+dt(df)`      | `cp~add(add.sd)+dt(df)`    |
+-------------------------+----------------+----------------------------+
| Cauchy\                 | `+dcauchy()`   | `cp~add(add.sd)+dcauchy()` |
| (t with df=1)           |                |                            |
+-------------------------+----------------+----------------------------+

Note that with the normal and t-related distributions `nlmixr2` will
calculate `cwres` and `npde` under the normal assumption to help
assess the goodness of the fit of the model.

Also note that the `+dnorm()` is mostly for testing purposes and will
slow down the estimation procedure in `nlmixr2`.  We suggest not
adding it (except for explicit testing).  When there are multiple
endpoint models that mix non-normal and normal distributions, the
whole problem is shifted to a log-likelihood method for estimation in
`nlmixr2`.

### Notes on additive + proportional models

There are two different ways to specify additive and proportional
models, which we will call combined1 and combined2, the same way that
Monolix calls the two distributions (to avoid between software
differences in naming).

The first, **combined1**, assumes that the additive and proportional
differences are on the standard deviation scale, or:

$$
y=f+(a+b\times f^c)\times\varepsilon
$$

The second, **combined2**, assumes that the additive and proportional
differences are combined on a variance scale:

$$
y=f+\left(\sqrt{a^2+b^2\times f^{2c}}\right)\times\varepsilon
$$

The default in `nlmixr2`/`rxode2` if not otherwise specified is
**combined2** since it mirrors how adding 2 normal distributions in
statistics will add their variances (not the standard deviations).
However, the **combined1** can describe the data possibly even better
than **combined2** so both are possible options in `rxode2`/`nlmixr2`.

### Distributions of known likelihoods

For residuals that are not related to normal, t-distribution or
cauchy, often the residual specification is of the form:

``` r
cmt ~ dbeta(alpha, beta)
```

Where the compartment specification is on the left handed side of the
specification. 

For generalized likelihood you can specify:

``` r
ll(cmt) ~ llik specification
```

### Ordinal likelihoods

Finally, ordinal likelihoods/simulations can be specified in 2 ways. The first is:

``` r
err ~ c(p0, p1, p2)
```

Here `err` represents the compartment and `p0` is the probability of
being in a specific category:

| Category | Probability |
|----------|-------------|
| 1        | p0          |
| 2        | p1          |
| 3        | p2          |
| 4        | 1-p0-p1-p2  |

It is up to the model to ensure that the sum of the `p` values are less than `1`. Additionally you can write an arbitrary number of categories in the ordinal model described above.

It seems a little off that `p0` is the probability for category `1` and sometimes scores are in non-whole numbers. This can be modeled as follows:

``` r
err ~ c(p0=0, p1=1, p2=2, 3)
```

Here the numeric categories are specified explicitly, and the probabilities remain the same:

| Category | Probability |
|----------|-------------|
| 0        | p0          |
| 1        | p1          |
| 2        | p2          |
| 3        | 1-p0-p1-p2  |

### General table of supported residual distributions

In general all the that are supported are in the following table
(available in `rxode2::rxResidualError`)

```{r, echo=FALSE, screenshot.force=FALSE}
library(rxode2)
library(DT)
datatable(rxResidualError, rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=TRUE) )
```

## Bugs and/or deficiencies

- The modulo operator `%%` is currently unsupported. 

## Note

The ODE specification mini-language is parsed with the help of the
open source tool \code{dparser}, Plevyak (2015).

## Example model

Below is a commented example to quickly show the capabilities of
`rxode2` syntax.

## Example

```{r}
f <- function() {
  ini({
    
  })
  model({
    # An rxode2 model specification (this line is a comment).

    if(comed==0) {  # concomitant medication (con-med)?
      F <- 1.0     # full bioavailability w.o. con-med
    } else {
      F <- 0.80    # 20% reduced bioavailability
    }

    C2 <- centr/V2  # concentration in the central compartment
    C3 <- peri/V3   # concentration in the peripheral compartment

    # ODE describing the PK and PD

    d/dt(depot) <- -KA*depot
    d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                      Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0)      <- 1
  })
}
```



## Interface and data handling between R and the generated C code

Users specify which variables are the dynamic system's state variables
via the `d/dt(identifier)` operator as part of the model specification,
and which are model parameters via the `params=` argument in `rxode2`
`solve()` method:

```
m1 <- rxode2(model = ode, modName = "m1")

# model parameters -- a named vector is required
theta <- 
   c(KA=0.29, CL=18.6, V2=40.2, Q=10.5, V3=297, Kin=1, Kout=1, EC50=200)

# state variables and their amounts at time 0 (the use of names is
# encouraged, but not required)
inits <- c(depot=0, centr=0, peri=0, eff=1)

# qd1 is an eventTable specification with a set of dosing and sampling 
# records (code not shown here)

solve(theta, event = qd1, inits = inits)
```


The values of these variables at pre-specified time points are saved
during model fitting/integration and returned as part of the fitted
values (see the function `et()`, to define a set of time points when
to capture the values of these variables) and returned as part of the
modeling output.

The ODE specification mini-language is parsed with the help of the
open source tool *DParser*, Plevyak (2015).
