---
title: "rxode2 piping and modifying"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Learn how to use the R pipe operator ( |> ) to modify rxode2 models,
  adding covariates, changing parameters, and updating equations.
---

```{r setup, include=FALSE}
library(learnr)
library(rxode2)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)

one.compartment <- function() {
  ini({
    tka <- 0.45; label("Ka")
    tcl <- 1; label("Cl")
    tv <- 3.45; label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d/dt(depot) = -ka * depot
    d/dt(center) = ka * depot - cl / v * center
    cp = center / v
    cp ~ add(add.sd)
  })
}
```

## Introduction to Piping

In `rxode2`, you can modify an existing model using the R pipe operator (`|>`). This is extremely useful for:
- Adding or removing covariate effects.
- Changing initial estimates without rewriting the whole model.
- Fixing or unfixing parameters.
- Experimenting with different model structures.

The two main functions used for piping are:
- `ini()`: To modify the initial estimates and parameters.
- `model()`: To modify the model equations.

## Modifying the Model with `model()`

When you use `model()` in a pipe, `rxode2` looks at the equations you provide and compares them to the existing model.

### Replacing a Line

If you provide a line that already exists (the left-hand side is the same), `rxode2` will replace that line.

```r
mod <- one.compartment |>
  model(ka <- exp(tka))
```
In the example above, `ka` was originally `exp(tka + eta.ka)`. After piping, the `eta.ka` term is removed.

### Adding a Line

If you provide a new line, it is typically appended to the model.

### Exercise: Modify a Model Line

Start with the `one.compartment` model and remove the between-subject variability from the volume `v`.

```{r model_pipe_ex, exercise=TRUE}
# Current v is: v <- exp(tv + eta.v)
# Modify it to be: v <- exp(tv)
new_mod <- one.compartment |>
  model(v <- exp(tv))

print(new_mod)
```

## Modifying Parameters with `ini()`

The `ini()` function in a pipe allows you to update parameter values, bounds, and labels.

### Changing Values

```r
mod <- one.compartment |>
  ini(tka = 0.5)
```

### Fixing and Unfixing

You can fix a parameter using `fix()` or by passing a logical value.

```r
mod <- one.compartment |>
  ini(tcl = fix(1.2))
```

### Exercise: Update Initial Estimates

Update the `one.compartment` model to set `tka` to 0.8 and fix `tv` at 3.5.

```{r ini_pipe_ex, exercise=TRUE}
new_mod <- one.compartment |>
  ini(tka = 0.8, tv = fix(3.5))

print(new_mod$iniDf)
```

## Automatic Detection and Covariates

`rxode2` tries to be smart about the variables you add:
- Names starting with `t`, `tv`, `pop`, `err` are assumed to be **population parameters**.
- Names starting with `eta`, `iiv`, `bsv` are assumed to be **random effects**.
- Other names are often assumed to be **covariates**.

### Adding a Covariate

```r
mod <- one.compartment |>
  model(cl <- exp(tcl + eta.cl + covClWt * WT))
```
In this case, `WT` is recognized as a covariate and `covClWt` as a new parameter.

### Exercise: Add a Weight Effect

Add a weight effect on clearance (`cl`) using the covariate `WT` and a new parameter `tcl_wt`.

```{r cov_pipe_ex, exercise=TRUE}
mod <- one.compartment |>
  model(cl <- exp(tcl + eta.cl + tcl_wt * log(WT/70)))

print(mod)
```

## Summary

Piping allows for a flexible, iterative workflow:
- `model(...)` updates equations.
- `ini(...)` updates parameters.
- `rxode2` automatically manages the internal structure of the UI object.
