---
editor_options: 
  markdown: 
    wrap: 72
---

This basic model specification consists of one or more statements
optionally terminated by semi-colons `;` and optional comments (comments
are delimited by `#` and an end-of-line).

A block of statements is a set of statements delimited by curly braces,
`{ ... }`.

Statements can be either assignments, conditional `if`/`else if`/`else`,
`while` loops (can be exited by `break`), special statements, or
printing statements (for debugging/testing).

Assignment statements can be:

-   **simple** assignments, where the left hand is an identifier (i.e.,
    variable)

-   special **time-derivative** assignments, where the left hand
    specifies the change of the amount in the corresponding state
    variable (compartment) with respect to time e.g., `d/dt(depot)`:

-   special **initial-condition** assignments where the left hand
    specifies the compartment of the initial condition being specified,
    e.g. `depot(0) = 0`

-   special model event changes including **bioavailability**
    (`f(depot)=1`), **lag time** (`alag(depot)=0`), **modeled rate**
    (`rate(depot)=2`) and **modeled duration** (`dur(depot)=2`). An
    example of these model features and the event specification for the
    modeled infusions the rxode2 data specification is found in [rxode2
    events
    vignette](https://nlmixr2.github.io/rxode2/articles/rxode2-event-types.html).

-   special **change point syntax, or model times**. These model times
    are specified by `mtime(var)=time`

-   special **Jacobian-derivative** assignments, where the left hand
    specifies the change in the compartment ode with respect to a
    variable. For example, if `d/dt(y) = dy`, then a Jacobian for this
    compartment can be specified as `df(y)/dy(dy) = 1`. There may be
    some advantage to obtaining the solution or specifying the Jacobian
    for very stiff ODE systems. However, for the few stiff systems we
    tried with LSODA, this actually slightly slowed down the solving.

Note that assignment can be done by `=`, `<-` or `~`.

When assigning with the `~` operator, the **simple assignments** and
**time-derivative** assignments will not be output. Note that with the
`rxode2` model functions assignment with `~` can also be overloaded with
a residual distribution specification.

Special statements can be:

-   **Compartment declaration statements**, which can change the default
    dosing compartment and the assumed compartment number(s) as well as
    add extra compartment names at the end (useful for multiple-endpoint
    nlmixr models); These are specified by `cmt(compartmentName)`

-   **Parameter declaration statements**, which can make sure the input
    parameters are in a certain order instead of ordering the parameters
    by the order they are parsed. This is useful for keeping the
    parameter order the same when using 2 different ODE models. These
    are specified by `param(par1, par2,...)`

An example model is shown below:

```         
   # simple assignment
   C2 <- centr/V2

   # time-derivative assignment
   d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3;
```

Expressions in assignment and `if` statements can be numeric or logical.

Numeric expressions can include the following numeric operators
`+, -, *, /, ^` and those mathematical functions defined in the C or the
R math libraries (e.g., `fabs`, `exp`, `log`, `sin`, `abs`).

You may also access the R's functions in the [R math
libraries](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Numerical-analysis-subroutines),
like `lgammafn` for the log gamma function.

The `rxode2` syntax is case-sensitive, i.e., `ABC` is different than
`abc`, `Abc`, `ABc`, etc.

### Identifiers

Like R, Identifiers (variable names) may consist of one or more
alphanumeric, underscore `_` or period `.` characters, but the first
character cannot be a digit or underscore `_`.

Identifiers in a model specification can refer to:

-   State variables in the dynamic system (e.g., compartments in a
    pharmacokinetics model).
-   Implied input variable, `t` (time), `tlast` (last time point), and
    `podo` (oral dose, in the undocumented case of absorption transit
    models).
-   Special constants like `pi` or [R's predefined
    constants](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Mathematical-constants).
-   Model parameters (e.g., `ka` rate of absorption, `CL` clearance,
    etc.)
-   Others, as created by assignments as part of the model
    specification; these are referred as *LHS* (left-hand side)
    variable.

Currently, the `rxode2` modeling language only recognizes system state
variables and "parameters", thus, any values that need to be passed from
R to the ODE model (e.g., `age`) should be either passed in the `params`
argument of the integrator function `rxSolve()` or be in the supplied
event data-set.

There are certain variable names that are in the `rxode2` event tables.
To avoid confusion, the following event table-related items cannot be
assigned, or used as a state but can be accessed in the rxode2 code:

-   `cmt`
-   `dvid`
-   `addl`
-   `ss`
-   `rate`
-   `id`

However the following variables are cannot be used in a model
specification:

-   `evid`
-   `ii`

Sometimes rxode2 generates variables that are fed back to rxode2.
Similarly, nlmixr2 generates some variables that are used in nlmixr
estimation and simulation. These variables start with the either the
`rx` or `nlmixr` prefixes. To avoid any problems, it is suggested to not
use these variables starting with either the `rx` or `nlmixr` prefixes.

## Logical Operators

Logical operators support the standard R operators `==`, `!=` `>=` `<=`
`>` and `<`. Like R these can be in `if()` or `while()` statements,
`ifelse()` expressions. Additionally they can be in a standard
assignment. For instance, the following is valid:

```         
cov1 = covm*(sexf == "female") + covm*(sexf != "female")
```

Notice that you can also use character expressions in comparisons. This
convenience comes at a cost since character comparisons are slower than
numeric expressions. Unlike R, `as.numeric` or `as.integer` for these
logical statements is not only not needed, but will cause an syntax
error if you try to use the function.

## Supported functions

All the supported functions in rxode2 can be seen with the
`rxSupportedFuns()`.

A brief description of the built-in functions are in the following
table:

```{r rxSyntaxFunctions, echo=FALSE, screenshot.force=FALSE}
.rxDocTable(rxSyntaxFunctions, caption="rxode2 supported functions")
```

Note that `lag(cmt) =` is equivalent to `alag(cmt) =` and not the same
as `= lag(wt)`

## Reserved keywords

There are a few reserved keywords in a rxode2 model. They are in the
following table:

```{r rxReservedKeywords, echo=FALSE, screenshot.force=FALSE}
.rxDocTable(rxReservedKeywords, caption="rxode2 supported keywords")
```

Note that `rxFlag` will always output `11` or `calc_lhs` since that is
where the final variables are calculated, though you can tweak or test
certain parts of `rxode2` by using this flag.

## Residual functions when using rxode2 functions

In addition to `~` hiding output for certain types of output, it also is
used to specify a residual output or endpoint when the input is an
`rxode2` model function (that includes the residual in the `model({})`
block).

These specifications are of the form:

``` r
var ~ add(add.sd)
```

Indicating the variable `var` is the variable that represents the
individual central tendencies of the model and it also represents the
compartment specification in the data-set.

You can also change the compartment name using the `|` syntax, that is:

``` r
var ~ add(add.sd) | cmt
```

In the above case `var` represents the central tendency and `cmt`
represents the compartment or `dvid` specification.

### Transformations

For normal and related distributions, you can apply the transformation
on both sides by using some keywords/functions to apply these
transformations.

```{r transformationCode, echo=FALSE, screenshot.force=FALSE}
.rxDocTable(dplyr::tribble(~"Transformation", ~"rxode2/nlmixr2 code",
                           "Box-Cox", "+boxCox(lambda)",
                           "Yeo-Johnson", "+yeoJohnson(lambda)",
                           "logit-normal", "+logitNorm(logit.sd, low, hi)",
                           "probit-normal", "+probitNorm(probid.sd, low, hi)",
                           "log-normal","+lnorm(lnorm.sd)"),
            caption="rxode2/nlmixr2 normal/t/cauchy transformations")
```

By default for the likelihood for all of these transformations is
calculated on the **untransformed** scale.

For bounded variables like logit-normal or probit-normal the low and
high values are defaulted to 0 and 1 if missing.

For models where you wish to have a proportional model on one of these
transformation you can replace the standard deviation with `NA`

To allow for more transformations, `lnorm()`, `probitNorm()` and
`logitNorm()` can be combined the variance stabilizing `yeoJohnson()`
transformation.

### Normal and t-related distributions

For the normal and t-related distributions, we wanted to keep the
ability to use skewed distributions additive and proportional in the
t/cauchy-space, so these distributions are specified differently in
comparison to the other supported distributions within `nlmixr2`:

```{r llikTable, echo=FALSE, screenshot.force=FALSE}
.rxDocTable(dplyr::tribble(
  ~"Distribution", ~"How to Add", ~"Example",
  "Normal (log-likelihood)", "+dnorm()", "cc ~ add(add.sd) + dnorm()",
  "T-distribution", "+dt(df)", "cc ~a dd(add.sd) + dt(df)",
  "Cauchy (t with df=1)", "+dcauchy()", "cc ~ add(add.sd) + dcauchy()"),
            caption="rxode2/nlmixr2 normal/t/cauchy log-liklihood specification")
```

Note that with the normal and t-related distributions `nlmixr2` will
calculate `cwres` and `npde` under the normal assumption to help assess
the goodness of the fit of the model.

Also note that the `+dnorm()` is mostly for testing purposes and will
slow down the estimation procedure in `nlmixr2`. We suggest not adding
it (except for explicit testing). When there are multiple endpoint
models that mix non-normal and normal distributions, the whole problem
is shifted to a log-likelihood method for estimation in `nlmixr2`.

### Notes on additive + proportional models

There are two different ways to specify additive and proportional
models, which we will call **combined1** and **combined2**, the same way
that Monolix calls the two distributions (to avoid between software
differences in naming).

The first, **combined1**, assumes that the additive and proportional
differences are on the standard deviation scale, or:

y=f+(a+b* f^c)*err

The second, **combined2**, assumes that the additive and proportional
differences are combined on a variance scale:

y=f+\[sqrt(a^2+b^2 *f^(2c))\]*err

The default in `nlmixr2`/`rxode2` if not otherwise specified is
**combined2** since it mirrors how adding 2 normal distributions in
statistics will add their variances (not the standard deviations).
However, the **combined1** can describe the data possibly even better
than **combined2** so both are possible options in `rxode2`/`nlmixr2`.

### Distributions of known likelihoods

For residuals that are not related to normal, t-distribution or cauchy,
often the residual specification is of the form:

``` r
cmt ~ dbeta(alpha, beta)
```

Where the compartment specification is on the left handed side of the
specification.

For generalized likelihood you can specify:

``` r
ll(cmt) ~ llik specification
```

### Ordinal likelihoods

Finally, ordinal likelihoods/simulations can be specified in 2 ways. The
first is:

``` r
err ~ c(p0, p1, p2)
```

Here `err` represents the compartment and `p0` is the probability of
being in a specific category:

| Category | Probability |
|----------|-------------|
| 1        | p0          |
| 2        | p1          |
| 3        | p2          |
| 4        | 1-p0-p1-p2  |

It is up to the model to ensure that the sum of the `p` values are less
than `1`. Additionally you can write an arbitrary number of categories
in the ordinal model described above.

It seems a little off that `p0` is the probability for category `1` and
sometimes scores are in non-whole numbers. This can be modeled as
follows:

``` r
err ~ c(p0=0, p1=1, p2=2, 3)
```

Here the numeric categories are specified explicitly, and the
probabilities remain the same:

| Category | Probability |
|----------|-------------|
| 0        | p0          |
| 1        | p1          |
| 2        | p2          |
| 3        | 1-p0-p1-p2  |

### General table of supported residual distributions

In general all the that are supported are in the following table
(available in `rxode2::rxResidualError`)

```{r rxResidualError, echo=FALSE, screenshot.force=FALSE}
.rxDocTable(rxResidualError, "rxode2/nlmixr2 esidual error specification")
```
