The ODE-based model specification may be coded inside four places: 

- Inside a `rxode2({})` block statements:

```{r modBracket}
library(rxode2)
mod <- rxode2({
  # simple assignment
  C2 <- centr/V2

  # time-derivative assignment
  d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3;
})
```

- Inside a `rxode2("")` string statement:

```{r modString}
mod <- rxode2("
  # simple assignment
  C2 <- centr/V2

  # time-derivative assignment
  d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3;
")
```

- In a file name to be loaded by rxode2:

```{r modFile} 
writeLines("
  # simple assignment
  C2 <- centr/V2

  # time-derivative assignment
  d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3;
", "modelFile.rxode2")
mod <- rxode2(filename='modelFile.rxode2')
unlink("modelFile.rxode2")
```
  
- In a model function which can be parsed by `rxode2`:

```{r modFun}
mod <- function() {
  model({
    # simple assignment
    C2 <- centr/V2

    # time-derivative assignment
    d/dt(centr) <- F*KA*depot - CL*C2 - Q*C2 + Q*C3;
  })
}

mod <- rxode2(mod) # or simply mod() if the model is at the end of the function
```

For the block statement, character string or text file an internal
`rxode2` compilation manager translates the ODE system into C,
compiles it and loads it into the R session.  The
call to `rxode2` produces an object of class `rxode2` which
consists of a list-like structure (environment) with various member
functions.

For the last type of model (a model function), a call to `rxode2`
creates a parsed `rxode2` ui that can be translated to the `rxode2`
compilation model.  This is the same type of function required for
`nlmixr2` estimation and will be focused on in the documentation.
