---
title: "rxode2 simulation"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Learn how to perform simulations using rxSolve, including single-subject
  solutions, population simulations with BSV, and simulations with uncertainty.
---

```{r setup, include=FALSE}
library(learnr)
library(rxode2)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)

pkModel <- function() {
  ini({
    tka  <- log(1.57)
    tcl  <- log(2.72)
    tv   <- log(31.07)
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

## Introduction to `rxSolve()`

Once you have defined a model, you use `rxSolve()` to calculate the predicted values over time. To solve a model, you need:
1. A **Model** (the UI object or function we created).
2. An **Event Table** (defining doses and sampling times).

## Creating Events with `et()`

The `et()` function is a convenient way to create event tables.

```r
ev <- et(amt = 100) |>       # Dose of 100 at time 0
  et(seq(0, 24, by = 0.5))   # Samples every 0.5h for 24h
```

### Exercise: Create an Event Table

Create an event table with:
- A dose of 250 mg at time 0.
- Observations every hour for 48 hours.

```{r et_ex, exercise=TRUE}
ev <- et(amt = 250) |>
  et(0:48)

print(ev)
```

## Solving for a Single Subject

By default, if you don't specify between-subject variability, `rxode2` solves for a single subject using the initial estimates.

```{r solve_single, exercise=TRUE}
ev <- et(amt = 100) |> et(0:24)
res <- rxSolve(pkModel, ev)
plot(res, cp)
```

## Population Simulation (`nSub`)

If your model has random effects (`eta` parameters), you can simulate a population of subjects by specifying `nSub`. `rxode2` will automatically draw random values for the `etas` from the specified distribution.

### Exercise: Simulate 100 Subjects

Simulate the `pkModel` for 100 subjects and plot the results.

```{r solve_pop, exercise=TRUE}
ev <- et(amt = 100) |> et(0:24)

# Use nSub = 100
res <- rxSolve(pkModel, ev, nSub = 100)

plot(res, cp)
```

## Simulation with Parameter Uncertainty (`nStud`, `thetaMat`)

In addition to between-subject variability, you can also simulate **parameter uncertainty** (uncertainty in the "fixed effects"). This is often used for clinical trial simulation.

You provide a variance-covariance matrix for the parameters (`thetaMat`) and specify the number of "studies" (`nStud`).

```{r solve_uncert, exercise=TRUE}
# Example Theta Variance-Covariance matrix
thetaVcov <- lotri::lotri({
  tka ~ 0.04
  tcl ~ 0.02
  tv  ~ 0.03
})

ev <- et(amt = 100) |> et(0:24)

# Simulate 10 virtual studies with 20 subjects each
res <- rxSolve(pkModel, ev, thetaMat = thetaVcov, nStud = 10, nSub = 20)

# The confint() function can summarize across studies
ci <- confint(res, "cp")
plot(ci)
```

## Summary

- Use `et()` to define your dosing and sampling.
- Use `rxSolve()` to run the simulation.
- `nSub` controls the number of subjects (BSV).
- `nStud` and `thetaMat` control the number of parameter draws (Uncertainty).
- Resulting objects can be easily plotted or summarized with `confint()`.
