---
title: "rxode2 covariates and data"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Learn how to include covariates in your rxode2 models, passing them as
  individual data (iCov) or time-varying data in event tables.
---

```{r setup, include=FALSE}
library(learnr)
library(rxode2)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)

modCov <- function() {
  ini({
    tka <- 0.5; tcl <- 1; tv <- 10
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl) * (WT/70)^0.75
    v  <- exp(tv)
    d/dt(center) <- - cl/v * center
    cp <- center/v
    cp ~ add(add.sd)
  })
}
```

## Review: Simulation and Model Functions

Before we add covariates, let's review how to run a basic population simulation.

### Exercise: Population Simulation

Using the `modCov` model (ignoring the `WT` part for a moment by setting it to 70), simulate 50 subjects with a 100 mg bolus dose.

```{r simulationReviewCov, exercise=TRUE}
# Set up event table ev (100 mg dose, 0-24h sampling)
ev <- et(amt = 100) |> et(0:24)

# Set WT to 70 in the event table to ignore the covariate effect for now
ev$WT <- 70

# Solve for 50 subjects
res <- rxSolve(modCov, ev, nSub = 50)

plot(res, cp)
```

## Introduction to Covariates

Covariates are external variables that influence your model parameters (e.g., body weight, age, or smoking status). In `rxode2`, any variable in your model that is NOT defined as a parameter in `ini()` and NOT calculated as a LHS variable is assumed to be a **covariate**.

## Individual Covariates (`iCov`)

If you have covariates that are constant for each individual (like baseline weight), you can pass them using the `iCov` argument in `rxSolve()`.

```r
# Create individual covariate data frame
myCovs <- data.frame(id = 1:5, WT = c(60, 70, 80, 90, 100))

# Solve with iCov
res <- rxSolve(modCov, ev, iCov = myCovs)
```

### Exercise: Simulate with Weight Covariate

Simulate 4 subjects with weights 60, 70, 80, and 90 kg.

```{r icovEx, exercise=TRUE}
ev <- et(amt = 100) |> et(0:24)

myCovs <- data.frame(id = 1:4, WT = c(60, 70, 80, 90))

res <- rxSolve(modCov, ev, iCov = myCovs)

plot(res, cp)
```

## Time-Varying Covariates

If a covariate changes over time (like heart rate or a continuous infusion), you can include it directly in your event table.

```r
ev <- et(amt = 100) |> et(0:24)
ev$WT <- 70  # Add WT to every row of the event table
```

### Covariate Interpolation

When the solver needs a value for a covariate at a time point that is not in your data, it uses interpolation. You can control this with the `covsInterpolation` argument:
- `"linear"`: Linear interpolation between points.
- `"locf"`: Last Observation Carried Forward (constant until the next point).
- `"nocb"`: Next Observation Carried Backward.

### Exercise: Time-Varying Weight

Suppose a subject starts at 70 kg and gains 10 kg over 24 hours.

```{r tvCovEx, exercise=TRUE}
ev <- et(amt = 100) |>
  et(0) |>
  et(24)

# Define WT at time 0 and time 24
ev$WT <- c(70, 80)

# Solve with linear interpolation
resLinear <- rxSolve(modCov, ev, covsInterpolation = "linear")
plot(resLinear, cp)
```

## Summary

- Covariates are automatically recognized if they are not defined in `ini()`.
- Use `iCov` for subject-level constants.
- Include columns in your event table for time-varying covariates.
- Choose an interpolation method (`linear`, `locf`, `nocb`) that suits your data.

## What's Next?

You've now covered the core workflow of `rxode2`: defining, modifying, simulating, and adding data to models. In the final tutorial, **[rxode2 advanced features](rxode2-advanced.Rmd)**, we'll explore some specialized tools that can speed up your work, such as analytical solutions for standard PK models.
