---
title: "Adaptive Dosing with rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(rxode2)
```

## Overview

`rxode2` can modify an individual's event history while the model is
being solved.  This allows adaptive dosing strategies such as:

- response-guided titration
- rescue doses
- infusion starts with fixed rate or fixed duration
- compartment reset, replacement, or proportional adjustment
- phantom/transit doses that update `podo()` and `tad()` without
  adding mass to the compartment
- adaptive observation scheduling

The examples in `tests/testthat/test-evid-push.R` cover the core
building blocks for these workflows.  This article shows how those
pieces fit together for pharmacometric tasks such as titration,
exposure matching, treatment interruption, and transit absorption.

## General pattern

Adaptive dosing in `rxode2` follows a simple pattern:

1. Put the decision logic inside the `model({})` block.
2. Make sure the solver visits the decision times.
3. Push a new event when the condition is met.

The push functions are evaluated at scheduled output times.  In
practice, this means you should include the assessment times in the
event table, often with `et(seq(...))`, and use `mtime()` when you want
the model to react at a specific planned time.

```{r adaptive-table, echo=FALSE}
paradigms <- data.frame(
  helper = c("evid_()", "bolus()", "infuse()", "infuseDur()", "reset()",
             "replace()", "multiply()", "phantom()", "obs()"),
  evid = c("0, 1, 3, 4, 5, 6, 7, >=100", "1", "1", "1", "3", "5", "6", "7", "0"),
  use = c(
    "General low-level interface for any supported pushed event",
    "Push bolus doses, optionally with ii/addl/ss",
    "Push fixed-rate infusions",
    "Push fixed-duration infusions",
    "Reset all compartments",
    "Set a compartment to a new amount",
    "Scale a compartment by a factor",
    "Start transit or phantom dosing without adding mass directly",
    "Push one or more observation rows after the current solve time"
  ),
  check.names = FALSE
)

knitr::kable(paradigms, align = "lcl")
```

For portability, the examples below pass the full helper argument list
explicitly when the helper supports `ii`, `addl`, or `ss`.

## A simple titration example with `bolus()`

The most common adaptive dosing use case is titration: assess response,
then add another dose if the current exposure is too low.

```{r bolus-titration}
titrateModel <- rxode2({
  d/dt(depot) <- -ka * depot
  d/dt(central) <- ka * depot - cl / v * central
  cp <- central / v
  # Check every 24 hours, add a bolus dose if needed
  if (t %% 24 == 0 && cp < 1) {
    bolus(50, depot, 0, 0, 0)
  }
})

titrateEvents <- et(amt = 100, time = 0) |>
  et(seq(0, 72, by = 1))

titratePars <- c(ka = 0.5, cl = 1.2, v = 20)
titrateSolve <- rxSolve(titrateModel, titratePars, titrateEvents)

plot(titrateSolve, cp)
```

Note that this patient is dosed at time zero, but not at 24 hours
since the concentration was not below 1.  However by 48 hours the
concentration fell below 1 so a dose of 50 is added.

As a word of caution, to administer a dose at time 48 that needs to be
seen by the solver. For example:

```{r bad-titration-unseen}
library(dplyr)
titrateEvents <- titrateEvents |>
  filter(time != 48)

titratePars <- c(ka = 0.5, cl = 1.2, v = 20)
titrateSolve <- rxSolve(titrateModel, titratePars, titrateEvents)

plot(titrateSolve, cp)
```

One way to overcome this is to use `mtime()`

```{r missing-time-but-still-dosed}

titrateModel <- rxode2({
  mtime(time48) <- 48 # always visit 48 hours
  d/dt(depot) <- -ka * depot
  d/dt(central) <- ka * depot - cl / v * central
  cp <- central / v
  # Check every 24 hours, add a bolus dose if needed
  if (t %% 24 == 0 && cp < 1) {
    bolus(50, depot, 0, 0, 0)
  }
})

titratePars <- c(ka = 0.5, cl = 1.2, v = 20)
titrateSolve <- rxSolve(titrateModel, titratePars, titrateEvents)

plot(titrateSolve, cp)
```

Of course the other is to include the sampling times in the solver (as
shown above).  You can also expose the events using the `obs()`
adaptive sampling helper:

```{r MissingWithObs}

titrateModel <- rxode2({
  if (time == 0) {
    obs(48) # only schedule at time 0 so that it doesn't schedule
            # multiple observations at 48 hours
  }
  d/dt(depot) <- -ka * depot
  d/dt(central) <- ka * depot - cl / v * central
  cp <- central / v
  # Check every 24 hours, add a bolus dose if needed
  if (t %% 24 == 0 && cp < 1) {
    bolus(50, depot, 0, 0, 0)
  }
})

titrateSolve <- rxSolve(titrateModel, titratePars, titrateEvents)

plot(titrateSolve, cp)
```

As a note, there are many required arguments for the `bolus()` dose
when using the `rxode2({})` solving form of the model.  However, these
are not required when you use the functional model:

```{r funModel}
titrateModel <- function() {
  ini({
    ka <- 0.5
    cl <- 1
    v <- 10
  })
  model({
    if (time == 0) {
      obs(48) # only schedule at time 0 so that it doesn't schedule
      # multiple observations at 48 hours
    }
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central
    cp <- central / v
    # Check every 24 hours, add a bolus dose if needed
    if (t %% 24 == 0 && cp < 1) {
      bolus(50) # or bolus(50, depot); default is cmt 1
    }
  })
}

titrateSolve <- rxSolve(titrateModel, titrateEvents)

plot(titrateSolve, cp)
```

Overall, this pattern is useful for:

- therapeutic drug monitoring
- concentration-guided rescue dosing
- model-informed dose escalation rules

If you need a repeating pushed regimen instead of a single titration
step, `bolus()` also supports `ii`, `addl`, and `ss`.

## Using `linCmt()` with adaptive dosing

Adaptive dosing works the same way in analytical linear compartment
models as it does in ODE models. You can reference the linear
compartment names (`depot`, `central`) directly in `bolus()`,
`infuse()`, and the other helpers.

In the `rxode2({})` solving form, supply the full set of positional
arguments to `bolus()`:

```{r lincmt-bolus-positional}
titrateModelLin <- rxode2({
  cp <- linCmt(ka, cl, v)
  # Check every 24 hours, add a bolus dose if needed
  if (t %% 24 == 0 && cp < 1) {
    bolus(50, depot, 0, 0, 0)
  }
})

titrateEventsLin <- et(amt = 100, time = 0) |>
  et(seq(0, 96, by = 1))

titratePars <- c(ka = 0.5, cl = 1.2, v = 20)
titrateSolveLin <- rxSolve(titrateModelLin, titratePars, titrateEventsLin)

plot(titrateSolveLin, cp)
```

In the functional model form you can use named arguments, and the
remaining `bolus()` arguments take their defaults:

```{r lincmt-bolus-functional}
titrateModelLin <- function() {
  ini({
    ka <- 0.5
    cl <- 1.2
    v <- 20
  })
  model({
    cp <- linCmt(ka, cl, v)
    # Check every 24 hours, add a bolus dose if needed
    if (t %% 24 == 0 && cp < 1) {
      bolus(50, cmt = depot)
    }
  })
}

titrateSolveLin <- rxSolve(titrateModelLin, titrateEventsLin)

plot(titrateSolveLin, cp)
```

If you already have an ODE model that uses adaptive dosing, you do not
have to rewrite it by hand. `odeToLin()` detects compatible 1-3
compartment linear PK systems and converts them to a `linCmt()` model,
preserving any in-model `bolus()`/`infuse()` calls (and renaming
non-standard compartment names to `depot`/`central` automatically):

```{r lincmt-odetolin}
titrateOde <- function() {
  ini({
    tka <- log(0.5)
    tcl <- log(1.2)
    tv  <- log(20)
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v  <- exp(tv)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central
    cp <- central / v
    if (t %% 24 == 0 && cp < 1) {
      bolus(50, cmt = depot)
    }
  })
}

# Convert explicitly ...
titrateLin <- odeToLin(titrateOde)
rxSolve(titrateLin, titrateEventsLin) |> plot(cp)

# ... or transparently at solve time with useLinCmt = TRUE
rxSolve(titrateOde, titrateEventsLin, useLinCmt = TRUE) |> plot(cp)
```

## Starting adaptive infusions

Two helper functions support adaptive infusion starts:

- `infuse(amt, rate, cmt, ii, addl, ss)` for fixed-rate infusions
- `infuseDur(amt, dur, cmt, ii, addl, ss)` for fixed-duration infusions

The next example starts a rescue infusion when concentration falls below
target at a planned decision time.

```{r infusion-titration}
infusionModel <- function(){
  ini({
    cl <- 1
    v <- 10
  })
  model({
    mtime(rescueAt) <- 24

    d/dt(central) <- -cl / v * central
    cp <- central / v

    if (t == rescueAt && cp < 1) {
      infuse(amt=50, rate=5, cmt=central)
    }
  })
}

infusionEvents <- et(amt = 100, time = 0, cmt = 1) |>
  et(seq(0, 48, by = 1))

# Here an infusion rescue is performed
infusionSolve <- rxSolve(infusionModel, infusionEvents)

plot(infusionSolve, cp)

# Here the bolus was high enough that there was no infusion rescue
infusionEvents <- et(amt = 1000, time = 0, cmt = 1) |>
  et(seq(0, 48, by = 1))

infusionSolve <- rxSolve(infusionModel, infusionEvents)

plot(infusionSolve, cp)

```

Use `infuse()` when the infusion pump rate is fixed and duration should
follow from `amt / rate`.  Use `infuseDur()` when the administration
window is fixed and the effective rate should be computed from the realized
drug amount and duration.

The real impact of the `infuse()` is that when `f` changes the amount
available (realized drug amount), the infusion duration increases.
When using `infuseDur()` the infusion rate changes, but not the duration.

## Reset, replace, and multiply

Not every adaptive intervention is a new dose.  Sometimes the state
itself needs to be reset or adjusted.

### `reset()`

`reset()` clears the system state.  This is useful for:

- washout periods
- treatment restarts
- restarting a simulation after protocol-defined interruption

### `replace(amt, cmt)`

`replace()` sets a compartment to a new amount.  This is useful for:

- forcing a compartment to a therapeutic target
- modeling device fills or reservoir changes
- imposing a measured value as a new state

### `multiply(amt, cmt)`

`multiply()` scales a compartment amount by a factor.  This is useful
for:

- proportional dose reduction or escalation
- dialysis-like fractional removal
- immediate bioavailability or depot scaling rules

```{r replace-multiply}

adjustModel <- rxode2({
  mtime(replaceAt) <- 12
  mtime(multAt) <- 24

  d/dt(depot) <- -ka * depot
  d/dt(central) <- ka * depot - cl / v * central
  cp <- central / v

  if (t == replaceAt && depot > 0) {
    replace(25, depot)
  }
  if (t == multAt && depot > 0) {
    multiply(0.5, depot)
  }
})

adjustEvents <- et(amt = 100, time = 0) |>
  et(seq(0, 36, by = 0.25))

adjustPars <- c(ka = 0.5, cl = 1, v = 10)
adjustSolve <- rxSolve(adjustModel, adjustPars, adjustEvents)

plot(adjustSolve, cp)

```

## Transit and phantom dosing with `phantom()`

`phantom()` is the adaptive analogue of `evid = 7`.  It updates
transit-related bookkeeping such as `podo()` and `tad()` without adding
the dose directly to the state variable as a bolus.

This is especially useful for transit absorption models, delayed onset
models, or situations where dose timing should update the model clock
without depositing mass immediately.

```{r phantom-example}

phantomModel <- function() {
  ini({
    k <- 0.2
  })
  model({
    mtime(startTransit) <- 12

    d/dt(depot) <- 0
    d/dt(x) <- -k * x
    x(0) <- 1
    pd <- podo(depot)
    td <- tad(depot)

    if (t == startTransit) {
      phantom(20, depot)
    }
  })
}

phantomEvents <- et(seq(0, 20, by = 0.5))
phantomSolve <- rxSolve(phantomModel, phantomEvents)

print(subset(as.data.frame(phantomSolve), time %in% 11:16,
             select = c(time, depot, x, pd, td)))

plot(phantomSolve, x)
```

Notice that `pd` and `td` update after the phantom event while the
depot state is not handled like an ordinary bolus dose.

## Adaptive observation scheduling with `obs()`

When the goal is simply to add future observation rows, `obs()` is a
more concise interface than `evid_(..., evid = 0, ...)`. Each argument
is interpreted as a time offset from the current solve time.

```{r obs-direct}
obsModel <- rxode2({
  mtime(extraSample) <- 8
  d/dt(x) <- -k * x

  if (t >= extraSample && t < extraSample + 0.5) {
    obs(0.5, 1, 2)
  }
})

obsEvents <- et(amt = 1, time = 0) |>
  et(c(0, 8, 12))

obsSolve <- rxSolve(obsModel, c(k = 0.2), obsEvents, inits = c(x = 1))
obsSolve[, c("time", "x")]
```

Use `obs()` when you want:

- adaptive follow-up sampling after a decision point
- multiple future observation rows from one trigger
- a simpler alternative to `evid_()` for observation-only pushes

## Direct control with `evid_()`

When one of the convenience helpers does not match the exact event you
need, use `evid_()` directly.  This gives full access to the
NONMEM-style event fields:

```{r evid-direct}
directModel <- rxode2({
  mtime(extraSample) <- 8
  d/dt(x) <- -k * x

  if (t >= extraSample && t < extraSample + 0.5) {
    evid_(t + 0.5, 0, 0, 1, 0, 0, 0, 0)
  }
})

directEvents <- et(amt = 1, time = 0) |>
  et(c(0, 8, 12))

directSolve <- rxSolve(directModel, c(k = 0.2), directEvents, inits = c(x = 1))
directSolve[, c("time", "x")]
```

The low-level `evid_()` interface is appropriate when you need:

- adaptive observation rows (`evid = 0`)
- explicit resets or reset-then-dose logic
- direct control over `rate`, `ii`, `addl`, and `ss`
- classic internal rxode2 evids


## Practical advice for titration workflows

### 1. Put decision times in the event table

Adaptive logic only runs when the solver is at a scheduled event or
output time.  If a titration visit is supposed to happen at 24 hours,
make sure the event table contains 24 hours.  For dosing regimens
every 24 hours you can use the modulo operator to check for dosing
times, that is `(time %% 24) == 0`.

### 2. Use `mtime()` or `obs()` for planned decision points

`mtime()` is a convenient way to mark protocol decision times inside the
model and makes the logic easier to read than repeating literal times.

The `mtime()` framework will only be observed once, while if you use
`obs()` make sure to add observations at the appropriate times.

### 3. Guard against repeated firing

A condition such as `if (cp < target)` can remain true across many
times.  Anchor it to a visit window or other one-time trigger so the
same adaptive action is not pushed repeatedly by accident.

### 4. Monitor `maxExtra`

Each pushed event counts against the `maxExtra` limit in `rxSolve()`.
If you accidentally create an event cascade, `rxode2` will stop with a
`maxExtra` error instead of silently generating an unbounded event
history.

### 5. Pick the helper that matches the pharmacology

- use `bolus()` for added doses
- use `infuse()` or `infuseDur()` for infusion starts
- use `reset()` for clearing the system
- use `replace()` when you know the new compartment amount. Helpful
  for calculating AUCs.
- use `multiply()` for proportional adjustment
- use `phantom()` for transit bookkeeping without direct deposition
- use `obs()` for observation-only follow-up rows

## Common pharmacometric applications

These adaptive dosing tools are useful in many pharmacometric settings:

- oncology dose holds, dose reductions, and re-escalation
- anti-infective therapeutic drug monitoring
- target-controlled rescue dosing
- model-based switching from loading dose to maintenance infusion
- transit absorption and delayed onset models
- protocol simulation with response-based dosing rules

## Summary

The adaptive event tools in `rxode2` let the model create new events at
solve time.  Together, `evid_()`, `bolus()`, `infuse()`,
`infuseDur()`, `reset()`, `replace()`, `multiply()`, `phantom()`, and
`obs()` cover the main adaptive dosing paradigms needed for titration,
interruption, rescue therapy, adaptive sampling, transit models, and
protocol simulation.

The tests in `tests/testthat/test-evid-push.R` provide concrete
regression coverage for these event types, and the same patterns can be
used in practical pharmacometric workflows.
