---
title: "Integrating User Defined Functions into rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## User Defined Functions

```{r setup}
library(rxode2)
```

When defining models you may have wished to write a small R function
or make a function integrate into `rxode2` somehow. This article
discusses 4 ways to do this:

- A R-based user function which can be loaded as a simple function or
  in certain circumstances translated to C to run more efficiently

- A C function that you define and integrate into code

- A user defined function that changes `rxode2` ui code by replacing
  the function with `rxode2` code.  This can be in the presence or
  absence of the data for simulation or estimation.

## R based user functions

A R-based user function is the most convenient to include in the ODE,
but is slower than what you could have done if it was written in `C` ,
`C++` or some other compiled language. This was requested [in
github](https://github.com/nlmixrdevelopment/RxODE/issues/162#issue-568886732)
with an appropriate example; However, I will use a very simple example
here to simply illustrate the concepts.

```{r fExample}

newAbs <- function(x) {
  if (x < 0) {
    -x
  } else {
    x
  }
}

f <- rxode2({
  a <- newAbs(time)
})

e <- et(-10, 10, length.out=40)
```

Now that the ODE has been compiled the R functions will be called
while solving the ODE. Since this is calling R, this forces the
parallization to be turned off since R is single-threaded. It also
takes more time to solve since it is shuttling back and forth between
R and C. Lets see how this very simple function performs:

```{r benmchmark1}
mb1 <- microbenchmark::microbenchmark(withoutC=suppressWarnings(rxSolve(f,e)))

library(ggplot2)
autoplot(mb1) + rxTheme()
```

Not terribly bad, even though it is shuffling between R and C.

You can make it a better by converting the functions to C:

```{r convertC}
# Create C functions automatically with `rxFun()`
rxFun(newAbs)
# Recompile to use the C functions
# Note it would recompile anyway if you didn't do this step,
# it just makes sure that it doesn't recompile every step in
# the benchmark
f <- rxode2({
  a <- newAbs(time)
})

mb2 <- microbenchmark::microbenchmark(withC=rxSolve(f,e, cores=1))

mb <- rbind(mb1, mb2)
autoplot(mb) + rxTheme() + xgxr::xgx_scale_y_log10()
print(mb)
```

The C version is almost twice as fast as the R version. You may have
noticed the conversion also created C versions of the first
derivative. This is done automatically and gives not just C versions
of function, but C versions of the derivatives and registers them with
`rxode2`. This allows the C versions to work with not only `rxode2`
but `nlmixr2` models.

This function was setup in advance to allow this type of
conversion. In general the derivatives will be calculated if there is
not a `return()` statement in the user defined function. This means
simply let R return the last value instead of explictly calling out
the `return()`. Many people prefer this method of coding.

Even if there is a `return` function, the function could be converted
to `C`. In the github issue, they used a function that would not
convert the derivatives:

```{r fConvNoDer}
# Light
f_R <- function(actRad, k_0, a_k) {
  photfac <- a_k * actRad + k_0
  if (photfac > 1) {
    photfac = 1
  }
  return(photfac)
}

rxFun(f_R)
```

While this is still helpful because some functions have early returns,
the `nlmixr2` models requiring derivatives would be calculated be
non-optimized finite differences when this occurs. While this gets
into the internals of `rxode2` and `nlmixr2` you can see this more
easily when calculating the derivatives:

```{r derExample}
rxFromSE("Derivative(f_R(actRad, k_0, a_k),k_0)")
```

Whereas the originally defined function `newAbs()` would use the new
derivatives calculated as well:

```{r derFromC}
rxFromSE("Derivative(newAbs(x),x)")
```

In some circumstances, the conversion to C is not possible, though you
can still use the R function.

There are some requirements for R functions to be integrated into the
rxode2 system:

-    The function must have a set number of arguments, variable
     arguments like `f(…)` are currently not allowed.

-   The function is given each argument as a single number, and the
    function should return a single number

If these requirements are met you can use the R function in
rxode2. Additional requirements for conversion to C include:

-   Any functions that you use within the R function must be
    understood and available to `rxode2`.

    -   Practically speaking if you have `fun2()` which refers to
        `fun1()`, `fun1()` must be changed to C code and available to
        `rxode2` before changing the function `fun2()` to C.

-   The functions can include `if`/`else` assignments or simple return
    statements (either by returning a value or having that value on a
    line by itself). Special R control structures and functions (like
    `for` and `lapply`) cannot be present.

-   The function cannot refer to any package functions

-   As mentioned, if the `return()` statement is present, the
    derivative C functions and `rxode2`'s derivative table is not
    updated.

## C based functions

You can add your own C functions directly into rxode2 as well using
`rxFun()`:

```{r rxFun}
fun <- "
 double fun(double a, double b, double c) {
   return a*a+b*a+c;
 }
" ## C-code for function

rxFun("fun", c("a", "b", "c"), fun)
```

If you wanted you could also use C functions or expressions for the derivatives by using the `rxD()` function:

```{r rxD}
rxD("fun", list(
   function(a, b, c) { # derivative of arg1: a
     paste0("2*", a, "+", b)
   },
   function(a, b, c) { # derivative of arg2: b
     return(a)
   },
   function(a, b, c) { # derivative of arg3: c
     return("0.0")
   }
))
```

Removing the function with `rxRmFun()` will also remove the derivative
table:

```{r rxDrm}
rxRmFun("fun")
```

## Functions to insert `rxode2` code into the current model

This replaces `rxode2` code in the current model with some expansion.
This can allow more R-like functions inside of the rxode2 ui models,
as well as adding approximating functions like polynomials, splines
and neural networks.

An example that allows more `R`-like functions is below:

```r
f <- function() {
  model({
    a <- rxpois(lambda=lam)
  })
}

# Which will evaluate into a standard rxode2 function that does not
# support named arguments (since it is translated to C)
f()

# Which is still true in the standard rxode2:

try(rxode2({
  a <- rxpois(lambda=lam)
}))
```

This is accomplished by a combination of two functions, which are
highly commented:

``` r
rxUdfUi.rxpois <- function(fun) {
  # Fun is the language object (ie quoted R object) to be evaluated or
  # changed in the code
  .fun <- fun
  # Since the `rxpois` function is built into the rxode2 we need to
  # have a function with a different conflicts.  In this case, I take
  # the function name (fun[[1]]), and prepend a ".", which follows
  # `rxode2`'s naming convention of un-exported functions.
  #
  # This next evaluation changes the expression function to .rxpois()
  .fun[[1]] <- str2lang(paste0(".", deparse1(fun[[1]])))
  # Since this is still a R expression, you can then evaluate the
  # function .rxpois to produce the proper code:
  eval(.fun)
}

# The above s3 method can be registered in a package or you can use
# the following code to register it in your session:
rxode2::.s3register("rxode2::rxUdfUi", "rxpois")

# This is the function that changes the code as needed
.rxpois <- function(lambda) {
  # The first part of this code tries to change the value into a
  # character.  This handles cases like rxpois(lambda=lam),
  # rxpois(lam), rxpois("lam").  It also tries to evaluate the
  # argument supplied to lambda in case it comes from a different
  # location.
  .lam <- as.character(substitute(lambda))
  .tmp <- try(force(lambda), silent=TRUE)
  if (!inherits(.tmp, "try-error")) {
    if (is.character(.tmp)) {
      .lam <- lambda
    }
  }
  # This part creates a list with the replacement text, in this case
  # it woulb be rxpois(lam) where there is no equals included, as
  # required by `rxode2`:
  list(replace=paste0("rxpois(", .lam, ")"))
}

```

In general the list that the function needs to return can have:

- `$replace` -- The text that will be replaced

- `$before` -- lines that will be placed in the model before the
  current function is found

- `$after` -- lines that are added in the model after the current
  function is found

- `$iniDf` -- the initial estimates `data.frame` for this problem.
  While calling this function, you can retrieve the initial conditions
  currently used parsing you can get the prior value with
  `rxUdfUiIniDf()` and then you can modify it inside the function and
  return the new `data.frame` in this list element.  This allows you
  add/delete initial estimates from the model as well as modify the
  model lines themselves.

- `$uiUseData` -- when `TRUE`, this instructs `rxode2` and
  `nlmixr2est` to re-parse this function in the presence of data, this
  means a bit more function setup will need to be done.

In addition to the `rxUdfUiIniDf()` you can get information about the
parser:

- `rxUdfUiParsing()` returns if the rxode2 ui function is being parsed
  currently (this allows a function to be overloaded as a `udf` for
  calling from `rxode2` as well as a function for modifying the
  model).

- `rxUdfUiNum()` during parsing the function you are calling (in the
  example above `rxpois()` can be called multiple times.  This gives
  the number of the function in the model in order (the first would
  give `1` the second, `2`, etc).  This can be used to create unique
  variables with functions like `rxIntToLetter()` or `rxIntToBase()`.

- `rxUdfUiIniLhs()` which gives the left-handed side of the equation
  where the function is found.  This is also a `R` language object.

- `rxUdfUiData()` which specifies the data that are being used to
  simulate, estimate, etc.

- `rxUdfUiEst()` which gives the estimation/simulation method that is
  being used with the model.  For example, with simulation it would be
  `rxSolve`.

### Using data for `rxode2` ui modification models

When you would like to use data to help inform arguments or
replacements, you will likely need to do the following steps:

- When data are not being processed, you need to put the function in
  an `rxode2` acceptable form, no named arguments, no strings, and
  only numbers or variables in the output.

- The number of arguments of this output needs to be declared in the
  `S3` method by adding the attribute `"nargs"` to method.  For
  example, the built in `linMod()` ui modification function uses:

```r
#' @export
rxUdfUi.linModD <- function(fun) {
  eval(fun)
}
attr(rxUdfUi.linModD, "nargs") <- 3L
```

- Since the number of arguments are specified, you can return the
  `linMod()` in the standardized form and allow it to be re-parsed
  when data are present; to do this use the return `$uiUseData`, you
  could customize further what happens by choosing extra steps for
  different "estimation" methods.
