---
title: "Integrating User Defined Functions into rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## User Defined Functions

```{r setup}
library(rxode2)
```

When defining models you may have wished to write a small R function or make a function integrate into `rxode2` somehow. This article discusses 2 ways to do this:

-   A R-based user function which can be loaded as a simple function or in certain circumstances translated to C to run more efficiently

-   A C function that you define and integrate into code

## R based user functions

A R-based user function is the most convenient to include in the ODE,
but is slower than what you could have done if it was written in `C` ,
`C++` or some other compiled language. This was requested [in
github](https://github.com/nlmixrdevelopment/RxODE/issues/162#issue-568886732)
with an appropriate example; However, I will use a very simple example
here to simply illustrate the concepts.

```{r fExample}
newAbs <- function(x) {
  if (x < 0) {
    -x
  } else {
    x
  }
}

f <- rxode2({
  a <- newAbs(time)
})

e <- et(-10, 10, length.out=40)
```

Now that the ODE has been compiled the R functions will be called
while solving the ODE. Since this is calling R, this forces the
parallization to be turned off since R is single-threaded. It also
takes more time to solve since it is shuttling back and forth between
R and C. Lets see how this very simple function performs:

```{r benmchmark1}
mb1 <- microbenchmark::microbenchmark(withoutC=suppressWarnings(rxSolve(f,e)))

library(ggplot2)
autoplot(mb1) + rxTheme()
```

Not terribly bad, even though it is shuffling between R and C.

You can make it a better by converting the functions to C:

```{r convertC}
# Create C functions automatically with `rxFun()`
rxFun(newAbs)
# Recompile to use the C functions
# Note it would recompile anyway if you didn't do this step,
# it just makes sure that it doesn't recompile every step in
# the benchmark
f <- rxode2({
  a <- newAbs(time)
})

mb2 <- microbenchmark::microbenchmark(withC=rxSolve(f,e, cores=1))

mb <- rbind(mb1, mb2)
autoplot(mb) + rxTheme() + xgxr::xgx_scale_y_log10()
print(mb)
```

The C version is almost twice as fast as the R version. You may have
noticed the conversion also created C versions of the first
derivative. This is done automatically and gives not just C versions
of function, but C versions of the derivatives and registers them with
`rxode2`. This allows the C versions to work with not only `rxode2`
but `nlmixr2` models.

This function was setup in advance to allow this type of conversion. In general the derivatives will be calculated if there is not a `return()` statement in the user defined function. This means simply let R return the last value instead of explictly calling out the `return()`. Many people prefer this method of coding.

Even if there is a `return` function, the function could be converted to `C`. In the github issue, they used a function that would not convert the derivatives:

```{r fConvNoDer}
# Light
f_R <- function(actRad, k_0, a_k) {
  photfac <- a_k * actRad + k_0
  if (photfac > 1) {
    photfac = 1
  }
  return(photfac)
}

rxFun(f_R)
```

While this is still helpful because some functions have early returns,
the `nlmixr2` models requiring derivatives would be calculated be
non-optimized finite differences when this occurs. While this gets
into the internals of `rxode2` and `nlmixr2` you can see this more
easily when calculating the derivatives:

```{r derExample}
rxFromSE("Derivative(f_R(actRad, k_0, a_k),k_0)")
```

Whereas the originally defined function `newAbs()` would use the new
derivatives calculated as well:

```{r derFromC}
rxFromSE("Derivative(newAbs(x),x)")
```

In some circumstances, the conversion to C is not possible, though you
can still use the R function.

There are some requirements for R functions to be integrated into the
rxode2 system:

-    The function must have a set number of arguments, variable
     arguments like `f(…)` are currently not allowed.

-   The function is given each argument as a single number, and the
    function should return a single number

If these requirements are met you can use the R function in
rxode2. Additional requirements for conversion to C include:

-   Any functions that you use within the R function must be
    understood and available to `rxode2`.

    -   Practically speaking if you have `fun2()` which refers to
        `fun1()`, `fun1()` must be changed to C code and available to
        `rxode2` before changing the function `fun2()` to C.

-   The functions can include `if`/`else` assignments or simple return
    statements (either by returning a value or having that value on a
    line by itself). Special R control structures and functions (like
    `for` and `lapply`) cannot be present.

-   The function cannot refer to any package functions

-   As mentioned, if the `return()` statement is present, the
    derivative C functions and `rxode2`'s derivative table is not
    updated.

## C based functions

You can add your own C functions directly into rxode2 as well using
`rxFun()`:

```{r rxFun}
fun <- "
 double fun(double a, double b, double c) {
   return a*a+b*a+c;
 }
" ## C-code for function

rxFun("fun", c("a", "b", "c"), fun)
```

If you wanted you could also use C functions or expressions for the derivatives by using the `rxD()` function:

```{r rxD}
rxD("fun", list(
   function(a, b, c) { # derivative of arg1: a
     paste0("2*", a, "+", b)
   },
   function(a, b, c) { # derivative of arg2: b
     return(a)
   },
   function(a, b, c) { # derivative of arg3: c
     return("0.0")
   }
))
```

Removing the function with `rxRmFun()` will also remove the derivative
table:

```{r rxDrm}
rxRmFun("fun")
```
