---
title: "A Guide to ODE Solvers in rxode2: Selection and Suitability for Pharmacometric Models"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## Introduction

In pharmacometrics, Ordinary Differential Equation (ODE) solvers are
the computational engines driving model simulations. Whether you are
running a simple one-compartment PK model, a complex
physiologically-based pharmacokinetic (PBPK) model, or a stiff
quantitative systems pharmacology (QSP) model, selecting the right
solver can drastically impact simulation speed, accuracy, and
numerical stability.

`rxode2` provides a rich set of 18 ODE solving methods, combining
classic solvers (like LSODA and DOP853) with modern adaptive,
fixed-step, implicit, and symplectic steppers from the C++
`Boost.odeint` library.

This guide provides an overview of the solvers available in `rxode2`,
explains key numerical concepts (stiffness, dense output, symplectic
integration), and recommends when to use each method.

---

## Summary of ODE Solvers in `rxode2`

The table below summarizes the solvers available in `rxode2` and their properties:

| Method | Type | Stiff-Safe | Dense Output | Key Options | Recommended Use Case |
|---|---|---|---|---|---|
| **`"liblsoda"`** *(Default)* | Adaptive (BDF / Adams) | Yes | No | `atol`, `rtol`, `maxsteps` | **Default choice for general PK/PD.** Thread-parallel across subjects. |
| **`"lsoda"`** | Adaptive (BDF / Adams) | Yes | No | `atol`, `rtol`, `maxsteps` | Fortran-based LSODA. Single-threaded. Supports user Jacobians. |
| **`"dop853"`** | Explicit Runge-Kutta 8(5,3) | No | Yes | `atol`, `rtol`, `dense` | High-order solver. Excellent for **smooth, non-stiff models** with rich sampling. |
| **`"rkf78"`** | Explicit Runge-Kutta-Fehlberg 7(8) | No | No | `atol`, `rtol`, `hmin` (initial dt) | Non-stiff models where high-accuracy adaptive step-sizes are desired. |
| **`"ck54"`** | Explicit Cash-Karp 5(4) | No | No | `atol`, `rtol`, `hmin` (initial dt) | Non-stiff models with moderate accuracy requirements. |
| **`"dop5"`** | Explicit Dormand-Prince 5(4) | No | Yes | `atol`, `rtol`, `dense` | Non-stiff models. Fast alternative to `dop853` for moderate accuracy. |
| **`"bs"`** | Bulirsch-Stoer | No | Yes | `atol`, `rtol`, `dense` | Smooth, non-stiff systems where very high precision is required. |
| **`"ros4"`** | Implicit Rosenbrock 4 | Yes | Yes | `atol`, `rtol`, `dense` | **Stiff systems** (e.g., target-mediated drug disposition). Supports dense output. |
| **`"iem"`** | Implicit Euler | Yes | No | `hmin` (fixed dt), `maxsteps` | Stiff systems where a **fixed integration step** is strictly required. |
| **`"rk4"`** | Explicit Runge-Kutta 4 | No | No | `hmin` (fixed dt), `maxsteps` | Fixed-step solver for debugging or comparing with historical code. |
| **`"em"`** | Explicit Euler | No | No | `hmin` (fixed dt), `maxsteps` | Simple 1st-order solver. Mainly for educational use or discrete comparisons. |
| **`"ab"`** | Adams-Bashforth | No | No | `hmin` (fixed dt), `order` (1-8) | Multi-step explicit solver. Useful for smooth non-stiff dynamics. |
| **`"abm"`** | Adams-Bashforth-Moulton | No | No | `hmin` (fixed dt), `order` (1-8) | Predictor-corrector multi-step explicit solver. |
| **`"mm"`** | Modified Midpoint | No | No | `hmin` (fixed dt), `order` (intermediate steps) | Explicit fixed-step mid-point formulation used internally by Bulirsch-Stoer. |
| **`"sem"`** | Symplectic Euler | No | No | `hmin` (fixed dt) | Hamiltonian mechanical systems. Padded for general/odd dimensions. |
| **`"sb3a"`** | Symplectic RKN SB3A McLachlan | No | No | `hmin` (fixed dt) | Hamiltonian mechanical systems. High-order Runge-Kutta-Nyström. |
| **`"sb3am4"`** | Symplectic RKN SB3A M4 McLachlan | No | No | `hmin` (fixed dt) | Hamiltonian mechanical systems. Alternative high-order RKN stepper. |
| **`"vv"`** | Velocity Verlet | No | No | `hmin` (fixed dt) | Hamiltonian mechanical systems. Second-order Verlet integrator. |

---

## Stiffness in Pharmacometrics

The choice of ODE solver should begin with assessing if your model is **stiff**.

### What is stiffness?

A system of differential equations is stiff if the solution has
components that vary at extremely different timescales. In
pharmacometrics, this commonly occurs when:

1. **Target-Mediated Drug Disposition (TMDD)**: The binding rate
   constant ($k_{on}$) is extremely fast, while elimination and
   distribution occur over hours or days.

2. **Systemic Feedback Loops**: Models with rapid signaling cascades
   (e.g. receptor phosphorylation) coupled with slow receptor
   synthesis/degradation.

### Stiff vs. Non-Stiff Solvers

- **Explicit solvers** (like `dop853`, `rk4`, `dop5`, `bs`, `em`)
  evaluate the derivatives using only the state at the current
  time. On stiff systems, they are forced to take micro-steps to
  prevent numerical instability, causing the solver to stall, hang, or
  fail.

- **Implicit solvers** (like `liblsoda`, `ros4`, `iem`) solve an
  implicit equation at each step, requiring the Jacobian matrix of the
  system. While each step is more expensive to calculate, they can
  take much larger steps through stiff regions.

For stiff systems, **`"liblsoda"`** (adaptive automatic switching) or
**`"ros4"`** (implicit Rosenbrock 4) should always be preferred; note
that the Rosenbrock 4 method is often slower and requires a symbolic
Jacobian calculation step (which is automatically done) which at times
may fail.

---

## Understanding Key Solver Options

### 1. Fixed-Step vs. Adaptive-Step

- **Adaptive Solvers** (e.g., `liblsoda`, `dop853`, `dop5`, `bs`,
  `ros4`, `rkf78`, `ck54`) dynamically adjust the step size ($dt$) to
  maintain the local integration error below the specified tolerances
  (`atol` and `rtol`). These are standard for almost all
  pharmacometric applications.

- **Fixed-Step Solvers** (e.g., `rk4`, `em`, `ab`, `abm`, `iem`,
  `sem`, `sb3a`, `sb3am4`, `vv`, `mm`) integrate using a step size
  strictly governed by the **`hmin`** parameter. If `hmin = 0`
  (default), `rxode2` assigns a default of `0.01` for `rk4` and
  `0.0001` for other fixed-step methods. If the number of steps
  exceeds `maxsteps`, the step size is automatically scaled up to
  prevent hangs.

### 2. Dense Output (`dense = TRUE`)

Normally, when integrating to an observation grid, adaptive solvers
must stop exactly at each observation time, which resets the step-size
selection algorithm and slows down the integration.

When **`dense = TRUE`** is set for supported solvers (`dop853`,
`dop5`, `bs`, `ros4`), the solver integrates across the interval at
its own optimal step size. It then uses cheap, continuous
interpolation polynomials to calculate the states at the requested
observation times. This is highly recommended for dense simulation
grids (e.g. generating smooth VPC lines) and can lead to a 10% to 20%
speedup.

### 3. The `order` Option

- For Adams-Bashforth (`"ab"`) and Adams-Bashforth-Moulton (`"abm"`),
  the `order` option (1 to 8, default 5) sets the number of historical
  steps used to extrapolate the next step.

- For the Modified Midpoint (`"mm"`) solver, `order` (default 5, must
  be `>= 1`) configures the number of intermediate sub-steps taken
  inside each interval.

---

## When to Use Symplectic Solvers

`rxode2` includes four symplectic integrators: `"sem"`, `"sb3a"`,
`"sb3am4"`, and `"vv"`.

Symplectic solvers are designed to conserve energy and phase-space
volume in Hamiltonian systems (conservative systems without friction
or dissipation, such as a harmonic oscillator or planetary orbits).

### Suitability for Pharmacometrics

Most pharmacometric compartment models represent physical mass
transfer with **dissipation** (e.g., drug clearance, decay,
metabolism). Because energy or mass is constantly leaving the system,
these are not Hamiltonian systems, and standard adaptive solvers (like
`liblsoda`) are mathematically much more appropriate.

However, symplectic solvers are highly useful in specific contexts:

- **Binding Dynamics Analogies**: Modeling conservative physical
  sub-systems or simulating idealized periodic feedback loops where
  long-term stability without numerical dampening is crucial.

- **Dimensionality Padding**: `rxode2` automatically handles general
  and odd-dimensional systems (which do not naturally split into
  coordinate-momentum pairs) by applying zero-derivative padding
  variables, enabling these solvers to integrate any arbitrary ODE
  model.

---

## Recommended Workflow for Solver Selection

1. **Start with `"liblsoda"` (default)**: It is robust, handles
   stiffness automatically, and scales efficiently across multiple CPU
   cores using OpenMP.

2. **If the system is non-stiff and you have a dense observation
   grid**: Try **`method = "dop853"`** or **`method = "dop5"`** with
   **`dense = TRUE`** to speed up simulations.

3. **If the system is highly stiff (e.g. complex TMDD/QSP)**: Use
   **`method = "ros4"`** (implicit Rosenbrock 4). If the model does
   not contain a manual analytical Jacobian, `rxode2` will
   automatically generate one for you via SymEngine when compiling.

---

## Session Information

```{r session-info}
sessionInfo()
```
