---
title: "Choosing ODE solvers in rxode2"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Choosing ODE solvers in rxode2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why solver choice matters in pharmacometrics

In pharmacometrics, the "best" ODE solver is rarely the highest-order
method on paper.  What usually matters more is whether the model is
stiff, whether dosing creates sharp discontinuities, whether output is
requested on a dense sampling grid, and whether you need reproducible
fixed-step behavior for sensitivity or comparison studies.

`rxode2` exposes a broad solver set because different model classes
benefit from different tradeoffs:

- classic PK or PK/PD models with mild stiffness often work well with
  `liblsoda`, `dop853`, or `dop54`
- TMDD, mechanistic PD, QSP, and transit-heavy systems often need
  stiffness-aware methods such as `liblsoda`, `ros4`, `ros43`,
  `sdirk43`, or `cvode`
- fixed-step methods such as `rk4`, `backwardEuler`, `gauss6`, or
  `radauiia5` are useful when you want controlled stepping between
  events, method-comparison experiments, or deterministic solver
  behavior for debugging

The `method` argument to `rxSolve()` selects the solver.  The four
historic and therefore most commonly used methods are:

| `method` | Algorithm | Best for |
|---|---|---|
| `"liblsoda"` | Thread-safe C LSODA | **Default.** Parallel multi-subject simulations |
| `"lsoda"` | FORTRAN LSODA | Single-threaded; numerical comparison with deSolve |
| `"dop853"` | Dormand--Prince 8th-order RK | Smooth non-stiff ODEs; dense output |
| `"indLin"` | Inductive linearization + matrix exponential | Linear/mildly-nonlinear systems; estimation loops |

All methods accept NONMEM-format event tables and handle dosing
discontinuities automatically.  The sections below first cover these
four methods with code examples, then give a practical pharmacometric
view and a complete method inventory split by stiffness and adaptivity.

## The four main methods

### `liblsoda` (default)

`liblsoda` is a thread-safe C rewrite of the classic LSODA algorithm.
LSODA (Livermore Solver for Ordinary Differential Equations with
Automatic Switching) automatically detects stiffness and switches
between Adams (non-stiff) and BDF (stiff) integration formulas.

Because it is thread-safe, `liblsoda` runs in parallel across subjects
via OpenMP when `cores > 1`:

```{r}
library(rxode2)

mod <- rxode2({
  d/dt(depot)  <- -ka * depot
  d/dt(center) <- ka * depot - (CL / V) * center
  Cp           <- center / V
})

set.seed(42)
ev <- et(amt = 100) |> et(0:72)
theta <- c(ka = 1, CL = 0.1, V = 10)
omega <- lotri(eta.ka ~ 0.04, eta.cl ~ 0.09, eta.v ~ 0.04)

sim <- rxSolve(mod, theta, ev,
               omega  = omega,
               nSub   = 1000,
               method = "liblsoda",
               cores  = 4)
```

Tolerance is controlled with `atol` and `rtol` (default `1e-8` and
`1e-6`).  For estimation, `nlmixr2` uses `liblsoda` internally because
it supports sensitivity equations.

### `lsoda`

The original FORTRAN LSODA from Hindmarsh (ODEPACK).  Not thread-safe,
so it cannot exploit OpenMP for parallel subjects.  For single-subject
use or when evaluating numerical agreement with deSolve (which also
wraps this FORTRAN code), `lsoda` is appropriate:

```{r}
sol <- rxSolve(mod, theta, ev, method = "lsoda")
```

### `dop853`

An explicit 8th-order Dormand--Prince Runge--Kutta method with 5th- and
3rd-order error estimates for step-size control.  It does not switch to
a stiff algorithm, so it is unsuitable for stiff systems.  It is,
however, very efficient for smooth problems and provides dense output
(accurate solutions at any requested time point, not just the grid).

```{r}
sol <- rxSolve(mod, theta, ev, method = "dop853")
```

For non-stiff problems it is often faster than LSODA because it avoids
the stiffness-detection overhead.

### `indLin`

Inductive linearization with matrix exponential integration.  Rather
than advancing one time step at a time, the ODE is iteratively
converted to a linear time-varying (LTV) system and solved exactly with
the matrix exponential.  Details and usage are covered in the companion
article
[Matrix Exponentials and Inductive Linearization in rxode2](rxode2-ind-lin.html).

```{r}
mmPK <- rxode2(
  {
    d/dt(depot)  <- -ka * depot
    Cp           <- center / V
    d/dt(center) <- ka * depot - Vmax * Cp / (Km + Cp)
  },
  indLin = TRUE
)

sol <- rxSolve(mmPK, c(ka = 1, Km = 0.5, Vmax = 0.2, V = 1),
               ev, method = "indLin")
```

## Stiffness and Jacobians

Stiff systems contain processes operating on widely different time
scales.  The default `liblsoda` handles stiffness automatically.  For
very stiff problems you can improve performance and stability by
providing an analytic Jacobian with `df(state)/dy(variable)`:

```{r}
vanPol <- rxode2({
  mu <- 1000
  d/dt(y)  <- dy
  d/dt(dy) <- mu * (1 - y^2) * dy - y
  df(y)/dy(dy)  <- 1
  df(dy)/dy(y)  <- -2 * dy * mu * y - 1
  df(dy)/dy(dy) <- mu * (1 - y^2)
  y(0)  <- 2
  dy(0) <- 0
})

sol <- rxSolve(vanPol, et(seq(0, 20, length.out = 500)), method = "lsoda")
```

rxode2 can also compute the Jacobian symbolically via the `symengine`
CAS using `calcJac = TRUE`:

```{r}
vanPol_auto <- rxode2(vanPol$symengineModelNoPrune, calcJac = TRUE)
```

## Practical pharmacometric starting points

| Modeling situation | Good rxode2 starting points | Why | Familiar names in other ecosystems |
|---|---|---|---|
| General PK/PD work, stiffness uncertain | `liblsoda`, `lsoda` | Automatically switches between non-stiff and stiff behavior; very good default for mixed event schedules | ODEPACK/deSolve `lsoda`; Julia `lsoda()`; MATLAB has no direct LSODA wrapper and usually splits this choice between `ode45` and `ode15s`; Phoenix NLME closest match is `Auto-Detect` (LSODA) |
| Smooth, mostly non-stiff PK/PD with many requested output times | `dop853`, `dop5`, `dop54`, `t54` | High-order explicit methods; efficient on smooth concentration-time problems | `dop54`/`dp54` is closest to MATLAB `ode45`; `t54` is the same method as Julia `Tsit5()`; `dop853` matches SciPy `DOP853`; Julia `DP5()` maps to `dop5`/`dop54`, `Vern7()` maps to `vern76`; Phoenix NLME analogs are `Non-Stiff (DOPRI5)` and `Non-Stiff (DVERK)` |
| Moderate stiffness: TMDD, indirect response with fast turnover, enzyme models | `ros4`, `ros43`, `sdirk43` | Stiff-aware one-step methods that often behave well across dosing discontinuities | MATLAB users can think `ode23s` / `ode15s`; Julia analogs are `Rosenbrock23()`, `Rodas4P()`, `Rodas5P()`; Phoenix NLME users usually compare against `Auto-Detect` (LSODA) or `Stiff` (LSODE/BDF) |
| Clearly stiff or larger systems where linear solver choice matters | `cvode` | CVODE gives Adams/BDF-style behavior with configurable linear solvers; supports `dense=TRUE` for efficient dense output using CVODE's native `CVodeGetDky` interpolation | Julia `CVODE_Adams()` / `CVODE_BDF()` in Sundials; MATLAB users usually compare against `ode15s`; Phoenix NLME analog is `Stiff` (LSODE/BDF) |
| Stiffness uncertain, want fast non-stiff speed with automatic fallback | `"t54+ros4"`, `"t54+ros43"`, `"vern76+ros43"`, `"vern76+sdirk43"` | AutoSwitch: starts with the non-stiff solver; detects stiffness mid-solve and retries the interval with the implicit solver; switches back if the region becomes non-stiff again | Closest to Julia `AutoTsit5(RosShamp4())`/`AutoVern7(Rodas4P())`; note Julia's `Rosenbrock23()` has no exact rxode2 match (use `ros4` as the lowest-order Rosenbrock-family secondary); no direct MATLAB or NONMEM analog |
| Fixed-step comparison runs or debugging around dose/event boundaries | `rk4`, `trapz`, `backwardEuler`, `gauss6`, `iiic6`, `radauiia5`, `ros6` | Forces a controlled step size and makes solver-to-solver comparisons easier | Julia has direct families such as `Heun()`, `ImplicitEuler()`, `GaussLegendre`, `LobattoIIIC`, `RadauIIA5()`; MATLAB has fewer built-in fixed-step counterparts |
| Mostly linear compartment structure with sparse nonlinear pieces | `indLin` | Uses the specialized inductive linearization pathway instead of a general ODE stepper | Best compared with matrix-exponential or linearization workflows, not with `ode45`/`ode15s` directly; Phoenix NLME analogs are `Matrix Exponent` and `Exponent Higham` |

## How rxode2 solvers split by stiffness and adaptivity

For pharmacometric work, the most useful split is:

1. adaptive and stiffness-aware or stiff-suitable

2. fixed-step and stiff-suitable

3. adaptive and primarily non-stiff

4. fixed-step and primarily non-stiff

5. pharmacometric special-case linearization

Some methods blur the boundary.  For example, `liblsoda` and `lsoda`
switch automatically between non-stiff and stiff formulas, and `cvode`
can behave more like a non-stiff Adams method or a stiff BDF method
depending on the problem.  In practice, pharmacometric users usually
reach for them when stiffness is possible.

### Adaptive, stiffness-aware, or stiff-suitable methods

These are the first methods to consider when a model has fast and slow
time scales, strong feedback, binding kinetics, or many hidden
turnover processes.

- `liblsoda`, `lsoda`: automatic stiffness detection and switching;
  still the safest "start here" choice for many PK/PD models

- `cvode`: Sundials CVODE backend; useful when you need a modern stiff
  solver with linear solver control (`dense`, `band`, `gmres`,
  `bicgstab`, `tfqmr`).  At present, `band` is kept for compatibility
  and aliases to the dense path.  `cvode` supports `dense=TRUE`: the
  BDF or Adams history is preserved across consecutive observation
  points by using `CV_ONE_STEP` + CVODE's native `CVodeGetDky`
  interpolation rather than reinitializing at each output time.  This
  is particularly beneficial for BDF, which maintains a multi-step
  history polynomial and a Jacobian factorization that would otherwise
  be discarded on every reinitialize.

- `ros4`, `ros43`: adaptive Rosenbrock methods for stiff or moderately
  stiff systems

- `sdirk43`: adaptive singly diagonally implicit Runge-Kutta method

These are closest in spirit to MATLAB `ode23s` / `ode15s` and to Julia
methods such as `Rosenbrock23()`, `Rodas4P()`, `Rodas5P()`, and the
Sundials `CVODE_*` wrappers.

### AutoSwitch composite methods

AutoSwitch methods let you combine a fast non-stiff primary solver with a
stiff fallback solver using a `"primary+stiff"` string:

```{r}
rxSolve(model, events, params, method = "t54+ros43")
rxSolve(model, events, params, method = "vern76+ros43")
rxSolve(model, events, params, method = "dop853+sdirk43")
```

Within each ODE interval (the segment between events/observations),
`rxode2` runs the primary (non-stiff) solver first.  If that probe
reports stiffness -- because the non-stiff solver failed (its adaptive
step size collapsed) or because DOP853's built-in dominant-eigenvalue
stiffness estimator fired -- the interval is re-solved from its start
with the stiff secondary.  This is a *reactive* scheme: rather than
computing a separate stiffness estimate at the composite level, it
delegates detection to the primary's own per-step estimator and to solve
success/failure.  Once consecutive stiff detections cross a threshold the
solver stops probing and uses the stiff method directly for that subject;
it returns to the primary after enough consecutive non-stiff intervals.

(Earlier versions of `rxode2` used a proactive estimate -- the Gershgorin
spectral radius of the Jacobian scaled by the interval length -- but on
the long inter-dose intervals typical of pharmacometric models that
over-estimated stiffness and switched too eagerly, so it was replaced by
the reactive scheme above.)

#### Why use AutoSwitch instead of `liblsoda`?

`liblsoda` already switches between Adams and BDF internally, so it is
already a good general-purpose default.  AutoSwitch complements it:

- **Speed on clearly non-stiff regions.**  The non-stiff primaries (`t54`,
  `vern76`, `dop853`, `bs32`, etc.) are typically faster than the Adams
  method inside `liblsoda` on smooth, non-stiff intervals because they
  use higher-order explicit stages optimized for accuracy per function
  evaluation rather than for very broad order-adaptation.

- **Explicit method choice.**  You decide which non-stiff solver to use as
  the primary (any thread-safe non-stiff method) and which stiff solver to
  use as the secondary (any implicit method except `cvode` and `lsoda`).
  This is useful when you already know from a preliminary solve that most
  of the dose interval is non-stiff and only specific regions -- typically
  shortly after dosing in TMDD or competitive binding models -- require
  implicit stepping.

- **Matches Julia's ecosystem.**  Pharmacometric models built and
  tested in Julia with `AutoTsit5(Rosenbrock23())` or
  `AutoVern7(Rodas5P())` can be matched closely by `"t54+ros43"` or
  `"vern76+ros43"` in `rxode2` (though the switching methods are different).

#### When `liblsoda` is likely better

- When you genuinely do not know whether any part of the model is stiff and
  you want a proven general-purpose default without tuning.
- When the model changes stiffness many times per dosing interval, making
  the switching overhead significant.
- When reproducibility across platforms is more important than per-method
  speed.

#### Choosing a stiff secondary: `ros4`, `ros43`, or `sdirk43`?

rxode2 has three main adaptive stiff secondaries:

- **`ros4`** -- Boost's Kaps-Rentrop 4th-order Rosenbrock W-method.  A-stable.
  This is the lightest-overhead Rosenbrock option and the closest available
  analog to Julia's lower-order Rosenbrock family.  rxode2 does **not**
  have a true 2nd/3rd-order Rosenbrock matching Julia's `Rosenbrock23()`;
  `ros4` is the lowest-order adaptive Rosenbrock-family method available.
- **`ros43`** -- libode's GRK4A, a 4th-order A-stable Rosenbrock-Kaps-Rentrop
  method with an embedded 3rd-order error estimate.  More sophisticated than
  `ros4`; analogous to Julia's `Rodas4P()`.
- **`sdirk43`** -- SDIRK 4(3), L-stable.  More robust for severely stiff
  problems (e.g., stiff oscillations, equilibrium reactions) where A-stability
  alone is not enough.

For AutoSwitch pairs, prefer `ros4` when you want low switching overhead,
`ros43` when you want reliable moderate-stiffness handling, and `sdirk43`
when the stiff regions are severe.

#### Recommended AutoSwitch pairs

| `rxode2` AutoSwitch method | Closest Julia analog | Character |
|---|---|---|
| `"t54+ros4"` | `AutoTsit5(RosShamp4())` | Tsit5 primary, lightweight Rosenbrock W-method secondary; closest to `AutoTsit5(Rosenbrock23())` but `ros4` is 4th order -- no exact 2nd-order Rosenbrock exists in rxode2 |
| `"t54+ros43"` | `AutoTsit5(Rodas4P())` | Tsit5 primary, GRK4A secondary; more robust stiff handling than `ros4` |
| `"t54+sdirk43"` | `AutoTsit5(SDIRK43())` | L-stable stiff secondary; best for severely stiff problems |
| `"vern65+ros4"` | -- | Verner 6(5) primary with lightweight stiff secondary |
| `"vern65+ros43"` | -- | Verner 6(5) primary; good accuracy-to-speed balance |
| `"vern76+ros4"` | `AutoVern7(RosShamp4())` | Higher-order primary with lighter stiff secondary |
| `"vern76+ros43"` | `AutoVern7(Rodas4P())` | Higher-order primary; efficient on smooth, low-error-tolerance simulations |
| `"vern76+sdirk43"` | `AutoVern7(SDIRK43())` | Higher-order primary with L-stable stiff backup |
| `"vern98+ros4"` | `AutoVern9(RosShamp4())` | Highest-order primary with lighter stiff secondary |
| `"vern98+ros43"` | `AutoVern9(Rodas4P())` | Highest-order primary; best on demanding tolerances |
| `"vern98+sdirk43"` | `AutoVern9(SDIRK43())` | High-order primary with L-stable stiff backup |
| `"dop853+ros4"` | -- | DOP853 primary; lower-overhead stiff fallback; supports `dense=TRUE` |
| `"dop853+ros43"` | -- | DOP853 primary; more robust stiff handling; supports `dense=TRUE` |
| `"bs32+ros4"` | `AutoBS3(RosShamp4())` | Low-order primary; useful for rough problems or coarser tolerances |

**Note on Julia's `Rosenbrock23()`:** Julia's `Rosenbrock23()` is a 2nd/3rd-order
Rosenbrock W-method (Shampine, 1982).  rxode2 does not have a 2nd/3rd-order
adaptive Rosenbrock.  The `ros4` method (Boost's Kaps-Rentrop 4th-order W-method)
is the lowest-order adaptive Rosenbrock-family secondary available, making
`"t54+ros4"` the closest practical analog to `AutoTsit5(Rosenbrock23())`.

#### Tuning AutoSwitch behavior

Six parameters control the switching logic, all passed directly to
`rxSolve()` or `rxControl()`:

| Parameter | Default | Meaning |
|---|---|---|
| `autoSwitchNonstifftol` | `0.9` | Stiffness ratio threshold (in non-stiff mode) above which the interval is considered stiff and retried |
| `autoSwitchStifftol` | `0.9` | Non-stiffness ratio threshold (in stiff mode) below which the switch-back counter is incremented |
| `autoSwitchMaxStiff` | `10L` | Consecutive stiff-detected intervals before permanently switching to the stiff solver |
| `autoSwitchMaxNonstiff` | `3L` | Consecutive non-stiff intervals (while in stiff mode) before switching back to the primary |
| `autoSwitchDtfac` | `2.0` | Factor applied to the suggested step size when switching to stiff; divided when switching back |
| `autoSwitchStiffFirst` | `FALSE` | Start with the stiff solver instead of the non-stiff primary |

For pharmacometric applications:

- The defaults work well for standard PK/PD and TMDD models.
- Setting `autoSwitchStiffFirst = TRUE` is useful when you expect the
  early post-dose period to be stiff (e.g., rapid binding or target
  saturation) and want to avoid the overhead of a failed non-stiff attempt
  on the first few intervals.
- Lowering `autoSwitchNonstifftol` (e.g., `0.5`) makes switching more
  aggressive; raising it toward `1.0` makes the solver more willing to
  stay with the non-stiff primary even in borderline intervals.

#### How rxode2 AutoSwitch compares to Julia's `AutoSwitch`

`rxode2`'s composite methods and Julia's `AutoTsit5`/`AutoVern7`/...
wrappers (the `AutoSwitch` algorithm in `OrdinaryDiffEq.jl`) share the
*same underlying stiffness signal* but apply it differently.

**The shared signal.**  Julia's `is_stiff` test computes a stiffness
ratio `stiffness = |eigen_est * dt / alg_stability_size(alg)|` and
declares the step stiff when this exceeds a tolerance (`nonstifftol`
= `stifftol` = `0.9` by default).  Here `eigen_est` is the stepper's
cheap estimate of the dominant Jacobian eigenvalue (for explicit
Runge-Kutta methods, Hairer's estimate from the last stages), `dt` is
the **actual step size**, and `stability_size` is the method's
stability-region radius.  DOP853's built-in stiffness test computes
exactly this quantity (Hairer & Wanner's `nstiff` heuristic), so when
`rxode2` probes an interval with `dop853`, the primary is using Julia's
`eigen_est * dt / stability_size` criterion internally.

**The differences.**

| | Julia `AutoSwitch` | rxode2 composite |
|---|---|---|
| Granularity | per accepted **step** | per ODE **interval** (event-to-event segment) |
| Decision style | **proactive** -- `is_stiff` chooses the algorithm for the *next* step | **reactive** -- run the primary, re-solve the interval with the stiff method only if it reports stiffness or fails |
| Stiffness source | the integrator's `eigen_est` (every stepper supplies one) | the primary's own per-step estimator (e.g. DOP853's `nstiff`) **or** an outright solve failure (step-size collapse) |
| Hysteresis | `maxstiffstep` / `maxnonstiffstep` / `dtfac` / `stiffalgfirst` | `autoSwitchMaxStiff` / `autoSwitchMaxNonstiff` / `autoSwitchDtfac` / `autoSwitchStiffFirst` (same defaults: `10` / `3` / `2` / `FALSE`) |

The practical upshot: because `rxode2` solves a whole inter-dose interval
as one unit, it does not re-pick the algorithm at every internal step the
way Julia does.  Instead it lets the non-stiff primary run (and detect
stiffness with the same Hairer estimate, or simply fail on a stiff
region), then re-solves that interval with the stiff secondary and uses a
short hysteresis so a persistently stiff subject stops paying for the
non-stiff probe.  A non-stiff primary that has no internal eigenvalue
estimate still triggers a switch through its step-size collapse, which is
why the reactive scheme works for every primary, not only `dop853`.

### Delay differential equations (DDEs)

`rxode2` solves delay differential equations through the same dense-output
machinery the composite methods use.  Writing `delay(state, T)` inside a
`d/dt()` equation evaluates an ODE compartment at the past time `t - T`
(Monolix `delay()` semantics):

```{r eval=FALSE}
dde <- rxode2({
  y(0) <- 1
  d/dt(y) <- -delay(y, 1)        # y'(t) = -y(t - 1)
})
rxSolve(dde, et(seq(0, 10, by = 0.1)))
```

The delayed value is interpolated from the solver's recorded dense
history, so it is obtained to the full order of the method.  Because this
requires dense output, a model that uses `delay()` is automatically
solved on a dense path and the default method becomes the AutoSwitch
composite `"dop853+ros4"` (which records the dop853 and ros4 dense
histories together and switches between them per segment).  `"dop853"`
and `"ros4"` may also be requested explicitly; non-dense methods raise an
error.  The step size is capped to the smallest delay so short delays
stay accurate, and forward sensitivities are generated for `delay()`
models so they can be fit with `nlmixr2`'s FOCEi.

For the full treatment -- worked examples, how the dense history is
recorded and interpolated, mixed delay + ODE models, and estimation --
see the dedicated article
[Delay differential equations in rxode2](rxode2-delay-differential-equations.html).

### Fixed-step, stiff-suitable, or implicit methods

These are useful when you want implicit stepping but you also want to
control the internal step size directly through `hmin`.

- `iem`: implicit Euler-style stepping for fixed-step studies
- `ros6`: fixed-step Rosenbrock method
- `backwardEuler`: first-order implicit baseline
- `gauss6`: fixed-step Gauss-Legendre IRK method
- `iiic6`: fixed-step Lobatto IIIC method
- `radauiia5`: fixed-step Radau IIA method
- `geng5`: fixed-step implicit method from the libode family

For pharmacometric model comparison work, these methods are especially
useful when you want to answer questions like "is the behavior caused
by stiffness or by adaptive step-size control?" rather than just "which
solver is fastest?"

### Adaptive, primarily non-stiff methods

These are the main choices for smooth PK, exposure-response, and
simulation workflows where the system is not strongly stiff between
events.

#### Most familiar adaptive non-stiff choices

- `dop853`: a strong default for smooth, non-stiff pharmacometric
  systems with dense sampling grids
- `dop5`: adaptive Dormand-Prince 5th-order family
- `dop54`, `dp54`: the closest match in naming and behavior to
  MATLAB `ode45`
- `f32`: closest in role to MATLAB `ode23`
- `bs`: adaptive Bogacki-Shampine family
- `vern65`, `v65e`; `vern76`, `v76e`; `vern98`, `v98e`: Verner
  families, conceptually closest to Julia's `Vern6`, `Vern7`, and
  `Vern9`

The full list of accepted names for these families appears in the
[Complete accepted solver names](#complete-accepted-solver-names-in-rxode2) section.

### Fixed-step, primarily non-stiff methods

These methods use `hmin` as the actual fixed step size (`rk4`, `trapz`,
`ab`, `abm`, and related methods -- see the complete names section for
the full list).  They are usually chosen for comparison studies,
controlled co-simulation, or educational/debugging purposes rather than
for routine production solves.

### Special pharmacometric pathway: `indLin`

`indLin` (see [The four main methods](#the-four-main-methods)) is worth
trying when:

- the model is close to a linear compartment system
- the nonlinear part is limited or localized
- event handling dominates cost more than the actual ODE algebra
- you want speedups for repeated simulation rather than a new solver
  family for a fundamentally nonlinear/stiff problem

## Adaptive versus fixed-step behavior in rxode2

In `rxode2`, this is the practical rule of thumb:

- for fixed-step methods, `hmin` is the step size you are asking the
  solver to use
- for adaptive methods, `hmin` is only the minimum step size, and
  `hmax` becomes important when events or dense output matter
- for dense output workflows, pass `dense=TRUE` to `rxControl()` or
  `rxSolve()`; the methods that support it are `dop853`, `dop5`,
  `bs`, `ros4`, and `cvode`

For pharmacometric workflows such as VPCs, trial simulation, and
posterior predictive checks, dense output can matter as much as the
integrator itself because it changes how often the expensive full
solver state has to be reconstructed at requested sample times.

## Solver tolerance

All numerical solvers respond to `atol` and `rtol`:

```{r}
sol <- rxSolve(mod, theta, ev,
               method = "liblsoda",
               atol   = 1e-10,
               rtol   = 1e-8)
```

Tighter tolerances increase accuracy at the cost of more function
evaluations.  The defaults (`atol = 1e-8`, `rtol = 1e-6`) match
NONMEM's recommended values for ADVAN8/ADVAN13 and are tighter than
MATLAB's `ode45` defaults (`RelTol = 1e-3`, `AbsTol = 1e-6`).

## MATLAB naming guide

MATLAB's ODE suite uses a small number of solver names.  The table below gives
the nearest `rxode2` equivalent for each.

| MATLAB solver | Closest rxode2 method(s) | Notes |
|---|---|---|
| `ode23` | `bs32` | Bogacki-Shampine 3(2) pair; same family |
| `ode45` | `dop54`, `dp54` | Dormand-Prince 5(4) FSAL; same algorithm as MATLAB |
| `ode78` | `f78` | Fehlberg 7(8) via Boost; same family |
| `ode89` | `v89`, `f89` | Verner/Fehlberg 8(9) pairs from rklib |
| `ode113` | `ab`, `abm` | Adams-Bashforth-Moulton predictor-corrector (variable order); `liblsoda` covers Adams/BDF switching |
| `ode15s` | `cvode`, `liblsoda` | BDF stiff solver; `cvode` is the closest explicit match |
| `ode23s` | `ros43` | Kaps-Rentrop Rosenbrock 4(3); structurally the same family as MATLAB `ode23s` |
| `ode23t` | `trapz` | Trapezoidal rule, fixed-step |
| `ode23tb` | `sdirk43` | SDIRK / TR-BDF2 stiff pair |
| `ode15i` | (none) | DAE solver; reformulate algebraic constraints as explicit ODEs before using rxode2 |

### Key differences from MATLAB

- **Parallel subjects**: rxode2 runs thousands of subjects in parallel
  via OpenMP.  MATLAB requires explicit vectorization or Parallel
  Computing Toolbox.
- **Compiled model**: rxode2 compiles the ODE to C once.  MATLAB's `ode45`
  calls an R/MATLAB function on every step unless you hand-code C
  with `mex`.
- **Event handling**: rxode2 accepts NONMEM-format event tables
  (infusions, steady state, lag times) natively.  MATLAB requires
  manual event functions.
- **Order**: rxode2's `dop853` uses 8th-order steps; `ode45` uses 5th.
  `dop853` is typically more efficient per function evaluation on
  smooth problems.
- **IndLin vs `ode45`**: Sharif et al. (2022) showed that IndLin with
  adaptive step size was approximately six times faster than `ode45` on
  a Michaelis--Menten example with comparable parameter estimate accuracy.

## Julia naming guide

Julia's DifferentialEquations.jl ecosystem has a large solver library with
explicit names for many classical methods.  The table below maps the most
commonly used Julia solvers to their rxode2 equivalents.

| Julia (DifferentialEquations.jl) | Closest rxode2 method(s) | Notes |
|---|---|---|
| `Euler()` | `euler` | Fixed-step explicit Euler, 1st-order (rklib) |
| `Midpoint()` | `midpoint` | Fixed-step explicit midpoint, 2nd-order (rklib) |
| `Heun()` | `heun`, `trapz` | Fixed-step Heun / explicit trapezoid, 2nd-order |
| `RK4()` | `rk4` | Classical 4th-order Runge-Kutta (Boost) |
| `SSPRK33()` | `ssp3`, `ssp33` | Strong Stability-Preserving RK3 |
| `SSPRK43()` | `ssp43` | SSP 4(3) adaptive pair (rklib) |
| `BS3()` | `bs32` | Bogacki-Shampine 3(2) adaptive (rklib); same as MATLAB `ode23` |
| `Tsit5()` | `t54` | Tsitouras 5(4) FSAL adaptive (rklib); same Butcher tableau |
| `DP5()` | `dop5`, `dop54`, `dp54` | Dormand-Prince 5th-order; same as MATLAB `ode45` |
| `TanYam7()` | `tmy7` | Tanaka-Muramatsu-Yamashita 7th-order (rklib) |
| `TsitPap8()` | `tp86` | Tsitouras-Papakostas 8(6) (rklib) |
| `DP8()` | `dop87`, `dp87` | Dormand-Prince 8(7) |
| `Vern6()` | `vern65`, `v65e` | Verner 6(5) efficient pair |
| `Vern7()` | `vern76`, `v76e` | Verner 7(6) efficient pair |
| `Vern8()` | `v78`, `dverk78` | Verner 7(8) pair |
| `Vern9()` | `vern98`, `v98e` | Verner 9(8) efficient pair |
| `Feagin10()` | `f108` | Feagin 10(8), 17 stages (rklib) |
| `Feagin12()` | `f1210` | Feagin 12(10), 25 stages (rklib) |
| `Feagin14()` | `f1412` | Feagin 14(12), 35 stages (rklib) |
| `ImplicitEuler()` | `backwardEuler`, `iem` | 1st-order fully implicit |
| `Trapezoid()` | `trapz` | Explicit trapezoidal rule, fixed-step |
| `Rosenbrock23()` | (none exact) | Julia's 2nd/3rd-order Rosenbrock W-method (Shampine 1982); rxode2 has no direct 2nd/3rd-order Rosenbrock -- `ros4` (Boost 4th-order W-method) is the lowest-order adaptive Rosenbrock-family analog available |
| `RosShamp4()` | `ros4` | Boost Kaps-Rentrop 4th-order Rosenbrock W-method; also the closest available analog to `Rosenbrock23()` (rxode2 has no 2nd/3rd-order Rosenbrock) |
| `Rodas4P()` | `ros43` | GRK4A from libode -- 4th-order A-stable Rosenbrock with 3rd-order error estimate |
| `Rodas5P()` | `ros43` | 5th-order in Julia; closest rxode2 analog remains `ros43` (4th-order) |
| `RadauIIA5()` | `radauiia5` | Radau IIA 5th-order, 3 stages |
| `GaussLegendre6()` | `gauss6` | Gauss-Legendre 6th-order |
| `LobattoIIIC6()` | `iiic6` | Lobatto IIIC 6th-order |
| `lsoda()` | `liblsoda`, `lsoda` | LSODA automatic Adams/BDF switching |
| `CVODE_Adams()` | `cvode` | SUNDIALS CVODE in Adams mode |
| `CVODE_BDF()` | `cvode` | SUNDIALS CVODE in BDF mode (default) |
| `AutoTsit5(Rosenbrock23())` | `"t54+ros4"` (closest; see note) | AutoSwitch: Tsitouras 5(4) + Boost Rosenbrock W-method; `ros4` is 4th order -- rxode2 has no 2nd/3rd-order Rosenbrock |
| `AutoTsit5(Rodas4P())` | `"t54+ros43"` | AutoSwitch: Tsitouras 5(4) with GRK4A (4th-order Rosenbrock) stiff fallback |
| `AutoTsit5(SDIRK43())` | `"t54+sdirk43"` | AutoSwitch: Tsitouras 5(4) with SDIRK 4(3) stiff fallback |
| `AutoVern6(Rosenbrock23())` | `"vern65+ros4"` (closest; see note) | AutoSwitch: Verner 6(5) with Boost Rosenbrock W-method |
| `AutoVern6(Rodas4P())` | `"vern65+ros43"` | AutoSwitch: Verner 6(5) with GRK4A stiff fallback |
| `AutoVern7(Rosenbrock23())` | `"vern76+ros4"` (closest; see note) | AutoSwitch: Verner 7(6) with Boost Rosenbrock W-method |
| `AutoVern7(Rodas4P())` | `"vern76+ros43"` | AutoSwitch: Verner 7(6) with GRK4A stiff fallback |
| `AutoVern7(SDIRK43())` | `"vern76+sdirk43"` | AutoSwitch: Verner 7(6) with SDIRK 4(3) stiff fallback |
| `AutoVern9(Rosenbrock23())` | `"vern98+ros4"` (closest; see note) | AutoSwitch: Verner 9(8) with Boost Rosenbrock W-method |
| `AutoVern9(Rodas4P())` | `"vern98+ros43"` | AutoSwitch: Verner 9(8) with GRK4A stiff fallback |
| `AutoVern9(SDIRK43())` | `"vern98+sdirk43"` | AutoSwitch: Verner 9(8) with SDIRK 4(3) stiff fallback |

Note: for many of the very high-order Runge-Kutta families in `rxode2`, the
closest naming culture is Julia's DifferentialEquations ecosystem rather than
MATLAB's built-in ODE suite.  In particular, `t54` (`Tsit5`), `tmy7`
(`TanYam7`), `tp86` (`TsitPap8`), and the three Feagin methods (`f108`,
`f1210`, `f1412`) have direct Julia name equivalents that have no MATLAB
counterpart.  The AutoSwitch composite methods are analogous to Julia's
`AutoTsit5`, `AutoVern7`, and `AutoVern9` wrappers, with the caveat that
rxode2 has no 2nd/3rd-order Rosenbrock matching Julia's `Rosenbrock23()` --
use `ros4` (4th-order Boost W-method) as the closest available lower-overhead
Rosenbrock-family secondary.

### Key differences from Julia

- **Language**: rxode2 is R-native; the model language feels like
  writing NONMEM or ordinary PK equations.  Julia models use Julia
  syntax and the ModelingToolkit.jl DSL.
- **Solver breadth**: Julia's ecosystem offers dozens of solvers.
  rxode2 provides a large set focused on the most useful methods for
  pharmacometrics.
- **Performance**: Julia achieves near-C speeds after JIT compilation.
  rxode2 compiles models to C ahead of time, reaching comparable speeds
  for repeated evaluations.  Julia has an advantage for single,
  long integrations; rxode2 has an advantage for large population
  simulations via OpenMP.
- **Sensitivity / gradients**: Julia's `sensealg` supports adjoint and
  forward sensitivity for large systems; rxode2 computes forward
  sensitivities (for nlmixr2 FOCEi) via symbolic differentiation,
  including for `delay()` models.
- **Automatic stiffness switching**: Julia's `AutoSwitch` re-picks the
  algorithm proactively at every step from its `eigen_est`; rxode2's
  `"primary+stiff"` composites switch reactively per interval using the
  same Hairer stiffness estimate (or a solve failure).  See *How rxode2
  AutoSwitch compares to Julia's `AutoSwitch`* under the AutoSwitch
  composite methods section above.
- **Interoperability**: The `JuliaCall` or `JuliaConnectoR` packages
  allow calling Julia's ODE solvers from R.

## NONMEM, Monolix, and Phoenix crosswalk

For pharmacometric users, the most useful comparison is often not
"which exact numerical method name matches?" but rather "which class of
models and solver behavior am I used to from another platform?"

### NONMEM ADVAN crosswalk

NONMEM's ADVAN family mixes three distinct ideas:

1. closed-form compartment models
2. matrix-exponential linear systems
3. general ODE/DAE solvers for nonlinear or stiff systems

That means the closest `rxode2` counterpart is sometimes **not** a
general ODE solver at all.

| NONMEM family | What NONMEM uses | Closest rxode2 comparison | Practical pharmacometric note |
|---|---|---|---|
| ADVAN1, ADVAN2, ADVAN3, ADVAN4, ADVAN10, ADVAN11, ADVAN12 | Closed-form PK model families | `linCmt()` first, then `indLin` or general ODE methods if you need custom structure | These are best compared to rxode2's analytic or near-analytic PK pathways, not to `dop853` or `cvode` |
| ADVAN5 | General linear model using a matrix exponential | `linCmt()` when the structure fits; otherwise `indLin` is usually the closest conceptual match | Good match for larger linear compartment systems or PBPK-like linear flows |
| ADVAN7 | General linear model with real eigenvalues, more efficient than ADVAN5 when that assumption holds | `linCmt()` / `indLin` | Again, this is closer to a linear algebra pathway than to a general adaptive ODE solver |
| ADVAN6 | General nonlinear model; NONMEM documents IMSL DVERK (Runge-Kutta-Verner 5/6) for nonstiff problems | `dop5`, `dop54`, `dop853`, or `liblsoda` | This is the closest NONMEM analog to rxode2's adaptive explicit non-stiff families |
| ADVAN8 | General nonlinear stiff model; NONMEM documents IMSL DGEAR (Gear) | `ros4`, `ros43`, `sdirk43`, `cvode`, or `liblsoda` | Best comparison when TMDD/QSP turnover or multi-scale kinetics make the system clearly stiff |
| ADVAN9 | Differential-algebraic / equilibrium-compartment model; NONMEM documents LSODI with BDF for stiff problems | `cvode` or stiff implicit methods in rxode2, but usually after rewriting the model into explicit ODE form if possible | Use this mental bucket for equilibrium or algebraic constraints rather than for ordinary PK models |
| ADVAN13, ADVAN14 | General nonlinear stiff/nonstiff equations in later NONMEM versions | `liblsoda` first(ADVAN13), then `cvode`(ADVAN14). You can also use `ros4`, or `dop853` depending on observed stiffness | These are the closest "modern general-purpose ODE" ADVANs to rxode2's default ODE workflows |
| ADVAN15 | Equilibrium-compartment extension of ADVAN13/14-style workflows | stiff implicit rxode2 workflows; sometimes custom model reformulation is the real comparison | Think "DAE-like/equilibrium-aware" rather than a one-name solver match |
| ADVAN16, ADVAN17 | Delay-equation models using RADAR5 | no direct ODE-solver equivalent in this vignette | This is beyond ordinary ODE comparison and should be treated separately |
| ADVAN18 | Delay-equation models using DDE solver | no direct ODE-solver equivalent in this vignette | Also outside the usual ODE-only solver menu |

The key practical mapping is:

- NONMEM analytic ADVANs -> `rxode2` `linCmt()` / `indLin`
- NONMEM ADVAN6 -> `rxode2` non-stiff adaptive solvers such as
  `dop54`, `dop5`, or `dop853`
- NONMEM ADVAN8 -> `rxode2` stiff-aware solvers such as `ros4`,
  `ros43`, `sdirk43`
- NONMEM ADVAN13 -> `liblsoda` general-purpose adaptive workflow
- NONMEM ADVAN14 -> `cvode` general-purpose adaptive workflows

#### Key differences from NONMEM

- **Open source**: rxode2 is GPL-3; NONMEM requires a commercial license.
- **Simulation speed**: rxode2 simulates thousands of subjects in
  seconds via OpenMP.  NONMEM simulation (`$SIM`) is single-threaded.
- **Estimation**: rxode2 itself is a simulator; population estimation is
  done through `nlmixr2`, which wraps rxode2.  NONMEM integrates
  estimation and simulation in one tool.
- **Tolerance**: NONMEM's `TOL` is a single scalar; rxode2 exposes
  separate `atol` and `rtol`.
- **Analytical solutions**: Both tools provide analytical compartment
  solutions.  NONMEM uses `ADVAN1--4`; rxode2 uses `linCmt()`, which
  supports 1--3 compartments and is compatible with stan-math gradients
  for FOCEi estimation in nlmixr2.
- **NONMEM datasets**: rxode2 accepts NONMEM-compatible event data
  frames directly (EVID 0--4).

#### Translating ADVAN5/7 rate constants with `matExp()`

NONMEM ADVAN5/7 models specify the rate matrix via `K(i,j)` scalars in
`$PK`, where `K(i,j)` is the transfer rate **from** compartment `i`
**to** compartment `j`.  rxode2's `matExp()` declaration uses the
**same source-first convention** (`k_from_to`), making translation
direct and avoiding all symbolic manipulation.  This is the most
numerically stable approach for linear compartmental models:

```{r}
threeComp <- rxode2({
  matExp()
  k_depot_central  = ka
  k_central_output = k20
  k_central_periph = k23
  k_periph_central = k32
  Cp = central / Vc
})
sol <- rxSolve(threeComp, params, ev, method = "indLin")
```

Compartments (`depot`, `central`, `periph`) are auto-detected from the
rate constant names; no explicit declarations are needed.

See the companion article
[Matrix Exponentials and Inductive Linearization](rxode2-ind-lin.html)
for a full description of `matExp()` syntax, the NONMEM index mapping,
and the `indLin(state)` forcing-term syntax.

#### Translating a full NONMEM control stream

The `nonmem2rx` package (part of the nlmixr2 ecosystem) can import
NONMEM control streams into rxode2/nlmixr2 models automatically.

```{r}
# nonmem2rx::nonmem2rx("run001.ctl")
```

### Monolix crosswalk

Monolix documents two ODE solver modes:

- `odeType = nonStiff`: Adams methods for non-stiff ODEs
- `odeType = stiff`: Backward Differentiation Formula (BDF) methods for
  stiff ODEs

That makes the practical `rxode2` comparison fairly direct:

| Monolix mode | Closest rxode2 comparison | Pharmacometric interpretation |
|---|---|---|
| nonStiff (Adams) | `ab`, `abm` for an Adams predictor-corrector match; `liblsoda` for automatic Adams/BDF switching; `dop54`, `dop5`, `dop853`, `t54` as practical non-stiff Runge-Kutta alternatives | Smooth PK/PD, transit, or exposure-response systems without severe time-scale separation |
| stiff (BDF) | `bdf`, `cvode`, `liblsoda`, `ros4`, `ros43`, or `sdirk43` | TMDD, target binding, indirect response with fast turnover, mechanistic PD, QSP-like stiffness |

Monolix's `nonStiff` mode uses Adams (predictor-corrector) methods internally.
The structurally closest rxode2 analogs are `ab` (Adams-Bashforth fixed-step)
and `abm` (Adams-Bashforth-Moulton fixed-step), or `liblsoda` which runs
Adams-style stepping in its non-stiff mode.  In practice, many pharmacometric
users who were happy with Monolix `nonStiff` will find that `liblsoda`, `dop54`,
`t54`, or `dop853` give equally good results on smooth PK/PD problems, with
the Runge-Kutta families often being faster on smooth outputs.

- if you were happy with Monolix `nonStiff` and want the closest Adams analog,
  try `ab` or `abm`; for general-purpose non-stiff use, try `liblsoda`,
  `dop54`, `t54`, or `dop853`

- if you needed Monolix `stiff`, start in `rxode2` with `bdf`, `liblsoda`,
  `cvode`, `ros4`, or `ros43`

#### Key differences from Monolix

- **Free and open source**: rxode2 and nlmixr2 are freely available on
  CRAN; Monolix requires a commercial license (free academic licenses
  exist).
- **Model language**: Monolix uses a proprietary model language
  (Mlxtran); rxode2 uses an R-embedded mini-language that is a superset
  of common PK notation.
- **Customization**: rxode2 models can call arbitrary C/C++ functions
  registered at the R level; Monolix has limited extensibility.
- **Ecosystem**: nlmixr2 integrates with ggPMX, xpose4, and vpc for
  diagnostics; Monolix has Datxplore and Pkanalix (could use PKNCA instead).
- **Simulation platform**: `rxode2` / `RxODE` underpins `simulx()`
  via the `RsSimulx` R package (PKPD simulations driven by Monolix
  parameter estimates).

### Phoenix WinNonlin / Phoenix NLME crosswalk

Phoenix WinNonlin's built-in compartmental tools are best compared by
**model family**.  Phoenix NLME, however, does publish its named ODE
choices, so those can be mapped more directly.

| Phoenix workflow / solver | Closest rxode2 comparison | Practical note |
|---|---|---|
| Phoenix WinNonlin built-in compartmental PK models | `linCmt()` or simple `rxode2()` PK models | Best match for standard 1-, 2-, and 3-compartment PK workflows |
| Phoenix NLME `Matrix Exponent` | `indLin` or `linCmt()` | Matrix-exponential solve for linear ODE systems with a constant Jacobian; usually the closest speed-oriented comparison |
| Phoenix NLME `Exponent Higham` | `indLin` | Alternative matrix-exponential path with the same basic pharmacometric role as `Matrix Exponent` |
| Phoenix NLME `Non-Stiff (DVERK)` | `vern65`, `v65e`, `dop5`, `dop54` | Explicit non-stiff Runge-Kutta/Verner family behavior |
| Phoenix NLME `Non-Stiff (DOPRI5)` | `dop54`, `dp54`, `dop5` | Closest Phoenix analog to the Dormand-Prince family most users associate with `ode45`-like behavior |
| Phoenix NLME `Auto-Detect` (LSODA) | `liblsoda`, `lsoda` | Automatic Adams/BDF switching; strongest direct name-level match in rxode2 |
| Phoenix NLME `Stiff` (LSODE/BDF) | `cvode`, `ros4`, `ros43`, `sdirk43` | Stiff-only pathway; closest to rxode2 stiff workflows rather than to one exact solver name |

Phoenix NLME also uses a three-level solving strategy:

1. closed-form for supported standard compartmental models
2. matrix exponentials for linear ODE systems with a constant Jacobian
3. numerical ODE solvers for nonlinear systems

So, for Phoenix users moving to `rxode2`, the practical sequence is:

1. if the Phoenix model is a standard compartmental PK model, compare
   first to `linCmt()`
2. if the Phoenix model used `Matrix Exponent` or `Exponent Higham`,
   compare first to `indLin`
3. if it used `Auto-Detect`, compare first to `liblsoda`
4. if it used `Stiff`, compare `cvode`, `ros4`, or `ros43`
5. if it used `Non-Stiff (DOPRI5)` or `Non-Stiff (DVERK)`, compare
   `dop54`, `dop5`, `vern65`, or `dop853`, `dverk65` or `dverk78`

## Summary

The table below compares rxode2 against other common pharmacometric
and scientific computing platforms across key features.

| Feature | rxode2 | MATLAB | NONMEM | Monolix | Julia (SciML) |
|---|---|---|---|---|---|
| Open source | Yes (GPL-3) | No | No | No (free academic) | Yes (MIT) |
| Language | R + C | MATLAB | NONMEM7 | Mlxtran | Julia |
| Stiff solver | LSODA (auto) | ode15s | ADVAN8/13 | ode23s | CVODE_BDF, Rodas5P |
| Non-stiff solver | dop853, LSODA Adams | ode45 | ADVAN6/9 | ode45 | Tsit5, DP8 |
| Analytical PK | linCmt (1-3 cmt) | Manual | ADVAN1-4 | PKanalix | ModelingToolkit |
| Matrix exponential | `matExp()` / `indLin` | expm() | ADVAN5/7 | -- | LinearAlgebra.exp |
| Parallel subjects | OpenMP (built-in) | PCT add-on | -- | -- | Threads.@threads |
| NONMEM datasets | Native | Manual | Native | Via import | Manual |
| Population estimation | nlmixr2 | -- | Native | Native | Turing.jl |
| Sensitivity equations | Forward (liblsoda) | -- | -- | -- | ForwardDiff, adjoint |

## Complete accepted solver names in rxode2

The accepted method names below cover the full current solver menu in
`rxode2`.

### AutoSwitch composite methods

Use the `"primary+stiff"` syntax; any thread-safe non-stiff primary
with any thread-safe implicit secondary (excludes `lsoda`, `liblsoda`,
`lsode`, `bdf` as primary).  Recommended pairs and Julia analogs are in
the [AutoSwitch composite methods](#autoswitch-composite-methods) section above.

### Adaptive, stiffness-aware, or stiff-suitable

- `liblsoda`, `lsoda`, `cvode`, `ros4`, `ros43`, `sdirk43`

### Fixed-step, stiff-suitable, or implicit

- `iem`, `ros6`, `backwardEuler`, `gauss6`, `iiic6`, `radauiia5`,
  `geng5`

### Fixed-step, primarily non-stiff

- `rk4`, `trapz`, `ssp3`, `ssp33`, `ab`, `abm`, `sem`, `sb3a`,
  `sb3am4`, `vv`, `mm`, `em`

### Pharmacometric linearization special case

- `indLin`

### DLSODE-based fixed-method solvers (single-threaded)

- `lsode`, `bdf`

These two methods use DLSODE (Hindmarsh 1983, ODEPACK) with a fixed method
flag -- unlike `lsoda`, they do **not** switch between Adams and BDF.
Both use non-reentrant Fortran COMMON blocks and therefore **always run on a
single core**, regardless of the `cores` setting.

- `lsode` -- DLSODE MF=10 (Adams, non-stiff), variable order 1-12; no Jacobian
  evaluation.  Corresponds to deSolve's `lsode(method="adams")`.
- `bdf` -- DLSODE MF=22 (BDF, stiff), variable order 1-5; internally generated
  dense Jacobian.  Corresponds to deSolve's `bdf()` / `lsode(method="bdf")`.

### Adaptive, primarily non-stiff: core explicit families

- `dop853`, `f78`, `ck54`, `dop5`, `bs`, `f32`, `rk43`,
  `dop54`, `dp54`, `vern65`, `v65e`, `vern76`, `v76e`,
  `dop87`, `dp87`, `vern98`, `v98e`

### Adaptive, primarily non-stiff: lower-order and SSP families

- `euler`, `midpoint`, `heun`, `ssp22`, `rk3`, `ssp53`,
  `bs32`, `ssp43`, `f45`

### Adaptive, primarily non-stiff: fourth- and fifth-order families

- `s4`, `r4`, `ls44`, `ls54`, `ssp54`, `s5`, `rk5`,
  `c5`, `l5`, `lk5a`, `lk5b`, `t54`, `s54`, `pp54`,
  `pp54b`, `bs54`, `ss54`

### Adaptive, primarily non-stiff: sixth- and seventh-order families

- `b6`, `rk7`, `dp65`, `c65`, `tp64`, `v65r`, `v65`,
  `dverk65`, `tf65`, `tp75`, `tmy7`, `tmy7s`, `v76r`,
  `ss76`, `v78`, `dverk78`

### Adaptive, primarily non-stiff: eighth-order and higher families

- `rk8_10`, `cv8`, `rk8_12`, `s10`, `z10`, `o10`, `h10`,
  `dp85`, `tp86`, `v87e`, `v87r`, `ev87`, `k87`,
  `f89`, `v89`, `t98a`, `v98r`, `s98`, `f108`,
  `c108`, `b109`, `s1110a`, `f1210`, `o129`, `f1412`

## What to try first

If you do not want to think about the entire menu, a good
pharmacometric workflow is:

1. start with `liblsoda`

2. if the system is smooth and clearly non-stiff, compare against
   `dop853` or `dop54`

3. if the system is stiff or nearly stiff, compare against `ros4`,
   `ros43`, `sdirk43`, or `cvode`

4. if stiffness is uncertain, try an AutoSwitch pair such as
   `"t54+ros4"`, `"t54+ros43"`, or `"vern76+ros43"` -- see the
   [AutoSwitch section](#autoswitch-composite-methods) for recommended
   pairs and Julia analogs

5. if you need reproducible fixed internal stepping for diagnosis, try
   `rk4`, `backwardEuler`, `gauss6`, or `radauiia5`

6. if the model is mostly linear between events, try `indLin`

That strategy is usually more valuable in pharmacometric practice than
trying to optimize across every available high-order RK family from
the start.

## References

- Sharif S, Hasegawa C, Duffull SB (2022). Exploring Inductive
  Linearization for simulation and estimation with an application to the
  Michaelis--Menten model. *J Pharmacokinet Pharmacodyn* 49:445--453.

- Shampine LF, Reichelt MW (1997). The MATLAB ODE suite. *SIAM J Sci
  Comput* 18(1):1--22.

- Hindmarsh AC (1983). ODEPACK, a systematized collection of ODE
  solvers. In Stepleman RS (ed), Scientific Computing, pp. 55--64.

- Rackauckas C, Nie Q (2017). DifferentialEquations.jl -- a performant
  and feature-rich ecosystem for solving differential equations in
  Julia. *J Open Res Softw* 5(1):15.

- Hairer E, Norsett SP, Wanner G (1993). Solving Ordinary Differential
  Equations I: Nonstiff Problems. 2nd ed. Springer.

- Beal S, Sheiner LB, Boeckmann A, Bauer RJ (2009). NONMEM Users
  Guides. (1989--2009). Icon Development Solutions, Ellicott City, MD.
