---
title: "Delay Differential Equations in rxode2"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## What is a delay differential equation?

An ordinary differential equation (ODE) describes how a system changes
based on its *current* state. A **delay differential equation (DDE)**
lets the rate of change also depend on a *past* state -- the system at
time `t - T`. Delays arise naturally in pharmacometrics (signaling and
maturation delays, delayed feedback, transit-like absorption) and in
ecology, physiology and control.

In `rxode2` a delayed state is written with `delay(state, T)`, which
evaluates the ODE compartment `state` at the earlier time `t - T`. The
semantics match the `delay()` function of Monolix.

The pharmacodynamic examples in this article (Examples 2 and 6) are taken
from the review by Koch, Krzyzanski, Perez-Ruixo and Schropp,
*Modeling of delays in PKPD: classification of delays and mathematical
properties* (J Pharmacokinet Pharmacodyn, 2014; see
[References](#references)), which classifies delay models in
pharmacometrics and gives the equations we reproduce below.

```{r library}
library(rxode2)
```

### A first example

The classic linear DDE

$$y'(t) = -y(t - 1), \qquad y(t) = 1 \text{ for } t \le 0$$

is written directly:

```{r classic}
dde <- function() {
  model({
    y(0) <- 1
    d/dt(y) <- -delay(y, 1)
  })
}

s <- rxSolve(dde, et(seq(0, 5, by = 0.1)))
```

Before the start of integration (`t <= 0` here) the history is the
constant initial condition, `y = 1`. After that, the solution is built
by the solver. We can check it against the exact "method of steps"
solution, which is piecewise polynomial:

$$y(t) = \begin{cases} 1 - t & 0 \le t \le 1 \\ \tfrac{t^2}{2} - 2t + \tfrac{3}{2} & 1 \le t \le 2 \end{cases}$$

```{r classic-check}
exact <- function(t) {
  ifelse(t <= 1, 1 - t, t^2 / 2 - 2 * t + 3 / 2)
}
max(abs(s$y[s$time <= 2] - exact(s$time[s$time <= 2])))
```

The agreement is at machine precision.

```{r classic-plot, fig.height=4}
plot(s, y)
```

## How delays use dense output

The key numerical challenge is that `delay(y, T)` needs the solution at
a past time `t - T` that is, in general, **not** one of the solver's
step boundaries or one of your observation times. `rxode2` answers this
the same way it speeds up dense-grid solving (see the companion article
[Dense output for fast dense-grid solves](rxode2-dense-output.html)):
every accepted step records the **dense-output polynomial** that
continuously interpolates the solution across that step.

For delay differential equations these polynomials are kept in a
per-subject history buffer as the integration advances. When the model
evaluates `delay(y, T)`, `rxode2`:

1. finds the recorded step that brackets the past time `t - T`, and
2. evaluates that step's dense polynomial at `t - T`.

Because this is the same interpolant the integrator uses for its own
output, the delayed value is obtained to the **full order of the
method** -- an 8th-order Dormand-Prince interpolant for `dop853`, and a
cubic Rosenbrock interpolant for `ros4`.

This has a few practical consequences, all handled automatically:

- **Delay models are solved on a dense path.** When a model uses
  `delay()`, `rxode2` enables dense output and switches the default
  method to the dense **AutoSwitch composite `"dop853+ros4"`**. A
  non-dense method (for example `lsoda`) cannot record the history and
  raises an error:

    ```{r non-dense, error=TRUE}
    rxSolve(dde, et(0:5), method = "lsoda")
    ```

- **Non-stiff and stiff regions are both handled.** The composite
  probes with the explicit `dop853` and falls back per segment to the
  implicit Rosenbrock `ros4` when a region turns stiff, so a delay
  model that is non-stiff in one region and stiff in another is solved
  efficiently in a single pass. A purely stiff delay model can also be
  solved with `method = "ros4"` directly.

- **The step size is capped to the smallest delay.** This keeps the
  lagged time `t - T` inside already-recorded history (rather than
  extrapolating off the current, not-yet-finished step), so short
  delays stay accurate.

- **Only the delayed states are recorded.** History is stored only for
  the compartments that some `delay()` actually looks back on, so a
  delay on one state in a large ODE system stays inexpensive.

You can still request a specific dense solver explicitly; `method =
"dop853"` (pure 8th-order) and `method = "ros4"` (stiff) both work.

## Example 2: a lifespan-based tumor-growth model

A more clinically relevant use of `delay()` is Koch et al.'s **lifespan
based tumor-growth model** (their Example 2), a delay reformulation of
the Simeoni tumor-growth model. Drug concentration `c(t)` (a one
-compartment oral PK model) drives proliferating tumor cells
`p = x4` into an apoptotic pool `x5`; apoptotic cells live for a fixed
lifespan `Td`, so cells that entered the pool one lifespan ago --
`delay(x4, Td)` -- leave it. A drug-free control `x3` grows in parallel:

```{r ex2, fig.height=4}
ex2 <- function() {
  ini({
    ka   <- 103.96 * 24
    ke   <- 0.1052 * 24
    V    <- 2.7882
    l0   <- 0.195
    l1   <- 0.245
    w0   <- 0.010
    kpot <- 0.007
    Td   <- 3.61        # apoptotic-cell lifespan (the delay)
  })
  model({
    ## one-compartment oral PK -> concentration c(t)
    d/dt(x1) <- -ka * x1
    d/dt(x2) <-  ka * x1 - ke * x2
    c    <- x2 / V
    cdel <- delay(x2, Td) / V                    # c(t - Td)
    ## lifespan tumor growth: x4 = proliferating p(t), x5 = apoptotic d(t)
    d/dt(x3) <- (2 * l0 * l1 * x3) / (l1 + 2 * l0 * x3)                 # drug-free control
    d/dt(x4) <- (2 * l0 * l1 * x4^2) / ((l1 + 2 * l0 * x4) * (x4 + x5)) - kpot * c * x4
    d/dt(x5) <-  kpot * c * x4 - kpot * cdel * delay(x4, Td)
    w  <- x4 + x5                                # observed treated tumor weight
    x3(0) <- w0
    x4(0) <- w0
    x5(0) <- 0
  })
}

## 100 mg orally at days 12-16, observe the tumor for 25 days
ev <- et(seq(0, 25, by = 0.1)) |> et(amt = 100, time = 12:16, cmt = "x1")
s2 <- rxSolve(ex2, ev)

plot(s2$time, s2$x3, type = "l", lwd = 2, xlab = "time (days)",
     ylab = "tumor weight", ylim = c(0, max(s2$x3)))
lines(s2$time, s2$w, lwd = 2, col = "red")
legend("topleft", c("control (drug-free)", "treated  w = p + d"),
       col = c("black", "red"), lwd = 2, bty = "n")
```

The treated tumor is suppressed by the dosing and then regrows once the
drug washes out, while the drug-free control grows monotonically -- the
signature of the Simeoni model. Note the two `delay(., Td)` terms share
one estimated lifespan `Td`, and the delay duration is a model parameter
rather than a constant.

The delay duration `T` is an arbitrary model expression -- it can be a
constant, a parameter, or a covariate -- and a model may use several
`delay()` terms, on the same state or on different states, each with
its own delay.

## Delays mixed with ordinary states

`delay()` composes freely with ordinary ODE compartments, dosing, and
covariates. Here a depot feeds a central compartment that has a delayed
auto-induction-like feedback term:

```{r mixed, fig.height=4}
mix <- function() {
  ini({
    ka  <- 1
    ke  <- 0.3
    kin <- 0.2
  })
  model({
    cen(0) <- 0
    d/dt(depot) <- -ka * depot
    d/dt(cen)   <-  ka * depot - ke * cen + kin * delay(cen, 2)
  })
}

ev <- et(amt = 100, cmt = "depot") |> et(seq(0, 24, by = 0.25))
s3 <- rxSolve(mix, ev)
plot(s3, cen)
```

## Example 6: a non-constant history with `past()`

By default the pre-history of a delayed state (its value for `t <= t0`)
is the constant initial condition. Some published models instead need a
**non-constant history function**. Koch et al.'s Example 6 is a
rheumatoid-arthritis model in which the cytokine GM-CSF `G(t)` is already
being over-produced during an *induction phase* before the observation
window, so its past is the exponential `G(t) = a * exp(b * t)` on the
interval `[-T, 0]`. The strongly delayed ankylosis score is driven by
`delay(G, T)`, which
during the early response reaches back into that induction-phase history.

`rxode2` provides this with a `past(state, T) <- expr` line, giving the
value of `state` for times before `t0`. It must reference the same delay
duration used in the matching `delay(state, T)`:

```{r ex6, fig.height=4}
ra <- function() {
  ini({
    a  <- 2       # induction-phase amplitude
    b  <- -0.25   # induction-phase rate
    Td <- 5       # delay
    k3 <- 1
    kg <- 0.4
    k4 <- 0.3
    k5 <- 0.1
  })
  model({
    G(0) <- a                                  # continuity: a*exp(0) = a
    d/dt(G) <- k3 - kg * G                      # development-phase cytokine turnover
    past(G, Td) <- a * exp(b * t)               # induction-phase history (Koch Eq. 63)
    I(0) <- 0
    d/dt(I) <- k4 * G - k4 * delay(G, Td)       # inflammation
    D(0) <- 0
    d/dt(D) <- k4 * delay(G, Td) - k5 * D       # bone destruction
    R1 <- I + D                                 # total arthritic score (TAS)
    R2 <- D                                     # ankylosis score (AKS)
  })
}

ev6 <- et(seq(0, 28, by = 0.25))
sra <- rxSolve(ra, ev6)
plot(sra, R2)
```

The `past()` history and the initial condition must agree at `t0` (here
`G(0) = a`, which equals `a * exp(b * 0)`), so the delayed state is
continuous. Using
the ordinary constant-history default here would hold `delay(G, Td)` at
`G(0)` and leave the ankylosis score flat until `t = Td`, missing the
induction-phase-driven early response entirely -- which is exactly why
this class of models needs a non-constant past.

## Estimation with nlmixr2 (sensitivities)

Gradient-based estimation (such as the FOCEi family in `nlmixr2`)
needs the sensitivities of the model with respect to the estimated
parameters. `rxode2` generates these for delay models too: the
forward-sensitivity equations gain the delayed term `delay(S, T)` (the
sensitivity of the delayed state), which reuses exactly the same dense
history machinery described above. Delay durations that themselves
depend on an estimated parameter are also supported. When a model has a
non-constant `past()` history, the sensitivity of that history with
respect to each estimated parameter is carried into the pre-history of
the corresponding sensitivity state, so the gradient (and analytic
Hessian) stay exact -- for instance the `a` and `b` in Example 6's
`past(G, Td) <- a * exp(b * t)` are estimable. As a result, a delay model
built with the `function` interface can be fit like any other `nlmixr2`
model.

## Notes and attribution

- Use `delay(state, T)` inside a `d/dt()` right-hand side. The first
  argument must be an ODE state (compartment) defined in the model; the
  second is the delay duration.
- Use `past(state, T) <- expr` to give a delayed state a non-constant
  pre-history (its value for `t <= t0`); `expr` is a function of `t` and
  the model parameters, `T` must match the state's `delay(state, T)`, and
  `expr` at `t0` should equal the state's initial condition. Without a
  `past()` line the history is the constant initial condition.
- The default solving method for delay models is the dense AutoSwitch
  composite `"dop853+ros4"`; `"dop853"` and `"ros4"` may be requested
  explicitly. Methods that cannot record dense history error out.
- The dense-output and delay-history machinery is adapted from the
  [`dde`](https://github.com/mrc-ide/dde) package by Rich FitzJohn and
  Wes Hinsley (Imperial College of Science, Technology and Medicine),
  following the dense-output approach of Hairer, Norsett and Wanner.

See also the function reference for `?delay` and the companion article
[Dense output for fast dense-grid solves](rxode2-dense-output.html).

## References {#references}

- Koch G, Krzyzanski W, Perez-Ruixo JJ, Schropp J. Modeling of delays in
  PKPD: classification of delays and mathematical properties.
  *J Pharmacokinet Pharmacodyn.* 2014;41(4):291-318.
  doi:10.1007/s10928-014-9368-y. (Examples 2 and 6 above.)
- Simeoni M, Magni P, Cammia C, De Nicolao G, Croci V, Pesenti E,
  Germani M, Poggesi I, Rocchetti M. Predictive pharmacokinetic
  -pharmacodynamic modeling of tumor growth kinetics in xenograft models
  after administration of anticancer agents. *Cancer Res.*
  2004;64(3):1094-1101. (Basis of the Example 2 tumor-growth model.)

## Session Information

```{r sessioninfo}
sessionInfo()
```
