---
title: "rxode2 and Simulx: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

[rxode2](https://nlmixr2.github.io/rxode2/) and
[Simulx](https://www.simulations-plus.com/software/monolix/simulx/)
are both tools for pharmacometric simulation, but they differ
substantially in design philosophy, licensing, and ecosystem
integration.

- **rxode2** is an open-source R package.  Models are defined in R
  using rxode2's own mini-language, compiled to a shared library, and
  solved within the R session.  It is fully scriptable and integrates
  with nlmixr2 for parameter estimation.

- **Simulx** is a commercial clinical trial simulation application
  (part of the MonolixSuite, distributed by Simulations Plus / Lixoft).
  It provides a graphical user interface (GUI) as its primary
  interaction mode, and an R scripting interface via the
  `lixoftConnectors` package.  Models are specified in mlxtran -- the
  same language used by Monolix for estimation -- so Monolix projects
  can be imported directly with no re-implementation.  A commercial or
  academic license is required.

## Shared capabilities

- ODE-based pharmacometric models
- Between-subject variability (BSV / omega)
- Residual error models
- Covariate distributions (continuous and categorical)
- Population simulation across many subjects
- Multiple dosing regimen groups
- Parameter uncertainty simulation
- VPC (visual predictive check) workflows
- Clinical trial scenario design and simulation

## Features rxode2 has that Simulx does not

### No license required

rxode2 is entirely open-source (GPL >= 3).  Simulx requires a
commercial license from Simulations Plus; an academic license is
available free of charge but must be requested and renewed.  rxode2
runs without any registration or license file on any machine with R
and a C compiler.

### Fully scriptable, reproducible workflows

rxode2 is a pure R package.  Every aspect of model definition,
simulation, and output processing is expressed in R code, making
workflows fully reproducible and straightforward to version-control,
review, and automate in CI/CD pipelines.

Simulx's primary interface is a GUI application.  Its R scripting
layer (`lixoftConnectors`) is a thin connector to the underlying
Simulx engine rather than a native R simulation library, so GUI state
and R scripts are partially coupled.

```{r}
library(rxode2)

# Entirely self-contained R script -- no GUI, no license file
mod <- rxode2({
  KA <- exp(tka + eta.ka)
  CL <- exp(tcl + eta.cl)
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})

et  <- et(amt = 100, time = 0) |> et(seq(0, 24, by = 1))
sim <- rxSolve(mod, et,
               params = c(tka = log(0.5), tcl = log(4), V = 10),
               omega  = lotri(eta.ka ~ 0.09, eta.cl ~ 0.09),
               nSub   = 100)
```

### Parameter estimation via nlmixr2

rxode2 models can be passed directly to nlmixr2 for population
parameter estimation (FOCE, FOCEi, SAEM, etc.) using the exact same
model object used for simulation.  Simulx is simulation-only;
estimation is performed in Monolix separately.

### OpenMP parallelism across subjects

rxode2's LSODA solver is implemented in thread-safe C, enabling
genuine parallel ODE solving across subjects via OpenMP within a
single R process.  Simulx uses a C++ solver internally and parallelises
across CPU cores through its own engine, but this is not directly
controlled from R.

### Model piping for incremental modification

rxode2 model objects support the native R pipe (`|>`) to clone and
modify a model incrementally without rewriting the full specification.

```{r}
base <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})

# Add population structure and residual error
full <- base |>
  model({
    KA <- exp(tka + eta.ka)
    CL <- exp(tcl + eta.cl)
  }, append = FALSE) |>
  ini({
    tka <- log(0.5); eta.ka ~ 0.09
    tcl <- log(4);   eta.cl ~ 0.09
    V   <- 10
    add.sd <- 0.5
  }) |>
  model(cp ~ add(add.sd), append = TRUE)

# Sensitivity run: fix BSV on KA
sens <- full |> ini(fix(eta.ka))
```

Simulx does not have an equivalent R-level model algebra; scenario
modifications are made through the GUI or via `lixoftConnectors`
function calls that manipulate named elements of a loaded project.

### Symbolic Jacobians and forward sensitivities

rxode2 automatically derives the symbolic Jacobian and forward-
sensitivity equations, used by nlmixr2's FOCEi algorithm for exact
gradient-based parameter estimation.  These extend to third-order
sensitivities and to *jump* (event) sensitivities -- exact derivatives
with respect to dosing parameters such as bioavailability, lag time,
infusion rate, duration, and amount.

### 1-3 compartment analytical solutions with exact gradients

`linCmt()` provides analytical solutions for one-, two-, and
three-compartment PK models with gradients via Stan math
auto-differentiation.  Simulx uses ODEs for all compartment models.
rxode2 can also detect a 1-3 compartment linear PK ODE system and
transparently solve it through this analytical path
(`rxSolve(..., useLinCmt = TRUE)`, the default), falling back to the ODE
solver if the converted model cannot be compiled.

### Multiple ODE solver backends

rxode2 offers several solver backends selectable per run: thread-safe
C LSODA (default), Fortran LSODA, DOP853 (explicit 8th-order
Runge-Kutta for non-stiff systems), a stiff Rosenbrock method (`ros4`),
and AutoSwitch composite methods (for example `method = "dop853+ros4"`)
that probe each segment with a non-stiff primary and reactively fall
back to a stiff secondary only where stiffness is detected.  Simulx uses
a single C++ LSODA-based solver.

### Delay differential equations

rxode2 solves delay differential equations through `delay(state, T)`,
which evaluates an ODE state at the past time `t - T` using dense-output
interpolation, with forward sensitivities generated automatically so the
models stay estimable in nlmixr2.

### Out-of-memory simulation and distributed parallelism

For very large trial simulations, passing a `file` prefix to
`rxSolve()` splits subjects into RAM-sized chunks (auto-sized with
`rxMemoryEstimate()`), solves each, and writes them to Parquet (or
`.rds`) behind a lightweight object that supports lazy `dplyr`/DuckDB
queries, so an out-of-memory result is never fully materialized.  Chunks
can be dispatched to `mirai` daemons (`rxSolve(..., parallel = n)`) for
distributed/HPC execution, all controlled directly from R.

```{r}
rxSolve(mod, params, ev, file = tempfile("rx"), parallel = 4)
```

### Runs anywhere without infrastructure

rxode2 runs on any machine with R -- laptops, CI servers, Docker
containers, or Shiny apps.  Simulx requires the MonolixSuite
installation and a valid license to be present on the machine.

## Features Simulx has that rxode2 does not

### GUI for interactive exploration

Simulx's primary strength is its point-and-click GUI.  The
**Exploration** workflow lets modelers interactively sweep parameter
values and dosing regimens for a typical individual with real-time
graphical output -- no code required.  This low-friction interface is
valuable for non-programmers and for rapid "what-if" exploration
before committing to a full simulation script.

rxode2 is code-only; interactive exploration requires writing R code
or building a custom Shiny interface.

### Direct Monolix project import

Because Simulx uses the same mlxtran model language as Monolix,
a Monolix estimation project can be imported into Simulx in one click:
the model equations, parameter estimates, omega and sigma matrices,
and covariate distributions are all carried over automatically with
zero re-implementation.

rxode2 uses a different model language.  Monolix models can be
converted via the `monolix2rx` package, which translates mlxtran to
rxode2 syntax and runs an automatic validation comparing PRED/IPRED
between the two tools.  Models that pass validation can be simulated
in rxode2 with confidence; models that do not pass require manual
review.

### Built-in scenario management and result display

Simulx provides a structured GUI workflow for defining named
simulation scenarios (groups with different doses, covariates, or
parameter values), running them, and visualizing results in
interactive tables and plots -- all within the application.

In rxode2 all scenario management, tabulation, and plotting is done
in R using general-purpose packages (e.g., `ggplot2`, `data.table`),
which gives more flexibility but requires more code.

### Standardized clinical trial simulation workflow

Simulx's GUI enforces a clear three-step workflow -- **Definition**
(parameters, treatments, outputs, covariates), **Exploration**
(typical individual), and **Simulation** (population) -- that guides
users toward a complete and well-structured simulation.  This
structured approach reduces the risk of ad hoc or incomplete
simulation setups.

rxode2 imposes no workflow structure; the user is responsible for
organizing their simulation code.

## Monolix-to-rxode2 import via `monolix2rx`

For users who estimate in Monolix and want to simulate in rxode2,
the `monolix2rx` package (part of the nlmixr2 ecosystem) converts a
Monolix project to an rxode2 model and automatically validates the
conversion by comparing PRED/IPRED against the original Monolix
output.  A passing validation gives confidence that the rxode2 model
is numerically equivalent to the Monolix model.

```{r}
library(monolix2rx)

# Convert a Monolix project and validate automatically
mod <- monolix2rx("path/to/run001.mlxtran")

# Simulate the converted model in rxode2
rxSolve(mod, event_table, nSub = 1000)
```

This bridges the Monolix/Simulx and rxode2/nlmixr2 ecosystems,
allowing estimation in Monolix and fast scripted simulation in rxode2.

## Summary table

| Feature | rxode2 | Simulx |
|---------|--------|--------|
| License | open-source (GPL >= 3) | commercial (free academic) |
| Primary interface | R code | GUI + `lixoftConnectors` R API |
| Model language | rxode2 mini-language | mlxtran |
| Monolix project import | `monolix2rx` (with auto-validation) | direct (same language) |
| Re-implementation risk | mitigated: `monolix2rx` auto-validates PRED/IPRED | none |
| Simulation engine | rxode2 C/Fortran solver | Simulx C++ solver |
| ODE solver options | LSODA (C/Fortran), DOP853, ros4, `"dop853+ros4"` | single C++ LSODA-based |
| 1-3 cmt analytical solution | `linCmt()` with gradients | ODE-based only |
| Automatic ODE -> `linCmt()` conversion | yes (`useLinCmt=TRUE`) | no |
| Delay differential equations (`delay()`) | yes | no |
| Symbolic Jacobians / sensitivities | yes (up to 3rd order + jump sensitivities) | no |
| Parameter estimation | via nlmixr2 | no (Monolix only) |
| OpenMP parallelism (R-controlled) | yes | engine-internal |
| Distributed / cluster parallelism | yes (`mirai`) | engine-internal |
| Out-of-memory / disk-backed solve | yes (`rxSolve(file=)`, Parquet/DuckDB) | no |
| Model piping (`\|>`) | yes | no |
| GUI / interactive exploration | no (code only) | yes |
| Built-in scenario management | no (use R code) | yes |
| Built-in result visualisation | no (use ggplot2 etc.) | yes |
| Monolix ecosystem integration | via monolix2rx | native |
| nlmixr2 ecosystem integration | native | no |
| Runs without installed application | yes | no |
| CRAN | yes | no (`lixoftConnectors` distributed by Lixoft) |

## References

- [Simulx product page](https://www.simulations-plus.com/software/monolix/simulx/)
- [Simulx documentation](https://monolixsuite.slp-software.com/simulx/2024R1/introduction-to-simulx)
- [MonolixSuite R functions (lixoftConnectors)](https://monolixsuite.slp-software.com/r-functions/2024R1)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [monolix2rx package](https://nlmixr2.github.io/monolix2rx/)
- [nlmixr2 package](https://nlmixr2.github.io/nlmixr2/)
