---
title: "rxode2 and NONMEM: A Simulation Workflow Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

NONMEM is the most widely used software for population pharmacokinetic
and pharmacodynamic modeling.  Its `$SIMULATION` step is a common tool
for generating simulated datasets used in visual predictive checks
(VPCs), clinical trial simulations, and model evaluation.

[rxode2](https://nlmixr2.github.io/rxode2/) is an open-source R
package that can perform the same population simulation tasks entirely
within R — without a NONMEM license — while offering a faster
iteration cycle and tighter integration with modern R workflows.

This article focuses specifically on the **simulation** use case.  For
parameter estimation, NONMEM can be used; rxode2 models
can be used for estimation via nlmixr2.

---

## The simulation workflow: a side-by-side overview

### NONMEM simulation workflow

A typical NONMEM simulation involves:

1. Writing or reusing a control stream with a `$SIMULATION` block.
2. Preparing an input dataset (NONMEM-format `.csv`).
3. Running NONMEM from the command line or via PsN/Wings.
4. Parsing the output `$TABLE` file back into R for analysis.

A minimal NONMEM simulation control stream looks like:

```
$PROB One-compartment PK simulation
$INPUT ID TIME AMT EVID CMT DV
$DATA  sim_input.csv IGNORE=@
$SUBROUTINE ADVAN2 TRANS2

$PK
  TVCL = THETA(1)
  TVV  = THETA(2)
  TVKA = THETA(3)
  CL   = TVCL * EXP(ETA(1))
  V    = TVV  * EXP(ETA(2))
  KA   = TVKA
  S2   = V

$ERROR
  IPRED = F
  W     = SIGMA(1,1)
  Y     = IPRED + W * EPS(1)

$THETA 4 20 1
$OMEGA 0.09 0.09
$SIGMA 0.04

$SIMULATION (12345) ONLYSIM SUBPROBLEM=200

$TABLE ID TIME IPRED Y NOPRINT ONEHEADER FILE=simtab.csv
```

The result is parsed back into R manually:

```{r}
sim_result <- read.csv("simtab.csv", skip = 1)
```

### rxode2 simulation workflow

The same simulation in rxode2 stays entirely within R:

```{r}
library(rxode2)

mod <- rxode2({
  KA  <- exp(tka)
  CL  <- exp(tcl + eta.cl)
  V   <- exp(tv  + eta.v)
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
  cp_obs      <- cp * (1 + prop.err)
})

theta <- c(tka = log(1), tcl = log(4), tv = log(20))
omega <- lotri(eta.cl ~ 0.09, eta.v ~ 0.09)
sigma <- matrix(0.04)

ev <- et(amt = 100, ii = 24, addl = 3) |>
  et(seq(0, 96, by = 1))

sim_result <- rxSolve(mod, ev,
                      params  = theta,
                      omega   = omega,
                      sigma   = sigma,
                      nSub    = 200,
                      nStud   = 1)
```

No external process, no file I/O, no parsing step.

---

## Features rxode2 has that direct NONMEM simulation does not

### No license required

rxode2 is open-source (GPL >= 3) and runs on any machine with R and a
C compiler — a laptop, a CI server, a Docker container, or a Shiny
app.  NONMEM requires a commercial license and a configured executable,
which typically restricts simulations to specific workstations or
servers.

### Interactive iteration speed

rxode2 compiles the model once (result is cached) and solves in-process
thereafter.  Iterating on parameters, event schedules, or sample sizes
takes milliseconds to seconds.  Each NONMEM simulation run incurs
startup, I/O, and table-writing overhead — typically seconds to
minutes even for simple models — making rapid exploration impractical.

```{r}
# Instant parameter sweep — no file I/O, no process launch
rxSolve(mod, ev, params = c(tka = log(1), tcl = log(4), tv = log(20)),
        omega = omega, nSub = 200)
rxSolve(mod, ev, params = c(tka = log(1), tcl = log(6), tv = log(20)),
        omega = omega, nSub = 200)
```

### OpenMP parallelism across subjects

rxode2's LSODA solver is thread-safe C, enabling parallel ODE solving
across subjects via OpenMP within a single R process.  Simulating
10,000 subjects uses all available cores automatically:

```{r}
getRxThreads()   # inspect active thread count

rxSolve(mod, ev, params = theta, omega = omega, nSub = 10000)
```

NONMEM's `$SIMULATION` step solves subjects sequentially within a
single run.  Parallelism requires submitting multiple NONMEM runs to
an HPC cluster (e.g., via PsN).

### Native R workflow — no file I/O

Simulation inputs (parameters, event tables, covariate datasets) and
outputs (data frames) are ordinary R objects.  Results pipe directly
into ggplot2, vpc, tidyverse, or any other R package:

```{r}
library(vpc)

sim_result |>
  vpc(obs_data = obs_data,
      obs_cols  = list(idv = "time", dv = "cp_obs"),
      sim_cols  = list(idv = "time", dv = "cp_obs"))
```

With NONMEM, each step — preparing the input CSV, running the model,
reading the table file — requires explicit file handling.

### Model piping for scenario analysis

rxode2 model objects support the native R pipe (`|>`) to clone and
modify a model without rewriting the control stream:

```{r}
# Add a covariate effect without rewriting the full model
mod_wt <- mod |>
  model({ CL <- exp(tcl + eta.cl + 0.75 * log(WT / 70)) },
        cov = "WT")

# Fix a parameter for a sensitivity run
mod_fixed_ka <- mod |> ini(tka = log(0.5), fix(tka))
```

In NONMEM, each scenario requires editing the control stream and
re-running.

### Residual error inside the model

Stochastic residual error (additive, proportional, combined, or
custom) is specified directly in the rxode2 model block alongside the
structural model.  The simulated and predicted values are both returned
in the same output data frame:

```{r}
mod_err <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
  cp_obs      <- cp * (1 + prop.err) + add.err   # combined error
})
```

### Importing a NONMEM model via nonmem2rx

`nonmem2rx` converts an existing NONMEM control stream into an rxode2
model and automatically validates it by comparing PRED/IPRED/IWRES
against the original NONMEM output:

```{r}
library(nonmem2rx)

mod <- nonmem2rx("run001.ctl")   # imports + auto-validates
```

After a passing validation, the rxode2 model can be used for all
subsequent simulations without re-running NONMEM.  Individual EBEs and
the variance-covariance matrix are imported alongside the model for
EBE-based and uncertainty simulations:

```{r}
# Simulate with parameter uncertainty from the NONMEM covariance step
rxSolve(mod, ev, thetaMat = mod$thetaMat, nStud = 500)

# Simulate known subjects with their individual ETAs
rxSolve(mod, ev, iCov = mod$etaData)
```

### Analytical 1–3 compartment solutions with gradients

`linCmt()` provides closed-form one-, two-, and three-compartment
solutions with Stan math auto-differentiation gradients — equivalent
to ADVAN1/2/3/4/11/12 but usable for both simulation and nlmixr2
estimation:

```{r}
mod_lincmt <- function() {
  ini({
    TCl <- 4;  eta.Cl ~ 0.09
    V   <- 10
    KA  <- 1
  })
  model({
    CL  <- TCl * exp(eta.Cl)
    cp  <- linCmt()
  })
}
```

### Multiple ODE solver backends

rxode2 provides three solver backends selectable per call:
thread-safe C LSODA (default), Fortran LSODA, and DOP853 (explicit
8th-order Runge-Kutta for non-stiff systems).  NONMEM's solver is
fixed to the ADVAN specified in the control stream.

```{r}
rxSolve(mod, ev, method = "dop853")   # non-stiff, fast
rxSolve(mod, ev, method = "lsoda")    # default adaptive solver
```

### Reproducible seeds in R

Random seeds for stochastic simulations are set with standard R
conventions and are fully reproducible across platforms:

```{r}
set.seed(12345)
rxSolve(mod, ev, params = theta, omega = omega, nSub = 200)
```

NONMEM's `$SIMULATION (seed)` syntax is platform-dependent: the same
seed may produce different results across NONMEM versions or operating
systems.

---

## Features NONMEM simulation has that rxode2 does not

### Exact fidelity to the estimation model

When you simulate in NONMEM using the same control stream as
estimation, there is zero risk of transcription error — the same
`$PK`, `$DES`, and `$ERROR` code runs in both contexts.

rxode2 requires the model to be expressed in its own mini-language or
imported via `nonmem2rx`.  `nonmem2rx` performs an automatic
PRED/IPRED/IWRES comparison against the original NONMEM output to
detect any discrepancies, but complex or heavily customised NONMEM
models may require manual adjustment when the automated check does not
pass.

### Full NONMEM ADVAN coverage

NONMEM ships ADVAN1–ADVAN13, covering all standard PK compartmental
models, steady-state solutions (ADVAN3 SS, ADVAN4 SS), and general
nonlinear ODE solving (ADVAN6, ADVAN8, ADVAN9).  rxode2 covers the
most common models but does not replicate every ADVAN:

| NONMEM ADVAN | Description | rxode2 equivalent |
|---|---|---|
| ADVAN1 | 1-cmt IV | `linCmt()` or ODE |
| ADVAN2 | 1-cmt oral | `linCmt()` or ODE |
| ADVAN3 | 2-cmt IV | `linCmt()` or ODE |
| ADVAN4 | 2-cmt oral | `linCmt()` or ODE |
| ADVAN5 | general linear | ODE |
| ADVAN6 | general nonlinear ODE | ODE |
| ADVAN7 | general linear (SS) | ODE (SS via event table) |
| ADVAN8 | general nonlinear ODE (SS) | ODE |
| ADVAN9 | general nonlinear ODE + SS | ODE |
| ADVAN10 | 1-cmt Michaelis-Menten | ODE |
| ADVAN11 | 3-cmt IV | `linCmt()` or ODE |
| ADVAN12 | 3-cmt oral | `linCmt()` or ODE |
| ADVAN13 | general nonlinear ODE (stiff) | ODE |

### `$MIXTURE` population mixture models

NONMEM's `$MIXTURE` block supports mixture distributions (discrete
subpopulations with different parameter distributions and associated
probabilities) as a first-class feature.  `nonmem2rx` can import
mixture models, but the automatic PRED/IPRED/IWRES validation is
skipped for mixture models, so correctness must be verified manually.

### Verbatim Fortran/C in model code

NONMEM allows arbitrary Fortran (or C via the `$SUBROUTINE` mechanism)
in `$PK`, `$DES`, and `$ERROR` blocks, enabling highly customised
models.  rxode2 supports custom C via `rxFun()` and `.extraC()` but
does not allow embedding arbitrary Fortran.

### Cluster submission via PsN

PsN (Perl-speaks-NONMEM) can distribute NONMEM simulation runs across
HPC cluster nodes.  rxode2's parallelism is always within a single R
process via OpenMP; distributed computing requires additional R
infrastructure (e.g., `future`, `batchtools`).

### Steady-state dosing (`SS=1`, `II`)

Both tools compute exact steady-state solutions for linear models.
NONMEM uses `SS=1` and `II` columns in the event record; rxode2
uses the same `SS` and `II` columns in a NONMEM-format dataset or the
`et()` event table.  For non-linear models, both tools simulate
repeated dosing numerically until convergence.

---

## Feature comparison: shared capabilities with different approaches

### NONMEM-compatible event datasets

Both tools accept NONMEM-format datasets with columns `ID`, `TIME`,
`AMT`, `EVID`, `CMT`, `RATE`, `II`, `ADDL`, `SS`.  rxode2 also
supports its own extended EVID codes (5 = replace, 6 = multiply,
7 = phantom/transit) with no NONMEM equivalent.

### Between-subject variability (omega)

Both tools sample ETAs from an omega matrix for population simulation.
rxode2 accepts the omega matrix directly in `rxSolve()`; NONMEM
specifies it in `$OMEGA` and draws automatically during `$SIMULATION`.

### Residual variability (sigma)

Both tools apply residual error via a sigma specification.  rxode2
specifies the error model inside the `model({})` block; NONMEM uses
`$ERROR` and `$SIGMA`.

### Inter-occasion variability (IOV)

Both tools support IOV.  In rxode2 it is specified in the `ini({})`
block with the `| occ` condition; NONMEM uses multiple `$OMEGA BLOCK`
entries keyed to an occasion variable.

### VPC workflows

Both tools are used for VPC generation.  Results from either can be
passed to the `vpc` R package or Xpose for VPC plots.

---

## Summary table

| Feature | rxode2 | NONMEM `$SIMULATION` |
|---------|--------|----------------------|
| License | open-source (GPL >= 3) | commercial license required |
| Simulation engine | compiled C / Fortran in-process | NONMEM external process |
| Simulation speed | sub-second to seconds | seconds to minutes per run |
| OpenMP within-process parallelism | yes | no (sequential subjects) |
| Cluster submission | via R (`future`, etc.) | via PsN |
| Model re-implementation needed | yes (or `nonmem2rx`) | no |
| Re-implementation validation | `nonmem2rx` auto-checks PRED/IPRED/IWRES | N/A |
| Full ADVAN coverage | partial (most common models) | yes (ADVAN1–13) |
| `$MIXTURE` models | partial (`nonmem2rx` import, no auto-validation) | yes |
| Verbatim Fortran in model | no | yes |
| Analytical 1–3 cmt solutions | `linCmt()` with gradients | ADVAN1/2/3/4/11/12 |
| Model piping / scenario analysis | yes (`\|>`) | edit control stream + re-run |
| Native R output (data frame) | yes | parse table file |
| File I/O required | no | yes |
| Reproducible seeds cross-platform | yes (R RNG) | version/OS dependent |
| NONMEM-format datasets | yes | yes |
| Extended EVIDs (replace, multiply) | yes (EVID=5/6/7) | no |
| EBE-based simulation | yes (`nonmem2rx` imports EBEs) | yes |
| Parameter uncertainty simulation | yes (`nonmem2rx` imports covariance) | yes (`$COVARIANCE`) |
| nlmixr2 estimation | yes | no (separate estimation step) |
| Symbolic Jacobians | yes | no |
| Multiple ODE solver backends | LSODA (C), LSODA (Fortran), DOP853 | fixed per ADVAN |
| Shiny / portable deployment | yes | no (requires NONMEM install) |

---

## References

- [NONMEM user guides](https://nonmem.iconplc.com/)
- [PsN (Perl-speaks-NONMEM)](https://uupharmacometrics.github.io/PsN/)
- [nonmem2rx package](https://nlmixr2.github.io/nonmem2rx/)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nlmixr2 package](https://nlmixr2.github.io/nlmixr2/)
- [vpc R package](https://cran.r-project.org/package=vpc)
