---
title: "rxode2 and deSolve: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

[deSolve](https://cran.r-project.org/package=deSolve) and
[rxode2](https://nlmixr2.github.io/rxode2/) are both R packages for
solving ordinary differential equations, but they occupy quite
different niches.

- **deSolve** is a general-purpose ODE/DAE/DDE solver library that
  accepts ODE right-hand sides written as plain R functions (or
  optionally compiled C/Fortran).  It exposes a broad collection of
  solvers (LSODA, LSODE, LSODES, vode, daspk, radau, rkF45, …) and is
  the de-facto standard for ODE solving in R across ecology, chemistry,
  epidemiology, and many other fields.

- **rxode2** is purpose-built for pharmacometric simulation and
  parameter estimation.  It parses a domain-specific mini-language into
  C code, compiles it, and solves with a thread-safe C LSODA
  implementation accelerated by OpenMP across subjects.  It integrates
  with nlmixr2 for population parameter estimation and accepts
  NONMEM-format event datasets directly.

Both are on CRAN.  deSolve is BSD-licensed; rxode2 is GPL (≥ 3).

---

## Shared capabilities

- Adaptive-step stiff/non-stiff ODE solving (LSODA family)
- Support for discontinuities / event handling
- Multiple solver algorithms
- Ability to use compiled C or Fortran for the RHS
- Output returned as matrix or data frame

---

## Features rxode2 has that deSolve does not

### Domain-specific mini-language compiled to C

rxode2 accepts a compact, readable notation that is automatically
compiled to C before solving.  No manual C or Fortran is required for
fast execution.

```{r}
library(rxode2)

mod <- rxode2({
  KA   <- exp(tka + eta.ka)
  CL   <- exp(tcl + eta.cl)
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})
```

In deSolve the equivalent is an R function (slow by default) or a
hand-written C function passed to `deSolve::ode()`:

```{r}
library(deSolve)

pk_ode <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    KA <- exp(tka)
    CL <- exp(tcl)
    list(c(
      ddepot = -KA * depot,
      dcentr =  KA * depot - CL / V * centr
    ))
  })
}
```

### NONMEM-compatible event datasets

rxode2 accepts standard NONMEM-format data frames (columns `ID`,
`TIME`, `AMT`, `EVID`, `CMT`, `RATE`, `II`, `ADDL`, etc.) directly.
Population datasets with hundreds of subjects and complex dosing
histories can be passed in a single call.

```{r}
# A NONMEM-style dataset works directly
rxSolve(mod, data = nm_dataset)
```

deSolve has no built-in concept of dosing events or population
datasets.  Dose administration must be implemented manually, either by
re-starting the integrator at each dose time or by using
`deSolve::events`.

### Pharmacometric event table (`et()`)

rxode2's `et()` function provides a fluent API for constructing dosing
and sampling schedules without writing a raw data frame:

```{r}
ev <- et(amt = 100, ii = 24, addl = 6) |>   # 7 doses, q24 h
  et(seq(0, 168, by = 1))                    # hourly observations

rxSolve(mod, ev, params = c(tka = -0.5, tcl = 1, V = 10))
```

deSolve supports events via a separate `events` list argument, but it
has no high-level API for common regimens (multiple dosing, addl,
infusions) and no integration with a population dataset structure.

### OpenMP parallelism across subjects

rxode2's LSODA solver is implemented in thread-safe C, allowing genuine
parallel ODE integration across subjects within a single R process using
OpenMP.  A 1000-subject simulation automatically uses all available
cores with no extra code.

```{r}
rxSolve(mod, ev, omega = omega_mat, nSub = 1000)  # parallel by default
```

deSolve integrates a single trajectory per call.  Parallelism across
subjects must be implemented by the user, typically with `parallel`,
`future`, or `foreach`.

### Between-subject variability and population simulation

rxode2 natively draws random effects from an omega matrix, applies them
per subject, and returns a combined data frame ready for analysis:

```{r}
omega <- lotri(eta.ka ~ 0.09, eta.cl ~ 0.09)

sim <- rxSolve(mod, ev,
               params  = c(tka = -0.5, tcl = 1, V = 10),
               omega   = omega,
               nSub    = 200)
```

deSolve has no concept of omega matrices, random effects, or
subjects.  Population simulation requires looping over subjects,
sampling etas manually, and combining output by hand.

### Residual error simulation

rxode2 supports additive, proportional, and combined residual error
models specified directly inside the model block, with the result
returned alongside the deterministic prediction:

```{r}
mod_err <- rxode2({
  KA <- exp(tka + eta.ka)
  CL <- exp(tcl + eta.cl)
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
  cp_obs      <- cp * (1 + prop.err) + add.err
})
```

deSolve returns only the deterministic ODE trajectory; residual error
must be added as a post-processing step.

### Analytical 1–3 compartment PK solutions with exact gradients

`linCmt()` provides closed-form solutions for one-, two-, and
three-compartment pharmacokinetic models.  Gradients are computed via
Stan math auto-differentiation, enabling exact gradient-based
estimation in nlmixr2.

```{r}
mod_lincmt <- function() {
  ini({
    TCl <- 4;  eta.Cl ~ 0.09
    V   <- 10
  })
  model({
    CL <- TCl * exp(eta.Cl)
    cp <- linCmt()   # 1-cmt analytical solution
  })
}
```

deSolve has no analytical PK solutions.

### Symbolic Jacobians and sensitivity equations

rxode2 can automatically derive the symbolic Jacobian of the ODE
system and forward-sensitivity equations, which are used by nlmixr2's
FOCEi algorithm for exact gradient-based population parameter
estimation.  deSolve does not support symbolic differentiation.

### Model piping for incremental modification

rxode2 model objects support R's native pipe (`|>`) to copy and
modify a model without rewriting the full specification:

```{r}
base <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})

full <- base |>
  model({ KA <- exp(tka + eta.ka); CL <- exp(tcl + eta.cl) },
        append = FALSE) |>
  ini({ tka <- log(0.5); eta.ka ~ 0.09; tcl <- log(4); eta.cl ~ 0.09; V <- 10 })
```

deSolve ODE functions are plain R functions; there is no model-object
API to pipe or modify.

### nlmixr2 integration for parameter estimation

The same rxode2 model object used for simulation can be passed
directly to nlmixr2 for population parameter estimation (FOCE, FOCEi,
SAEM, Laplace, etc.).  deSolve is a simulation-only package; parameter
estimation requires third-party tools (e.g., `FME`, manual wrappers).

```{r}
library(nlmixr2)

fit <- nlmixr(mod_lincmt, theo_sd, est = "focei")
```

### Phantom / transit dose event type

rxode2 supports EVID=7 for phantom / transit doses — a concept with no
direct equivalent in deSolve.

### deSolve-format event data frames accepted by rxode2

Both deSolve and rxode2 support replace and multiply state events, and
rxode2 can accept deSolve-style event data frames directly.  deSolve
uses a data frame with columns `var`, `time`, `value`, and `method`
(values `"add"`, `"mult"`, `"rep"`):

```{r}
eventdat <- data.frame(
  var    = c("v1", "v2", "v2", "v1"),
  time   = c(1,    1,    5,    9),
  value  = c(1,    2,    3,    4),
  method = c("add", "mult", "rep", "add")
)
```

rxode2 translates this automatically to its internal EVID codes
(`"add"` → EVID=1, `"mult"` → EVID=6, `"rep"` → EVID=5) via
`etTrans()`, so the same event data frame can be used with either
package:

```{r}
# deSolve usage
deSolve::ode(y0, times, func = derivs, parms = NULL,
             events = list(data = eventdat))

# rxode2 usage — accepts the same data frame via et()
e <- et(eventdat) |> et(seq(0, 10, by = 0.1))
rxSolve(mod, e)
```

The rxode2 EVID equivalents are:

| rxode2 EVID | deSolve method | Meaning |
|-------------|----------------|---------|
| 1  | `"add"` | Add value to compartment |
| 5  | `"rep"` | Replace compartment amount |
| 6  | `"mult"` | Multiply compartment amount |
| 7  | (none)  | Phantom / transit dose |

### Time-after-dose tracking

rxode2 provides built-in functions for tracking time after the most
recent dose — `tad()`, `tad0()`, `tad(cmt)`, and `tad0(cmt)` — usable
directly inside model equations.  deSolve has no equivalent; the user
must track this manually.

---

## Features deSolve has that rxode2 does not

### Large solver library

deSolve ships the broadest selection of ODE/DAE/DDE solvers available
in R:

| Solver | Type |
|--------|------|
| `lsoda` | adaptive stiff/non-stiff |
| `lsode` | stiff (backward diff. / Adams) |
| `lsodes` | stiff, sparse Jacobian |
| `lsodar` | lsoda + root finding |
| `vode` | variable-coefficient stiff |
| `daspk` | differential-algebraic |
| `radau` | implicit Runge-Kutta, stiff |
| `rkF45` | explicit Runge-Kutta, non-stiff |
| `rk4` | classic fixed-step RK4 |
| `euler` | explicit Euler (teaching) |

rxode2 offers three backends: thread-safe C LSODA (default), Fortran
LSODA, and DOP853.  Most pharmacometric models are stiff and well
served by LSODA, but rxode2 does not currently offer sparse Jacobian
solvers, DAE solvers, or delay-differential equation solvers.

### Delay differential equations (DDEs)

deSolve's `dede()` function solves delay differential equations where
the RHS depends on the history of the state vector.  rxode2 does not
support DDEs.

```{r}
# deSolve only
dde_model <- function(t, y, parms, lag) {
  list(c(-parms["r"] * y[1] * (1 - lagvalue(t - lag, 1) / parms["K"])))
}

deSolve::dede(y0, times, func = dde_model, parms = parms,
              lag = 1.5, method = "dde23")
```

### Differential-algebraic equations (DAEs)

`daspk()` solves DAE systems where some equations are algebraic
constraints rather than differential equations.  rxode2 supports
algebraic (ODE-free) models but does not solve mixed ODE/algebraic
systems as a DAE.

### Sparse Jacobian support

`lsodes()` exploits a sparse Jacobian structure, important for large
systems (hundreds of state variables) where the Jacobian is mostly
zero.  rxode2's solvers use dense Jacobians.

### Root finding / event detection

`lsodar()` finds roots of user-defined functions during integration,
enabling automatic detection of threshold-crossing events (e.g., a
state variable hitting zero).  rxode2 event handling is based on
pre-specified event times rather than continuous root finding.

### Complete solver control

deSolve exposes every solver tuning parameter (absolute and relative
tolerances per variable, maximum step count, Jacobian method, band
widths for banded Jacobians, etc.) directly to the user.  rxode2
exposes the most commonly needed options via `rxControl()` but does
not expose solver internals to the same degree.

### General-purpose use outside pharmacometrics

deSolve imposes no domain-specific structure.  Any system of ODEs —
ecological, chemical, mechanical, epidemiological — can be solved by
writing a function with the signature `f(t, y, parms)`.  rxode2's
feature set (dosing events, omega matrices, NONMEM datasets) is
designed specifically for pharmacometric workflows and is unnecessary
overhead for general ODE work.

### Solving a model in pure R (no compilation)

deSolve can solve any R-function ODE right away — no compilation step.
This makes it simple for quick explorations, teaching, or prototyping
where compilation latency matters:

```{r}
# deSolve: immediately solvable, no compilation
out <- deSolve::ode(
  y     = c(depot = 100, centr = 0),
  times = seq(0, 24, by = 0.5),
  func  = pk_ode,
  parms = c(tka = -0.5, tcl = 1, V = 10)
)
```

rxode2 compiles the model to C on first use (results are cached), so
there is a one-time latency before the first solve.

---

## Performance comparison

For a single-subject, single-trajectory solve, deSolve with a compiled
C RHS and rxode2 are broadly comparable.  The main performance
differences arise at scale:

| Scenario | deSolve | rxode2 |
|----------|---------|--------|
| Single trajectory, R RHS | baseline | ~10–100× faster (compiled C) |
| Single trajectory, C RHS | fast | comparable |
| Population (n = 1000) | requires manual parallelism | parallel by default (OpenMP) |
| Iterative estimation calls | no built-in support | fast (pre/post overhead separated) |

For large population simulations or when the model will be used for
parameter estimation, rxode2 is substantially faster in practice due to
compiled C code and OpenMP parallelism.

---

## Summary table

| Feature | rxode2 | deSolve |
|---------|--------|---------|
| License | GPL ≥ 3 | BSD |
| CRAN | yes | yes |
| Model language | compiled C (mini-language) | R function (or C/Fortran) |
| Compilation step | yes (cached) | no (for R RHS) |
| LSODA (thread-safe C) | yes | yes (Fortran) |
| DOP853 solver | yes | yes (`rkF45` / `ode45`) |
| DAE solver | no | yes (`daspk`) |
| DDE solver | no | yes (`dede`) |
| Sparse Jacobian solver | no | yes (`lsodes`) |
| Root finding / threshold | no | yes (`lsodar`) |
| OpenMP parallelism | yes (built-in) | no (manual) |
| Population simulation (IIV) | yes (omega matrices) | no (manual) |
| Residual error simulation | yes (in model) | no (manual) |
| NONMEM-format datasets | yes | no |
| `et()` dosing API | yes | no |
| Multiple dosing (addl, ii) | yes | manual |
| Infusion rate/duration | yes | manual |
| Analytical 1–3 cmt PK | `linCmt()` | no |
| Symbolic Jacobians | yes | no |
| Parameter estimation | via nlmixr2 | via `FME` or manual |
| Model piping (`\|>`) | yes | no |
| NONMEM model import | `nonmem2rx` | no |
| Time-after-dose | `tad()`, `tad(cmt)` | manual |
| Replace / multiply state events | EVID=5 / EVID=6 | `method = "rep"` / `"mult"` |
| deSolve-format event data frames | yes (via `etTrans()`) | yes (native) |
| EVID=7 phantom / transit dose | yes | no |

---

## References

- [deSolve CRAN page](https://cran.r-project.org/package=deSolve)
- [deSolve vignette: "Solving Initial Value Differential Equations in R"](https://cran.r-project.org/web/packages/deSolve/vignettes/deSolve.pdf)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nlmixr2 package](https://nlmixr2.github.io/nlmixr2/)
- [FME: Flexible Modelling Environment](https://cran.r-project.org/package=FME)
