---
title: "Separating phase 1 and phase 3 residual error"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
```

## Why one residual error model is often not enough

A pooled population PK analysis usually combines studies that were not
run under the same conditions.  Phase 1 data come from rich sampling
in healthy volunteers at a single site, with a validated assay and
controlled sample handling.  Phase 3 data come from sparse sampling in
patients across many sites, with nominal rather than recorded times,
more variable sample handling, and sometimes a different assay
version.

The structural model is the same for both -- the drug does not change
between phases -- but the *noise around the prediction* is not.  Fitting
one residual error model to the pooled data splits the difference: it
overstates the noise on the phase 1 observations and understates it on
the phase 3 ones.  Anything that reads the residual error then
inherits that compromise, including the width of a simulated
prediction interval and the weighting each observation gets during
estimation.

The fix is to keep one prediction and give it two residual error
models, one per phase.

## Two endpoints on one prediction

An `rxode2` error line may be named with `| <name>` after the error
expression.  The name is the endpoint, and the same model variable may
be used by more than one of them:

```{r model}
pk <- function() {
  ini({
    tka <- 0.45; label("Log Ka")
    tcl <- 1;    label("Log Cl")
    tv  <- 3.45; label("Log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    phase1.sd <- 0.2; label("Phase 1 additive error (mg/L)")
    phase3.sd <- 0.7; label("Phase 3 additive error (mg/L)")
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(phase1.sd) | phase1
    cp ~ add(phase3.sd) | phase3
  })
}

pk <- pk()
print(pk)
```

There is one `cp`, one set of ODEs and one set of structural
parameters.  What differs is which standard deviation is applied, and
that is chosen per observation record.

The `$multipleEndpoint` table shows how a record selects its endpoint:

```{r multipleEndpoint}
pk$multipleEndpoint
```

So an observation with `cmt="phase1"` (equivalently `dvid=1`) gets
`phase1.sd`, and one with `cmt="phase3"` (`dvid=2`) gets `phase3.sd`.

### What rxode2 does with the shared variable

Almost everything downstream of the model -- the simulated residual
draw, the residual parameter lookup, `ar()` state -- is keyed on the
endpoint's prediction variable.  Two endpoints named `cp` would
collide there, so `rxode2` gives each of them a generated variable of
its own and records the mapping:

```{r alias}
pk$endpointAlias
pk$predDf[, c("cond", "var", "dvid", "cmt")]
```

You never write those lines and they never appear in your `model({})`
block, in `modelExtract()`, or in the solved output.  They are only
worth knowing about because `$predDf$var` is where they show up.

Before this was supported the aliases had to be written by hand:

```r
cp.phase1 <- cp
cp.phase3 <- cp
cp.phase1 ~ add(phase1.sd) | phase1
cp.phase3 ~ add(phase3.sd) | phase3
```

That model still works and still means the same thing -- it is just no
longer necessary.

## Simulating each phase

Label the observation records with the phase they belong to.  A rich
phase 1 profile and a sparse phase 3 trough/peak design:

```{r events}
ph1 <- et(amt=100) |>
  et(seq(0.25, 24, by=0.5), cmt="phase1")

ph3 <- et(amt=100) |>
  et(c(1, 4, 12, 24), cmt="phase3")
```

Solving each design gives `ipredSim` (the individual prediction) and
`sim` (the prediction plus residual error):

```{r solve}
set.seed(42)
s1 <- rxSolve(pk, ph1, nSub=100, addDosing=FALSE)
s3 <- rxSolve(pk, ph3, nSub=100, addDosing=FALSE)
```

The residual spread differs by exactly the amount the model says it
should:

```{r residSd}
c(phase1 = sd(s1$sim - s1$ipredSim),
  phase3 = sd(s3$sim - s3$ipredSim))
```

```{r plotPhase, fig.width=7, fig.height=4}
d <- rbind(
  data.frame(phase="phase1", time=s1$time, resid=s1$sim - s1$ipredSim),
  data.frame(phase="phase3", time=s3$time, resid=s3$sim - s3$ipredSim))

ggplot(d, aes(x=resid, fill=phase)) +
  geom_density(alpha=0.5) +
  xlab("sim - ipredSim (mg/L)") +
  theme_bw()
```

### Both phases in one simulation

The two endpoints are just two compartments, so a single event table
can carry both.  This is what a pooled dataset looks like: the same
subject contributes rich early samples and sparse late ones, and each
record is scored against its own error model.

```{r bothPhases}
both <- et(amt=100) |>
  et(seq(0.25, 12, by=0.5), cmt="phase1") |>
  et(c(24, 36, 48), cmt="phase3")

set.seed(42)
sBoth <- as.data.frame(rxSolve(pk, both, nSub=50, addDosing=FALSE))

# CMT 3 is the phase1 endpoint and CMT 4 the phase3 endpoint, as
# $multipleEndpoint showed above
sBoth$phase <- ifelse(sBoth$CMT == 3, "phase1", "phase3")
tapply(sBoth$sim - sBoth$ipredSim, sBoth$phase, sd)
```

The two endpoints draw their residuals independently, so a record in
one phase never inherits the other phase's noise.

## Predicting a trial with one phase's noise

The most common reason to separate the two is to *predict* with only
one of them.  When you simulate a future phase 3 trial, the noise you
want on the simulated observations is the phase 3 noise -- using the
pooled value would give an interval that is too narrow, and using the
phase 1 value one that is far too narrow.

Because the endpoint is chosen per record, this is just a matter of
which compartment the observation records name:

```{r futureTrial, fig.width=7, fig.height=4}
future <- et(amt=100) |>
  et(seq(0, 48, by=1), cmt="phase3")

set.seed(1)
sFuture <- rxSolve(pk, future, nSub=500, addDosing=FALSE)

ci <- confint(sFuture, "sim", level=0.90)

plot(ci, ylab="Simulated concentration (mg/L)")
```

Swapping `cmt="phase3"` for `cmt="phase1"` in that event table gives
the same central tendency with the tighter phase 1 interval, which is
the right answer for predicting a replicate phase 1 study and the
wrong one for a phase 3 trial.

If you want the prediction with no residual error at all, use
`ipredSim`, which is the same for both endpoints.

## Different error structures per phase

The phases do not have to share an error *structure*, only a
prediction.  An assay that is precise near the LLOQ but proportional
higher up, versus one that is additive throughout, is written directly:

```{r mixedStructure}
pk2 <- pk |>
  model(cp ~ prop(phase3.prop) | phase3)

print(pk2)
```

Since both endpoints have the same left-hand side, the `| phase3` is
what tells piping which one to replace; without it the model piping
reports the available conditions instead of guessing.  The same name
removes an endpoint:

```{r dropPhase}
pk3 <- pk2 |>
  model(-phase3)

print(pk3)
```

Transformations are per endpoint too, so `cp ~ lnorm(phase3.sd) |
phase3` gives phase 3 a log-normal residual while phase 1 stays
additive.

## Estimating both from pooled data

Nothing here is simulation-only.  Give `nlmixr2` a pooled dataset with
a `dvid` (or `cmt`) column naming the phase and it estimates both
standard deviations at once:

```r
library(nlmixr2)

# `dat` is the pooled analysis dataset with a `dvid` column whose
# values are "phase1" and "phase3" on the observation records
fit <- nlmixr2(pk, dat, "focei")
fit$parFixedDf["phase1.sd", ]
fit$parFixedDf["phase3.sd", ]
```

Each observation is weighted by the residual error of the study it
came from, so the rich phase 1 profiles no longer inflate the phase 3
error and the sparse phase 3 records no longer inflate the phase 1
error.

## Notes

- The name after `|` is a compartment name.  In the data it can be
  supplied as a character `cmt` column, a character `dvid` column, or
  the numeric `dvid` shown by `$multipleEndpoint`.
- Endpoints sharing a prediction must each be named.  Two unnamed
  endpoints on one variable are rejected when the model is built,
  because no data record could tell them apart.
- The split is not limited to two, and not limited to study phase --
  assay version, matrix (plasma versus dried blood spot), and site are
  all handled the same way.
