---
title: "rxode2 and PKPDsim: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

Both [rxode2](https://nlmixr2.github.io/rxode2/) and
[PKPDsim](https://insightrx.github.io/PKPDsim/) are open-source R
packages for ODE-based pharmacokinetic-pharmacodynamic simulation.
They occupy a similar niche and share many capabilities, but differ in
model language, solver design, and intended use cases.

- **rxode2** uses its own domain-specific mini-language (`d/dt(cmt)`),
  compiles it to C, and solves using a thread-safe C LSODA
  implementation with OpenMP parallelism across subjects.  It
  integrates tightly with nlmixr2 for parameter estimation via
  symbolic Jacobians.

- **PKPDsim** uses C++ array notation (`dAdt[1] = ...`) for ODE
  definitions, compiled via `new_ode_model()`, and solves using the
  Boost::odeint C++ library.  It separates a fast inner solver
  (`sim_core()`) from the full-featured outer function (`sim()`) and
  has a built-in literature model library and model-package export
  system.  It was developed by InsightRX with a focus on clinical dose
  individualization.

Both are on CRAN and are MIT or similarly permissive open-source
licenses.

## Shared capabilities

- ODE-based PK/PD simulation
- Between-subject variability (BSV / IIV)
- Residual error models
- Time-varying covariates
- Population simulation across many subjects
- Complex dosing regimens (bolus, infusion, oral)
- Integration with nlmixr2 for parameter estimation
- C/C++ compiled model code for speed

## Features rxode2 has that PKPDsim does not

### Readable domain-specific model language

rxode2 uses a natural ODE notation that closely mirrors mathematical
writing:

```{r}
library(rxode2)

mod <- rxode2({
  KA   <- exp(tka + eta.ka)
  CL   <- exp(tcl + eta.cl)
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})
```

PKPDsim uses C++ array indexing (`A[1]`, `dAdt[1]`) which is more
verbose and requires familiarity with zero-based vs one-based indexing
conventions:

```{r}
library(PKPDsim)

mod <- new_ode_model(
  code = "
    dAdt[1] = -KA * A[1];
    dAdt[2] =  KA * A[1] - (CL/V) * A[2];
  ",
  obs  = list(cmt = 2, scale = "V"),
  dose = list(cmt = 1)
)
```

### NONMEM-compatible event datasets

rxode2 accepts standard NONMEM-format data frames (with `EVID`, `AMT`,
`CMT`, `TIME`, `ID`, `RATE`, etc.) directly as event tables.  This
makes it straightforward to reuse datasets from NONMEM workflows or
produce datasets compatible with multiple tools.

PKPDsim uses its own `new_regimen()` API for dosing, which is clean
and simple but is not directly interchangeable with NONMEM-format
datasets.

### OpenMP parallelism across subjects

rxode2's LSODA solver is implemented in thread-safe C, enabling genuine
parallel ODE solving across subjects via OpenMP within a single R
process.  PKPDsim loops over subjects in R, calling the Boost::odeint
C++ solver for each subject sequentially.

As PKPDsim's own documentation notes, rxode2 is substantially faster
than PKPDsim in iterative contexts (e.g., population estimation or
optimal design) because rxode2 separates pre- and post-processing from
the solver by default, while PKPDsim's `sim()` includes this overhead
on every call.  PKPDsim's `sim_core()` mitigates this but requires the
user to manually manage the separation.

```{r}
# rxode2: pre/post processing is separated by default
# Iterative calls re-use the compiled model with minimal overhead
rxSolve(mod, et, params = p)   # fast in any context

# PKPDsim: use sim_core() for iterative use to avoid repeated overhead
design <- sim(mod, regimen = r, return_design = TRUE)
sim_core(design = design, ode = mod)   # fast inner loop only
```

### Model piping for incremental modification

rxode2 model objects support the native R pipe (`|>`) to clone and
modify a model without rewriting the full specification.

```{r}
base <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
  cp          <- centr / V
})

full <- base |>
  model({ KA <- exp(tka + eta.ka); CL <- exp(tcl + eta.cl) },
        append = FALSE) |>
  ini({ tka <- log(0.5); eta.ka ~ 0.09; tcl <- log(4); eta.cl ~ 0.09; V <- 10 }) |>
  model(cp ~ add(add.sd), append = TRUE) |>
  ini(add.sd = 0.5)
```

PKPDsim has no equivalent R-level model modification API; changes
require editing the `new_ode_model()` call.

### Symbolic Jacobians and exact gradients for estimation

rxode2 automatically derives the symbolic Jacobian and forward-
sensitivity equations used by nlmixr2's FOCEi algorithm for exact
gradient-based estimation.  These now extend to third-order
sensitivities and to *jump* (event) sensitivities -- exact derivatives
with respect to dosing parameters such as bioavailability, lag time,
infusion rate, duration, and amount.  PKPDsim's nlmixr2 integration uses
finite-difference gradients.

### Multiple ODE solver backends

rxode2 provides several solver backends selectable per run: thread-safe
C LSODA (default), Fortran LSODA, DOP853 (explicit 8th-order
Runge-Kutta for non-stiff systems), a stiff Rosenbrock method (`ros4`),
and AutoSwitch composite methods (for example `method = "dop853+ros4"`)
that probe each segment with a non-stiff primary and reactively fall
back to a stiff secondary only where stiffness is detected.  PKPDsim
uses a single Boost::odeint backend with a configurable step size.

### Delay differential equations

rxode2 solves delay differential equations through `delay(state, T)`,
which evaluates an ODE state at the past time `t - T` (Monolix `delay()`
semantics) using dense-output interpolation, with forward sensitivities
generated automatically so the models stay estimable in nlmixr2.
PKPDsim does not solve delay differential equations.

### Automatic ODE-to-`linCmt()` conversion

When a model's ODE system is a 1-3 compartment linear PK system, rxode2
detects this and transparently solves it with the fast analytical
`linCmt()` path (`rxSolve(..., useLinCmt = TRUE)`, the default), falling
back to the ODE solver if the converted model cannot be compiled.

### Out-of-memory simulation and distributed parallelism

For populations too large to hold in RAM, passing a `file` prefix to
`rxSolve()` splits subjects into RAM-sized chunks (auto-sized with
`rxMemoryEstimate()`), solves each, and writes them to Parquet (or
`.rds`) behind a lightweight object that supports lazy `dplyr`/DuckDB
queries so the full result is never materialized.  Chunks can be
dispatched to `mirai` daemons (`rxSolve(..., parallel = n)`) for
distributed/HPC execution, on top of the within-call OpenMP threading.

```{r}
rxSolve(mod, params, ev, file = tempfile("rx"), parallel = 4)
```

## Features available in both tools (with different approaches)

### 1-3 compartment analytical solutions

Both tools provide closed-form (non-ODE) solutions for 1-, 2-, and
3-compartment linear PK systems.  rxode2's `linCmt()` computes gradients
via Stan math auto-differentiation, giving nlmixr2's FOCEi exact
derivatives.  PKPDsim provides ADVAN-style analytical equations
(`sim(..., analytical = "...")`, based on Abuhelwa et al. 2015) for fast
closed-form solving.

### Literature model libraries

Both tools provide curated libraries of published PK/PD models, though
they are organized differently.

PKPDsim ships models as self-contained installable R packages via its
own distribution system:

```{r}
# See what models are available
available_default_literature_models()

# Install and use a literature model
install_default_literature_model("pk_busulfan_mccune")
library(pkbusulfanmccune)

sim(
  ode        = pkbusulfanmccune::model(),
  parameters = pkbusulfanmccune::parameters(),
  regimen    = new_regimen(amt = 100, n = 4, interval = 6)
)
```

rxode2 models from the literature are available via the `nlmixr2lib`
package (part of the nlmixr2 ecosystem), which provides a collection
of published models as rxode2 model functions ready for simulation or
estimation:

```{r}
library(nlmixr2lib)

# List available models
modeldb

# Load and simulate a model directly
mod <- modeldb$one.compartment
rxSolve(mod, et(amt = 100, time = 0) |> et(seq(0, 24, by = 1)))
```

### Inter-occasion variability (IOV)

Both tools support inter-occasion variability, with different syntax.

In rxode2, IOV is specified in the `ini({})` block using the `| occ`
condition syntax, then included in the model equation:

```{r}
mod <- function() {
  ini({
    tcl    <- log(4)
    eta.cl ~ 0.09          # IIV on CL
    iov.cl ~ 0.04 | occ    # IOV on CL, keyed by occasion variable `occ`
  })
  model({
    CL <- exp(tcl + eta.cl + iov.cl)
    d/dt(centr) <- -(CL / V) * centr
  })
}
mod <- rxode2(mod)
```

PKPDsim specifies IOV via the dedicated `iov` argument to
`new_ode_model()`:

```{r}
mod <- new_ode_model(
  code = "dAdt[1] = -(CL/V)*A[1];",
  iiv  = list(CL = "exp", V = "exp"),
  iov  = list(CL = "exp")
)
```

## Features PKPDsim has that rxode2 does not

### Model export as installable R packages

PKPDsim can export a compiled model as a proper R package (with
versioning, documentation, and default parameters) via the `package`
argument to `new_ode_model()`.  This makes it easy to share a specific
model version as a reproducible, installable unit.

```{r}
new_ode_model(
  code       = "dAdt[1] = -(CL/V)*A[1]",
  obs        = list(cmt = 1, scale = "V"),
  parameters = list(CL = 5, V = 50),
  package    = "myPKmodel",
  version    = "1.0.0"
)
```

rxode2 models can be included in packages (see the `rxUse()` vignette)
but there is no single-function model-package export workflow.

### Event-specific C++ hooks (`pk_code`, `dose_code`)

`new_ode_model()` accepts `pk_code` (C++ code executed at every event)
and `dose_code` (C++ code executed only at dose events).  These allow
model-specific logic -- such as parameter recomputation at dose time or
post-dose state resets -- to be embedded directly in the compiled model.

```{r}
mod <- new_ode_model(
  code      = "dAdt[1] = -(CL/V)*A[1];",
  pk_code   = "CL = TVCL * pow(WT/70, 0.75);",   # recompute at every event
  dose_code = "A[1] = A[1] + oral_dose_fraction * dose;"
)
```

rxode2 handles event-specific behavior through model equations and
the event table but does not expose separate C++ hooks at the event
level.

### Mixture models as a first-class argument

PKPDsim supports mixture models (subpopulations with different
parameter values and associated probabilities) directly via the
`mixture` argument to `new_ode_model()`:

```{r}
mod <- new_ode_model(
  code    = "dAdt[1] = -(CL/V)*A[1];",
  mixture = list(
    CL = list(values = c(2, 10), probability = 0.8)
  )
)
```

rxode2 supports within-model mixture via the `mix()` function, but it
is not a single dedicated argument at the model-definition level.

## Converting a PKPDsim model to rxode2 / nlmixr2

PKPDsim ships a built-in conversion function, `pkpdsim_to_nlmixr()`,
that translates a compiled PKPDsim model object into an nlmixr2-
compatible function with `ini({})` and `model({})` blocks.  Because
nlmixr2 uses rxode2 as its model representation layer, the resulting
object is simultaneously a valid rxode2 model that can be passed to
`rxSolve()` for simulation or to any nlmixr2 estimator for parameter
estimation.

```{r}
library(PKPDsim)
library(nlmixr2)

# 1. Build the PKPDsim model as usual
mod <- new_ode_model(
  code = "
    dAdt[1] = -KA * A[1];
    dAdt[2] =  KA * A[1] - (CL / V) * A[2];
  ",
  obs  = list(cmt = 2, scale = "V"),
  dose = list(cmt = 1)
)

# 2. Convert to an nlmixr2 / rxode2 function
f <- pkpdsim_to_nlmixr(
  model         = mod,
  parameters    = list(KA = 1, CL = 5, V = 50),
  omega         = c(0.1, 0.05, 0.1),   # lower-diagonal BSV matrix
  res_var       = list(prop = 0.1, add = 0.1),
  log_transform = TRUE
)
```

The returned object `f` is a plain R function containing `ini({})` and
`model({})` blocks.  It can be used directly for:

**Simulation with rxode2:**

```{r}
library(rxode2)

ev <- et(amt = 100, time = 0) |> et(seq(0, 24, by = 0.5))
rxSolve(f, ev)
```

**Population estimation with nlmixr2:**

```r
fit <- nlmixr2(f, data = your_dataset, est = "focei")
```

The `fixed` argument to `pkpdsim_to_nlmixr()` accepts a character
vector of parameter names to exclude from estimation (i.e. fix them at
their initial values).  The `ini_code`, `model_code`, and
`model_par_code` arguments allow manual overrides of the generated
nlmixr2 code blocks when the automatic translation needs adjustment.

See the [PKPDsim estimation with nlmixr2 vignette](https://insightrx.github.io/PKPDsim/articles/estimation-nlmixr.html)
and the [`pkpdsim_to_nlmixr()` reference page](https://insightrx.github.io/PKPDsim/reference/pkpdsim_to_nlmixr.html)
for further details.

## Summary table

| Feature | rxode2 | PKPDsim |
|---------|--------|---------|
| License | GPL >= 3 | MIT |
| CRAN | yes | yes |
| ODE language | rxode2 mini-language (`d/dt(cmt)`) | C++ array notation (`dAdt[1]`) |
| Solver | LSODA (C/Fortran), DOP853, ros4, `"dop853+ros4"` | Boost::odeint C++ |
| OpenMP parallelism | yes (thread-safe solver) | no (R loop over subjects) |
| Distributed / cluster parallelism | yes (`mirai`) | no |
| Out-of-memory / disk-backed solve | yes (`rxSolve(file=)`, Parquet/DuckDB) | no |
| Iterative-use speed | fast by default (separated pre/post) | use `sim_core()` explicitly |
| NONMEM-format datasets | yes | no (`new_regimen()` API) |
| Delay differential equations (`delay()`) | yes | no |
| Automatic ODE -> `linCmt()` conversion | yes (`useLinCmt=TRUE`) | no |
| 1-3 cmt analytical solution | `linCmt()` with exact gradients | ADVAN-style closed-form (`analytical=` in `sim()`), finite-diff gradients |
| Symbolic Jacobians / sensitivities | yes (up to 3rd order + jump sensitivities) | no (finite differences) |
| nlmixr2 estimation | yes (exact gradients via FOCEi) | yes (finite-difference gradients) |
| Model piping (`\|>`) | yes | no |
| Literature model library | yes (`nlmixr2lib`) | yes (`install_default_literature_model()`) |
| Export model as R package | via `rxUse()` (manual) | built-in (`package` argument) |
| `pk_code` / `dose_code` C++ hooks | no | yes |
| Inter-occasion variability (IOV) | yes (`iov.cl ~ 0.2 \| occ` in `ini({})`) | yes (`iov` argument) |
| Mixture models | `mix()` function | `mixture` argument |
| Time-varying covariates | yes | yes (`new_covariate()`) |
| Between-subject variability | yes | yes (`iiv` argument) |
| Residual error | yes | yes (`ruv` argument) |

## References

- [PKPDsim website](https://insightrx.github.io/PKPDsim/)
- [PKPDsim speed vignette](https://insightrx.github.io/PKPDsim/articles/speed.html)
- [PKPDsim estimation with nlmixr2](https://insightrx.github.io/PKPDsim/articles/estimation-nlmixr.html)
- [`pkpdsim_to_nlmixr()` reference](https://insightrx.github.io/PKPDsim/reference/pkpdsim_to_nlmixr.html)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nlmixr2 package](https://nlmixr2.github.io/nlmixr2/)
