---
title: "Providing Custom C/C++ Functions to rxode2 from a Package"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette explains how to write an R package that provides custom
C or C++ functions for use inside `rxode2` (and `nlmixr2`) models.
Compared to the [inline approach](Integrating-User-Defined-Functions-into-rxode2.html),
packaging the C code lets you:

- provide proper documentation, tests, and versioning for the
  functions that are not yet provided in `nlmixr2`/`rxode2`.
- supply exact analytic derivatives so that `nlmixr2` optimization
  works without falling back to finite differences.

The worked example implements a simple Michaelis-Menten
saturable-elimination function

$$
\text{mm}(C, V_{\max}, K_m) = \frac{V_{\max} \cdot C}{K_m + C}
$$

and its three partial derivatives.  The same pattern generalises to
any number of custom functions.

## How rxode2 calls compiled functions

When rxode2 compiles a model it generates a C source file.  Inside
that file, every external function the model uses is resolved at
load time with a call like

```c
mm = (rxode2_fn3) R_GetCCallable("mypkg", "mm");
```

`R_GetCCallable` is R's standard mechanism for C-level inter-package
calls.  For it to succeed, your package must register the symbol with
`R_RegisterCCallable` during its own initialisation.

The mapping from a model expression (`mm(Cp, Vmax, Km)`) to the
generated `R_GetCCallable` call is held in a data frame that rxode2
maintains at run-time.  You add a row to that data frame in `.onLoad`
using `rxode2::rxode2parseAssignTranslation()`.

## Package skeleton

```
mypkg/
├── DESCRIPTION
├── NAMESPACE
├── R/
│   └── zzz.R          # .onLoad / .onUnload hooks
└── src/
    ├── myfuns.c        # the numeric C functions
    └── init.c          # R_init_mypkg: registers callables
```

### DESCRIPTION

```
Package: mypkg
Version: 0.1.0
Imports: rxode2
LinkingTo: rxode2
```

`LinkingTo: rxode2` gives your C code access to the rxode2 headers,
including the `rxode2_fn*` typedef family used in the translation
table.

### NAMESPACE

```
useDynLib(mypkg, .registration = TRUE)
import(rxode2)
importFrom(rxode2, rxD, rxode2parseAssignTranslation, rxode2parseGetTranslation)
```

These can also be generated with roxygen2 comments:

```r
#' @useDynLib mypkg, .registration=TRUE
#' @import rxode2
#' @importFrom rxode2 rxD rxode2parseAssignTranslation rxode2parseGetTranslation
```

## Writing the C functions

### `src/myfuns.c`

```c
/* Standard R headers ---------------------------------------- */
#define STRICT_R_HEADERS
#include <R.h>
#include <Rinternals.h>
#include <R_ext/RS.h>

/* ----------------------------------------------------------
   Michaelis-Menten:  mm(C, Vmax, Km) = Vmax*C / (Km+C)
   ---------------------------------------------------------- */
double mm(double C, double Vmax, double Km) {
  return Vmax * C / (Km + C);
}

/* Partial derivatives --------------------------------------- */

/* d(mm)/dC   = Vmax*Km / (Km+C)^2 */
double mm_dC(double C, double Vmax, double Km) {
  double den = Km + C;
  return Vmax * Km / (den * den);
}

/* d(mm)/dVmax = C / (Km+C) */
double mm_dVmax(double C, double Vmax, double Km) {
  (void)Vmax;
  return C / (Km + C);
}

/* d(mm)/dKm  = -Vmax*C / (Km+C)^2 */
double mm_dKm(double C, double Vmax, double Km) {
  double den = Km + C;
  return -Vmax * C / (den * den);
}
```

Rules every rxode2-compatible C function must follow:

- Return type is `double`.
- All arguments are `double` scalars, not vectors.
- No R API calls inside the body — pure numeric C.

### `src/init.c` — registering the callables

```c
#include <R.h>
#include <Rinternals.h>
#include <R_ext/Rdynload.h>
#include <stdlib.h>   /* NULL */

/* Forward-declare functions from myfuns.c */
extern double mm      (double, double, double);
extern double mm_dC   (double, double, double);
extern double mm_dVmax(double, double, double);
extern double mm_dKm  (double, double, double);

void R_init_mypkg(DllInfo *dll) {
  /* R_registerRoutines makes symbols available to .C()/.Call() from R.
     This is optional for rxode2 integration but good practice. */
  static const R_CMethodDef cMethods[] = {
    {"mm",       (DL_FUNC) &mm,       3},
    {"mm_dC",    (DL_FUNC) &mm_dC,    3},
    {"mm_dVmax", (DL_FUNC) &mm_dVmax, 3},
    {"mm_dKm",   (DL_FUNC) &mm_dKm,  3},
    {NULL, NULL, 0}
  };
  R_registerRoutines(dll, cMethods, NULL, NULL, NULL);
  R_useDynamicSymbols(dll, FALSE);

  /* R_RegisterCCallable is REQUIRED for rxode2 integration.
     It makes each symbol retrievable from other packages' C code
     via R_GetCCallable("mypkg", "<name>").
     rxode2's generated model C files use exactly this mechanism. */
  R_RegisterCCallable("mypkg", "mm",       (DL_FUNC) &mm);
  R_RegisterCCallable("mypkg", "mm_dC",    (DL_FUNC) &mm_dC);
  R_RegisterCCallable("mypkg", "mm_dVmax", (DL_FUNC) &mm_dVmax);
  R_RegisterCCallable("mypkg", "mm_dKm",   (DL_FUNC) &mm_dKm);
}
```

`R_RegisterCCallable` and `R_registerRoutines` serve different
purposes:

| Function | Who uses it | Purpose |
|---|---|---|
| `R_registerRoutines` | R itself | `.C()` / `.Call()` from R code |
| `R_RegisterCCallable` | Other packages' compiled C | `R_GetCCallable(pkg, sym)` |

rxode2 uses `R_GetCCallable`, so `R_RegisterCCallable` is the
critical call for rxode2 integration.

## Registering the functions with rxode2

### `R/zzz.R`

``` r
## Cache the original translation table so .onUnload can restore it.
.origTranslation <- new.env(parent=emptyenv())
.origTranslation$value <- NULL


.onLoad <- function(libname, pkgname) {

  ## 1. Add a row to rxode2's function translation table.
  ##
  ##    Column meanings:
  ##      rxFun       - name used in rxode2 model code
  ##      fun         - C variable name in the generated model source
  ##      type        - rxode2 function-pointer typedef (call convention)
  ##      package     - first argument to R_GetCCallable()
  ##      packageFun  - second argument to R_GetCCallable()
  ##      argMin      - minimum number of accepted arguments
  ##      argMax      - maximum number of accepted arguments
  ##      threadSafe  - 1 if safe to call from OpenMP threads, 0 otherwise
  ##
  ##    Common function-pointer typedefs (rxode2_model_shared.h):
  ##      rxode2_fn    double (*)(double)
  ##      rxode2_fn2   double (*)(double, double)
  ##      rxode2_fn3   double (*)(double, double, double)
  ##      ...
  ##      // up to 10 args with development rxode2 (> 5.0.2)
  ##      rxode2_fn10  double (*)(double, double, ..., double)
  ##
  ##    mm() takes three double args  ->  type = "rxode2_fn3"

  .origTranslation$value <- rxode2::rxode2parseGetTranslation()

  .newRow <- data.frame(
    rxFun      = c("mm", "mm_dC", "mm_dVmax", "mm_dKm"),
    fun        = c("mm", "mm_dC", "mm_dVmax", "mm_dKm"),
    type       = c("rxode2_fn3", "rxode2_fn3", "rxode2_fn3", "rxode2_fn3"),
    package    = c("mypkg", "mypkg", "mypkg", "mypkg"),
    packageFun = c("mm", "mm_dC", "mm_dVmax", "mm_dKm"),
    argMin     = 3L,
    argMax     = 3L,
    threadSafe = 1L,
    stringsAsFactors = FALSE
  )

  rxode2::rxode2parseAssignTranslation(rbind(.origTranslation, .newRow))

  ## 2. Register analytic derivatives for nlmixr2 gradient computation.
  ##
  ##    rxD(name, list-of-derivative-functions)
  ##
  ##    Each function receives the same argument names as mm() and must
  ##    return a CHARACTER string containing the C expression for the
  ##    partial derivative with respect to that argument.

  rxode2::rxD("mm", list(

    ## d(mm)/dC
    function(C, Vmax, Km) {
      paste0("(", Vmax, "*", Km, ")/((", Km, "+", C, ")*(", Km, "+", C, "))")
    },

    ## d(mm)/dVmax
    function(C, Vmax, Km) {
      paste0("(", C, ")/(", Km, "+", C, ")")
    },

    ## d(mm)/dKm
    function(C, Vmax, Km) {
      paste0("-(", Vmax, "*", C, ")/((", Km, "+", C, ")*(", Km, "+", C, "))")
    }
  ))

  invisible()
}

.onUnload <- function(libpath) {
  ## Restore the original translation table.
  ## This is important: if the DLL is unloaded, any model that tries to
  ## call R_GetCCallable("mypkg", "mm") would crash.
  if (!is.null(.origTranslation$value)) {
    try(rxode2::rxode2parseAssignTranslation(.origTranslation$value), silent = TRUE)
  }
  ## Remove the derivative entry as well.
  try(rxode2::rxRmFun("mm"), silent = TRUE)
}
```

### How it works end-to-end

When a user writes `rxode2({ dA1/dt <- -mm(Cp, Vmax, Km) })`:

1. Once registered, rxode2 generates C code containing:
   ```c
   mm = (rxode2_fn3) R_GetCCallable("mypkg", "mm");
   ```
2. When the compiled model DLL is loaded, that call resolves to the
   `mm` function exported from `mypkg` via `R_RegisterCCallable`.
4. Calls to `mm()` inside the ODE system execute directly as C with
   no R overhead and full OpenMP thread safety.

## Using the custom function in a model

``` r
library(rxode2)
library(mypkg)   # triggers .onLoad -> rxode2parseAssignTranslation + rxD

mod <- rxode2({
  Cp     <- A1 / Vc
  dA1/dt <- -mm(Cp, Vmax, Km)   # Michaelis-Menten elimination
})

# Because you registered the mm() function to have 3 arguments, rxode2 will
# produce a syntax error if you try to call it with the wrong number of arguments:

try(rxode2({
  Cp     <- A1 / Vc
  dA1/dt <- -mm(Cp, Vmax, Km)   # Michaelis-Menten elimination
}))


ev  <- et(amt = 100) |> et(0, 24, by = 0.5)
sim <- rxSolve(mod, ev, params = c(Vc = 10, Vmax = 5, Km = 1))
plot(sim, Cp ~ time)
```

`nlmixr2` estimation works without additional changes because the
analytic derivatives registered via `rxD()` supply exact gradients.

## Verifying registration

```{r verify}
## Check that mm() is in the translation table
trans <- rxode2::rxode2parseGetTranslation()
trans[trans$rxFun == "mm", ]

## Check that derivatives are registered (should return C expressions,
## not NA or finite-difference fallbacks)
rxode2::rxFromSE("Derivative(mm(C, Vmax, Km), C)")
rxode2::rxFromSE("Derivative(mm(C, Vmax, Km), Vmax)")
rxode2::rxFromSE("Derivative(mm(C, Vmax, Km), Km)")
```

## Summary checklist

| Step | File | What to do |
|------|------|------------|
| 1 | `src/myfuns.c` | Write `double fun(double, ...)` functions |
| 2 | `src/init.c` | Call `R_RegisterCCallable("mypkg", "fun", ...)` for each function |
| 3 | `src/init.c` | Optionally call `R_registerRoutines` + `R_useDynamicSymbols` |
| 4 | `NAMESPACE` | `useDynLib(mypkg, .registration = TRUE)` |
| 5 | `R/zzz.R` | Call `rxode2parseAssignTranslation(rbind(old, newRow))` in `.onLoad` |
| 6 | `R/zzz.R` | Call `rxD("fun", list(...))` in `.onLoad` |
| 7 | `R/zzz.R` | Restore old translation table + call `rxRmFun` in `.onUnload` |

## Further reading

- [Integrating User Defined Functions into rxode2](Integrating-User-Defined-Functions-into-rxode2.html) -- the inline / R-function approach
- [Writing R Extensions S5.4](https://cran.r-project.org/doc/manuals/r-release/R-exts.html) -- `R_RegisterCCallable`, `R_GetCCallable`
- `?rxode2parseAssignTranslation`, `?rxode2parseGetTranslation`, `?rxD`
