---
title: "Sticky variables in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
```

In `rxode2`, variables defined on the left-hand side (LHS) of an equation are typically re-evaluated at every time point or integration step. However, there are scenarios where you want a variable to "stick" to its previous value unless a certain condition is met. This is particularly useful for calculating non-compartmental analysis (NCA) type metrics like $C_{max}$, $T_{max}$, $C_{min}$, and $T_{min}$ during a simulation.

## What are Sticky Variables?

A sticky variable is a variable that retains its value from the previous time point if it is not explicitly updated in the current time point. 

In `rxode2`, a variable becomes "sticky" if:

1.  It is checked with `is.na()` before being assigned.
2.  It is used in an expression before it is assigned in the current evaluation block.

When `rxode2` identifies a variable as sticky, it initializes it to `NA` for each individual and ensures its value persists across integration steps and time points.

## Calculating $C_{max}$ and $T_{max}$

To calculate $C_{max}$ and $T_{max}$, you need to keep track of the maximum concentration seen so far and the time it occurred. 

Here is how you can do this in an `rxode2` model:

```{r cmax_model}
m <- rxode2({
  v  <- 10
  cl <- 1
  ka <- 1
  d/dt(depot) <- -ka * depot
  d/dt(centr) <- ka * depot - cl/v * centr
  cp <- centr / v
  
  # Initialize sticky variables if they are NA
  if (is.na(C_max)) {
    C_max <- 0
    T_max <- 0
  }
  
  # Update C_max and T_max
  if (cp > C_max) {
    C_max <- cp
    T_max <- time
  }
})
```

In this model, `C_max` and `T_max` are checked with `is.na(C_max)`. The first time the model is evaluated for a subject, `C_max` will be `NA`, so it gets initialized to 0. In subsequent evaluations (at different time points), `C_max` will carry its value from the previous step.

## Comprehensive Example: $C_{max}$, $T_{max}$, $C_{min}$, and $T_{min}$

Let's look at a more complete example that calculates both peaks and troughs.

```{r comprehensive}
m <- rxode2({
  v  <- 10
  cl <- 1
  ka <- 1
  d/dt(depot) <- -ka * depot
  d/dt(centr) <- ka * depot - cl/v * centr
  cp <- centr / v
  
  if (is.na(C_max)) {
    C_max <- 0
    C_min <- Inf
    T_max <- 0
    T_min <- Inf
  }
  
  if (cp > C_max) {
    C_max <- cp
    T_max <- time
  } 
  
  if (cp < C_min) {
    C_min <- cp
    T_min <- time
  }
})

ev <- et(amt=100) |>
  et(seq(0, 24, length.out=100))

s <- rxSolve(m, ev)

tail(s)
```

## Sticky Variables for Time After Dose (TAD)

Another common use case is calculating the time since the last dose. While `rxode2` provides internal ways to handle this, you can also implement it using sticky variables and the `is.na(amt)` check (which is true when no dose is administered at the current time).

```{r tad_example}
m <- rxode2({
  cl <- 1
  v <- 10
  d/dt(centr) <- -cl/v * centr
  cp <- centr/v
  
  if (!is.na(amt)) {
    # A dose is happening
    tdose <- time
  }
  
  # tad is sticky because tdose is sticky
  tad <- time - tdose
})

ev <- et(amt=100, time=0) |>
  et(amt=100, time=12) |>
  et(seq(0, 24, by=1))

s <- rxSolve(m, ev)
head(s, 15)
```

## How it works internally

When `rxode2` parses the model, it identifies which variables are "left-handed side" (LHS) variables. Normally, these are reset to `NA` or `0` at each new time step to avoid bleeding values from one time point to another. 

However, when a variable is identified as sticky, `rxode2` changes this behavior:
- It allocates space to store the variable's value for each individual.
- It does NOT reset the value between time points for the same individual.
- It resets the value to `NA` only when a new individual starts (or at the beginning of a new `rxSolve` call).

This allows you to implement complex logic that depends on the history of the simulation without needing to use ODE states for everything.

## Summary

Sticky variables provide a powerful way to perform state-dependent calculations within your `rxode2` models. By using `is.na()` to initialize these variables, you can ensure they persist throughout the simulation for each individual, enabling real-time calculation of NCA metrics, time-since-event markers, and more.
