---
title: "Solve-time hooks for package developers"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Most `rxode2` models are fully specified by their parameters, data and initial
conditions.  Some downstream tools, though, need to change *what a solve sees* from
outside the model text -- for example to inject a block of externally-owned
parameters (neural-network weights, a lookup table) on every solve, or to add a
forcing term to a state derivative.  `rxode2` exposes a small set of hooks for this.
They fall into two groups:

* **R-level:** `rxForcedPars()` -- override parameter values on a model, carried
  with the model itself.
* **C-level:** the *par-loader* and *dydt forcing* hooks, plus the
  **function-pointer table** that lets a downstream package call `rxode2`'s C
  entry points.  These are for package authors writing compiled code.

This article is a companion to
[Providing Custom C/C++ Functions to rxode2](rxode2-pkg-exported-funs.html): that
one adds new *functions* callable inside a model; this one changes the *parameters
and derivatives* a solve uses.

## `rxForcedPars()`: forced parameters on a model

`rxForcedPars()` sets a named vector of parameter values that override `params`/data
and the model initial estimates on **every** solve of a ui.  The values are stored
on the ui (hidden from the printed model, registered as a `sticky` item), so they
travel with it -- through model piping and into any `nlmixr2` fit built from it.
That keeps a model self-contained: a fit that carries, say, trained weights
re-solves, predicts and simulates with those weights and no external state.

```{r}
ui <- function() {
  model({
    d/dt(center) <- -(cl/v) * center
    cp <- center/v
  })
}
ui <- rxode2(ui)
rxForcedPars(ui) <- c(cl = 1.2, v = 8)   # force cl, v on every solve
rxForcedPars(ui)                          # getter
rxForcedPars(ui) <- NULL                  # clear
```

Names that are not model parameters are ignored at solve time.  Forced parameters
are written into every subject/simulation column at solve setup -- the *same
injection point* the par-loader hook (below) uses, so a registered loader can still
override them if needed.

## The par-loader hook (C)

For a parameter block that is not a fixed vector but is *computed* -- owned by a
plugin and refreshed each solve -- register a **par-loader**.  It runs once per
solve, single-threaded, right after `rxode2` lays out the global parameter matrix
(`gpars`) and before the parallel integration, so the injected values are in place
for the (thread-safe) solve.

```c
/* signature (inst/include/rxode2.h) */
typedef void (*t_rxParLoader)(rx_solve *rx, double *gpars, int npars, int ncols);
void rxRegisterParLoader(t_rxParLoader cb);
void rxRemoveParLoader(t_rxParLoader cb);
```

`gpars` is the flat `ncols x npars` parameter matrix (column `c`'s `par_ptr` is
`&gpars[c*npars]`).  A loader writes its block into every column:

```c
static void myLoader(rx_solve *rx, double *gpars, int npars, int ncols) {
  for (int c = 0; c < ncols; ++c) {
    double *par_ptr = gpars + (size_t)c * npars;
    for (int k = 0; k < BLOCK_N; ++k) par_ptr[base + k] = myBlock[k];  /* base = block start */
  }
}
```

Register it in `.onLoad()` (via a `.Call` wrapper) and `rxRemoveParLoader()` in
`.onUnload()`.  The block must occupy real `par_ptr` slots: declare its parameters
with `param(...)` (or as covariates) so `rxode2` reserves contiguous positions,
resolve `base` from the model's solve parameter order once, and cache it.
`rxInjectedPars()` reports what a loader changed on the last solve.

### Named loaders: dispatch to the right model only

A loader registered with `rxRegisterParLoader()` runs on **every** solve, so it must
guard itself, and it can still corrupt an unrelated model if its guards are loose.
Prefer a **named** loader instead:

```c
rxRegisterParLoaderNamed("myPkg:myLoader", myLoader);   /* runs only when flagged */
```

A named loader runs *only* while a model flags its name.  Flag the injector on the
model that needs it with `rxParLoader()` (a sticky ui item, like `rxForcedPars()`):

```{r}
rxParLoader(ui) <- "myPkg:myLoader"   # only this loader fires for ui's solves
```

`rxSolve()` sets that flag active for the model's solve and clears it afterward, so
the loader never touches any other model.  For solves that bypass the ui path
(estimation internals, plain `rxode2()` models), a package can set the active flag
directly around its own solve batch and clear it after.  This is how `nlmixr2nn`
keeps its neural-network weight loader from leaking into unrelated fits.

## The dydt forcing hook (C)

To add a term to a state derivative that the model text cannot express (e.g. a
plugin's contribution computed in C), register a **dydt-force** callback.  The
generated model calls it at the end of `dydt`, so the added forcing is integrated
like any other RHS term:

```c
/* signature (inst/include/rxode2.h) */
typedef void (*t_rxDydtForce)(int *neq, double t, double *y, double *dydt);
void rxRegisterDydtForce(t_rxDydtForce cb);
void rxRemoveDydtForce(t_rxDydtForce cb);
```

```c
static void myForce(int *neq, double t, double *y, double *dydt) {
  dydt[stateIdx] += myForcingTerm(t, y);   /* on top of the model's own RHS */
}
```

A model with no registered forcing is unaffected (the call is a null-pointer
no-op), so this is safe to leave installed.

## The function-pointer table

The hooks above are C callables *inside* `rxode2`.  Because CRAN discourages
`R_RegisterCCallable` ABI coupling across packages, `rxode2` shares its C entry
points through a positional **function-pointer table** rather than by name:

* `rxode2::.rxode2ptrs()` returns the table (an external-pointer list) built by
  `_rxode2_rxode2Ptr()` in `src/init.c`.
* A downstream package installs it into its own globals with `iniRxodePtrs()` /
  `iniRxodePtrs0()` from the header `inst/include/rxode2ptr.h`, calling
  `.Call("_myPkg_iniRxodePtrs", rxode2::.rxode2ptrs())` in `.onLoad()`.

Each table slot is a fixed index; new entries are appended (so the layout stays
backward compatible).  `rxRegisterParLoader` / `rxRegisterDydtForce` are exposed
this way, so a plugin can register its hooks from its own compiled code.  See the
`rxode2` source (`src/init.c`, `inst/include/rxode2ptr.h`) for the current table,
and the `nlmixr2nn` package for a worked consumer that uses the par-loader to inject
neural-network weights.

## See also

* [Providing Custom C/C++ Functions to rxode2](rxode2-pkg-exported-funs.html) --
  adding model-callable functions.
* `nlmixr2est`: *"Extending nlmixr2est"* -- creating estimation methods,
  intercepting estimations, and contributing to the objective, which build on
  these solve hooks.
