---
title: "Event Tables after 5.0.2: Architecture and Benchmark"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Overview

This vignette compares the current event table handling with the old
event table in 5.0.2.  The focus is the event table layer (`rxEt`),
because that is where the branch makes its largest architectural
change.

At a high level, the old 5.0.2 event table handling keeps more of the
event table implementation in compiled code, especially for
sequence/repeat style operations.  The new architecture moves a
large part of event table construction and manipulation into R, keeps
the solver and event translation in C/C++, and uses a lazy,
environment-backed `rxEt` object as a data-frame proxy that defers
full materialization until it is needed.  By contrast, `main` works
much more like an actual realized data frame throughout the
event-table workflow.

The comparison below reflects:

- new event table handling (at commit `3ee936842`)
- old event table handling (at commit `3b8b35cce`)

## Architectural decisions on new event table structure

The branch makes five important design choices.

### 1. Pure-R event table construction

The public `et()` API remains the same, but much more of the work now
happens in R:

- `et.default()` resolves event-table arguments and compatibility aliases
- `.newRxEt()` creates the new object
- `.etBuildMethods()` attaches the attached backward compatible
  methods like `$add.sampling()`.
- `.etExpandAddl()` expands `addl` records in R
- `etSeq()`, `etRbind()`, and `etRep()` are implemented in R instead of
  depending on the older exported Rcpp helpers

This reduces the amount of cross-language glue needed for ordinary event
table authoring.

### 2. Lazy, environment-backed state

The new `rxEt` object is a data-frame subclass with its mutable state
in an attached environment accessed through `.rxEtEnv()`.  Internally
the branch stores event data as ID-indexed chunks and only creates the
full canonical data frame when `as.data.frame()`, printing, or solver
handoff requires it.  This saves time in the common multi-subject
solving scenarios seen in pharmacometrics.

That design intentionally shifts work away from repeated construction
and combination and toward materialization.  In other words, the old
5.0.2 event handling is a realized data frame with methods attached,
while the new event handling behaves more like a data-frame proxy that
eventually materializes to a realized event table.

### 3. Materialize only when needed

`.etMaterialize()` is the key function creating the full event table.
It converts the chunked representation into the canonical event-table
data frame, applies defaults, sorts records, and restores units.  The
branch uses this as a single place where lazy state becomes concrete.

This makes behavior easier to reason about, but it also means the
first forced materialization can be more expensive than having the
whole event table as a data.frame object in the first place.  However
there is a speed-up by not having to sort by id and time, instead
materializng the chunks of IDs that are stored in the event table.

### 4. Using data-frame/vctrs/dplyr methods for data-like behavior

`R/etVctrs.R` adds `vec_proxy()`, `vec_restore()`, `vec_cast()`, and
related helpers for `rxEt`.  That is a significant shift in
philosophy: the event table becomes a better-behaved data-frame
subtype instead of a proper data-frame.

### 5. Keep C/C++ where it still matters

The new event handling does **not** move everything into R.  Event
translation and solver preparation still flow through the C/C++ side,
especially `src/etTran.cpp` and `src/rxData.cpp`.  In other words,
this is not a "remove compiled code" rewrite; it is a "move
authoring/manipulation logic into R and keep translation/solve logic
compiled" rewrite for the new event system

## Main functions in the new event system

```{r}
main_functions <- data.frame(
  function_name = c(
    ".newRxEt()",
    ".rxEtEnv()",
    ".etBuildMethods()",
    "et.default()",
    ".etMaterialize()",
    ".etExpandAddl()",
    "etSeq() / etRbind() / etRep()",
    ".etFixCmtForSolve()",
    "vec_proxy.rxEt() / vec_restore.rxEt()"
  ),
  role = c(
    "Create the new env-backed rxEt shell.",
    "Recover the mutable event-table environment.",
    "Implement the mutable method API such as add.dosing() and add.sampling().",
    "Resolve public et() calls, aliases, seq()-style forms, IDs, and imports.",
    "Materialize the lazy chunk store into canonical event-table rows.",
    "Expand addl dosing records in pure R.",
    "Compose, bind, and repeat event tables without the older Rcpp sequence helpers.",
    "Normalize character cmt values before solver handoff.",
    "Round-trip rxEt through vctrs/dplyr-style reconstruction."
  ),
  old_equivalent = c(
    "No direct equivalent; builds older rxEt/EventTable state through the C++ path.",
    "relies on compiled rxEt detection/state access more heavily.",
    "More behavior lived in the older EventTable/Rcpp implementation.",
    "Same public entry point, but with a more eager and more compiled implementation.",
    "reaches a realized table earlier in the workflow.",
    "relied more on compiled event helpers.",
    "used exported Rcpp helpers such as etSeq_ and etRep_.",
    "Not needed in the same way before the new lazy/data-frame path exposed more mixed input cases.",
    "No comparable vctrs restore layer on main."
  ),
  stringsAsFactors = FALSE
)
knitr::kable(main_functions, col.names = c("Current branch function", "Role", "Rough equivalent on `main`"))
```

## Motivation compared with old event handling

The branch makes sense if the goal is to optimize for correctness,
maintainability, and interoperability of the event-table layer.

Compared with the old event table handling, the motivations are:

1. **Shrink the Rcpp surface area for event authoring.**  Fewer exported
   compiled helpers makes debugging and incremental fixes easier.
2. **Represent event tables as a data-frame subtype.**  That
   makes `rxEt` behave more naturally with modern R tooling.
3. **Delay work until the user actually needs a realized table.**  This
   favors construction-heavy pipelines.
4. **Keep the solver path compiled.**  The branch does not give up the
   performance benefits of compiled translation and solving where they
   matter most.

In practice, that is why the branch has been able to absorb many bug
fixes in the event-table layer without changing the public tests.

## Relative strengths and weaknesses

```{r}
strengths <- data.frame(
  aspect = c(
    "Construction model",
    "Composability",
    "Interoperability",
    "Maintainability",
    "Materialization cost",
    "Repeat-heavy workflows",
    "Solve handoff overhead"
  ),
  new = c(
    "Lazy chunked representation is well suited to pipelines that keep appending or combining events.",
    "Pure-R etSeq()/etRbind() are easier to reason about and modify.",
    "Much stronger vctrs/data.frame/tibble behavior.",
    "More logic is visible and testable in R.",
    "Usually slower because full rows are built later and in one step.",
    "Currently slower in the benchmark below.",
    "Slightly slower after warm-up in the benchmark below."
  ),
  old = c(
    "More eager/compiled representation; less lazy bookkeeping.",
    "Compiled helpers are fast when the operation already matches the older model well.",
    "Weaker integration with newer data-frame reconstruction patterns.",
    "Harder to debug because more behavior crosses the R/C++ boundary.",
    "Usually faster because the table is already closer to realized form.",
    "Currently faster in the benchmark below.",
    "Currently a little faster in the benchmark below."
  ),
  stringsAsFactors = FALSE
)
knitr::kable(strengths, col.names = c("Aspect", "new", "old"))
```

In short:

- The new implementation is stronger when event tables are still being
  *built* or *combined* which is the most common use-case for `et()` event
  tables.

## Benchmark design

The benchmark intentionally measures common event-table workflows instead
of isolated micro-ops:

- `add_dosing_sampling`: build with the mutable method API
- `vector_build`: build large event tables from vectorized `et()` calls
- `sequence_tables`: combine many event tables with `c()`
- `repeat_table`: repeat a regimen with `rep()`
- `materialize_table`: force a full `as.data.frame(..., all=TRUE)`
- `solve_handoff`: solve a simple PK model after the event table is built
- `issue_724_large_sampling`: append sampling records for a large
  subject-by-dose dataset based on Issue #724

The measurements below were collected in the same Linux environment from
which this vignette was written, after warming the package once in each
branch.  They should be read as *relative* comparisons, not absolute
promises for every machine or every regimen.

```{r}
benchmark_results <- data.frame(
  scenario = c(
    "add_dosing_sampling",
    "vector_build",
    "sequence_tables",
    "repeat_table",
    "materialize_table",
    "solve_handoff",
    "issue_724_large_sampling"
  ),
   new_median_s = c(0.1275, 0.0104, 0.3130, 0.4240, 0.00199, 0.0260, 1.9210),
   old_median_s = c(0.1055, 0.1396, 0.6485, 0.1640, 0.00003, 0.0200, 80.2900),
   stringsAsFactors = FALSE
)

benchmark_results$winner <- ifelse(
  benchmark_results$new_median_s < benchmark_results$old_median_s,
  "refactor-et",
  "main"
)
benchmark_results$relative_result <- c(
  "old ~1.2x faster",
  "new ~13.4x faster",
  "new ~2.1x faster",
  "old ~2.6x faster",
  "old ~66x faster",
  "old ~1.3x faster",
  "new ~41.8x faster"
)

knitr::kable(
  benchmark_results,
  col.names = c("Scenario", "`refactor-et` median (s)", "`main` median (s)", "Winner", "Relative result")
)
```

## Interpreting the benchmark

The benchmark is consistent with the architectural choices:

### Where the new event table performs better

1. **Large vectorized construction**

   The lazy chunk-based representation is very effective when many event
   rows are appended before the table is forced into canonical form.

2. **Sequencing many tables**

   The pure-R sequence/bind logic avoids some of the overhead of the
   older compiled sequence path and seems to benefit from operating on
   chunked event-table state.

3. **Large appended datasets (Issue #724)**

   This is one of the clearest cases motivating the branch.  The
   workload in [Issue #724](https://github.com/nlmixr2/rxode2/issues/724)
   constructs a large subject-by-dose dataset and then appends a dense
   sampling schedule.  On `main`, the event table behaves more like an
   already-realized data frame, so repeatedly growing it becomes very
   expensive at larger sizes.  On `refactor-et`, the event table remains
   a proxy backed by chunked state until later, so the same workflow is
   much cheaper.

### Where the old method performs better

1. **Mutable method API (`add.dosing()`/`add.sampling()`)**

   The branch keeps compatibility, but the new path currently pays a
   small overhead relative to the older implementation.

2. **`rep()`**

   Repetition is still a strong case for the older compiled approach.

3. **Forced materialization**

   This is the clearest cost of the branch.  Since `refactor-et`
   postpones realization, the eventual `as.data.frame(..., all=TRUE)`
   step is much more expensive.

4. **Final solve handoff**

   Once the event table is already built, `main` retains a modest edge in
   the benchmarked solve path.

## Large-scale benchmark from Issue #724

[Issue #724](https://github.com/nlmixr2/rxode2/issues/724) is especially
useful because it shows the effect of event-table architecture on a
workload that becomes much larger than the smaller benchmark scenarios
above.

The reported workload is:

1. Build a subject-by-dose data set with 3 dose groups
2. Create one event table row per subject for dosing
3. Append a dense sampling grid from 0 to 168 hours by 0.5 hours

On the original `main`-branch implementation, the elapsed time rose
rapidly with subject count:

```{r}
issue_724_scaling <- data.frame(
  n_subj = c(100, 200, 300, 400, 500, 600, 800, 1000),
  old_elapsed_s = c(0.31, 2.28, 7.38, 13.1, 20.1, 28.7, 51.0, 80.29)
)

knitr::kable(
  issue_724_scaling,
  col.names = c("Subjects per dose group", "`main` elapsed (s)")
)
```

For the exact 1000-subject case from the issue, a later run on this
new event table code reported:

```{r}
issue_724_comparison <- data.frame(
  branch = c("old", "new"),
  elapsed_s = c(80.29, 1.921),
  source = c(
    "Issue #724 original report",
    "Issue #724 follow-up comment after the et() rewrite"
  ),
  stringsAsFactors = FALSE
)

issue_724_comparison$relative <- c(
  "baseline",
  "about 41.8x faster than main"
)

knitr::kable(
  issue_724_comparison,
  col.names = c("Branch", "Elapsed (s)", "Source", "Relative result")
)
```

This benchmark is more compelling than a small microbenchmark because it
exercises exactly the pattern that motivated the refactor:

- The old event table keeps the event table closer to a realized data frame, so very
  large repeated row growth becomes increasingly expensive

- The new event table keeps the event table as a proxy backed by chunked state
  and delays full data-frame realization until a boundary like printing,
  `as.data.frame()`, or solver handoff

That does **not** mean the new event table wins every benchmark.  The earlier
table still shows that forced materialization and some final execution
paths can remain faster on the old event table.  But Issue #724 demonstrates why the
branch architecture can produce very large wins on bigger event-table
authoring workloads.

## Overall assessment

Since the goal is **a cleaner, more maintainable, more interoperable event
table architecture**, the new event handling code is a stronger design.  The
branch makes the event-table layer easier to understand in R, easier to
test, and faster in the construction/composition workloads that fit its
lazy design.


## Session information

```{r}
sessionInfo()
```
