---
title: "ETA and EPS Resampling in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(lotri)
```

## Overview

rxode2 supports resampling of between-subject variability (ETAs) and
residual variability (EPSs) within the model code itself using
`simeta()` and `simeps()`.  This allows you to implement rejection
sampling, truncated distributions, or any custom sampling logic
directly in the model definition — without post-processing the
simulated output.

This is analogous to the `SIMETA` and `SIMEPS` functions available
in NONMEM, though the implementation and the context in
which they are used differs slightly between the tools.

## Resampling ETAs with `simeta()`

A common use case is constraining a covariate derived from a
lognormal distribution to a physiologically plausible range.  For
example, simulating body weight from a lognormal distribution but
only accepting values between 60 and 80 kg:

```{r simeta-example}
mod <- rxode2({
  wt  <- 70 * exp(eta.wt)   # individual weight (kg)
  i   <- 0                   # iteration counter (for safety)
  while ((wt < 60) || (wt > 80)) {
    i <- i + 1
    if (i > 100) break       # prevent infinite loops
    simeta()                 # resample all ETAs for this subject
    wt <- 70 * exp(eta.wt)
  }
})
```

When `simeta()` is called, all ETA values for the current subject are
re-drawn from the supplied omega matrix.  The model then re-evaluates
the downstream expressions.  Because the resampling is done at the C
level, parallel solving with OpenMP is fully supported.

```{r simeta-solve}
set.seed(1234)

e <- et(seq(0, 24, by = 4), id = 1:100)

f <- rxSolve(
  mod, e,
  omega = lotri(eta.wt ~ 0.25)   # SD ~ 0.5 on log scale: wide variability
)

summary(f$wt)
```

All simulated weights fall within the 60–80 kg window:

```{r simeta-check}
stopifnot(all(f$wt >= 60), all(f$wt <= 80))
range(f$wt)
```

### Interaction with `nStud`

`simeta()` works correctly when simulating multiple studies
(`nStud`).  Each study draws its own omega matrix (if `dfSub` is
specified for a scaled chi-squared draw), and `simeta()` resamples
within the study-specific omega:

```{r simeta-nstud}
set.seed(42)

f2 <- rxSolve(
  mod, e,
  omega  = lotri(eta.wt ~ 0.25),
  nStud  = 5,
  dfSub  = 40
)

range(f2$wt)
```

## Resampling EPSs with `simeps()`

`simeps()` re-draws all EPS (residual variability) values for the
current observation record.  A practical use case is ensuring a
simulated observation is non-negative without post-hoc truncation:

```{r simeps-example}
mod_conc <- rxode2({
  C <- 0 + err            # hypothetical: mean zero + residual noise
  i <- 0
  while (C < 0) {
    simeps()              # resample all EPSs for this record
    C <- 0 + err
    i <- i + 1
    if (i > 10) break
  }
})
```

```{r simeps-solve}
set.seed(10)

e2 <- et(0:9)

f3 <- rxSolve(mod_conc, e2, sigma = lotri(err ~ 1))

# Without resampling some records would be negative; with simeps() all are >= 0
all(f3$C >= 0)
```

### Preserving original draws when the condition is already met

When the initial draw already satisfies the condition, `simeps()` is
never called and the original value is preserved.  This means the
marginal distribution of accepted draws is the same as the original
distribution restricted to the feasible region, without any hidden
bias from extra calls:

```{r simeps-preserve}
# Solve without simeps as reference
mod_ref <- rxode2({
  C <- 0 + err
  i <- 0
})

set.seed(10)
f_ref <- rxSolve(mod_ref, e2, sigma = lotri(err ~ 1))

set.seed(10)
f_resamp <- rxSolve(mod_conc, e2, sigma = lotri(err ~ 1))

# Records where no resampling was needed (i == 0) should match exactly
both <- merge(f_ref, f_resamp[f_resamp$i == 0, ], by = "time")
stopifnot(isTRUE(all.equal(both$C.x, both$C.y)))
```

## Differences from a pure post-hoc truncation

Using `simeta()` / `simeps()` inside the model has two advantages over
filtering or replacing values after `rxSolve()`:

1. **Self-consistency**: downstream model quantities (e.g., PK
   parameters computed from body weight) always reflect the accepted
   draw — no risk of a filtered row having inconsistent derived
   variables.
2. **Parallelism**: resampling happens inside the C solver loop,
   so OpenMP parallelism across subjects is fully preserved.
