---
title: "rxode2 and NMsim: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

[rxode2](https://nlmixr2.github.io/rxode2/) and
[NMsim](https://nmautoverse.github.io/NMsim/) both help
pharmacometricians simulate pharmacokinetic and pharmacodynamic models
from R, but they take fundamentally different approaches.

- **rxode2** is a self-contained R/C/Fortran ODE solver.  Models are
  written (or imported) in R, compiled to a shared library, and solved
  entirely within the R session.  No external software is required.

- **NMsim** is an R interface that orchestrates NONMEM.  It takes an
  existing NONMEM estimation control stream, modifies it into a
  simulation control stream, runs NONMEM as an external process, and
  returns the results to R.  A licensed NONMEM installation is
  required.

This fundamental difference drives most of the trade-offs described
below.

## Shared capabilities

- NONMEM-compatible datasets (EVID, AMT, CMT, TIME, ID, …)
- Simulation of new subjects with sampled ETAs
- Simulation of typical subjects (all ETAs = 0)
- Covariate sampling and prediction-interval generation
- VPC (visual predictive check) workflows
- Simulation from empirical Bayes estimates (individual ETAs from a
  prior NONMEM estimation run)
- Parameter uncertainty simulation (e.g., from bootstrap runs or a
  covariance step)
- Multiple models simulated in a single call (e.g., all bootstrap
  replicates)
- Dataset creation helpers (`NMcreateDoses()`/`NMaddSamples()` in
  NMsim; `et()` in rxode2)

## Features rxode2 has that NMsim does not

### No NONMEM license required

rxode2 is entirely open-source (GPL ≥ 3) and self-contained.
Simulation runs without any external software.  NMsim requires a
working, licensed NONMEM installation configured via
`NMdataConf(path.nonmem = ...)` before any simulation can be
performed.

### Interactive iteration speed

Because rxode2 compiles and solves models in-process, the turnaround
for a simulation is typically sub-second to a few seconds for most
models.  NMsim must launch NONMEM as an external process for every
simulation run.  NONMEM's startup, I/O, and table-writing overhead
means each NMsim call takes at least several seconds and often
minutes, even for simple models.  This makes rxode2 much more
practical for exploratory iteration, sensitivity analyses, and
interactive model development.

```{r}
library(rxode2)

# Compile once; re-solve in milliseconds
mod <- rxode2({
  C2 <- centr / V
  d/dt(centr) <- -CL * C2
})
et <- et(seq(0, 24, by = 1))
# Parameter sweeps are instant
rxSolve(mod, et, params = c(CL = 2, V = 10))
rxSolve(mod, et, params = c(CL = 5, V = 10))
```

### OpenMP parallelism across subjects

rxode2's LSODA solver is implemented in thread-safe C, allowing
parallel ODE solving across subjects via OpenMP within a single R
process.  NMsim relies on NONMEM for its computation, which can be
submitted to an HPC cluster (via PSN) but does not offer
within-process parallelism from R.

```{r}
# rxode2 automatically uses multiple cores for multi-subject sims
getRxThreads()   # see how many threads are active
```

### Model piping for incremental modification

rxode2 model objects support the native R pipe operator (`|>`) to
clone and modify a model without rewriting the full specification.

```{r}
base <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
})

# Add population structure incrementally
full <- base |>
  model({
    KA <- exp(tka + eta.ka)
    CL <- exp(tcl + eta.cl)
    V  <- exp(tv)
  }, append = FALSE) |>
  ini({
    tka <- log(0.5); eta.ka ~ 0.09
    tcl <- log(4);   eta.cl ~ 0.09
    tv  <- log(20)
  })

# Fix a parameter for a sensitivity run
sens <- full |> ini(fix(eta.ka))
```

NMsim's `modify` argument provides some control over the simulation
control stream, but it operates at the text level on an already-
estimated NONMEM model rather than providing an R-level model algebra.

### Parameter estimation via nlmixr2

rxode2 models can be passed directly to nlmixr2 for population
parameter estimation (FOCE, FOCEi, SAEM, etc.) using the same model
object.  NMsim is simulation-only; estimation is always done in
NONMEM separately.

### Symbolic Jacobians and forward sensitivities

rxode2 automatically derives the symbolic Jacobian of the ODE system
and forward-sensitivity equations, which are used by nlmixr2's FOCEi
algorithm for exact gradient-based estimation.  This capability is
irrelevant to NMsim, which delegates all numerical work to NONMEM.

### 1–3 compartment analytical solutions with exact gradients

`linCmt()` in rxode2 provides analytical solutions for one-, two-,
and three-compartment models with gradients computed via Stan math
auto-differentiation.  mrgsolve and NMsim do not provide this
independently of NONMEM's own ADVAN routines.

### Multiple ODE solver backends

rxode2 offers three solver backends selectable per run: a thread-safe
C LSODA (default), a Fortran LSODA reference implementation, and
DOP853 (explicit 8th-order Runge-Kutta for non-stiff systems).
NMsim uses whichever ADVAN is specified in the NONMEM control stream.

### Runs anywhere without infrastructure

rxode2 simulations run on any machine with R and a C compiler — a
laptop, a CI server, a Docker container, or Shiny.  NMsim requires
a NONMEM license and a configured executable path, which typically
restricts it to specific workstations or servers.

## Features NMsim has that rxode2 does not

### Zero model re-implementation risk

NMsim's core value proposition is that it simulates the *exact*
NONMEM model used for estimation — the same control stream, the same
ADVAN, the same `$PRED`/`$DES`/`$ERROR` code.  There is no
translation step, so there is no risk of introducing bugs when
converting the model to a different syntax.

rxode2 requires the model to be re-expressed in rxode2's own
mini-language or imported via `nonmem2rx`.  Importantly, `nonmem2rx`
performs an automatic validation step after import: it re-runs the
simulation with the converted rxode2 model and compares PRED/IPRED/IWRES
against the original NONMEM output, reporting whether the two agree
within a tolerance.  This means you know immediately whether the
re-implementation is correct, rather than discovering discrepancies
later.  Complex or highly customised NONMEM models may still require
manual adjustment when the automated comparison does not pass.

```{r}
library(NMsim)

# Simulate the exact estimation model — no reimplementation
simres <- NMsim(
  file.mod = "path/to/run001.mod",
  data     = data.sim
)
```

### Full NONMEM feature coverage

Because NMsim runs NONMEM itself, it supports the entire NONMEM
feature set: all ADVAN routines (ADVAN1–ADVAN13), `$MIXTURE` models,
any `$PRED`/`$DES` code, inter-occasion variability, `$PRIOR`, and
any NONMEM-compatible extension.  rxode2's own solver covers the
most common pharmacometric models but does not replicate every NONMEM
feature; notably:

- `$MIXTURE` population mixture models: `nonmem2rx` can import them,
  but the automatic PRED/IPRED/IWRES validation step is skipped for
  mixture models, so correctness of the import must be verified
  manually.
- Models that rely heavily on NONMEM-specific verbatim Fortran or
  C code in `$DES` have no direct equivalent.

### Cluster submission via PSN

`NMexec()` (the NMsim companion for running estimation) can submit
NONMEM jobs to an HPC cluster using PsN.  For large bootstrap or
uncertainty simulations NMsim can similarly distribute NONMEM runs
across cluster nodes.  rxode2's parallelism is always within a single
R process via OpenMP.

## Features available in both tools (with different approaches)

### EBE-based simulation of known subjects

Both tools can simulate known subjects using their empirical Bayes
estimates (individual ETAs) under a new dosing regimen without
re-estimating the model.

NMsim does this natively from any NONMEM run via
`method.sim = NMsim_EBE`, which retrieves the individual ETAs directly
from NONMEM output tables.

`nonmem2rx` imports the individual EBEs alongside the model, storing
them in `$etaData` of the imported model object, so they can be passed
directly to `rxSolve()` without any manual extraction step.

```{r}
# NMsim approach
simres <- NMsim(
  file.mod   = "path/to/run001.mod",
  data       = data.sim,
  method.sim = NMsim_EBE
)

# rxode2 approach (after nonmem2rx import)
library(nonmem2rx)
mod <- nonmem2rx("path/to/run001.mod")
rxSolve(mod, data.sim, iCov = mod$etaData)
```

### Parameter uncertainty from the NONMEM covariance step

Both tools support simulation with parameter uncertainty derived from
NONMEM's `$COVARIANCE` step.

NMsim uses `method.sim = NMsim_VarCov` to draw parameter sets from
the variance-covariance matrix and run one NONMEM instance per draw,
or `method.sim = NMsim_NWPRI` for a `$PRIOR NWPRI` specification.

`nonmem2rx` imports the NONMEM variance-covariance matrix alongside
the model, so draws can be generated immediately and passed to
`rxSolve()` without any manual extraction step.  Draws can also come
from a bootstrap, nlmixr2's uncertainty output, or a user-supplied
covariance matrix.

## Complementary use: NMsim and rxode2 together

The two tools are often complementary rather than competing.  A
common workflow is:

1. **Estimate** in NONMEM.
2. **Import and validate** with `nonmem2rx`: the package converts the
   control stream to an rxode2 model *and* automatically compares its
   PRED/IPRED/IWRES against the original NONMEM output.  A passing
   validation confirms the rxode2 model is numerically equivalent to
   the NONMEM model.
3. **Simulate** with rxode2 for fast exploratory analyses, sensitivity
   analyses, and trial-design work — with confidence from the
   validation step.
4. **Simulate** with NMsim when exact NONMEM fidelity is required
   for regulatory submissions, when the automated `nonmem2rx`
   validation does not pass, or when the model uses NONMEM features
   with no rxode2 equivalent.

## Summary table

| Feature | rxode2 | NMsim |
|---------|--------|-------|
| NONMEM required | no | yes (licensed) |
| Simulation engine | rxode2 C/Fortran solver | NONMEM |
| Model re-implementation needed | yes (or via `nonmem2rx`) | no |
| Re-implementation risk | mitigated: `nonmem2rx` auto-validates PRED/IPRED/IWRES vs NONMEM | none |
| Simulation speed | sub-second to seconds | seconds to minutes |
| Full NONMEM feature set | partial | yes |
| `$MIXTURE` models | `nonmem2rx` imports; no auto-validation | yes |
| EBE-based simulation | yes — `nonmem2rx` imports EBEs into `$etaData` | yes — `NMsim_EBE` |
| Uncertainty from covariance step | yes — `nonmem2rx` imports covariance matrix for draws | yes — `NMsim_VarCov` / `NMsim_NWPRI` |
| Bootstrap uncertainty | manual (supply draws) | built-in (simulate all bootstrap models) |
| OpenMP within-process parallelism | yes | no |
| Cluster submission | no | yes (via PSN) |
| Model piping (`\|>`) | yes | no |
| Parameter estimation | via nlmixr2 | no (NONMEM only) |
| Symbolic Jacobians / sensitivities | yes | no |
| 1–3 cmt analytical solutions | `linCmt()` with gradients | ADVAN1/2/3/4/11/12 via NONMEM |
| Multiple ODE solver backends | LSODA (C), LSODA (Fortran), DOP853 | NONMEM ADVAN |
| Runs without licensed software | yes | no |
| CRAN | yes | yes |

## References

- [NMsim website](https://nmautoverse.github.io/NMsim/)
- [NMsim introductory vignette](https://nmautoverse.github.io/NMsim/articles/NMsim-intro.html)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nonmem2rx package](https://nlmixr2.github.io/nonmem2rx/)
- [nlmixr2 package](https://nlmixr2.github.io/nlmixr2/)
