---
title: "Visual Predictive Checks (VPC) with rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
```

## What is a VPC?

A visual predictive check (VPC) is a simulation-based diagnostic that
compares the distribution of observed data to the distribution of
model-predicted data.  The observed data are overlaid on prediction
intervals (e.g. 5th, 50th, 95th percentiles) computed from many
simulated replicates of the study design.  A well-fitting model
produces prediction intervals that envelop the observed percentiles.

## Setting up the model and observed data

We use a two-compartment PK model estimated from the `theo_sd` dataset
(single-dose theophylline) to illustrate the workflow.

```{r model}
pkModel <- function() {
  ini({
    tka  <- log(1.57)
    tcl  <- log(2.72)
    tv   <- log(31.07)
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka  <- exp(tka + eta.ka)
    cl  <- exp(tcl + eta.cl)
    v   <- exp(tv  + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp  <- center / v
    cp  ~ add(add.sd)
  })
}
```

For illustration we use the built-in `theo_sd` dataset as observed data:

```{r obsdata}
obs <- nlmixr2data::theo_sd
```

## Running replicated simulations

A VPC requires simulating the study design many times (typically
100–1000 replicates).  Each replicate samples new ETAs and residual
error.

```{r simulate, message=FALSE}
nRep <- 200   # use >= 500 for a publication-quality VPC

set.seed(1234)
simDat <- rxSolve(
  pkModel,
  obs,
  nStud  = nRep,          #  per replicate
  returnType = "data.frame",
  keep="WT"
)
```

## Computing prediction intervals

Compute the 5th, 50th, and 95th percentiles of `cp` at each nominal
time point across all replicates:

```{r piCalc}
library(data.table)
simDT <- as.data.table(simDat)

# Observed percentiles
obsDT <- as.data.table(obs[obs$EVID == 0, ])
obsPctl <- obsDT[, .(
  p05 = quantile(DV, 0.05, na.rm = TRUE),
  p50 = quantile(DV, 0.50, na.rm = TRUE),
  p95 = quantile(DV, 0.95, na.rm = TRUE)
), by = TIME]

# Simulated prediction intervals
simPctl <- simDT[, .(
  p05 = quantile(cp, 0.05, na.rm = TRUE),
  p50 = quantile(cp, 0.50, na.rm = TRUE),
  p95 = quantile(cp, 0.95, na.rm = TRUE)
), by = time]
setnames(simPctl, "time", "TIME")
```

## Plotting the VPC

```{r vpcPlot, fig.width=7, fig.height=4}
ggplot() +
  # Simulated prediction interval ribbon
  geom_ribbon(data = simPctl,
              aes(x = TIME, ymin = p05, ymax = p95),
              fill = "steelblue", alpha = 0.3) +
  # Simulated median line
  geom_line(data = simPctl,
            aes(x = TIME, y = p50),
            colour = "steelblue", linewidth = 1) +
  # Observed percentile lines
  geom_line(data = obsPctl,
            aes(x = TIME, y = p05),
            colour = "black", linetype = "dashed") +
  geom_line(data = obsPctl,
            aes(x = TIME, y = p50),
            colour = "black", linewidth = 1) +
  geom_line(data = obsPctl,
            aes(x = TIME, y = p95),
            colour = "black", linetype = "dashed") +
  # Raw observed data
  geom_point(data = obs[obs$EVID == 0, ],
             aes(x = TIME, y = DV),
             alpha = 0.4, size = 1) +
  labs(x = "Time (h)", y = "Theophylline concentration (mg/L)",
       title = "VPC — 5th/50th/95th percentiles",
       subtitle = "Shaded: simulated PI; dashed lines: observed percentiles") +
  theme_bw()
```

## Stratified VPC

For covariate-stratified VPCs, split both observed and simulated data
by the covariate before computing percentiles, then use `facet_wrap()`:

```{r stratified, fig.width=7, fig.height=5}
# Create a weight group (above/below median)
medWt <- median(obs$WT)
obs$wtGrp    <- ifelse(obs$WT >= medWt, "WT >= median", "WT < median")
simDat$wtGrp <- ifelse(simDat$WT >= medWt, "WT >= median", "WT < median")

simDT2  <- as.data.table(simDat)
obsDT2  <- as.data.table(obs[obs$EVID == 0, ])

simPctl2 <- simDT2[, .(
  p05 = quantile(sim, 0.05, na.rm = TRUE),
  p50 = quantile(sim, 0.50, na.rm = TRUE),
  p95 = quantile(sim, 0.95, na.rm = TRUE)
), by = .(time, wtGrp)]
setnames(simPctl2, "time", "TIME")

obsPctl2 <- obsDT2[, .(
  p05 = quantile(DV, 0.05, na.rm = TRUE),
  p50 = quantile(DV, 0.50, na.rm = TRUE),
  p95 = quantile(DV, 0.95, na.rm = TRUE)
), by = .(TIME, wtGrp)]

ggplot() +
  geom_ribbon(data = simPctl2,
              aes(x = TIME, ymin = p05, ymax = p95),
              fill = "steelblue", alpha = 0.3) +
  geom_line(data = simPctl2,
            aes(x = TIME, y = p50),
            colour = "steelblue", linewidth = 1) +
  geom_line(data = obsPctl2,
            aes(x = TIME, y = p50),
            colour = "black", linewidth = 1) +
  geom_point(data = obsDT2,
             aes(x = TIME, y = DV),
             alpha = 0.3, size = 1) +
  facet_wrap(~wtGrp) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Stratified VPC by weight group") +
  theme_bw()
```

## Using the `vpc` package

The `vpc` CRAN package provides a purpose-built VPC function that
handles binning, confidence intervals on the prediction intervals, and
several plot options.

```{r vpcPkg, eval=FALSE}
library(vpc)

vpc(
  sim  = simDat,
  obs  = obs[obs$EVID == 0, ],
  obs_cols = list(id = "ID", dv = "DV", idv = "TIME"),
  sim_cols = list(id = "id",  dv = "sim", idv = "time"),
  pi   = c(0.05, 0.95),
  ci   = c(0.05, 0.95)
)
```

## Tips

- Use at least 500 replicates for stable 5th/95th percentile estimates.
- For log-scale PK data add `scale_y_log10()` to the ggplot.
- For censored data (`CENS=1`) use `vpc_cens()` from the `vpc`
  package.
- For time-to-event models use `vpc_tte()`.
- Binning strategy (equal-width, equal-count, or user-defined)
  strongly affects VPC appearance; the `vpc` package provides several
  options via the `bins` argument.
