---
title: "Dual Absorption Models with splitBolus()"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## Background

Some drugs exhibit two distinct absorption phases after oral dosing. A classic
example is a modified-release tablet where an immediate-release coat dissolves
rapidly in the stomach while a matrix core releases drug more slowly in the
intestine. Another example is a drug with meaningful gastric absorption in
addition to the usual intestinal absorption. Either way, the plasma concentration
versus time profile shows a secondary rise or shoulder that a simple
first-order depot model cannot capture.

The traditional workaround is to split the dose *in the dataset*: add a second
row with the same dose time but directed at a different compartment, and scale
each row's amount by the fraction going through each pathway. This ties the
model structure to the data format and forces dataset edits whenever the model
changes.

`splitBolus()` is a model-only directive in rxode2 that moves this split into
the **model** block. Bolus doses arriving at the source compartment are
transparently redirected to one or more target compartments. The dataset never
changes.

## Setup: a shared oral-dosing event table

```{r setup}
library(rxode2)

ev <- et(amt = 100, time = 0, cmt = "depot", ii = 24, addl = 6) |>
  et(seq(0, 168, by = 1))
```

This event table represents a week of once-daily 100 mg oral doses with
hourly observations. Both models below will use `ev` unchanged.

## Single first-order absorption model

First to show we start with a single one-compartment oral absorption model:

```{r single-model}
modSingle <- function() {
  ini({
    tka <- log(1.2)
    tcl <- log(6)
    tv  <- log(40)
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v  <- exp(tv)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central
    cp <- central / v
  })
}

resSingle <- rxSolve(modSingle, ev)
plot(resSingle, cp)
```

## Dual absorption model with `splitBolus`

To add a second absorption pathway, we change **only the model**. The event
table `ev` stays exactly the same.

`splitBolus(depot, depot, depotSlow)` tells rxode2 that any bolus dose
directed at `depot` should instead be applied to `depot` and
`depotSlow`.  Both target compartments receive the full original
amount, so bioavailability modifiers `f(depot)` and `f(depotSlow)` are
used to partition the dose between the two pathways; The
`lag(depotSlow)` is to show how much time the second part of the dose
started to release.

```{r dual-model}
modDual <- function() {
  ini({
    tka1 <- log(0.4)
    tka2 <- log(0.2)
    fa   <- 0.7
    tcl  <- log(6)
    tv   <- log(40)
  })
  model({
    ka1 <- exp(tka1)
    ka2 <- exp(tka2)
    cl  <- exp(tcl)
    v   <- exp(tv)

    splitBolus(depot, depot, depotSlow)
    f(depot) <- fa
    f(depotSlow) <- 1 - fa
    lag(depotSlow) <- 6

    d/dt(depot) <- -ka1 * depot
    d/dt(depotSlow) <- -ka2 * depotSlow
    d/dt(central)   <- ka1 * depot + ka2 * depotSlow - cl / v * central
    cp <- central / v
  })
}

resDual <- rxSolve(modDual, ev)
plot(resDual, cp)
```

The parameters here are:

| Parameter | Meaning |
|-----------|---------|
| `tka1`    | Log of the fast absorption rate constant (hr^-1) |
| `tka2`    | Log of the slow absorption rate constant (hr^-1) |
| `fa`      | Fraction of the dose entering the fast pathway |
| `tcl`     | Log of clearance (L/hr) |
| `tv`      | Log of central volume (L) |

## Comparing the two concentration profiles

```{r compare, message=FALSE}
library(ggplot2)

dfSingle <- as.data.frame(resSingle)[, c("time", "cp")]
dfSingle$model <- "Single absorption"

dfDual <- as.data.frame(resDual)[, c("time", "cp")]
dfDual$model <- "Dual absorption"

dfAll <- rbind(dfSingle, dfDual)

ggplot(dfAll, aes(time, cp, color = model)) +
  geom_line() +
  labs(x = "Time (hr)", y = "Concentration (mg/L)", color = NULL) +
  theme_bw()
```

The dual absorption profile shows a shoulder on the ascending limb of each
dose interval, driven by the slow pathway continuing to deliver drug after the
fast pathway has been nearly exhausted.

## Key points

- The event table `ev` is identical for both models. No dataset edits are
  needed to switch from single to dual absorption.

- `splitBolus(depot, depot, depotSlow)` intercepts every bolus arriving
  at `depot` and redirects it to `depot` and `depotSlow`. The source
  compartment `depot` itself never accumulates drug.
- `f(depot) <- fa` and `f(depotSlow) <- 1 - fa` ensure the dose is
  partitioned so that the total amount absorbed equals the original dose (when
  `fa` is between 0 and 1).
- The same approach scales to population simulations: supply an `iCov` or
  `omega` matrix to `rxSolve()` and the dataset requires no changes.

- More than two pathways are possible-simply list additional target
  compartments in `splitBolus()` and add corresponding `f()` modifiers.
