---
title: "Simulation with Parameter Uncertainty in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
```

## Why simulate with parameter uncertainty?

Population PK/PD parameter estimates carry uncertainty expressed by
their variance-covariance matrix (from the covariance step of the
estimator) or by bootstrap replicates.  Propagating this uncertainty
into simulations produces confidence intervals on predicted
concentrations, exposure metrics, or target attainment probabilities
that reflect both between-subject variability (BSV) *and* uncertainty
in the model parameters.

## Model and parameter setup

```{r model}
pkModel <- function() {
  ini({
    tka  <- log(1.57);  label("log Ka")
    tcl  <- log(2.72);  label("log CL")
    tv   <- log(31.07); label("log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

## Variance-covariance matrix for the fixed effects (thetas)
## In practice this comes from your estimation software output
thetaVcov <- lotri::lotri({
  tka ~ 0.04
  tcl ~ c(0.01, 0.02)
  tv  ~ 0.03
})

## Extend the model with an AUC compartment via model piping.
## d/dt(auc) <- cp accumulates concentration over time, so auc at
## the last observation time equals AUC_0-t directly.
pkModelAuc <- pkModel |>
  model(d/dt(auc) <- cp, append = TRUE)
```

## Built-in simulation with parameter uncertainty

Pass `thetaMat` (the theta variance-covariance matrix) and `nStud`
(number of virtual studies, i.e. parameter draws) directly to
`rxSolve()`.  Internally rxode2 samples `nStud` parameter vectors from
a multivariate normal distribution and simulates `nSub` subjects for
each draw.

```{r sim, message=FALSE}
rxSetSeed(42)

et <- et(amt = 100, time = 0) |>
  et(seq(0, 24, by = 0.5))

sim <- rxSolve(
  pkModelAuc,
  et,
  thetaMat = thetaVcov,   # uncertainty matrix for thetas
  nStud    = 500,          # virtual studies (parameter draws)
  nSub     = 20            # subjects per study
)
```

## Confidence interval on the concentration-time profile

Call `confint()` on the solved object to summarize across both
parameter draws and subjects.  When `nStud > 1` rxode2 uses the study
dimension to construct confidence bands around the percentile curves.

```{r ciProfile, fig.width=7, fig.height=4}
ci <- confint(sim, "cp", level = 0.95)
plot(ci)
```

The shaded ribbon spans the 2.5th to 97.5th percentile of the
population prediction (combining BSV and parameter uncertainty); the
solid line is the median.  The `ci` argument to `confint()` controls
the width of the confidence bands placed around those percentiles:

```{r ciProfileNarrow, fig.width=7, fig.height=4}
## 90% interval with 95% confidence bands around it
ci90 <- confint(sim, "cp", level = 0.90, ci = 0.95)
plot(ci90)
```

## Uncertainty in a summary exposure metric (AUC)

Extract the per-study, per-subject AUC from the simulation and
summarize across subjects within each study to obtain the distribution
of study-level median AUC.

```{r aucDist, fig.width=6, fig.height=4, message=FALSE}
library(data.table)

dt <- as.data.table(sim)

## auc at time == 24 is AUC_0-24 directly (no trapezoid rule needed)
aucByStudy <- dt[time == 24, .(medAUC = median(auc)), by = sim.id]

## Point estimate and 90% CI across draws
quantile(aucByStudy$medAUC, c(0.05, 0.50, 0.95))

ggplot(aucByStudy, aes(x = medAUC)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = quantile(aucByStudy$medAUC, c(0.05, 0.95)),
             linetype = "dashed") +
  labs(x = "Median AUC per study (mg*h/L)",
       y = "Count",
       title = "Distribution of median AUC across parameter draws",
       subtitle = "Dashed lines: 5th and 95th percentiles") +
  theme_bw()
```

## Forest plot for covariate effects with uncertainty

Simulate each covariate scenario with the same parameter uncertainty,
then compare the AUC ratio relative to a reference.

```{r forestSetup, message=FALSE}
## Helper: simulate one scenario and return median AUC per study
simScenAuc <- function(params, label) {
  s <- rxSolve(
    pkModelAuc, et,
    params   = params,
    thetaMat = thetaVcov,
    nStud    = 500,
    nSub     = 10
  )
  dt <- as.data.table(s)
  ## auc at the last time point is AUC_0-24 directly
  med <- dt[time == 24, .(medAUC = median(auc)), by = sim.id]
  med[, scenario := label]
  med
}

set.seed(42)
scenarios <- list(
  reference  = c(tka = log(1.57), tcl = log(2.72),       tv = log(31.07)),
  mild_ri    = c(tka = log(1.57), tcl = log(2.72 * 0.80), tv = log(31.07)),
  moderate_ri = c(tka = log(1.57), tcl = log(2.72 * 0.50), tv = log(31.07))
)

aucAll <- rbindlist(Map(simScenAuc, scenarios, names(scenarios)))
```

```{r forestPlot, fig.width=7, fig.height=4}
## AUC ratio relative to reference, median + 90% CI
refAuc <- aucAll[scenario == "reference", medAUC]

forest <- aucAll[scenario != "reference",
  .(
    ratio    = median(medAUC) / median(refAuc),
    lo       = quantile(medAUC / median(refAuc), 0.05),
    hi       = quantile(medAUC / median(refAuc), 0.95)
  ),
  by = scenario
]
forest[, scenario := factor(scenario,
  levels = c("mild_ri", "moderate_ri"),
  labels = c("Mild RI (CL -20%)", "Moderate RI (CL -50%)"))]

ggplot(forest, aes(x = ratio, y = scenario)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.2) +
  geom_vline(xintercept = 1,        linetype = "dashed") +
  geom_vline(xintercept = c(0.8, 1.25), linetype = "dotted", colour = "red") +
  scale_x_continuous(limits = c(0.3, 2.0),
                     breaks = c(0.5, 0.8, 1.0, 1.25, 1.5)) +
  labs(x = "AUC ratio vs reference",
       y = NULL,
       title = "Forest plot: covariate effects on AUC",
       subtitle = "Points: median ratio; bars: 90% CI from parameter uncertainty\nDotted red: 0.8-1.25 no-effect boundary") +
  theme_bw()
```
