---
title: "Out-of-memory solving with rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
```

## Why chunked solving?

Population pharmacokinetic/pharmacodynamic simulations can involve tens or
hundreds of thousands of subjects.  Each subject requires solver workspace,
output arrays, and covariate storage, so the combined memory footprint grows
linearly with the number of subjects.  A 100,000-subject simulation with a
two-compartment model and rich sampling can easily require 10–50 GB of RAM —
more than many workstations or cloud instances provide.

rxode2 addresses this by **chunked (out-of-memory) solving**: subjects are
split into batches, each batch is solved independently, and the result of each
batch is written to a temporary file (parquet or rds).  The final return value
is a lightweight `rxSolveOom` object that reads chunks on demand rather than
holding the entire dataset in RAM.

> For running chunks in parallel across multiple CPU cores or compute nodes,
> see the [Parallel rxode2 solving with mirai](rxode2-mirai-parallel.html)
> vignette.

## The model and event table

Throughout this vignette we use a one-compartment oral PK model with between-
subject variability on clearance and absorption rate.  Parameters are defined
using `ini()` and `model()` blocks — the standard rxode2 UI approach.

```{r model}
pkModel <- function() {
  ini({
    tka      <- log(0.5);  label("Typical log Ka (1/h)")
    tcl      <- log(4);    label("Typical log CL (L/h)")
    tv       <- log(70);   label("Typical log V (L)")
    eta.ka   ~ 0.09
    eta.cl   ~ 0.04
    prop.err <- 0.15
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp           <- center / v
    cp ~ prop(prop.err)
  })
}
```

The event table includes 10 subjects.  With subject IDs in the event table,
rxode2 draws one set of etas per subject from the `eta.ka ~ 0.09` and
`eta.cl ~ 0.04` variances defined in `ini()`.

```{r events}
ev <- et(amt = 100, ii = 24, until = 72) |>
  et(seq(0, 72, by = 4)) |>
  et(id = 1:10)
```

## Estimating memory usage

Before solving you can inspect the expected memory footprint with
`rxMemoryEstimate()`.  Pass the event table (as a data frame) and the compiled
model; the function calls the same allocator formulas used internally by
`rxSolve()`.

```{r mem-estimate}
mod <- rxode2(pkModel)
rxMemoryEstimate(as.data.frame(ev), model = mod)
```

The output lists every major allocation (ODE state buffers, event-time arrays,
output data, etc.) as well as the fraction of total and free RAM the solve
would consume.  The `freeRamBytes` line is the key guard: if `total` exceeds
90 % of free RAM, rxode2 will refuse to proceed without `file`.

## Automatic OOM detection

When rxode2 detects that a solve would use more than 90 % of free RAM it stops
immediately and prints a message pointing to the `file` argument:

```r
# This would error on a machine with insufficient RAM:
rxSolve(pkModel, ev)
#> Error: Solve requires 12.3 GB but only 6.1 GB appears free.
#> Re-run with rxSolve(..., file = 'path/prefix') to solve in chunks.
```

## Chunked solving with `rxSolve()`

Pass `file` to opt in to chunked solving.  rxode2 will:

1. Auto-compute a chunk size from available free RAM (or use `chunkSize` if
   you supply it).
2. Solve each chunk with a separate `rxSolve()` call.
3. Write each chunk's result to `<file>_chunk_NNNNN.parquet` (or `.rds` if
   the `arrow` package is unavailable).
4. Save a manifest to `<file>_manifest.rds`.
5. Return an `rxSolveOom` object.

```{r oom-solve}
res <- rxSolve(pkModel, ev,
               file      = tempfile("rxChunk"),
               chunkSize = 3)   # 3 subjects per chunk
res
```

The print method reports how many chunks were created and the total row count.
With `chunkSize = 3` and 10 subjects, rxode2 creates four chunks (3 + 3 + 3 + 1).

## Accessing results

### Materialise the full data frame

```{r materialise}
df <- as.data.frame(res)
head(df)
```

This reads and `rbind`s every chunk file.  For very large results keep the
`rxSolveOom` object in place and only materialise columns you actually need.

### Lazy column access

```{r lazy-col}
cp_all <- res$cp
length(cp_all)
```

`res$col` reads only the requested column from each chunk, avoiding loading
columns you do not need.

### Row and dimension queries

```{r dims}
nrow(res)
dim(res)
```

## Faking low memory in demos and tests

The easiest way to demonstrate or test the chunked path on a small dataset is
to set `chunkSize` directly — no need to modify system RAM.  With
`chunkSize = 2` on 10 subjects you get 5 chunks of 2 subjects each:

```{r fake-oom}
res_small <- rxSolve(pkModel, ev,
                     file      = tempfile("rxDemo"),
                     chunkSize = 2)
res_small
```

In production code, omit `chunkSize` entirely and rxode2 auto-computes an
appropriate chunk size from `rxMemoryEstimate()`.

## File-backed event tables

For datasets so large that even the event table does not fit in RAM, save it to
a file first and wrap the path with `rxEventTableFile()`:

```{r et-file, eval = requireNamespace("arrow", quietly = TRUE)}
et_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(as.data.frame(ev), et_path)

et_file <- rxEventTableFile(et_path, format = "parquet")

res_file <- rxSolve(pkModel, et_file,
                    file      = tempfile("rxFile"),
                    chunkSize = 3)
res_file
```

`rxEventTableFile()` also accepts `"csv"`, `"fst"`, and `"rds"` formats.

## Persistent output files

By default `file` points to a `tempfile()` prefix, so the chunk files and
manifest are cleaned up when R's garbage collection removes the temp directory.
To keep results across sessions, point `file` at a directory in your project:

```r
res_persist <- rxSolve(pkModel, ev,
                       file      = "output/my_sim",
                       chunkSize = 5000)
```

Files written will be `output/my_sim_chunk_00001.parquet`, ...,
`output/my_sim_manifest.rds`.  Reload the manifest later:

```r
manifest <- readRDS("output/my_sim_manifest.rds")
```

## Working with Arrow datasets

The combination of chunked solving and Arrow datasets means rxode2 can handle
populations that are **too large to fit in memory at any stage** — not just
during the solve, but also when post-processing the results.  Chunk files stay
on disk; Arrow reads only what your query needs.

When the `arrow` package is installed (which it must be for parquet chunk
files), rxode2 exposes two Arrow-native accessors on the `rxSolveOom` object.

### Lazy Arrow Dataset — querying without loading everything

`as.arrow()` (an rxode2 generic; arrow itself has no such function)
opens all chunk files as a single lazy `arrow::Dataset`.  No rows are read
from disk until you explicitly collect them.  You can push filters, column selections, and aggregations down to the
Arrow engine so that only the result — not the full dataset — is brought into R
memory:

```{r arrow-dataset, eval = requireNamespace("arrow", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)}
ds <- as.arrow(res)
ds
```

Because `ds` is lazy you can chain dplyr verbs and only call `dplyr::collect()`
at the end to materialize the (small) result:

```{r arrow-dplyr, eval = requireNamespace("arrow", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)}
# Compute the peak concentration per subject without loading all rows.
# Arrow cannot translate argmax expressions like time[which.max(cp)], so
# tmax comes from a lazy self-join: keep the rows where cp hits the
# per-subject maximum.
cmax_tbl <- ds |>
  dplyr::group_by(id) |>
  dplyr::summarise(cmax = max(cp, na.rm = TRUE))

pk_summary <- ds |>
  dplyr::inner_join(cmax_tbl, by = "id") |>
  dplyr::filter(cp == cmax) |>
  dplyr::select(id, cmax, tmax = time) |>
  dplyr::collect() |>
  dplyr::arrange(id)

pk_summary
```

This is the key advantage over materialising with `as.data.frame()`: for a
100,000-subject simulation that would require 40 GB of RAM as an R data frame,
the summary above reads each chunk sequentially and returns only one row per
subject — a result that fits easily in memory even when the full dataset does
not.

Other useful patterns:

```{r arrow-dplyr2, eval = requireNamespace("arrow", quietly = TRUE) && requireNamespace("dplyr", quietly = TRUE)}
# Keep only observations after the first dose interval
late_obs <- ds |>
  dplyr::filter(time > 24) |>
  dplyr::select(id, time, cp) |>
  dplyr::collect()

head(late_obs)
```

### In-memory Arrow Table

`arrow::as_arrow_table()` reads and concatenates all chunk files into a single
`arrow::Table` in columnar memory.  Use this when the full result fits in RAM
but you want the speed of Arrow columnar operations rather than R's row-oriented
`rbind()`:

```{r arrow-table, eval = requireNamespace("arrow", quietly = TRUE)}
tbl <- arrow::as_arrow_table(res)
tbl
```

### Materialisation via Arrow

When arrow is available, `as.data.frame()`, `tibble::as_tibble()`, and
`data.table::as.data.table()` all route through `as_arrow_table()` internally,
so they benefit from Arrow's concatenation speed rather than R's `rbind()`:

```{r arrow-materialise, eval = requireNamespace("arrow", quietly = TRUE)}
df  <- as.data.frame(res)       # Arrow path when arrow installed
```

```{r arrow-tibble, eval = requireNamespace("arrow", quietly = TRUE) && requireNamespace("tibble", quietly = TRUE)}
tib <- tibble::as_tibble(res)   # also Arrow-backed
```

## Parallel chunk solving

Each chunk is independent and can be solved on a separate CPU or compute node
using the `mirai` package.  Pass `parallel` to control the number of worker
processes:

```r
res_par <- rxSolve(pkModel, ev,
                   file      = tempfile("rxPar"),
                   chunkSize = 3,
                   parallel  = 2)
```

See the [Parallel rxode2 solving with mirai](rxode2-mirai-parallel.html)
vignette for a full walkthrough of daemon setup, performance tuning, and
custom parallel workflows.

## Session information

```{r session}
sessionInfo()
```
