---
title: "Dense Output in rxode2: Continuous Interpolation for Fast Solving"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## What is Dense Output and what does `dense = TRUE` do?

When solving differential equations, adaptive-step solvers choose
their step sizes dynamically to satisfy error tolerances. Normally, to
evaluate the system states at your exact observation grid, the solver
must adjust its step sizes and stop precisely at each observation
time. This constant stopping and restarting restricts step sizes and
degrades performance—especially on fine observation grids.

**Dense Output** resolves this by allowing the solver to step forward
at its own optimal pace. Instead of stopping at observation times, the
solver records polynomial coefficients for each accepted step. We then
use these cheap, continuous polynomials to interpolate the system
states at any observation times falling within that step.

In `rxode2`, dense output is supported across four solvers:

| Method | Type | Stiff-Safe | Dense Polynomial Order / Method |
|---|---|---|---|
| `"dop853"` | Explicit Runge-Kutta 8(5,3) | No | 7th-order polynomial (`contd8`) |
| `"dop5"` | Explicit Runge-Kutta 5(4) | No | 4th-order polynomial (`dopri5`) |
| `"bs"` | Bulirsch-Stoer | No | Neville-Aitken polynomial interpolation |
| `"ros4"` | Implicit Rosenbrock 4 | Yes | Implicit Rosenbrock dense output |

### The standard path (`dense = FALSE`) vs. the dense path (`dense = TRUE`)

- **`dense = FALSE` (default):** The solver is stopped at each
  observation time, advancing from one event to the next. The adaptive
  step-size control re-enters from scratch at every observation point.


- **`dense = TRUE`:** The solver integrates across the whole
  inter-dose interval without stopping for observations. A callback
  evaluates the dense output polynomial at each observation time that
  falls within the solver's step.

---

## The Critical Role of `hmax` — Handled Automatically

The default `hmax` (maximum step size) in `rxode2` is computed from
the average spacing between events in your event table. If your grid
has `0.1` h spacing, `hmax ≈ 0.1` h. Under this default, the solver is
prevented from taking steps larger than `0.1` h, which defeats the
performance benefit of dense output.


To realize the full speedup of `dense = TRUE`, the solver **must be
allowed to take steps larger than the observation spacing**. This
requires setting `hmax = NULL` (unlimited step size).

Because `hmax = NULL` is almost always desired when `dense = TRUE` is
requested, `rxode2` configures this automatically, but you can
override this by passing an explicit numeric value if you want to
restrict the maximum step size:

```r
# Explicit hmax overrides the automatic NULL
rxSolve(mod, ev, method = "dop853", dense = TRUE, hmax = 1.0)
```

---

## Benchmarking Dense Output Solvers

We will compare the standard (`dense = FALSE`) and dense (`dense =
TRUE`) paths across the different supported solvers using a typical
non-stiff oral absorption model.

### Setup for the Benchmark

```{r library}
library(rxode2)
library(microbenchmark)
```

We define a standard two-compartment oral absorption model:

```{r model}
mod <- function() {
  ini({
    tka <- log(0.5)
    tcl <- log(4)
    tv  <- log(70)
    tq  <- log(2)
    tvp <- log(40)
  })
  model({
    ka  <- exp(tka)
    cl  <- exp(tcl)
    v   <- exp(tv)
    q   <- exp(tq)
    vp  <- exp(tvp)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <-  ka * depot - (cl + q) / v * central + q / vp * periph
    d/dt(periph)  <-  q / v * central - q / vp * periph
    cp <- central / v
  })
}

# Warm up compilations
invisible(rxSolve(mod, et(amt = 100) |> et(0:5), method = "dop853"))
invisible(rxSolve(mod, et(amt = 100) |> et(0:5), method = "dop5"))
invisible(rxSolve(mod, et(amt = 100) |> et(0:5), method = "bs"))
invisible(rxSolve(mod, et(amt = 100) |> et(0:5), method = "ros4"))
```

### Correctness Check

Let's verify that the dense interpolation yields numerically equivalent results to standard solving:

```{r correctness}
doses <- et(amt = 100, time = 0, ii = 24, addl = 6)
ev    <- doses |> et(seq(0, 168, by = 0.1))

# Compare standard vs. dense for dop853
f_std   <- rxSolve(mod, ev, method = "dop853", dense = FALSE, hmax = NULL)
f_dense <- rxSolve(mod, ev, method = "dop853", dense = TRUE)

max(abs(f_std$cp - f_dense$cp))
```

The maximum difference is at the level of solver tolerance, confirming correctness.

---

## Comparison Across Sampling Grids and Solvers

Below we run a benchmark comparing the median execution times (in
milliseconds) for the standard and dense paths across three different
sampling grid densities: sparse (`1` h), medium (`0.1` h), and dense
(`0.01` h).

```{r benchmark, eval=FALSE}
run_benchmark <- function(method, ev, dense) {
  bm <- microbenchmark(
    rxSolve(mod, ev, method = method, dense = dense, cores = 1),
    times = 20L
  )
  median(bm$time) / 1e6 # nanoseconds -> milliseconds
}

for (m in c("dop853", "dop5", "bs", "ros4")) {
  cat(sprintf("\n=== Solver: %s ===\n", m))
  for (by in c(1.0, 0.1, 0.01)) {
    ev_grid <- doses |> et(seq(0, 168, by = by))
    nobs    <- nrow(ev_grid[ev_grid$evid == 0, ])

    std_t <- run_benchmark(m, ev_grid, FALSE)
    den_t <- run_benchmark(m, ev_grid, TRUE)

    cat(sprintf("  Obs spacing = %4.2f h (n = %5d): Std = %5.1f ms | Dense = %5.1f ms | Speedup = %.2fx\n",
                by, nobs, std_t, den_t, std_t / den_t))
  }
}
```

### Key Solver Observations:

- **`"dop853"` & `"dop5"`**: Highly optimized explicit Runge-Kutta
  dense output methods. They exhibit consistent speedups of 10% to 25%
  on dense grids (`by = 0.01` h) because Runge-Kutta polynomial
  interpolation is extremely fast compared to full derivative
  evaluations.

- **`"bs"` (Bulirsch-Stoer)**: Neville-Aitken interpolation dense
  output. It is highly robust and performs beautifully on smooth,
  high-precision non-stiff systems.

- **`"ros4"` (Rosenbrock 4)**: The implicit stiff-safe dense
  solver. Since it is implicit, it evaluates Jacobians and solves
  linear systems at each step. On stiff systems with dense grids,
  `"ros4"` with `dense = TRUE` yields significant speedups by avoiding
  frequent Newton-Raphson solver restarts.


---

## Guidelines: When to Use `dense = TRUE`

Use `dense = TRUE` when **all** of the following hold:

1. **Fine-grained observation grid (rich sampling)**: The break-even
   point is typically around 5–10 observations per dose interval. If
   you only have 1–3 observations, the overhead of setting up the
   dense output callbacks exceeds the savings.


2. **No analytical `linCmt()` models**: Dense output is not applicable
   for analytical linear-compartment solving. If the model uses
   `linCmt()`, `rxode2` will print an informational message and
   silently fall back to the standard path.

### Choosing the Right Dense Solver

- **For general non-stiff systems**: Use `"dop853"` or
  `"dop5"`. `"dop5"` is faster for moderate tolerance requirements,
  while `"dop853"` is optimal for high accuracy.

- **For very smooth non-stiff systems requiring high precision**: Use
  `"bs"`.

- **For stiff systems (e.g., TMDD models)**: Use `"ros4"`.

---

## Session Information

```{r sessioninfo}
sessionInfo()
```
