---
title: "Matrix Exponentials and Inductive Linearization in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Background

Nonlinear ordinary differential equations (ODEs) arise naturally in
pharmacokinetic--pharmacodynamic (PKPD) models.  Consider a drug that
undergoes Michaelis--Menten (saturable) elimination: the clearance
depends on the current concentration, making the ODE nonlinear.
Standard time-stepping solvers (LSODA, Runge--Kutta) handle this by
advancing one small step at a time, using only local information about
the solution.

**Inductive Linearization** (IndLin) takes a different route.  Any ODE
system can be written in the standard form

```
dy/dt = f(t, y) + A(t, y) * y,     y(t0) = y0
```

where `A` collects everything that is linear in the states and `f`
collects the rest.  IndLin turns this into a *linear time-varying* (LTV)
system by evaluating `A` at the previous iteration's trajectory:

```
dy^[n]/dt = f(t) + A(t, y^[n-1]) * y^[n],     y(t0) = y0
```

Starting from the plug-in value `y^[0](t) = 0` for all `t`, each pass
`n = 1, 2, ...` produces a better approximation of the true solution.
Because the system is now linear, each pass can be integrated with the
**matrix exponential** (ME) via eigenvalue decomposition (EVD),

```
y^[n](t) = V * diag(exp(lambda * t)) * V^-1 * y0
```

where `V` are the eigenvectors and `lambda` the eigenvalues of the rate
matrix `K`.

The method was introduced by Duffull and Hegarty (2014) and developed
further by Hasegawa and Duffull (2018); the EVD pairing, the convergence
stopping rule, and the adaptive step size used here are described in
[Sharif, Hasegawa and Duffull
(2022)](https://doi.org/10.1007/s10928-022-09813-z), which is the source
of the Michaelis--Menten example used throughout this article.

### Key properties

- For a purely **linear** system, `A` does not depend on `y` at all, the
  iteration is finished after one pass, and the matrix exponential gives
  the exact solution.
- For a **nonlinear** system the method is approximate, controlled by how
  often the linearization is refreshed.
- The linearized system is solved over an interval rather than at a
  single point, so no local step-size control is needed inside the
  interval.

## Three things named `indLin`

The name shows up in three related but distinct places, so it is worth
separating them before going further:

| Name | What it is | What it does |
|---|---|---|
| `indLin(model)` | An R function | Rewrites a `d/dt()` model into matrix-exponential (`matExp()`) form and returns the model code as text |
| `indLin(state) <- expr` | A model statement | Adds a time-varying forcing term `f(t)` to one compartment inside a `matExp()` model |
| `method = "indLin"` | An `rxSolve()` option | Solves with the matrix-exponential integrator instead of a time-stepping solver |

The rest of this article covers each in turn.

## What `indLin()` does -- a simple example

`indLin()` is the symbolic half of the method.  Give it a model written
with `d/dt()` and it returns the same model written as a rate matrix:

```{r}
library(rxode2)

oralOde <- function() {
  ini({
    ka <- 1
    cl <- 0.1
    v <- 1
  })
  model({
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - (cl / v) * central
    cp <- central / v
  })
}

cat(indLin(oralOde))
```

which prints

```
matExp()
cmt(depot)
cmt(central)
k_depot_central = ka
k_central_output = cl/v
cp=central/v
```

Read that against the standard form above.  `indLin()` differentiated
each right-hand side with respect to each state to get

```
A = [ -ka       0    ]      f = [ 0 ]
    [  ka   -cl/v    ]          [ 0 ]
```

and then reported `A` through its *micro-constants* instead of as a
matrix: the off-diagonal `A[central, depot] = ka` becomes
`k_depot_central`, and the column sum that is lost to the outside world,
`-(A[depot, depot] + A[central, depot])`, becomes `k_central_output`.
The diagonal is implied, so it never has to be written down.  Nothing
else about the model changes -- the output line `cp = central/v` is
carried through untouched.

The interesting case is a nonlinear one.  Michaelis--Menten elimination
gives:

```{r}
mmOde <- function() {
  ini({
    ka <- 1
    km <- 0.5
    vmax <- 0.2
    v <- 1
  })
  model({
    d/dt(depot) <- -ka * depot
    cp <- central / v
    d/dt(central) <- ka * depot - vmax * cp / (km + cp)
  })
}

cat(indLin(mmOde))
```

```
matExp()
cmt(depot)
cmt(central)
k_depot_central = ka
k_central_output = vmax/(v*(km + central/v))
cp=central/v
```

The structure is identical, but `k_central_output` now *contains a
state*.  That is exactly the `A(t, y)` of the standard form: it is a
first-order rate constant only once you decide what value of `central`
to plug into it.  Choosing the previous iterate for that plug-in value is
the whole idea of inductive linearization, and it reproduces the rate
matrix of Sharif et al. (2022), Eq. 8:

```
K = [ -ka                          0                    ]
    [  ka    -(1/v) * vmax / (km + C^[n-1])              ]
```

You rarely have to call `indLin()` yourself: `rxSolve(..., method =
"indLin")` on an ODE model runs this conversion internally and solves the
converted model.  Calling it directly is how you inspect the
factorization, or how you obtain a `matExp()` model you want to edit and
keep.

## The iteration, by hand

The conversion above is only half the story.  To see what the iteration
does, here is IndLin written out in plain R for the Michaelis--Menten
model, using EVD for the matrix exponential exactly as in the reference:

```{r}
ka <- 1
km <- 0.5
vmax <- 0.2
v <- 1
dose <- 3
tGrid <- seq(0, 30, by = 0.05)

## rate matrix K, with the previous iterate's concentration plugged in
kMatrix <- function(cPrev) {
  matrix(c(-ka, ka,
           0, -vmax / (v * (km + cPrev))),
         nrow = 2L)
}

## one exact step of a frozen (time-invariant) system, via EVD
evdStep <- function(kMat, y0, dt) {
  ev <- eigen(kMat)
  Re(ev$vectors %*% diag(exp(ev$values * dt)) %*% solve(ev$vectors, y0))
}

## one inductive-linearization pass over the whole interval
indLinPass <- function(cPrev) {
  y <- c(dose, 0)
  cNew <- numeric(length(tGrid))
  for (i in seq_along(tGrid)[-1]) {
    y <- evdStep(kMatrix(cPrev[i - 1L]), y, tGrid[i] - tGrid[i - 1L])
    cNew[i] <- y[2] / v
  }
  cNew
}

## y^[0](t) = 0 for all t, then iterate to the stopping rule
cCur <- rep(0, length(tGrid))
for (n in 1:20) {
  cNext <- indLinPass(cCur)
  relErr <- max(abs(cNext - cCur) / ifelse(cNext == 0, 1, cNext))
  message(sprintf("iteration %2d  max relative change %.3e", n, relErr))
  cCur <- cNext
  if (relErr < 1e-6) break
}
```

```
iteration  1  max relative change 1.000e+00
iteration  2  max relative change 8.443e-01
...
iteration 13  max relative change 2.953e-05
iteration 14  max relative change 4.347e-06
iteration 15  max relative change 5.879e-07
```

Fifteen passes and the successive iterates stop moving; comparing against
LSODA on the same grid gives agreement to about 0.6%, the discretization
error of freezing `K` across each 0.05 h step.  The stopping rule here is
the one proposed in Sharif et al. (2022), Eq. 11: stop when the maximum
absolute relative change between successive iterates falls below a
tolerance.

```{r}
library(ggplot2)

ref <- rxSolve(mmOde, et(amt = 3) |> et(tGrid), method = "liblsoda")

compare <- data.frame(
  time = rep(tGrid, 2L),
  cp = c(cCur, ref$cp),
  solver = rep(c("IndLin (by hand)", "LSODA"), each = length(tGrid))
)

ggplot(compare, aes(time, cp, colour = solver, linetype = solver)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)", colour = "Solver",
       linetype = "Solver")
```

## How rxode2 solves it

rxode2 does not run the fixed-point loop above.  Instead it marches
forward and *relinearizes as it goes*: over each substep the rate matrix
is evaluated once at the current state, the matrix exponential advances
the states exactly for that frozen matrix, and the next substep starts
from the updated state.  This is the same linearization, applied one
substep at a time rather than iterated over the whole interval.

The practical consequence is that accuracy is controlled by **how long a
substep is allowed to be**, and that is what `hmax` sets:

```{r}
solveMm <- function(...) {
  rxSolve(mmOde, et(amt = 3) |> et(c(0.1, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12, 16, 24, 30)), ...)
}

refMm <- solveMm(method = "liblsoda")

maxRelErr <- function(sol) max(abs(sol$cp - refMm$cp) / refMm$cp)

maxRelErr(solveMm(method = "indLin"))
maxRelErr(solveMm(method = "indLin", hmax = 0.1))
maxRelErr(solveMm(method = "indLin", hmax = 0.01))
```

| `hmax` | Maximum relative error vs LSODA |
|---|---|
| default (the spacing of your own sampling times) | 7.1e-01 |
| `1` | 2.6e-01 |
| `0.1` | 1.3e-02 |
| `0.01` | 1.3e-03 |
| `0.001` | 1.4e-04 |

> **Set `hmax` for nonlinear models.**  The default `hmax` is tied to the
> spacing of your own sampling and dosing times, which has nothing to do
> with how fast the rate matrix changes; with the sparse sampling grid
> above that default is badly too coarse.  Always set an explicit `hmax`
> when solving a state-dependent model with `method = "indLin"`, and
> confirm the answer is stable when you halve it.

For a **linear** model none of this applies: the rate matrix is constant,
every substep length gives the same answer, and the matrix exponential is
exact.

```{r}
oralMe <- function() {
  ini({
    ka <- 1
    cl <- 0.1
    v <- 1
  })
  model({
    matExp()
    cmt(depot)
    cmt(central)
    k_depot_central <- ka
    k_central_output <- cl / v
    cp <- central / v
  })
}

ev <- et(amt = 100, ii = 12, until = 48) |> et(seq(0, 72, by = 0.5))

meSol <- rxSolve(oralMe, ev, method = "indLin")
odeSol <- rxSolve(oralOde, ev, method = "liblsoda")

## differences are at machine precision (~3e-13)
max(abs(meSol$cp - odeSol$cp))
```

## Matrix exponential algorithms

Three ME algorithms are available via `indLinMatExpType`:

| Value | Algorithm | Notes |
|---|---|---|
| `"expokit"` | Sidje (1998) Krylov/Pade | **Default.** Good for sparse or large matrices; controlled by `indLinPhiTol` and `indLinPhiM` |
| `"Al-Mohy"` | Al-Mohy and Higham (2009) scaling and squaring | Accurate; order controlled by `indLinMatExpOrder` |
| `"arma"` | RcppArmadillo `expmat` | Fastest for small dense matrices |

```{r}
almohySol <- rxSolve(mmOde, ev, method = "indLin", hmax = 0.01,
                     indLinMatExpType = "Al-Mohy", indLinMatExpOrder = 8L)

armaSol <- rxSolve(mmOde, ev, method = "indLin", hmax = 0.01,
                   indLinMatExpType = "arma")
```

For the expokit method, two additional controls are available:

- `indLinPhiTol` (default `1e-7`): tolerance for the exponential-matrix
  action computation.
- `indLinPhiM` (default `0L`, meaning automatic): maximum Krylov basis
  size.

## Controlling the linearization strategy

When an ODE term is the product of two or more states (`y * z`), there is
no unique way to split it into `A * y`: the term can be assigned to `y`,
to `z`, or shared.  Two session-level settings control the choice made by
`indLin()`.  The van der Pol oscillator, whose `mu * (1 - y^2) * dy` term
mixes both states, shows the difference.

```{r}
vanPol <- function() {
  ini({
    mu <- 1
  })
  model({
    y(0) <- 2
    dy(0) <- 0
    d/dt(y) <- dy
    d/dt(dy) <- mu * (1 - y^2) * dy - y
  })
}
```

### `rxIndLinStrategy()`

- `"curState"` (default): prefer the state whose derivative is being
  computed; fall back to the first state found when it is absent.
- `"split"`: divide the term equally among all states involved.

```{r}
cat(indLin(vanPol))
```

```
matExp()
cmt(y)
cmt(dy)
k_y_dy = -1
k_y_output = 1
k_dy_y = 1
k_dy_output = -(1 + mu - y^2*mu)
y(0)=2
dy(0)=0
```

```{r}
rxIndLinStrategy("split")
cat(indLin(vanPol))
rxIndLinStrategy("curState") # restore the default
```

```
matExp()
cmt(y)
cmt(dy)
k_y_dy = -1+(-1/2)*mu*dy*Rx_pow_di(y, 2)/y
k_y_output = -(-1 + (-1/2)*y*mu*dy)
k_dy_y = 1
k_dy_output = -(1 + mu + (-1/2)*y^2*mu)
y(0)=2
dy(0)=0
```

### `rxIndLinState()`

For fine-grained control, name the state that should carry the linear
factor for each ODE.  Sending the whole nonlinear term into `d/dt(y)`
leaves `k_dy_output` free of any state, so that row of the rate matrix
becomes constant:

```{r}
rxIndLinState(list(y = "dy", dy = "y"))
cat(indLin(vanPol))
rxIndLinState(NULL) # clear preferences
```

```
matExp()
cmt(y)
cmt(dy)
k_y_dy = -1-mu*dy*Rx_pow_di(y, 2)/y
k_y_output = -(-1 - y*mu*dy)
k_dy_y = 1
k_dy_output = -(1 + mu)
y(0)=2
dy(0)=0
```

## Writing the rate matrix directly with `matExp()`

Everything above went through `symengine` to factor the ODEs
symbolically.  For complex expressions that step can fail or produce an
unexpected factorization.  `matExp()` skips it entirely: you write the
rate constants, and rxode2 builds the rate matrix from them at compile
time.  This is:

- **The most stable path** -- no CAS factorization, no symengine
  dependency, no expression simplification.
- **Familiar to NONMEM users** -- it mirrors ADVAN5/7 `K(i,j)` notation
  with the same source-first index convention.
- **Self-documenting** -- every rate constant name says what it connects.

### `matExp()` model syntax

| Statement | Meaning |
|---|---|
| `matExp()` | Declare this model uses matrix exponential integration. **Required.** Cannot be combined with `d/dt()`. |
| `k_from_to <- expr` | First-order rate from compartment `from` to compartment `to`. |
| `k.from.to <- expr` | Dot-separator alternative (identical meaning). |
| `k_cmt_output <- expr` | Elimination from `cmt` to outside the system (like NONMEM's `K(i,0)`). |
| `k_from_to_nd <- expr` | Non-depleting transfer: only `to` gains; `from` is unchanged. |
| `indLin(state) <- expr` | Add a time-varying forcing term to `state` (the inhomogeneous part). |
| `cmt(name)` | Optionally declare a compartment explicitly. |
| `other <- expr` | Standard derived-variable assignment (output calculation). |

Compartments are **auto-detected** from the `k_from_to` names, and the
diagonal entries of the rate matrix are computed automatically from the
outflow terms.

### Declaring compartment order with `cmt()`

Without `cmt()` declarations, compartment order in the rate matrix -- and
therefore the CMT numbers used in the event table -- follows the order in
which names first appear in the `k_from_to` assignments.  Use `cmt()`
before the rate constants to pin the numbering, for example to match a
NONMEM control stream where the depot is CMT=1:

```{r}
mmComp <- function() {
  ini({
    ka <- 1
    km <- 0.5
    vmax <- 0.2
    v <- 1
  })
  model({
    matExp()
    cmt(depot)   ## CMT = 1
    cmt(central) ## CMT = 2
    cp <- central / v
    k_depot_central <- ka ## absorption: depot -> central
    ## MM elimination as an effective first-order rate on the AMOUNT:
    ## vmax * cp / (km + cp) = [vmax / (v * (km + cp))] * central
    k_central_output <- vmax / (v * (km + cp))
  })
}
```

Compartments that appear in rate constants but have no `cmt()`
declaration are appended after the declared ones in first-appearance
order, so you can fix the positions that matter and let the rest be
auto-ordered.  `cmt()` declarations are purely for ordering; they do not
change the model mathematics.

### Rules and restrictions

```{r}
## ERROR: matExp() cannot be combined with d/dt()
rxode2({
  matExp()
  d/dt(depot) <- -ka * depot
})
#> :ERR: Matrix exponential models cannot be used with any ODEs

## ERROR: self-transfer is not allowed
rxode2({
  matExp()
  k_depot_depot <- 0.1
})
#> :ERR: transfer from a compartment to itself (e.g., k_cmt1_cmt1) is not allowed

## ERROR: indLin() without matExp()
rxode2({
  cmt(depot)
  indLin(depot) <- 0.5
})
#> :ERR: indLin() cannot be used without matExp() defined in the model
```

### Michaelis--Menten elimination

Rate expressions can reference derived variables that depend on the
current state.  The system is then no longer linear, and the substep
argument from the section above applies -- set `hmax`:

```{r}
ev <- et(amt = 3) |> et(seq(0, 30, by = 0.25))

mmSol <- rxSolve(mmComp, ev, method = "indLin", hmax = 0.01)

ggplot(as.data.frame(mmSol), aes(time, cp)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Michaelis-Menten PK solved with matExp() + indLin")
```

rxode2 infers `depot` and `central` as compartments from the rate
constant names.  The `output` label is reserved for the elimination sink,
analogous to compartment 0 in NONMEM.

### Two-compartment PK with a peripheral compartment

```{r}
twoCmt <- function() {
  ini({
    ka <- 1
    cl <- 2
    vc <- 10
    q <- 1
    vp <- 20
  })
  model({
    matExp()
    k_depot_central <- ka   ## absorption
    k_central_output <- cl / vc ## central elimination
    k_central_periph <- q / vc  ## distribution to peripheral
    k_periph_central <- q / vp  ## redistribution from peripheral
    cp <- central / vc
  })
}

twoCmtSol <- rxSolve(twoCmt, et(amt = 100) |> et(seq(0, 48, by = 0.25)),
                     method = "indLin")

ggplot(as.data.frame(twoCmtSol), aes(time, cp)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)")
```

### Adding a forcing term with `indLin()`

`indLin(state) <- expr` supplies the `f(t)` of the standard form: a term
that enters a compartment without being proportional to any state.  This
covers endogenous production, constant inputs that are not in the dosing
event table, and steady-state baselines.  A turnover response model is
the smallest useful example -- production `kin` is a forcing term, loss
`kout * resp` is a rate constant:

```{r}
turnover <- function() {
  ini({
    kin <- 5
    kout <- 0.1
  })
  model({
    matExp()
    cmt(resp)
    k_resp_output <- kout   ## first-order loss
    indLin(resp) <- kin     ## zero-order production
  })
}

turnoverSol <- rxSolve(turnover, et(seq(0, 100, by = 1)), method = "indLin")

## baseline approaches kin/kout = 50
ggplot(as.data.frame(turnoverSol), aes(time, resp)) +
  geom_line() +
  geom_hline(yintercept = 5 / 0.1, linetype = 2) +
  labs(x = "Time (h)", y = "Response")
```

### Comparison with NONMEM ADVAN5/7

NONMEM's general linear solver (ADVAN5 for non-steady-state, ADVAN7 for
steady-state) specifies rate constants as `K(i,j)` scalars in `$PK`,
where **`K(i,j)` is the first-order transfer rate from compartment `i` to
compartment `j`** (source first, destination second).  rxode2's
`matExp()` uses the **same source-first convention**: `k_from_to`.

| NONMEM | rxode2 `matExp()` | Meaning |
|---|---|---|
| `K12 = ka` | `k_cmt1_cmt2 <- ka` | depot (1) to central (2) |
| `K20 = ke` | `k_cmt2_output <- ke` | central (2) elimination |
| `K23 = Q/V2` | `k_cmt2_cmt3 <- q / v2` | central to peripheral |
| `K32 = Q/V3` | `k_cmt3_cmt2 <- q / v3` | peripheral to central |

Differences from NONMEM:

- **Compartment naming**: NONMEM numbers compartments (1, 2, 3, ...);
  rxode2 uses descriptive names (`depot`, `central`, `periph`).
- **Compartment declarations**: NONMEM requires explicit `COMP`
  declarations; rxode2 auto-detects from rate constant names.
- **Elimination sink**: NONMEM uses `K(i,0)` (compartment 0 = outside);
  rxode2 uses `k_cmt_output` (`output` is the reserved sink label).
- **Diagonal**: both compute the diagonal (self-decay) entries
  automatically from the listed outflow rates.
- **Non-depleting inputs**: rxode2 supports `k_from_to_nd` for
  zero-depletion transfers; NONMEM has no direct equivalent.
- **Forcing terms**: rxode2's `indLin(state) <- expr` adds an
  inhomogeneous term; in NONMEM this requires explicit ODEs (ADVAN6+).

### When to use `matExp()` vs `indLin()`

| Situation | Recommendation |
|---|---|
| Linear compartmental PK/PD with known rate constants | `matExp()` -- most stable, no symengine |
| Model migrated from NONMEM ADVAN5/7 | `matExp()` -- direct `K(i,j)` translation |
| Nonlinear ODE (Michaelis--Menten, etc.) | Write `d/dt()` and let `method = "indLin"` convert it |
| Exploratory model with changing ODE structure | `d/dt()` -- faster iteration |
| Complex nonlinear ODE where the factorization is wrong | `matExp()`, if you can state the rate matrix analytically |

## When to use inductive linearization

| Scenario | Recommendation |
|---|---|
| Multi-subject population simulation, linear PK | `method = "liblsoda"` (default) is typically fastest overall |
| Purely linear system, single subject | `method = "indLin"` gives an exact, closed-form solution |
| Mildly nonlinear system (Michaelis--Menten saturation) | `method = "indLin"` with an explicit `hmax` |
| Heavily nonlinear or chaotic ODE | Time-stepping solvers (`liblsoda`, `dop853`) are safer |
| Stiff system | `liblsoda` or `lsoda`; EVD can struggle when eigenvalues span many orders of magnitude |

### Advantages

1. **Exact for linear systems**: no discretization error at all, at any
   step size.
2. **Interval-based**: the linearized system is solved across an
   interval rather than stepped through it, so no local error control is
   needed inside the interval.
3. **Estimation warm-start**: during iterative parameter estimation the
   previous parameter vector's solution can seed the next linearization
   (the "smart update" of Sharif et al. 2022), reducing the number of
   passes needed.

### Limitations

1. EVD does not exist for all square matrices (defective matrices).
2. For large systems the matrix operations become expensive; EVD scales
   as O(n^3).
3. Accuracy on nonlinear models depends entirely on `hmax`, which has no
   automatic error control behind it.
4. The implementation is under active development; results may change
   between versions.

## References

- Duffull SB, Hegarty G (2014). An inductive approximation to the
  solution of systems of nonlinear ordinary differential equations in
  pharmacokinetics-pharmacodynamics. *J Theor Comput Sci* 1(4):1000119.
  <https://doi.org/10.4172/jtco.1000119>

- Hasegawa C, Duffull SB (2018). Exploring inductive linearization for
  pharmacokinetic-pharmacodynamic systems of nonlinear ordinary
  differential equations. *J Pharmacokinet Pharmacodyn* 45(1):35--47.
  <https://doi.org/10.1007/s10928-017-9527-z>

- Sharif S, Hasegawa C, Duffull SB (2022). Exploring Inductive
  Linearization for simulation and estimation with an application to the
  Michaelis--Menten model. *J Pharmacokinet Pharmacodyn* 49:445--453.
  <https://doi.org/10.1007/s10928-022-09813-z>

- Fidler M, Schoemaker R, Wilkins J et al (2020). Using RxODE for
  inductive linearization ODE solving. *CPT Pharmacometrics Syst
  Pharmacol* 9(S1):S22. <https://doi.org/10.1002/psp4.12497>

- Moler C, Van Loan C (2003). Nineteen dubious ways to compute the
  exponential of a matrix, twenty-five years later. *SIAM Rev*
  45(1):3--49. <https://doi.org/10.1137/S00361445024180>

- Al-Mohy AH, Higham NJ (2009). A new scaling and squaring algorithm for
  the matrix exponential. *SIAM J Matrix Anal Appl* 31(3):970--989.

- Sidje RB (1998). Expokit: a software package for computing matrix
  exponentials. *ACM Trans Math Softw* 24(1):130--156.

- Beal S, Sheiner LB, Boeckmann A, Bauer RJ (2009). NONMEM Users Guides
  (1989--2009). Icon Development Solutions, Ellicott City, MD.
  (ADVAN5/7 documentation for K(i,j) rate-constant specification.)
