---
title: "rxode2 and mrgsolve: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

Both rxode2 and mrgsolve are open-source R packages for ODE-based
pharmacometric simulation.  They share the same fundamental goal and
many capabilities, but differ in design philosophy and in some specific
features.

This article documents the feature landscape as accurately as possible.
Both packages are moving targets; details may change as each evolves.

## Features shared by both tools

Before comparing differences it helps to establish how much the two
tools have in common:

- NONMEM-compatible data sets (EVID, AMT, CMT, TIME, ID, …)
- ODE solving with adaptive step-size solvers (LSODA-family)
- Analytical PK compartment models (though parameterization differs)
- Between-subject variability (omega) and residual variability (sigma)
- Event tables with complex dosing regimens
- `simeta()` / `simeps()` for within-model ETA/EPS resampling
- Custom C/C++ functions registered at the R or package level
- Custom header files included in the compiled model
- Algebraic (ODE-free) prediction models
- NONMEM dataset translation tooling
- Population simulation across many subjects
- Parallel execution (OpenMP or otherwise)

## Features rxode2 has that mrgsolve does not

### Symbolic Jacobians and forward sensitivities

rxode2 can automatically derive the symbolic Jacobian of the ODE
system and forward-sensitivity equations.  These are used directly by
nlmixr2's FOCEi algorithm for parameter estimation without finite
differences.  mrgsolve does not currently support symbolic
differentiation.

### 1–3 compartment analytical solutions with exact gradients

`linCmt()` in rxode2 supports one-, two-, and three-compartment
analytical solutions.  Gradients are computed via Stan math
auto-differentiation, enabling exact gradient-based estimation in
nlmixr2.  mrgsolve's `$PKMODEL` block supports one and two
compartments only, and does not provide analytical gradients.

```{r}
mod <- function() {
  ini({
    TCl <- 4;
    eta.Cl ~ 0.09
    V   <- 10
    Q   <- 2;
    V2 <- 50
    V3  <- 200;
    Q2 <- 0.5
  })
  model({
    CL <- TCl * exp(eta.Cl)
    cp <- linCmt(CL, V, Q, V2, V3, Q2)           # 3-cmt solved automatically from parameter names
  })
}
```

Mrgsolve equivalent would require explicit ODE equations for the
three-compartment case.

### Multiple ODE solvers

rxode2 ships three solver backends that can be selected per run:

| Solver | Type | Use case |
|--------|------|----------|
| LSODA (C, thread-safe) | adaptive stiff/non-stiff | default |
| Fortran LSODA | adaptive stiff/non-stiff | reference |
| DOP853 | explicit Runge-Kutta | non-stiff, fast |

mrgsolve uses a single C++ LSODA backend.

```{r}
rxSolve(mod, ev, method = "dop853")   # explicit RK8 solver
rxSolve(mod, ev, method = "lsoda")    # default
```

### NONMEM model import

`nonmem2rx` converts NONMEM control streams directly into rxode2 model
objects, including parameter estimates, omega/sigma matrices, and
covariate relationships.  mrgsolve has a community package
`nonmem2mrgsolve` (<https://github.com/Andy00000000000/nonmem2mrgsolve>)
that performs a similar translation, but it is not on CRAN and is not
maintained by the mrgsolve core team.  `nonmem2rx` is on CRAN and is
maintained as part of the nlmixr2 ecosystem.

```{r}
library(nonmem2rx)

# Import a NONMEM run directly into an rxode2 model
mod <- nonmem2rx("run001.ctl")
```

### nlmixr2 integration for parameter estimation

rxode2 models can be passed directly to nlmixr2 for population
parameter estimation (FOCE, FOCEi, SAEM, etc.).  The same model
object is used for both simulation and estimation.  mrgsolve is
a simulation-only tool.

### Thread-safe C LSODA enabling OpenMP parallelism

rxode2's LSODA is implemented in thread-safe C, allowing genuine
parallel ODE solving across subjects via OpenMP.  mrgsolve parallelism
relies on forked processes or future-based backends.

### Model piping

rxode2 model objects support R's native pipe operator (`|>`) to
incrementally build or modify models without rewriting the full
specification.  `model()` and `ini()` pipes copy the model,
apply the change, and return a new object — leaving the original
untouched.

```{r}
library(rxode2)

base <- rxode2({
  d/dt(depot) <- -KA * depot
  d/dt(centr) <- KA * depot - CL / V * centr
})

# Add population parameters and a covariate in one pipeline
full <- base |>
  model({
    KA <- exp(tka + eta.ka)
    CL <- exp(tcl + eta.cl + cov.wt * WT)
    V  <- exp(tv)
  }, append = FALSE, cov = "WT") |>
  ini({
    tka <- log(0.5);  eta.ka ~ 0.09
    tcl <- log(4);    eta.cl ~ 0.09
    tv  <- log(20)
  })
```

The `append` argument controls where new model lines are inserted
(`FALSE` = prepend, `TRUE` = append).  The `cov` argument declares
covariate columns so they are not treated as missing parameters.

Additional convenience pipes allow targeted changes:

```{r}
# Change a single initial estimate without touching the rest of the model
mod2 <- full |> ini(tka = log(1))

# Fix a parameter
mod3 <- full |> ini(fix(eta.ka))

# Remove a covariate relationship
mod4 <- full |> model(CL <- exp(tcl + eta.cl))
```

mrgsolve models are defined as self-contained text blocks or files;
there is no equivalent incremental piping API.

### R functions available inside model code

rxode2 allows R function defined in the calling environment (session,
script, or package) to be automatically available for use inside an
rxode2 model.  No special block or registration step is needed —
rxode2 finds the function by name in the parent environment and calls
it at solve time.

```{r eval=TRUE}
library(rxode2)

# Plain R function defined anywhere in the session
Hill <- function(C, Emax, EC50) Emax * C / (EC50 + C)

mod <- rxode2({
  effect <- Hill(cp, Emax, EC50)   # calls the R function above
})
```

For performance, `rxFun()` can translate such R functions to C
automatically (see the user-functions vignette).

mrgsolve allows function calls but requires the plugin `mrgx` and the following steps in the `mrgx` environment:

- Add a function to the `mrgx` environment

- Call the function using `Rcpp`.

## Features mrgsolve has that rxode2 does not

### `$GLOBAL` block

mrgsolve allows global C++ declarations (variables, typedefs, helper
structs) that persist across the model's C++ functions via a
`$GLOBAL` block.  rxode2 has no direct equivalent at the model
specification level; shared state must be handled through rxFun or
package-level C code.

```cpp
// mrgsolve only
$GLOBAL
static int call_count = 0;
```

### `$ENV` block
mrgsolve's `$ENV` block places a R environment inside the model file,
evaluated into a dedicated environment at compile time; those R
objects are then accessible from C++ via the `mrgx` plugin.  These can
also be used as functions and then Rcpp can call them.


### `$EVENT` block

mrgsolve's `$EVENT` block (v1.5.2+) executes just before `$TABLE`,
allowing programmatic dose creation inside the model file with the
result visible in output on the same record.  rxode2 handles
modeled events through the event table API outside the model
definition.



### `nm-vars` plugin for verbatim NONMEM-style ODE syntax

mrgsolve's `nm-vars` plugin lets you write `A(1)`, `DADT(1)`, `F1`,
`R1` etc. inside ODE and PK blocks — effectively copy-pasting NONMEM
`$DES`/`$PK` code with minimal changes.  This is a syntax convenience
within a *new* mrgsolve model file, not a full NONMEM import.  rxode2
uses its own ODE notation (`d/dt(cmt)`) and `nonmem2rx` handles the
translation programmatically.

### External library plugins

mrgsolve's `$PLUGIN` block can link model C++ code against:

- `Rcpp` — full Rcpp header access inside the model
- `BH` — Boost headers
- `RcppArmadillo` — Armadillo linear algebra

rxode2's extension system (via `rxFun()`, `.extraC()`, or package
registration) provides custom C/C++ function injection but does not
expose Rcpp, Boost, or Armadillo directly inside model code.

### `EVID=8` replacement event

mrgsolve supports `EVID=8` for a compartment replacement (set
compartment to a given value at a point in time).  rxode2 uses
`EVID=5` for the equivalent operation.  Both tools support the
concept; only the numeric code differs.

## Features that exist in both tools but are often claimed by AI as mrgsolve-only

### ETA/EPS resampling

Both tools provide `simeta()` and `simeps()` with essentially
identical semantics.  See the dedicated rxode2 vignette
[ETA and EPS Resampling in rxode2](rxode2-eta-eps-resampling.html)
for worked examples.

### Automatic time-after-dose tracking

mrgsolve's `tad` plugin automatically tracks time after the most
recent dose in a variable accessible within the model.

In rxode2 this requires assigning the either `tad()`/`tad0()` or `tad(cmt)`/`tad0(cmt)`
(for a specific compartment) to a variable in the code to use it in
the model.

For the functions without `0` appended, values before dosing occur are
`NA`; on the other hand, when `0` is appended times before dosing are
zero.


### Custom C functions with header support

rxode2 supports custom C functions through `rxFun()` and inline C
strings, or through `.extraC()` which accepts either a file path
(generating a `#include`) or raw C code.  The MD5 of any included
file contributes to the model's cache key, so a changed header
triggers recompilation automatically.

```{r}
# Include a local header file; model recompiles if the file changes
.extraC("mymodels/special_functions.h")

mod <- rxode2({
  cp <- mySpecialPK(dose, CL, V)
})
```

Full package-level registration (including analytic derivatives for
nlmixr2) is described in the vignette
[Providing Custom C/C++ Functions to rxode2 from a Package](rxode2-pkg-exported-funs.html).

### Custom R code / model expansion

rxode2's `rxUdfUi` S3 dispatch system allows R functions to rewrite
model code at parse time — inserting lines, modifying initial
estimates, or expanding a single function call into multi-line model
code.  Built-in examples include `linCmt()`, `linMod()`, and
distribution functions like `rxpois()`.  This is described in
[Integrating User Defined Functions into rxode2](Integrating-User-Defined-Functions-into-rxode2.html).

```{r}
# Register a model-expansion function
rxUdfUi.myFun <- function(fun) { eval(fun) }
rxode2::.s3register("rxode2::rxUdfUi", "myFun")
```

### Algebraic (ODE-free) prediction models

Both tools support purely algebraic models without any ODEs.  In
rxode2 you simply omit `d/dt()` statements:

```{r eval=TRUE}
library(rxode2)

mod_pred <- rxode2({
  ipre <- 10 * exp(-ke * t)
})

et  <- et(seq(0, 24, length.out = 20))
rxSolve(mod_pred, et, params = c(ke = 0.5))
```

This is equivalent to NONMEM's `$PRED` block or mrgsolve's `$PRED`
block.

### NONMEM-compatible datasets

Both tools accept NONMEM-format datasets with only minor differences
in EVID coding (`EVID=5` in `rxode2` vs `EVID=8` in `mrgsolve` for
replacement events; note replacement is not supported by NONMEM).

## Compilation speed

Compilation speed between rxode2 and mrgsolve is comparable in
practice.  rxode2 generates C (not C++) which compiles faster per
translation unit, but includes an additional grammar parsing step.
mrgsolve generates C++ which allows richer language features.
Benchmarks on equivalent models show similar wall-clock times.

## Summary table

| Feature | rxode2 | mrgsolve |
|---------|--------|----------|
| ODE solving | LSODA (C, thread-safe), Fortran LSODA, DOP853 | C++ LSODA |
| Parallel solving (OpenMP) | yes (thread-safe solver) | process-based |
| 1–3 cmt analytical solution | `linCmt()` with gradients | `$PKMODEL` 1–2 cmt, no gradients |
| Symbolic Jacobian / sensitivities | yes | no |
| Parameter estimation | via nlmixr2 | simulation only |
| Model piping (`\|>`) | yes (`model()`, `ini()` pipes) | no |
| NONMEM model import | `nonmem2rx` (CRAN, nlmixr2 ecosystem) | `nonmem2mrgsolve` (GitHub only, community) |
| `simeta()` / `simeps()` | yes | yes |
| Custom C functions | `rxFun()`, `.extraC()`, package registration | `$GLOBAL`, `$PLUGIN` |
| R functions usable inside model | yes (any function in calling env) | yes (`$ENV` block, `mrgx` plugin) |
| Algebraic (no-ODE) models | yes (omit `d/dt`) | `$PRED` block |
| Custom header files | `.extraC("file.h")` with MD5 cache | `$INCLUDE` with MD5 cache |
| Direct C++ in model | no | yes (`$MAIN`, `$ODE`, `$TABLE`) |
| `$GLOBAL` persistent C++ state | no | yes |
| `$EVENT` block | no | yes |
| Rcpp/Boost/Armadillo in model | no | via `$PLUGIN` |
| `nm-vars` verbatim NONMEM syntax | no (use `nonmem2rx`) | yes |
| `tad` auto time-after-dose | manual calculation | `$PLUGIN tad` |
| EVID=5 (replacement) | yes | no (uses EVID=8) |
| EVID=8 (replacement) | no (uses EVID=5) | yes |

## References

- [nlmixr2 blog: mrgsolve vs rxode2](https://blog.nlmixr2.org/blog/2022-10-13-mrgsolve-vs-rxode2/)
- [mrgsolve user guide](https://mrgsolve.org/user-guide/)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nonmem2rx package](https://nlmixr2.github.io/nonmem2rx/)
