---
title: "The rxode2 event bus"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
has_bus <- exists("rxEventListen", asNamespace("rxode2"))
knitr::opts_chunk$set(eval = has_bus)
```

```{r no-bus, eval = !has_bus, echo = FALSE, results = "asis"}
cat("> **Note:** the installed `rxode2` does not have the event bus, so the",
    "code in this article is shown but not run.\n")
```

## Overview

`rxode2` has a small *event bus*: a publish/subscribe mechanism that tells
other code when a top-level simulation, fit or result is finished.  A package
that finishes an operation *emits* an event; any number of *listeners*
registered with `rxEventListen()` are called with the event name and its
payload.

The bus is what [nlmixr2log](https://github.com/nlmixr2/nlmixr2log) uses to
record every fit, simulation, plot and saved object of a session without the
user changing their code.  Any package or script can listen, though, and any
package can emit.

The bus is built around one rule: **only the outermost user-facing operation
reports**.  A fit runs many `rxSolve()` calls, a bootstrap runs many fits, and a
logger wants exactly one entry for what the user asked for -- not one per
internal step.

## The API

| Function | What it does |
|----------|--------------|
| `rxEventListen(id, fun)` | Register the listener `fun` under `id`.  Registering the same `id` again replaces the old listener. |
| `rxEventUnlisten(id)` | Remove the listener `id`. |
| `rxEventListeners()` | The ids of the registered listeners. |
| `rxEventEmit(event, ..., fun = NULL)` | Emit `event` with the named payload fields in `...`.  Delivered only at depth 0. |
| `rxEventScope(expr)` | Evaluate `expr` one level deeper; nothing emitted inside it is delivered. |
| `rxEventDepth()` | The current scope depth (0 means "outermost"). |
| `rxEventSeq()` | The number of events delivered in this process so far. |
| `.rxEventEnter()` | Go one level deeper (for emitters; pair it with `.rxEventExit()`). |
| `.rxEventExit(event = NULL, ..., fun = NULL)` | Go one level back up and, when `event` is given, emit it -- so it is delivered only when this was the outermost scope. |
| `.rxEventCall(call, fun = NULL)` | Normalize a recorded call (used for the `call` payload field; see below). |

A listener is a function `function(event, ...)`.  It is called with the event
name as its first argument and every payload field as a named argument, so it
can pick the fields it needs by name and ignore the rest.

Listeners are protected from each other and the caller is protected from
them:

* a listener that errors gives a warning naming the listener and the event; the
  other listeners still run and the operation that emitted is not affected;
* listeners run inside a scope, so anything a listener triggers (for example a
  listener that solves a model) is silent -- a listener can never cause another
  event;
* with no listener registered, `rxEventEmit()` costs a single length check.

## Listening

A recording listener keeps every event it receives:

```{r listen}
mod <- rxode2({
  d/dt(depot) <- -ka * depot
  d/dt(center) <- ka * depot - cl / v * center
  cp <- center / v
})
ev <- et(amt = 100) |> et(0:24)
p <- c(ka = 1, cl = 1, v = 10)

rec <- list()
rxEventListen("demo", function(event, ...) {
  rec[[length(rec) + 1L]] <<- list(event = event, payload = list(...))
})
rxEventListeners()

s <- rxSolve(mod, p, ev)
length(rec)
rec[[1]]$event
names(rec[[1]]$payload)
rec[[1]]$payload$kind
rec[[1]]$payload$call
```

The `call` field is the call the user typed, not the value of every argument.
When the call was built by `do.call()` its argument values are inlined; single
constants (such as `nSub = 10`) are kept, and larger values are replaced by `` `<value>` `` (or, when there are more than five of them, by a
single `` `<...>` ``) so a recorded call never embeds a large data set:

```{r do-call}
s <- do.call(rxSolve, list(mod, p, ev))
rec[[2]]$payload$call
```

When you are done, remove the listener:

```{r unlisten}
rxEventUnlisten("demo")
rxEventListeners()
```

To log a whole session to disk (fits, simulations, plots, saved objects, with
their code and data), use nlmixr2log rather than writing your own listener.

## The depth rule

Every emitting function enters one shared *operation scope* and emits on its
way out.  An event is delivered only when no scope is active, that is when
`rxEventDepth()` is 0.  Because all packages share the same scope, nesting
works across packages:

* `rxSolve()` called directly emits `solveComplete`;
* the `rxSolve()` calls inside `nlmixr2()` are silent, and `nlmixr2()` emits
  one `fitComplete`;
* the fits inside `bootstrapFit()` are silent, and `bootstrapFit()` emits one
  event for the bootstrap.

You can silence any code yourself with `rxEventScope()`:

```{r scope}
rec <- list()
rxEventListen("demo", function(event, ...) rec[[length(rec) + 1L]] <<- event)

f <- function() rxEventScope(rxSolve(mod, p, ev))
s <- f()
length(rec)

s <- rxSolve(mod, p, ev)
unlist(rec)
```

`rxSolve()` emits nothing when it does not produce a solved result: calling it
with no model (to build an `rxControl()` object) or when it errors.  The depth
is restored on error, since scopes are left with `on.exit()`:

```{r errors}
ctl <- rxControl()
## cl and v are missing, so this solve fails
out <- capture.output(res <- try(rxSolve(mod, c(ka = 1), ev), silent = TRUE))
inherits(res, "try-error")
unlist(rec)
rxEventDepth()
```

### Worker processes: `RXODE2_EVENT_DEPTH`

A fit or driver may run its work in other R processes (for example with
`future`, `mirai` or `callr`).  Those processes have their own copy of the bus
and would otherwise start at depth 0 and report every internal fit.  To keep
them silent, the depth is mirrored in the environment variable
`RXODE2_EVENT_DEPTH`.  A process started inside a scope inherits the variable,
and `rxode2` reads it when it is loaded, so the worker starts at the same depth
and emits nothing:

```{r env-var}
Sys.getenv("RXODE2_EVENT_DEPTH")
rxEventScope(Sys.getenv("RXODE2_EVENT_DEPTH"))
rxEventScope(rxEventScope(rxEventDepth()))
```

At depth 0 the variable is unset.  A worker pool that is started *before* the
scope (and reused) does not inherit it; work sent to such a pool is silent only
if the code run on the worker enters its own scope.

### Counting events: `rxEventSeq()`

`rxEventSeq()` counts the events delivered in this process.  Events dropped
because of the depth rule, and events emitted with no listener registered, are
not counted.  Taking the value before an operation and comparing it afterward
tells you whether anything reported in between -- nlmixr2save uses this to
link an assignment `fit := nlmixr2(...)` to the fit it produced:

```{r seq}
s0 <- rxEventSeq()
s <- rxSolve(mod, p, ev)
rxEventSeq() - s0
s <- rxEventScope(rxSolve(mod, p, ev))
rxEventSeq() - s0
rxEventUnlisten("demo")
```

## Events emitted across nlmixr2

The event names and payload fields are a convention shared by the nlmixr2
packages.  `rxode2` itself only emits `solveComplete`; the rest come from the
packages listed (each needs the version with event-bus support, and does
nothing with an `rxode2` that lacks the bus).

| Event | Payload fields | Emitted by |
|-------|----------------|------------|
| `solveComplete` | `result`, `object`, `call`, `kind` | rxode2 (`rxSolve()`); nlmixr2est (`simulate()`, `predict()`, `nlmixr2(est = "rxSolve")`, `vpcSim()`, `augPred()`) |
| `fitComplete` | `fit`, `object`, `call`, `objName`, `source` | nlmixr2est (`nlmixr2()`); nlmixr2extra (drivers returning a new fit); nlmixr2scm (the final model of `runSCM()`) |
| `fitUpdate` | `fit`, `original`, `name`, `what`, `inPlace` | nlmixr2est (`addCwres()`, `addNpde()`, `addTable()`, `setOfv()`, a deferred objective computed on access); nlmixr2extra (`preconditionFit()`, `bootstrapFit()`); nlmixr2targets (`what = "ui"`) |
| `fitResult` | `fit`, `result`, `kind`, `call` | nlmixr2plot (`plot()`, `vpcPlot()`, augPred plots; also `data`); nlmixr2extra (bootstrap, profile, lasso, searches, multistart, ...); nlmixr2scm (`runSCM()`); nlmixr2sir (`runSIR()`) |
| `assignStart` | `name` | nlmixr2save (`:=`, before it evaluates anything) |
| `assign` | `name`, `value`, `call`, `envir`, `cached`, `sha`, `random`, `seed`, `since` | nlmixr2save (`:=` on success, `saveFit()`); nlmixr2targets (the final fit of a target, named by `tar_name()`) |
| `assignAbort` | `name` | nlmixr2save (`:=` that errored) |

What the fields mean:

* **`solveComplete`**: `result` is the solved object, `object` the model (or
  fit) that was solved, `call` the recorded call and `kind` how it was solved
  (`"rxSolve"`, `"simulate"`, `"predict"`, `"vpcSim"`, `"augPred"`;
  `simulate(fit)` reports `"rxSolve"`).  Because
  `object` is the fit for `simulate(fit)`, a logger can link a simulation to its
  fit.
* **`fitComplete`**: `fit` is the new fit, `object` what it was estimated from
  (a model, or the input fit for a driver), `objName` the variable name the user
  passed (or `NULL`), and `source` what produced it: `"fit"` for `nlmixr2()`,
  otherwise the driver (for example `"linearize"`, `"multistart"`, `"scm"`).
* **`fitUpdate`**: `fit` is the changed fit, `original` the fit passed in,
  `what` what changed (`"cwres"`, `"npde"`, `"table"`, `"ofv"`, `"ui"`,
  `"covariance"`, ...), and `inPlace` whether the user's object now holds the
  change (then `name` is its variable name, when known).  A function that
  returns its input unchanged (for example `addCwres()` on a fit that already
  has CWRES) emits nothing.
* **`fitResult`**: a result that is not a fit, attached to the fit it came
  from: `result` (a plot, or a summary in which embedded fits are reduced to
  their objective function and parameter table) and `kind` (for example
  `"plot"`, `"vpcPlot"`, `"bootstrap"`, `"profile"`, `"sir"`).
* **`assignStart` / `assign` / `assignAbort`**: bracket an assignment so a
  logger can tie the events in between to a variable name.  `cached` is `TRUE`
  when the value was restored from the nlmixr2save cache, `random` and `seed`
  describe a random-number-dependent value, and `since` is the `rxEventSeq()`
  value at `assignStart`.

A listener should treat fields as optional and ignore fields it does not
know: new fields may be added, and other packages may emit their own events.

## Writing an emitter

### Feature-detect the bus

So that your package keeps working with an `rxode2` that has no bus, do not
call the bus functions directly.  Look them up when they are needed, and do
nothing when they are missing.  The nlmixr2 packages use small wrappers like
these:

```{r wrappers}
.myEventBus <- function() {
  exists("rxEventEmit", envir = asNamespace("rxode2"), inherits = FALSE)
}
.myEventEnter <- function() {
  if (.myEventBus()) getExportedValue("rxode2", ".rxEventEnter")()
  invisible()
}
.myEventExit <- function(event = NULL, ..., fun = NULL) {
  if (.myEventBus()) getExportedValue("rxode2", ".rxEventExit")(event, ..., fun = fun)
  invisible()
}
.myEventScope <- function(expr) {
  if (!.myEventBus()) return(expr)
  .myEventEnter()
  on.exit(.myEventExit(), add = TRUE)
  force(expr)
}
```

### A user-facing function

Enter the scope at the start of the function and leave it in `on.exit()`,
emitting the event from there so it is emitted after the function has
returned its value (`returnValue()`) and is not emitted on error
(`returnValue()` is then `NULL`):

```{r emitter}
mySimulation <- function(model, params, events) {
  force(model)
  .myEventEnter()
  .call <- sys.call()
  on.exit({
    .ret <- returnValue()
    if (inherits(.ret, "rxSolve")) {
      .myEventExit("solveComplete", result = .ret, object = model, call = .call,
                   kind = "mySimulation", fun = "mySimulation")
    } else {
      .myEventExit()
    }
  }, add = TRUE)
  ## the solves in here are silent
  .s <- lapply(1:3, function(i) rxSolve(model, params, events))
  .s[[3]]
}

rec <- list()
rxEventListen("demo", function(event, kind = NULL, ...) {
  rec[[length(rec) + 1L]] <<- paste(event, kind)
})
s <- mySimulation(mod, p, ev)
unlist(rec)
```

Points to keep in mind:

* **Always leave the scope.**  Pair every `.rxEventEnter()` with an
  `.rxEventExit()` in `on.exit(..., add = TRUE)`; a scope left open silences
  the rest of the session.
* **Force arguments before entering.**  R evaluates arguments lazily: if the
  user writes `myFun(nlmixr2(model, data, "focei"))`, that fit runs when the
  argument is first used.  Inside your scope it would be silent, so `force()`
  the argument (or use it) before `.rxEventEnter()`.
* **Silence internal work with a scope.**  Code that is not a user-facing
  operation of its own -- evaluating something to build a cache key, a helper
  that refits a model, a parallel map -- should run inside `.myEventScope()` or
  `rxEventScope()`.
* **Emit one event per user call.**  When an operation produces two things (a
  summary and a new best fit, say), emit the main event from `.rxEventExit()`
  and the second with `rxEventEmit()` afterward, still from the `on.exit()`
  code; you are then at depth 0 if your caller was.
* **Record the call with `fun`.**  Passing `fun = "myFun"` makes the head of
  the recorded `call` `myFun` even when the user called it through
  `do.call()` or a function object.

### S3 generics

For a generic, enter the scope in the generic itself, around `UseMethod()`.
All methods are then covered, and a method that calls another solve or fit is
silent.  This is what `rxSolve()` does:

```{r generic}
mySolve <- function(object, ...) {
  .myEventEnter()
  .call <- sys.call()
  on.exit({
    .ret <- returnValue()
    if (inherits(.ret, "rxSolve")) {
      .myEventExit("solveComplete", result = .ret, object = object, call = .call,
                   kind = "mySolve", fun = "mySolve")
    } else {
      .myEventExit()
    }
  }, add = TRUE)
  UseMethod("mySolve")
}
mySolve.default <- function(object, ...) rxSolve(object, ...)

rec <- list()
s <- mySolve(mod, p, ev)
unlist(rec)
rxEventUnlisten("demo")
```

The `on.exit()` handler set in a generic runs when the method returns, and
`returnValue()` there is the method's value.  Put any early returns that
should not emit (like `rxSolve()`'s `rxControl()` shortcut) *before*
`.rxEventEnter()`.

## See also

* `?rxEventListen` for the reference documentation.
* [nlmixr2log](https://github.com/nlmixr2/nlmixr2log), the logger built on
  these events.
