---
title: "Delay Differential Equations in rxode2"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## What is a delay differential equation?

An ordinary differential equation (ODE) describes how a system changes
based on its *current* state. A **delay differential equation (DDE)**
lets the rate of change also depend on a *past* state -- the system at
time `t - T`. Delays arise naturally in pharmacometrics (signaling and
maturation delays, delayed feedback, transit-like absorption) and in
ecology, physiology and control.

In `rxode2` a delayed state is written with `delay(state, T)`, which
evaluates the ODE compartment `state` at the earlier time `t - T`. The
semantics match the `delay()` function of Monolix.

```{r library}
library(rxode2)
```

### A first example

The classic linear DDE

$$y'(t) = -y(t - 1), \qquad y(t) = 1 \text{ for } t \le 0$$

is written directly:

```{r classic}
dde <- rxode2({
  y(0) <- 1
  d/dt(y) <- -delay(y, 1)
})

s <- rxSolve(dde, et(seq(0, 5, by = 0.1)))
```

Before the start of integration (`t <= 0` here) the history is the
constant initial condition, `y = 1`. After that, the solution is built
by the solver. We can check it against the exact "method of steps"
solution, which is piecewise polynomial:

$$y(t) = \begin{cases} 1 - t & 0 \le t \le 1 \\ \tfrac{t^2}{2} - 2t + \tfrac{3}{2} & 1 \le t \le 2 \end{cases}$$

```{r classic-check}
exact <- function(t) {
  ifelse(t <= 1, 1 - t, t^2 / 2 - 2 * t + 3 / 2)
}
max(abs(s$y[s$time <= 2] - exact(s$time[s$time <= 2])))
```

The agreement is at machine precision.

```{r classic-plot, fig.height=4}
plot(s, y)
```

## How delays use dense output

The key numerical challenge is that `delay(y, T)` needs the solution at
a past time `t - T` that is, in general, **not** one of the solver's
step boundaries or one of your observation times. `rxode2` answers this
the same way it speeds up dense-grid solving (see the companion article
[Dense output for fast dense-grid solves](rxode2-dense-output.html)):
every accepted step records the **dense-output polynomial** that
continuously interpolates the solution across that step.

For delay differential equations these polynomials are kept in a
per-subject history buffer as the integration advances. When the model
evaluates `delay(y, T)`, `rxode2`:

1. finds the recorded step that brackets the past time `t - T`, and
2. evaluates that step's dense polynomial at `t - T`.

Because this is the same interpolant the integrator uses for its own
output, the delayed value is obtained to the **full order of the
method** -- an 8th-order Dormand-Prince interpolant for `dop853`, and a
cubic Rosenbrock interpolant for `ros4`.

This has a few practical consequences, all handled automatically:

- **Delay models are solved on a dense path.** When a model uses
  `delay()`, `rxode2` enables dense output and switches the default
  method to the dense **AutoSwitch composite `"dop853+ros4"`**. A
  non-dense method (for example `lsoda`) cannot record the history and
  raises an error:

    ```{r non-dense, error=TRUE}
    rxSolve(dde, et(0:5), method = "lsoda")
    ```

- **Non-stiff and stiff regions are both handled.** The composite
  probes with the explicit `dop853` and falls back per segment to the
  implicit Rosenbrock `ros4` when a region turns stiff, so a delay
  model that is non-stiff in one region and stiff in another is solved
  efficiently in a single pass. A purely stiff delay model can also be
  solved with `method = "ros4"` directly.

- **The step size is capped to the smallest delay.** This keeps the
  lagged time `t - T` inside already-recorded history (rather than
  extrapolating off the current, not-yet-finished step), so short
  delays stay accurate.

- **Only the delayed states are recorded.** History is stored only for
  the compartments that some `delay()` actually looks back on, so a
  delay on one state in a large ODE system stays inexpensive.

You can still request a specific dense solver explicitly; `method =
"dop853"` (pure 8th-order) and `method = "ros4"` (stiff) both work.

## A pharmacodynamic example: delayed feedback growth

The delayed logistic (Hutchinson's equation) models a population whose
growth is regulated by its size one delay-period in the past. For a
long enough delay it produces sustained oscillations:

```{r hutch, fig.height=4}
hutch <- rxode2({
  r   <- 0.5
  K   <- 10
  tau <- 1
  N(0) <- 2
  d/dt(N) <- r * N * (1 - delay(N, tau) / K)
})

s2 <- rxSolve(hutch, et(seq(0, 40, by = 0.5)))
plot(s2, N)
```

The delay duration `T` is an arbitrary model expression -- it can be a
constant, a parameter, or a covariate -- and a model may use several
`delay()` terms, on the same state or on different states, each with
its own delay.

## Delays mixed with ordinary states

`delay()` composes freely with ordinary ODE compartments, dosing, and
covariates. Here a depot feeds a central compartment that has a delayed
auto-induction-like feedback term:

```{r mixed, fig.height=4}
mix <- rxode2({
  ka  <- 1
  ke  <- 0.3
  kin <- 0.2
  cen(0) <- 0
  d/dt(depot) <- -ka * depot
  d/dt(cen)   <-  ka * depot - ke * cen + kin * delay(cen, 2)
})

ev <- et(amt = 100, cmt = "depot") |> et(seq(0, 24, by = 0.25))
s3 <- rxSolve(mix, ev)
plot(s3, cen)
```

## Estimation with nlmixr2 (sensitivities)

Gradient-based estimation (such as the FOCEi family in `nlmixr2`)
needs the sensitivities of the model with respect to the estimated
parameters. `rxode2` generates these for delay models too: the
forward-sensitivity equations gain the delayed term `delay(S, T)` (the
sensitivity of the delayed state), which reuses exactly the same dense
history machinery described above. Delay durations that themselves
depend on an estimated parameter are also supported. As a result, a
delay model built with the `function` interface can be fit like any
other `nlmixr2` model.

## Notes and attribution

- Use `delay(state, T)` inside a `d/dt()` right-hand side. The first
  argument must be an ODE state (compartment) defined in the model; the
  second is the delay duration.
- The default solving method for delay models is the dense AutoSwitch
  composite `"dop853+ros4"`; `"dop853"` and `"ros4"` may be requested
  explicitly. Methods that cannot record dense history error out.
- The dense-output and delay-history machinery is adapted from the
  [`dde`](https://github.com/mrc-ide/dde) package by Rich FitzJohn and
  Wes Hinsley (Imperial College of Science, Technology and Medicine),
  following the dense-output approach of Hairer, Norsett and Wanner.

See also the function reference for `?delay` and the companion article
[Dense output for fast dense-grid solves](rxode2-dense-output.html).

## Session Information

```{r sessioninfo}
sessionInfo()
```
