---
title: "Simulation with Parameter Uncertainty in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
```

## Why simulate with parameter uncertainty?

Population PK/PD parameter estimates carry uncertainty expressed by
their variance-covariance matrix (from the covariance step of the
estimator) or by bootstrap replicates.  Propagating this uncertainty
into simulations produces confidence intervals on predicted
concentrations, exposure metrics, or target attainment probabilities
that reflect both between-subject variability (BSV) *and* uncertainty
in the model parameters.

## Model and parameter setup

```{r model}
pkModel <- function() {
  ini({
    tka  <- log(1.57);  label("log Ka")
    tcl  <- log(2.72);  label("log CL")
    tv   <- log(31.07); label("log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

## Assumed variance-covariance matrix for the fixed effects (thetas)
## In practice this comes from your estimation software output
thetaVcov <- matrix(
  c(0.04,  0.01,  0.00,
    0.01,  0.02,  0.00,
    0.00,  0.00,  0.03),
  nrow = 3,
  dimnames = list(c("tka","tcl","tv"), c("tka","tcl","tv"))
)

thetaMean <- c(tka = log(1.57), tcl = log(2.72), tv = log(31.07))
```

## Drawing parameter sets from the covariance matrix

`rxRmvn()` samples multivariate-normal parameter vectors from the
theta mean and covariance matrix:

```{r draws}
set.seed(42)
nDraw <- 500

thetaDraws <- rxRmvn(nDraw, thetaMean, thetaVcov)
head(thetaDraws)
```

## Simulating across parameter draws

For each draw, simulate a population of subjects and compute a
summary statistic of interest (here, AUC over 24 h for a 100 mg dose).

```{r simUncertainty, message=FALSE}
et <- et(amt = 100, time = 0) |>
  et(seq(0, 24, by = 0.5))

# Simulate all draws at once by passing a matrix of parameters
# Each row of thetaDraws is one parameter-uncertainty replicate;
# within each replicate rxSolve simulates nSub subjects with BSV.
simRes <- rxSolve(
  pkModel,
  et,
  params  = thetaDraws,
  nSub    = 20,           # subjects per draw
  returnType = "data.frame"
)
```

## Confidence interval on concentration-time profile

Summarise across both parameter draws and subjects to get the overall
prediction interval:

```{r ciProfile, fig.width=7, fig.height=4}
library(data.table)
dt <- as.data.table(simRes)

ci <- dt[, .(
  p05 = quantile(cp, 0.05),
  p50 = quantile(cp, 0.50),
  p95 = quantile(cp, 0.95)
), by = time]

ggplot(ci, aes(x = time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = p50), colour = "steelblue", linewidth = 1) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Concentration-time profile with parameter uncertainty",
       subtitle = "Shaded: 90% CI combining parameter uncertainty and BSV") +
  theme_bw()
```

## Uncertainty in a summary exposure metric (AUC)

```{r aucDist, fig.width=6, fig.height=4}
## Approximate AUC_0-24 with the trapezoidal rule for each sim replicate
auc <- dt[, .(AUC = sum(diff(time) * (cp[-length(cp)] + cp[-1]) / 2)),
          by = .(sim.id, id)]

aucByDraw <- auc[, .(medAUC = median(AUC)), by = sim.id]

ggplot(aucByDraw, aes(x = medAUC)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = quantile(aucByDraw$medAUC, c(0.05, 0.95)),
             linetype = "dashed") +
  labs(x = "Median AUC per draw (mg·h/L)",
       y = "Count",
       title = "Distribution of median AUC across parameter draws",
       subtitle = "Dashed lines: 5th and 95th percentiles") +
  theme_bw()
```

## Forest plot for covariate effects

A forest plot shows how a covariate shifts exposure relative to a
reference individual.  Simulate the reference subject and each
covariate scenario across parameter draws.

```{r forestPlot, fig.width=7, fig.height=5}
# Build covariate scenarios
scenarios <- data.frame(
  label  = c("Reference (WT=70)", "WT = 40 kg", "WT = 100 kg",
             "Mild RI (CL -20%)", "Moderate RI (CL -50%)"),
  tka    = log(1.57),
  tcl    = c(log(2.72), log(2.72), log(2.72),
             log(2.72 * 0.80), log(2.72 * 0.50)),
  tv     = log(31.07)
)

simScen <- rxSolve(
  pkModel,
  et,
  params     = scenarios,
  nSub       = 1,
  returnType = "data.frame"
)

aucScen <- as.data.table(simScen)[, .(
  AUC = sum(diff(time) * (cp[-length(cp)] + cp[-1]) / 2)
), by = .(sim.id)]
aucScen$label <- scenarios$label

refAUC <- aucScen$AUC[aucScen$label == "Reference (WT=70)"]
aucScen$ratio <- aucScen$AUC / refAUC

ggplot(aucScen[aucScen$label != "Reference (WT=70)", ],
       aes(x = ratio, y = label)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.8, 1.25), linetype = "dotted", colour = "red") +
  scale_x_continuous(limits = c(0.3, 1.8),
                     breaks = c(0.5, 0.8, 1.0, 1.25, 1.5)) +
  labs(x = "AUC ratio vs reference",
       y = NULL,
       title = "Forest plot — covariate effects on AUC",
       subtitle = "Dotted red lines: 0.8–1.25 no-effect boundary") +
  theme_bw()
```

## Using `rxSimThetaOmega()`

`rxSimThetaOmega()` is a convenience function that jointly samples
thetas and omegas from their respective distributions in a single
call, which is useful when you also want to propagate omega
uncertainty:

```{r rxSimThetaOmega}
## Simulate 200 sets of (theta, omega) from their distributions
thetaOmegaSim <- rxSimThetaOmega(
  params  = c(tka = log(1.57), tcl = log(2.72), tv = log(31.07)),
  lower   = c(-Inf, -Inf, -Inf),
  upper   = c(Inf, Inf, Inf),
  sigma   = thetaVcov,
  n       = 200,
  omega   = lotri(eta.ka ~ 0.6, eta.cl ~ 0.09, eta.v ~ 0.1),
  omegaDf = 10   # degrees of freedom for the inverse-Wishart draw
)
head(thetaOmegaSim)
```

Pass the resulting data frame directly to `rxSolve()` as the `params`
argument for fully uncertainty-propagated simulations.
