---
title: "Parallel rxode2 solving with mirai"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
```

> This vignette builds on
> [Out-of-memory solving with rxode2](rxode2-oom-solve.html), which introduces
> chunked solving in detail.  Read that vignette first if you are new to the
> `file` / `chunkSize` arguments.

## Two parallelism layers in rxode2

rxode2 has two independent parallelism mechanisms:

| Layer | Mechanism | Granularity | When active |
|-------|-----------|-------------|-------------|
| 1 | OpenMP threads | Subject-level, within one `rxSolve()` call | Always; controlled by `rxControl(cores=)` |
| 2 | mirai daemons | Chunk-level, across separate processes | When `parallel > 0` |

**Layer 1** is always on and works well for most use cases.  **Layer 2** adds
process-level parallelism on top: each chunk is an independent `rxSolve()` call
dispatched to a separate background process (daemon).

## When mirai parallelism helps

mirai chunk-level parallelism is most beneficial when:

- Each chunk requires enough RAM that the solve is I/O-bound on a single core.
- You have more physical cores than OpenMP threads per solve (e.g., a 48-core
  machine solving 4-thread chunks gives 12 chunks in parallel).
- You want to distribute chunks across compute nodes on a cluster or HPC.
- You want true memory isolation between chunks so one chunk's peak usage does
  not interfere with another.

If your population fits comfortably in RAM and OpenMP is already saturating your
cores, Layer 1 alone is usually sufficient.

## Model and event table

We reuse the same one-compartment oral PK model from the
[OOM vignette](rxode2-oom-solve.html), defined with `ini()` and `model()`
blocks.

```{r model}
pkModel <- function() {
  ini({
    tka      <- log(0.5);  label("Typical log Ka (1/h)")
    tcl      <- log(4);    label("Typical log CL (L/h)")
    tv       <- log(70);   label("Typical log V (L)")
    eta.ka   ~ 0.09
    eta.cl   ~ 0.04
    prop.err <- 0.15
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp           <- center / v
    cp ~ prop(prop.err)
  })
}

ev <- et(amt = 100, ii = 24, until = 72) |>
  et(seq(0, 72, by = 4)) |>
  et(id = 1:10)
```

## Parallel chunked solving with `rxSolve()`

Add `parallel = n` alongside `file` and (optionally) `chunkSize`.  rxode2
calls `mirai::daemons(n)` internally, dispatches one chunk per daemon, and
shuts the daemons down when done.

```{r parallel-solve, eval = requireNamespace("mirai", quietly = TRUE)}
res <- rxSolve(pkModel, ev,
               file      = tempfile("rxPar"),
               chunkSize = 3,
               parallel  = 2)
res
```

```{r parallel-df, eval = requireNamespace("mirai", quietly = TRUE)}
head(as.data.frame(res))
```

`chunkSize = 3` gives four chunks of up to 3 subjects; `parallel = 2` means
two daemons solve chunks concurrently (chunks 1 and 2 run together, then 3
and 4).

### Persistent daemons

When you call `rxSolve()` with `parallel > 0`, rxode2 creates **and destroys**
daemons around the solve.  If you run many consecutive chunked solves, spinning
daemons up and down for each one adds overhead.  Instead, start them yourself
before the first solve:

```{r persistent-daemons, eval = requireNamespace("mirai", quietly = TRUE)}
mirai::daemons(2)

res1 <- rxSolve(pkModel, ev,
                file      = tempfile("rxPar1"),
                chunkSize = 3,
                parallel  = 2)

res2 <- rxSolve(pkModel, ev,
                file      = tempfile("rxPar2"),
                chunkSize = 3,
                parallel  = 2)

mirai::daemons(0)  # shut down when done
```

When daemons are already running rxode2 reuses them rather than creating new
ones.

## Custom mirai workflows

Not every parallel use case fits the `file` pattern.  You can dispatch
arbitrary `rxSolve()` calls across mirai daemons for independent scenarios such
as sensitivity analyses or dose-response simulations.

```{r custom-mirai, eval = requireNamespace("mirai", quietly = TRUE)}
doses <- c(50, 100, 200)

mirai::daemons(3)

tasks <- lapply(doses, function(d) {
  mirai::mirai(
    {
      library(rxode2)
      ev_d <- et(amt = d, ii = 24, until = 72) |>
        et(seq(0, 72, by = 4)) |>
        et(id = 1:10)
      as.data.frame(rxSolve(pkModel, ev_d))
    },
    pkModel = pkModel,
    d       = d
  )
})

results <- lapply(tasks, `[`)   # block until each task finishes (m[] collects)

mirai::daemons(0)

# Each element of results is a data frame for one dose level
sapply(results, nrow)
```

This pattern is useful for:

- **Sensitivity analysis** -- vary one parameter at a time across daemons.
- **Dose-response** -- simulate many dose levels simultaneously.
- **Bootstrap replicates** -- each daemon processes an independent replicate
  data set.

## Reproducibility with seeds

rxode2's OOM path uses its own internal seed engine (`rxSetSeed()` /
`rxGetSeed()`), which is separate from R's `set.seed()`.  To make a chunked
parallel solve reproducible, call `rxSetSeed()` **before** `rxSolve()`:

```{r seeds, eval = requireNamespace("mirai", quietly = TRUE)}
rxSetSeed(42)
res_seed <- rxSolve(pkModel, ev,
                    file      = tempfile("rxSeed"),
                    chunkSize = 3,
                    parallel  = 2)

head(as.data.frame(res_seed)[, c("id", "time", "cp")])
```

When `parallel` is set, rxode2 pre-draws **all** subjects' etas in one pass
before splitting into chunks.  This means the eta draws match those of an
equivalent single-call `rxSolve()` at the same rxode2 seed -- chunk boundaries
do not affect the random-number sequence.

For custom mirai workflows, seed each task independently:

```r
tasks <- lapply(seq_along(reps), function(i) {
  mirai::mirai(
    {
      library(rxode2)
      rxSetSeed(seed)
      ev_i <- et(id = reps[[i]]) |> et(amt = 100, ii = 24, until = 72) |>
        et(seq(0, 72, by = 4))
      as.data.frame(rxSolve(pkModel, ev_i))
    },
    pkModel = pkModel, reps = reps, seed = base_seed + i, i = i
  )
})
```

## Performance tips

| Guideline | Reasoning |
|-----------|-----------|
| Chunk size ~ total_subjects / (2 x ndaemons) | Keeps daemons busy without under-utilizing them |
| Start with `parallel = 4` for 4-core machines | Match physical cores; OpenMP within each chunk uses 1 thread by default in daemon processes |
| Avoid `chunkSize = 1` | Per-chunk mirai overhead outweighs any parallelism gain |
| Profile before tuning | `system.time(rxSolve(...))` is the fastest way to find the sweet spot |

For cluster / HPC use, see `mirai::daemons()` with the `url` and `remote`
arguments; the rxode2 `parallel` path works unchanged with remote daemons.

## Session information

```{r session}
sessionInfo()
```
