---
title: "DOP853 Dense Output: When and How to Use It"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE}
options(cli.unicode = FALSE, crayon.enabled = FALSE)
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
options(width = 80, cli.width = 80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH = 80)
```

## What is DOP853 and what does `dense = TRUE` do?

rxode2 provides four ODE solvers:

| method | type | thread-parallel across subjects |
|---|---|---|
| `"liblsoda"` (default) | adaptive stiff/non-stiff | yes |
| `"lsoda"` | adaptive stiff/non-stiff (Fortran) | no |
| `"dop853"` | explicit Runge-Kutta 8(5,3) | yes |
| `"indLin"` | matrix-exponential (linear ODE only) | yes |

**DOP853** is an explicit Runge-Kutta method of order 8(5,3) by Dormand and
Prince. It is excellent for smooth, non-stiff problems and has an important
internal feature: a 7th-order *dense* (continuous) polynomial output via its
`contd8` function. This polynomial exactly matches the solver's accepted step
and can be evaluated cheaply at any time within that step.

DOP853 is fully thread-parallel across subjects via OpenMP (both standard and
dense paths). Its internal context (`dop853_ctx_t`) is stack-allocated per
call, so there is no shared mutable state between threads.

The `dense` argument in `rxSolve()` / `rxControl()` exposes this feature:

- **`dense = FALSE` (default):** rxode2 calls `dop853()` once per observation
  time, advancing the solver from each event to the next. Each call re-enters
  the adaptive step-size control from scratch.

- **`dense = TRUE`:** rxode2 makes one `dop853()` call per inter-dose segment.
  A callback fires after each accepted step and evaluates the 7th-order
  polynomial at every observation time that falls within that step. The solver
  therefore steps across the whole interval at its own pace, uninhibited by the
  observation grid.

## The critical role of `hmax` — and why rxode2 handles it automatically

The default `hmax` is computed from the average spacing between events in the
event table. For a grid with 0.1 h spacing, `hmax ≈ 0.1 h`. Since DOP853
cannot take a step larger than `hmax`, both the standard and the dense path
are already limited to one step per 0.1 h interval — so `dense = TRUE` with
the default `hmax` provides essentially no speedup.

The benefit of `dense = TRUE` is only realised when the solver is **allowed to
take steps larger than the observation spacing**, which requires
`hmax = NULL` (unlimited step size).

Because `hmax = NULL` is almost always the right choice when `dense = TRUE` is
requested, rxode2 sets it automatically and informs you with a message:

```{r hmax-auto, eval=FALSE}
# rxode2 automatically sets hmax=NULL and prints a message:
rxSolve(mod, ev, method = "dop853", dense = TRUE)
#> dop853 dense=TRUE: setting hmax=NULL so the solver can take steps
#> larger than the observation spacing
```

You can still override this by passing an explicit numeric value:

```{r hmax-override, eval=FALSE}
# explicit hmax overrides the automatic NULL — dense=TRUE respects it
rxSolve(mod, ev, method = "dop853", dense = TRUE, hmax = 1.0)
```

## Setup for the benchmark

```{r library}
library(rxode2)
library(microbenchmark)
```

We use a two-compartment oral absorption model — a typical non-stiff PK system.

```{r model}
mod <- function() {
  ini({
    tka <- log(0.5)
    tcl <- log(4)
    tv  <- log(70)
    tq  <- log(2)
    tvp <- log(40)
  })
  model({
    ka  <- exp(tka)
    cl  <- exp(tcl)
    v   <- exp(tv)
    q   <- exp(tq)
    vp  <- exp(tvp)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <-  ka * depot - (cl + q) / v * central + q / vp * periph
    d/dt(periph)  <-  q / v * central - q / vp * periph
    cp <- central / v
  })
}

# Warm up compilation
invisible(rxSolve(mod, et(amt = 100) |> et(0:5), method = "dop853"))
```

## Correctness check

Before benchmarking, confirm that `dense = TRUE` gives numerically equivalent
results:

```{r correctness}
doses <- et(amt = 100, time = 0, ii = 24, addl = 6)
ev    <- doses |> et(seq(0, 168, by = 0.1))

f_std   <- rxSolve(mod, ev, method = "dop853", dense = FALSE, hmax = NULL)
f_dense <- rxSolve(mod, ev, method = "dop853", dense = TRUE)

max(abs(f_std$cp - f_dense$cp))
```

The maximum difference is at the level of the solver tolerance (`rtol` x
solution magnitude), confirming the two paths are numerically equivalent.

## Benchmark: sparse vs. dense sampling grids

We compare three observation spacings. For each, we benchmark the standard
path (`dense = FALSE`) and the dense path (`dense = TRUE`). Because rxode2
automatically sets `hmax = NULL` when `dense = TRUE`, no extra argument is
needed to unlock large steps.

```{r benchmark}
run <- function(ev, dense, hmax_val = NULL, n = 30L) {
  microbenchmark(
    rxSolve(mod, ev, method = "dop853", dense = dense,
            hmax = hmax_val, cores = 1),
    times = n
  )$time / 1e6  # nanoseconds -> milliseconds
}

for (by in c(1, 0.1, 0.01)) {
  ev   <- doses |> et(seq(0, 168, by = by))
  nobs <- nrow(ev[ev$evid == 0, ])
  std  <- median(run(ev, FALSE))
  den  <- median(run(ev, TRUE))
  cat(sprintf(
    "obs spacing = %4.2f h  (n_obs = %5d):  standard = %5.1f ms  ",
    by, nobs, std))
  cat(sprintf("dense = %5.1f ms  speedup = %.2fx\n", den, std / den))
}
```

Typical results on this machine:

| Obs spacing | n\_obs | Standard (ms) | Dense (ms) | Speedup |
|---|---|---|---|---|
| 1.00 h | 168 | ~80 | ~70 | ~1.1x |
| 0.10 h | 1681 | ~85 | ~80 | ~1.1x |
| 0.01 h | 16801 | ~175 | ~160 | ~1.1x |

The gains are modest but consistent for this two-compartment model. Larger
gains appear when the ODE right-hand side is more expensive to evaluate (more
compartments or custom functions), because then the ratio of RHS evaluations
saved versus polynomial evaluations added shifts further in favour of the dense
path.

## How `hmax` interacts with the speedup

This benchmark shows the effect of an explicit `hmax` override explicitly
(suppressing messages with `suppressMessages()`):

```{r hmax-benchmark}
ev <- doses |> et(seq(0, 168, by = 0.1))

bm <- microbenchmark(
  std_default_hmax  = suppressMessages(
    rxSolve(mod, ev, method = "dop853", dense = FALSE)),
  dense_auto_hmax   = suppressMessages(
    rxSolve(mod, ev, method = "dop853", dense = TRUE)),
  dense_explicit_hmax = suppressMessages(
    rxSolve(mod, ev, method = "dop853", dense = TRUE, hmax = 0.1)),
  times = 30L
)
print(bm, unit = "ms", order = "median")
```

`dense_auto_hmax` (automatic `hmax = NULL`) is faster than
`dense_explicit_hmax` (hmax pinned to the obs spacing), confirming that the
automatic NULL is the correct default for the dense path.

## When to use `dense = TRUE`

Use `dense = TRUE` when **all** of the following hold:

1. **Non-stiff system.** DOP853 is an explicit method; it stalls on stiff
   problems. Use `liblsoda` (the default) if the system is stiff. See the
   [stiff systems vignette](rxode2-stiff.html).

2. **More than a handful of observations per dose interval.** With only 1-3
   observations per interval the per-callback overhead of the dense path
   exceeds the savings. The break-even point is roughly 5-10 observations per
   interval.

3. **No `linCmt()` calls.** Dense output is not yet implemented for the
   analytical linear-compartment solver. If the model contains `linCmt()`,
   rxode2 emits a message and silently falls back to the standard path.

Note: `hmax` is handled automatically. When `dense = TRUE` is set, rxode2
sets `hmax = NULL` and prints a message. You do not need to specify
`hmax = NULL` yourself, though you may override it with an explicit value if
your problem requires a step-size limit.

## When NOT to use `dense = TRUE`

- **Stiff systems.** Use the default `liblsoda`.

- **Models with `linCmt()`.** The dense path is silently bypassed; setting
  `dense = TRUE` has no effect and prints an informational message.

- **Very sparse sampling (< ~5 obs per interval).** The polynomial callback
  overhead outweighs the savings; the standard path is at least as fast.

## Summary

```r
# Rule of thumb — hmax=NULL is set automatically
rxSolve(mod, ev, method = "dop853", dense = TRUE)
```

is the formulation to try when:

- the system is non-stiff,
- the event table has a fine-grained observation grid (many obs per dose interval), and
- maximum solving speed is desired.

For typical pharmacometric models with standard rich-sampling designs, the
combination of `method = "dop853"` + `dense = TRUE` provides a 10-20%
wall-clock improvement over the standard DOP853 path. The automatic
`hmax = NULL` ensures the solver can take steps as large as accuracy permits.
Both DOP853 paths are OpenMP-parallel across subjects, so the dense path
retains the same multi-subject scaling as the standard path. The default
`liblsoda` solver remains faster for stiff or moderately stiff systems.

| Setting | Best for |
|---|---|
| `liblsoda` (default) | Any system; stiff-safe; OpenMP parallel |
| `dop853` | Non-stiff; moderate sampling; OpenMP parallel |
| `dop853, dense=TRUE` | Non-stiff; dense obs grid; OpenMP parallel; no `linCmt()` |

## Session information

```{r sessioninfo}
sessionInfo()
```
