---
title: "rxode2 and mrgsolve: A Feature Comparison"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

Both rxode2 and mrgsolve are open-source R packages for ODE-based
pharmacometric simulation.  They share the same fundamental goal and
many capabilities, but differ in design philosophy and in some specific
features.

This article documents the feature landscape as accurately as possible.
Both packages are moving targets; details may change as each evolves.

## Features shared by both tools

Before comparing differences it helps to establish how much the two
tools have in common:

- NONMEM-compatible data sets (EVID, AMT, CMT, TIME, ID, ...)
- ODE solving with adaptive step-size solvers (LSODA-family)
- Analytical PK compartment models (though parameterization differs)
- Between-subject variability (omega) and residual variability (sigma)
- Event tables with complex dosing regimens
- Adaptive / dynamic dosing from inside the model (mrgsolve via
  `evtools`; rxode2 in versions after 5.0.2)
- `simeta()` / `simeps()` for within-model ETA/EPS resampling
- Custom C/C++ functions registered at the R or package level
- Custom header files included in the compiled model
- Algebraic (ODE-free) prediction models
- NONMEM dataset translation tooling
- Population simulation across many subjects
- Parallel execution (rxode2 built-in; mrgsolve via the
  `mrgsim.parallel` extension)
- Out-of-memory / disk-backed simulation of larger-than-RAM populations
  (rxode2 built-in; mrgsolve via the `mrgsim.ds` extension)

### Adaptive dosing inside the model

Both tools now support model-driven adaptive dosing, but the user-facing
API is different.

mrgsolve implements this through the `evtools` plugin
(<https://mrgsolve.org/dynamic-dosing/evtools.html>,
<https://mrgsolve.org/blog/posts/evtools.html>,
<https://mrgsolve.org/blog/posts/2024-new-1-4-0-evtools.html>).  The
workflow is C++-oriented: enable `$PLUGIN evtools`, write event logic in
`$EVENT`, and call helpers such as `evt::bolus(self, amt, cmt)`,
`evt::infuse(self, amt, cmt, rate)`, `evt::replace(self, amt, cmt)`, or
`evt::reset(self)`.  For more complex workflows, mrgsolve also provides
mutable event objects plus an `evt::regimen` class that can change dose
amount, rate, interval, and duration as the simulation progresses.

rxode2 now provides the same general capability directly in the rxode2
model language.  This is newer than rxode2 5.0.2.  Instead of a
separate plugin and a C++ event block, adaptive actions are written
inside `model({})` using native statements such as `evid_()`,
`bolus()`, `infuse()`, `infuseDur()`, `reset()`, `replace()`,
`multiply()`, `phantom()`, and `obs()`.

| Topic | rxode2 | mrgsolve |
|------|------|------|
| Activation | built into the model syntax | opt in with `$PLUGIN evtools` |
| Where logic lives | `model({})` | `$EVENT` (preferred) or other model blocks |
| Event API style | domain-specific rxode2 statements | C++ helpers in `evt::` namespace |
| Adaptive observation rows | `obs()` or `evid_(..., 0, ...)` | handled through event objects / event API rather than a dedicated `obs()` helper |
| Regimen mutation | explicit push statements when conditions are met | event objects and `evt::regimen` support mutable ongoing regimens |

In practice, rxode2's interface is usually shorter and closer to the
pharmacometric intent, while mrgsolve's evtools framework is lower-level
and more programmable from the C++ side.  Both approaches can express
titration, rescue doses, dose holds, delayed starts, interval changes,
replacement/reset operations, and other dynamic regimen updates.

## Large-scale simulation: parallelism and out-of-memory solving

Simulating very large populations raises two separate problems: making
the solve fast enough (parallelism) and keeping the result small enough
to fit in memory (out-of-memory storage).  rxode2 addresses both inside
`rxSolve()` itself.  For mrgsolve the same two problems are solved by
two dedicated community extension packages -- `mrgsim.parallel`
(<https://github.com/kylebaron/mrgsim.parallel>) for parallel execution
and `mrgsim.ds` (<https://github.com/p-emex/mrgsim.ds>) for disk-backed
"out-of-memory" datasets.

### Parallel solving

rxode2 parallelizes at two levels.

At the *thread* level, rxode2's LSODA and DOP853 backends are
implemented in thread-safe C, so a single `rxSolve()` call spreads the
per-subject ODE solves across OpenMP threads automatically.  No data
splitting, forking, or extra package is involved -- it is controlled by
the `cores` argument (defaulting to the value of `getRxThreads()`).

```{r}
# Subjects are solved in parallel across OpenMP threads in one call
rxSolve(mod, ev, cores = 4)
```

At the *process* level, rxode2 (since 5.1.1) can dispatch chunks of
subjects to `mirai` daemons, which enables distributed and HPC/cluster
execution.  This is the same mechanism used by the out-of-memory path
below (the `parallel` argument gives the number of daemons).

mrgsolve's own C++ LSODA runs single-threaded within a process, so it
cannot spread subjects across threads the way rxode2's OpenMP solver
does.  The `mrgsim.parallel` package fills this gap by splitting the
data set with `chunk_by_id()` / `chunk_by_row()` and simulating the
chunks in parallel through the `future` or `parallel` (forking /
`mclapply`) ecosystems, then recombining the pieces into a single
in-memory data frame (`future_mrgsim_d()`, `mc_mrgsim_d()`,
`future_mrgsim_ei()`, ...).

| Topic | rxode2 | mrgsolve (`mrgsim.parallel`) |
|------|------|------|
| Within-subject-loop threading | OpenMP in the C solver (`cores=`) | not available (single-threaded C++ solver) |
| Cross-process parallelism | `mirai` daemons (`parallel=`) | `future` / `parallel` (fork, multisession) |
| Data splitting | automatic (per subject) | manual (`chunk_by_id()`, `chunk_by_row()`) |
| Extra package needed | no (built into `rxSolve()`) | yes (`mrgsim.parallel`) |
| Result assembly | in memory, or streamed to disk (see below) | recombined in memory |

Because rxode2's threading and its `mirai` chunking compose, a large run
can use OpenMP *within* each chunk while `mirai` distributes the chunks
across processes or nodes.

### Out-of-memory (larger-than-RAM) solving

A population simulation can easily produce a result that does not fit in
RAM.  rxode2 handles this directly: pass a `file` prefix to `rxSolve()`
and it splits the subjects into chunks, solves each chunk, writes the
chunk to a Parquet file (falling back to `.rds` when the `arrow` package
is not installed), and returns a lightweight `rxSolveOom` object that
reads chunks lazily on demand rather than holding the whole result in
memory.

```{r}
res <- rxSolve(mod, params, ev,
               file      = tempfile("rx"), # enables the out-of-memory path
               chunkSize = 5000,           # subjects per chunk (optional)
               parallel  = 4)              # mirai daemons (optional)
```

If `chunkSize` is omitted, rxode2 auto-computes it from the currently
available RAM using `rxMemoryEstimate()`, so the chunks are sized to
fit.  The returned object behaves like an ordinary solved object -- it
prints the `$params` and `$inits`, supports `$column` extraction,
`head()`, `nrow()`, `ncol()` / `dim()`, and the usual `as.data.frame()`
/ `as_tibble()` / `as.data.table()` coercions -- but none of these read
more of the data than they need to.

For genuinely out-of-memory results, the chunks can be queried lazily so
that filtering and aggregation are pushed down to the on-disk Parquet
files and the full result is never materialized:

```{r}
library(dplyr)

res |>
  as.arrow() |>            # a lazy arrow Dataset over the chunks
  filter(time > 24) |>
  group_by(id) |>
  summarise(cmax = max(cp)) |>
  collect()                        # only the small summary comes into RAM
```

A DuckDB query layer is used under the hood for `head()`, single-column
extraction, and schema access when DuckDB is installed.  The
storage/query engine can be pinned with the `rxode2.oom.backend` option
(`"auto"`, `"duckdb"`, `"arrow"`, or `"rds"`); a requested engine that
is not installed degrades gracefully (`duckdb` -> `arrow` -> `rds`), and
the option is forwarded to the parallel `mirai` workers.

The *input* side can be kept out of memory too.  `rxEventTableFile()`
references an event table that lives on disk (Parquet, CSV, `fst`, or
`rds`); each chunk reads only its own subjects' rows, so even the event
data set never has to be loaded in full.

```{r}
big_ev <- rxEventTableFile("events.parquet", id = "id")
rxSolve(mod, params, big_ev, file = tempfile("rx"))
```

mrgsolve's `mrgsim.ds` package takes a very similar approach on the
output side: `mrgsim_ds()` stores results as Apache Arrow / Parquet
datasets in `tempdir()`, returns a lightweight pointer object, and lets
you apply `dplyr` verbs over an Arrow (or DuckDB) backend before
`collect()`-ing.  It also adds a parallel-safe model loader
(`modlib_ds()`), combination (`reduce_ds()`), and persistent
save/read (`save_ds()` / `read_ds()`).

The two designs are close in spirit; the main differences are:

| Topic | rxode2 | mrgsolve (`mrgsim.ds`) |
|------|------|------|
| Packaging | built into `rxSolve(file=)` | separate `mrgsim.ds` package + `mrgsim_ds()` |
| On-disk format | Parquet (Arrow), `.rds` fallback with no Arrow | Parquet (Arrow) |
| Lazy query engines | DuckDB and Arrow/`dplyr` pushdown | Arrow/`dplyr` and DuckDB |
| Automatic chunk sizing | yes, from free RAM (`rxMemoryEstimate()`) | user manages replicate/chunking |
| Backend selection | `rxode2.oom.backend` option, graceful degrade | Arrow-based |
| Persisted alongside data | per-subject `$params` and `$inits` | simulation output |
| Out-of-memory *input* | yes (`rxEventTableFile()`) | -- |
| Composes with parallelism | yes (`parallel=` `mirai` daemons + OpenMP) | via `mrgsim.parallel` |

`rxSolveChunked()` is a thin convenience wrapper around the same
machinery, but the `rxSolve(..., file=)` form above is the recommended
interface.

## Features rxode2 has that mrgsolve does not

### Symbolic Jacobians and forward sensitivities

rxode2 can automatically derive the symbolic Jacobian of the ODE
system and forward-sensitivity equations.  These are used directly by
nlmixr2's FOCEi algorithm for parameter estimation without finite
differences.  mrgsolve does not currently support symbolic
differentiation.

The sensitivity machinery has grown well beyond first order.  rxode2
now generates exact, finite-difference-free forward sensitivities up to
third order (`d^3 state / ds1 ds2 ds3`), which supply the analytic
FOCEi/FOCE covariance Hessian terms that previously required numeric
differencing.

It also generates *jump* (event) sensitivities: the derivative of the
solution with respect to dosing parameters such as bioavailability (`F`),
lag time (`alag`), infusion rate, duration, and dose amount, capturing
the discontinuities these introduce at dose times (based on the
`EventSensitivities` approach).  These are available for ODE, matrix
exponential, and `linCmt()` models -- up to third order for the ODE and
matrix-exponential cases -- and are selected through the
`eventSens` build option.  mrgsolve has no equivalent.

### Matrix exponential and inductive linearization with analytic gradients

rxode2 provides matrix-exponential and inductive-linearization solving
paths for linear (and locally linearized) systems, specified with a
NONMEM-like interface and translated automatically from ODE syntax.
Both paths carry analytic gradients computed by symbolic
differentiation, so they can be used inside gradient-based estimation
rather than only for forward simulation.  mrgsolve does not offer these
solving modes.

### 1-3 compartment analytical solutions with exact gradients

`linCmt()` in rxode2 supports one-, two-, and three-compartment
analytical solutions.  Gradients are computed via Stan math
auto-differentiation, enabling exact gradient-based estimation in
nlmixr2.  mrgsolve's `$PKMODEL` block supports one and two
compartments only, and does not provide analytical gradients.

```{r}
mod <- function() {
  ini({
    TCl <- 4
    eta.Cl ~ 0.09
    V <- 10
    Q <- 2
    V2 <- 50
    V3 <- 200
    Q2 <- 0.5
  })
  model({
    CL <- TCl * exp(eta.Cl)
    cp <- linCmt(CL, V, Q, V2, V3, Q2) # 3-cmt solved automatically from parameter names
  })
}
```

Mrgsolve equivalent would require explicit ODE equations for the
three-compartment case.

### Multiple ODE solvers

rxode2 ships several solver backends that can be selected per run:

| Solver | Type | Use case |
|--------|------|----------|
| LSODA (C, thread-safe) | adaptive stiff/non-stiff | default |
| Fortran LSODA | adaptive stiff/non-stiff | reference |
| DOP853 | explicit Runge-Kutta | non-stiff, fast |
| ros4 | Rosenbrock (implicit) | stiff |
| `"dop853+ros4"` | AutoSwitch composite | mixed stiff/non-stiff (see below) |

mrgsolve uses a single C++ LSODA backend.

```{r}
rxSolve(mod, ev, method = "dop853") # explicit RK8 solver
rxSolve(mod, ev, method = "lsoda") # default
```

rxode2 additionally offers **AutoSwitch composite** methods, written as
`"primary+secondary"` (for example `method = "dop853+ros4"`).  These
probe each integration segment with a fast non-stiff primary (`dop853`)
and reactively fall back to a stiff secondary (a Rosenbrock/implicit
method such as `ros4`, for which an analytical Jacobian is generated
automatically) only where stiffness is detected.  A problem that is
non-stiff in one region and stiff in another is thus solved efficiently
in a single pass, in both the standard and dense-output paths.

```{r}
rxSolve(mod, ev, method = "dop853+ros4") # non-stiff probe, stiff fallback
```

### Delay differential equations

rxode2 supports delay differential equations (DDEs) through the
`delay(state, T)` model function, which evaluates an ODE state at the
past time `t - T` (the same semantics as Monolix's `delay()`).  Delayed
values are interpolated from the solver's dense output (the 8th-order
Dormand-Prince interpolant), so they are obtained to the full accuracy
of the integration, and dense history is recorded only for the states
that are actually looked back on.  Delay models default to the dense
AutoSwitch composite `"dop853+ros4"`, and forward sensitivities are
generated for `delay()` models so they can be estimated with
gradient-based methods in nlmixr2.  mrgsolve does not solve delay
differential equations.

```{r}
mod_dde <- rxode2({
  d/dt(depot)  <- -ka * depot
  d/dt(central) <- ka * depot - ke * delay(central, tau) # delayed feedback
})
```

### Automatic ODE-to-`linCmt()` conversion

When a model's ODE system is a 1-3 compartment linear PK system,
rxode2 can detect this and transparently solve it with the fast
analytical `linCmt()` path instead of the numerical ODE solver.  This
happens automatically at solve time (`rxSolve(..., useLinCmt = TRUE)`,
the default), with the detected PK parameters passed explicitly to
`linCmt()` so the parameterization is inferred even when parameters are
declared only in `ini()`.  If a converted model cannot be compiled the
original ODE model is used instead, so the conversion never breaks an
otherwise-valid solve.  The reverse direction, `linToOde()`, expands a
`linCmt()` model back to explicit ODEs.  mrgsolve keeps analytical
(`$PKMODEL`) and ODE (`$ODE`) models as distinct, user-selected forms.

### NONMEM model import

`nonmem2rx` converts NONMEM control streams directly into rxode2 model
objects, including parameter estimates, omega/sigma matrices, and
covariate relationships and is on CRAN.  `mrgsolve` has a community package
`nonmem2mrgsolve` (<https://github.com/Andy00000000000/nonmem2mrgsolve>)
that performs a similar translation, that is not on CRAN.

Additionally, the [`amp.sim` package](https://github.com/LeidenAdvancedPKPD/amp.sim)
(available on CRAN) can translate NONMEM models to several R simulation
frameworks, including both `rxode2` (using `nonmem2rx`) and `mrgsolve`.

```{r}
library(nonmem2rx)

# Import a NONMEM run directly into an rxode2 model
mod <- nonmem2rx("run001.ctl")
```

### nlmixr2 integration for parameter estimation

rxode2 models can be passed directly to nlmixr2 for population
parameter estimation (FOCE, FOCEi, SAEM, etc.).  The same model
object is used for both simulation and estimation.  mrgsolve is
a simulation-only tool.

### Thread-safe C LSODA enabling OpenMP parallelism

rxode2's LSODA (and DOP853) are implemented in thread-safe C, allowing
genuine parallel ODE solving across subjects via OpenMP within a single
`rxSolve()` call, whereas mrgsolve's single-threaded C++ solver relies
on forked processes or future-based backends (via `mrgsim.parallel`).
See [Parallel solving](#parallel-solving) above for the full comparison,
including rxode2's `mirai`-based cluster/HPC path.

### Model piping

rxode2 model objects support R's native pipe operator (`|>`) to
incrementally build or modify models without rewriting the full
specification.  `model()` and `ini()` pipes copy the model,
apply the change, and return a new object -- leaving the original
untouched.

```{r}
library(rxode2)

base <- rxode2({
  d / dt(depot) <- -KA * depot
  d / dt(centr) <- KA * depot - CL / V * centr
})

# Add population parameters and a covariate in one pipeline
full <- base |>
  model(
    {
      KA <- exp(tka + eta.ka)
      CL <- exp(tcl + eta.cl + cov.wt * WT)
      V <- exp(tv)
    },
    append = FALSE,
    cov = "WT"
  ) |>
  ini({
    tka <- log(0.5)
    eta.ka ~ 0.09
    tcl <- log(4)
    eta.cl ~ 0.09
    tv <- log(20)
  })
```

The `append` argument controls where new model lines are inserted
(`FALSE` = prepend, `TRUE` = append).  The `cov` argument declares
covariate columns so they are not treated as missing parameters.

Additional convenience pipes allow targeted changes:

```{r}
# Change a single initial estimate without touching the rest of the model
mod2 <- full |> ini(tka = log(1))

# Fix a parameter
mod3 <- full |> ini(fix(eta.ka))

# Remove a covariate relationship
mod4 <- full |> model(CL <- exp(tcl + eta.cl))
```

mrgsolve models are defined as self-contained text blocks or files;
there is no equivalent incremental piping API.

### R functions available inside model code

rxode2 allows R function defined in the calling environment (session,
script, or package) to be automatically available for use inside an
rxode2 model.  No special block or registration step is needed --
rxode2 finds the function by name in the parent environment and calls
it at solve time.

```{r eval=TRUE}
library(rxode2)

# Plain R function defined anywhere in the session
Hill <- function(C, Emax, EC50) Emax * C / (EC50 + C)

mod <- rxode2({
  effect <- Hill(cp, Emax, EC50) # calls the R function above
})
```

For performance, `rxFun()` can translate such R functions to C
automatically (see the user-functions vignette).

mrgsolve allows function calls but requires the plugin `mrgx` and the following steps in the `mrgx` environment:

- Add a function to the `mrgx` environment

- Call the function using `Rcpp`.

## Features mrgsolve has that rxode2 does not

### `$GLOBAL` block

mrgsolve allows global C++ declarations (variables, typedefs, helper
structs) that persist across the model's C++ functions via a
`$GLOBAL` block.  rxode2 has no direct equivalent at the model
specification level; shared state must be handled through rxFun or
package-level C code.

```cpp
// mrgsolve only
$GLOBAL
static int call_count = 0;
```

### `$ENV` block
mrgsolve's `$ENV` block places a R environment inside the model file,
evaluated into a dedicated environment at compile time; those R
objects are then accessible from C++ via the `mrgx` plugin.  These can
also be used as functions and then Rcpp can call them.


### `$EVENT` block

mrgsolve's `$EVENT` block (v1.5.2+) executes just before `$TABLE`,
allowing programmatic dose creation inside the model file with the
result visible in output on the same record.  rxode2 now also supports
modeled adaptive dosing, but it does not use a separate `$EVENT` block;
instead, event-push statements are written directly in `model({})`.
So the difference is the presence of a dedicated block and a C++
event-object API, not the existence of adaptive dosing itself.



### `nm-vars` plugin for verbatim NONMEM-style ODE syntax

mrgsolve's `nm-vars` plugin lets you write `A(1)`, `DADT(1)`, `F1`,
`R1` etc. inside ODE and PK blocks -- effectively copy-pasting NONMEM
`$DES`/`$PK` code with minimal changes.  This is a syntax convenience
within a *new* mrgsolve model file, not a full NONMEM import.  rxode2
uses its own ODE notation (`d/dt(cmt)`) and `nonmem2rx` handles the
translation programmatically.

### External library plugins

mrgsolve's `$PLUGIN` block can link model C++ code against:

- `Rcpp` -- full Rcpp header access inside the model
- `BH` -- Boost headers
- `RcppArmadillo` -- Armadillo linear algebra

rxode2's extension system (via `rxFun()`, `.extraC()`, or package
registration) provides custom C/C++ function injection but does not
expose Rcpp, Boost, or Armadillo directly inside model code.

### `EVID=8` replacement event

mrgsolve supports `EVID=8` for a compartment replacement (set
compartment to a given value at a point in time).  rxode2 uses
`EVID=5` for the equivalent operation.  Both tools support the
concept; only the numeric code differs.

## Features that exist in both tools but are often claimed by AI as mrgsolve-only

### ETA/EPS resampling

Both tools provide `simeta()` and `simeps()` with essentially
identical semantics.  See the dedicated rxode2 vignette
[ETA and EPS Resampling in rxode2](rxode2-eta-eps-resampling.html)
for worked examples.

### Automatic time-after-dose tracking

mrgsolve's `tad` plugin automatically tracks time after the most
recent dose in a variable accessible within the model.

In rxode2 this requires assigning the either `tad()`/`tad0()` or `tad(cmt)`/`tad0(cmt)`
(for a specific compartment) to a variable in the code to use it in
the model.

For the functions without `0` appended, values before dosing occur are
`NA`; on the other hand, when `0` is appended times before dosing are
zero.


### Custom C functions with header support

rxode2 supports custom C functions through `rxFun()` and inline C
strings, or through `.extraC()` which accepts either a file path
(generating a `#include`) or raw C code.  The MD5 of any included
file contributes to the model's cache key, so a changed header
triggers recompilation automatically.

```{r}
# Include a local header file; model recompiles if the file changes
.extraC("mymodels/special_functions.h")

mod <- rxode2({
  cp <- mySpecialPK(dose, CL, V)
})
```

Full package-level registration (including analytic derivatives for
nlmixr2) is described in the vignette
[Providing Custom C/C++ Functions to rxode2 from a Package](rxode2-pkg-exported-funs.html).

### Custom R code / model expansion

rxode2's `rxUdfUi` S3 dispatch system allows R functions to rewrite
model code at parse time -- inserting lines, modifying initial
estimates, or expanding a single function call into multi-line model
code.  Built-in examples include `linCmt()`, `linMod()`, and
distribution functions like `rxpois()`.  This is described in
[Integrating User Defined Functions into rxode2](Integrating-User-Defined-Functions-into-rxode2.html).

```{r}
# Register a model-expansion function
rxUdfUi.myFun <- function(fun) {
  eval(fun)
}
rxode2::.s3register("rxode2::rxUdfUi", "myFun")
```

### Algebraic (ODE-free) prediction models

Both tools support purely algebraic models without any ODEs.  In
rxode2 you simply omit `d/dt()` statements:

```{r eval=TRUE}
library(rxode2)

mod_pred <- rxode2({
  ipre <- 10 * exp(-ke * t)
})

et <- et(seq(0, 24, length.out = 20))
rxSolve(mod_pred, et, params = c(ke = 0.5))
```

This is equivalent to NONMEM's `$PRED` block or mrgsolve's `$PRED`
block.

### NONMEM-compatible datasets

Both tools accept NONMEM-format datasets with only minor differences
in EVID coding (`EVID=5` in `rxode2` vs `EVID=8` in `mrgsolve` for
replacement events; note replacement is not supported by NONMEM).

## Compilation speed

Compilation speed between rxode2 and mrgsolve is comparable in
practice.  rxode2 generates C (not C++) which compiles faster per
translation unit, but includes an additional grammar parsing step.
mrgsolve generates C++ which allows richer language features.
Benchmarks on equivalent models show similar wall-clock times.

## Converting between rxode2 and mrgsolve

The [`rxode2mrgsolvebridge`](https://cran.r-project.org/package=rxode2mrgsolvebridge)
package (available on CRAN) provides bidirectional conversion between
rxode2 and mrgsolve model code.  It works through an intermediate
representation so that the same translation logic handles both
directions.

### rxode2 to mrgsolve

```{r}
library(rxode2mrgsolvebridge)

rx_code <- "
  KA = THETA_KA
  KE = THETA_KE
  d/dt(A)  = -KA * A
  d/dt(Cp) = KA * A - KE * Cp
"

convert_rxode2_to_mrgsolve(
  rx_model       = rx_code,
  theta_file     = "params.R",
  output_basename = "model",
  out_dir        = tempdir()
)
```

This writes a self-contained mrgsolve model file to `out_dir`.

### mrgsolve to rxode2

```{r}
convert_mrgsolve_to_rxode2(
  mrgsolve_file  = "model.cpp",
  output_basename = "model",
  out_dir        = tempdir()
)
```

This produces an rxode2 model code file and a matching theta parameter
file in `out_dir`.

### Lower-level utilities

The package also exposes lower-level helpers for parsing and rendering
when you need finer control:

| Function | Purpose |
|----------|---------|
| `parse_rxode2_model()` | Parse rxode2 code into intermediate representation |
| `parse_mrgsolve_model()` / `parse_mrgsolve_file()` | Parse mrgsolve code or file into intermediate representation |
| `ir_to_mrgsolve()` | Render intermediate representation as mrgsolve code |
| `ir_to_rxode2()` | Render intermediate representation as rxode2 code |
| `export_mrgsolve_model()` | Write mrgsolve model to file |
| `export_rxode2_files()` | Write rxode2 model and theta files |

Note that the conversion covers model equations and parameter
declarations; simulation-level features with no equivalent in the
target engine (e.g., rxode2 symbolic Jacobians, mrgsolve `$GLOBAL`
state) are either dropped or require manual adjustment after
conversion.

## Summary table

| Feature | rxode2 | mrgsolve |
|---------|--------|----------|
| ODE solving | LSODA (C, thread-safe), Fortran LSODA, DOP853 | C++ LSODA |
| AutoSwitch composite solvers (`"dop853+ros4"`) | yes | no |
| Delay differential equations (`delay()`) | yes (dense output, with sensitivities) | no |
| Parallel solving | OpenMP threads (`cores=`) + `mirai` (`parallel=`) | process-based via `mrgsim.parallel` |
| Out-of-memory / disk-backed solve | built in (`rxSolve(file=)`, Parquet/DuckDB) | via `mrgsim.ds` |
| 1-3 cmt analytical solution | `linCmt()` with gradients | `$PKMODEL` 1-2 cmt, no gradients |
| Automatic ODE -> `linCmt()` conversion | yes (`useLinCmt=TRUE`) | no |
| Symbolic Jacobian / sensitivities | yes (up to 3rd order) | no |
| Event / jump sensitivities (F, alag, rate, dur, amt) | yes | no |
| Matrix exponential / inductive linearization with gradients | yes | no |
| Parameter estimation | via nlmixr2 | simulation only |
| Model piping (`\|>`) | yes (`model()`, `ini()` pipes) | no |
| NONMEM model import | `nonmem2rx` (CRAN), `amp.sim` (CRAN) | `nonmem2mrgsolve` (GitHub), `amp.sim` (CRAN) |
| `simeta()` / `simeps()` | yes | yes |
| Custom C functions | `rxFun()`, `.extraC()`, package registration | `$GLOBAL`, `$PLUGIN` |
| R functions usable inside model | yes (any function in calling env) | yes (`$ENV` block, `mrgx` plugin) |
| Algebraic (no-ODE) models | yes (omit `d/dt`) | `$PRED` block |
| Custom header files | `.extraC("file.h")` with MD5 cache | `$INCLUDE` with MD5 cache |
| Direct C++ in model | no | yes (`$MAIN`, `$ODE`, `$TABLE`) |
| `$GLOBAL` persistent C++ state | no | yes |
| `$EVENT` block | no | yes |
| Rcpp/Boost/Armadillo in model | no | via `$PLUGIN` |
| `nm-vars` verbatim NONMEM syntax | no (use `nonmem2rx`) | yes |
| `tad` auto time-after-dose | yes `tad()`, `tad0()`, `tad(cmt)`, `tad0(cmt)` | `$PLUGIN tad` |
| EVID=5 (replacement) | yes | no (uses EVID=8) |
| EVID=6 (multiply)    | yes | no |
| EVID=7 (phantom/transit) | yes | no |
| EVID=8 (replacement) | no (uses EVID=5) | yes |

## References

- [nlmixr2 blog: mrgsolve vs rxode2](https://blog.nlmixr2.org/blog/2022-10-13-mrgsolve-vs-rxode2/)
- [mrgsolve user guide](https://mrgsolve.org/user-guide/)
- [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/)
- [nonmem2rx package](https://nlmixr2.github.io/nonmem2rx/)
- [nonmem2mrgsolve](https://github.com/Andy00000000000/nonmem2mrgsolve)
- [amp.sim package](https://github.com/LeidenAdvancedPKPD/amp.sim)
- [mrgsim.parallel (parallel mrgsolve)](https://github.com/kylebaron/mrgsim.parallel)
- [mrgsim.ds (out-of-memory mrgsolve)](https://github.com/p-emex/mrgsim.ds)
- [EventSensitivities (event/jump sensitivities)](https://github.com/dkaschek/EventSensitivities)
- [rxode2mrgsolvebridge package (CRAN)](https://cran.r-project.org/package=rxode2mrgsolvebridge)
- [rxode2mrgsolvebridge source (GitLab)](https://gitlab.com/fiha1/rxode2-mrgsolve-bridge)
