---
title: "Summarizing simulations with confint()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
```

`confint()` turns a solved `rxode2` object into the percentiles of the
simulated values at each time, and -- when the simulation can support it
-- a confidence band around each of those percentiles.  This article
covers what it returns, which arguments control it, and why the returned
columns change with the size and structure of the simulation.

## A model to summarize

```{r model}
pkModel <- function() {
  ini({
    tka  <- log(1.57)
    tcl  <- log(2.72)
    tv   <- log(31.07)
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

## uncertainty in the fixed effects, as it would come from the
## covariance step of an estimation run
thetaVcov <- lotri::lotri({
  tka ~ 0.04
  tcl ~ c(0.01, 0.02)
  tv  ~ 0.03
})

ev <- et(amt = 100, time = 0) |>
  et(seq(0, 24, by = 0.5))
```

## Two levels of summary, and two shapes of output

There are two different things you may want from a simulation:

- the **percentiles of the population**: at each time, the 2.5th, 50th
  and 97.5th percentile of the simulated individuals.  This describes
  the spread of the *individuals*.

- a **confidence band around each of those percentiles**: how well each
  percentile itself is known.  This describes the uncertainty in the
  *summary*, and it is what the shaded ribbons of a visual predictive
  check show.

`confint()` returns the first as a `rxSolveConfint1` object, with the
percentile in `p1` and its value in `eff`, and the second as a
`rxSolveConfint2` object, with the percentile in `p1` and its band in
the `p2.5`, `p50` and `p97.5` columns.

`level` sets the population percentiles.  `ci` sets the width of the
band placed around them, and defaults to `level`:

```{r ciArgs, eval = FALSE}
## 90% of the individuals, with a 95% band around each of those three curves
confint(sim, "cp", level = 0.90, ci = 0.95)
```

## Where the band comes from

The band needs replicate estimates of the percentile.  `confint()` gets
them in one of two ways, and this is what decides which of the two
shapes you get back.

### `nStud > 1`: replicate studies

This is the case the band is really meant for.  With `thetaMat` and
`nStud`, each study is a separate draw of the fixed effects, so the
percentiles are computed within each study and the band is the quantile
of those study-level percentiles.  The band then reflects the parameter
uncertainty, which is usually what is wanted.

```{r nStud, message = FALSE}
rxSetSeed(42)

simStud <- rxSolve(
  pkModel,
  ev,
  thetaMat = thetaVcov,  # uncertainty in the thetas
  nStud    = 500,        # virtual studies (one parameter draw each)
  nSub     = 20          # subjects per study
)

ciStud <- confint(simStud, "cp", level = 0.95)

class(ciStud)[1]

head(ciStud)
```

```{r nStudPlot, fig.width = 7, fig.height = 4}
plot(ciStud)
```

The solid line is the median individual, the ribbon around it is the
95% confidence band on that median, and the outer ribbons are the same
band on the 2.5th and 97.5th percentiles.

Note that `nSub = 20` subjects per study is *not* a problem here: the
band comes from the 500 studies, not from the 20 subjects.

### One study of at least 2500 individuals: sub-sampling

With a single study there are no replicates to work with, so
`confint()` makes some: the individuals are split into
`round(sqrt(n))` equal sub-samples, the percentiles are computed in
each, and the band is the quantile across sub-samples.  This is the
sampling variability of the percentile estimate itself, and it does
**not** include any parameter uncertainty.

`confint()` only does this when there are at least 2500 individuals,
which is the point at which the sub-samples are large enough
(`sqrt(2500) = 50` sub-samples of 50) for their percentiles to mean
something.

```{r big, message = FALSE}
rxSetSeed(42)

simBig <- rxSolve(pkModel, ev, nSub = 2500)

ciBig <- confint(simBig, "cp", level = 0.95)

class(ciBig)[1]

head(ciBig)
```

Below 2500 individuals there is nothing to build a band from, so
`confint()` says so and returns the percentiles alone:

```{r small}
rxSetSeed(42)

simSmall <- rxSolve(pkModel, ev, nSub = 2400)

ciSmall <- confint(simSmall, "cp", level = 0.95)

class(ciSmall)[1]

head(ciSmall)
```

This is the one surprising part of the interface: **the same
`confint()` call returns different columns for `nSub = 2400` and
`nSub = 2500`**.  The threshold is a property of the simulation, not of
the call.

### Asking for the percentiles only

If you want the simple percentiles regardless of how large the
simulation is -- for instance because downstream code expects fixed
column names -- pass `ci = FALSE`:

```{r noCi, message = FALSE}
ciNone <- confint(simBig, "cp", level = 0.95, ci = FALSE)

class(ciNone)[1]

head(ciNone)
```

`ci = FALSE` also pools across studies when there are several, so it
gives the percentiles of all `nStud * nSub` simulated individuals
together.

## The two bands are not the same quantity

It is worth comparing the two directly.  Both simulations below have
10000 individuals; only the second one carries parameter uncertainty.

```{r compare, message = FALSE}
rxSetSeed(42)
simPop <- rxSolve(pkModel, ev, nSub = 10000)
ciPop <- confint(simPop, "cp", level = 0.95)

rxSetSeed(42)
simUnc <- rxSolve(pkModel, ev, thetaMat = thetaVcov,
                  nStud = 500, nSub = 20)
ciUnc <- confint(simUnc, "cp", level = 0.95)

## width of the band on the median at 4 h
subset(as.data.frame(ciPop), time == 4 & p1 == 0.5, c("p2.5", "p50", "p97.5"))
subset(as.data.frame(ciUnc), time == 4 & p1 == 0.5, c("p2.5", "p50", "p97.5"))
```

The sub-sampled band is much narrower: with the parameters held fixed,
10000 individuals pin the population median down tightly.  The band from
the replicate studies is wider because the thetas themselves are only
known to within `thetaVcov`.  If you want a band that means "how well do
we know this curve", simulate with `thetaMat`/`nStud`; the sub-sampled
band answers the narrower question "how much would this curve move if we
re-drew the individuals".

## Stratifying with `by`

Columns carried through the solve with `keep=` can be used to split the
summary:

```{r by, message = FALSE}
evDose <- rbind(
  data.frame(et(amt = 100, time = 0) |> et(seq(0, 24, by = 2)) |>
               et(id = 1:100), dose = 100),
  data.frame(et(amt = 300, time = 0) |> et(seq(0, 24, by = 2)) |>
               et(id = 101:200), dose = 300)
)

rxSetSeed(42)
simBy <- rxSolve(pkModel, evDose, keep = "dose")

ciBy <- confint(simBy, "cp", level = 0.95, by = "dose")

head(ciBy)
```

## Means instead of percentiles

`mean = TRUE` reports the mean and its confidence interval with
`meanProbs()` rather than the empirical quantiles, and
`mean = "binom"` uses `binomProbs()` for a 0/1 variable such as a
target-attainment indicator:

```{r mean, message = FALSE}
head(confint(simSmall, "cp", level = 0.95, mean = TRUE))
```

## Summary

| Simulation | `confint()` returns |
|---|---|
| any, with `ci = FALSE` | percentiles only (`p1`, `eff`) |
| `nStud > 1` | percentiles per study, band across studies (`p1`, `p2.5`, `p50`, `p97.5`) |
| one study, `>= 2500` individuals | percentiles per sub-sample, band across sub-samples |
| one study, `< 2500` individuals | percentiles only, with a message |

See also the [parameter uncertainty
article](rxode2-parameter-uncertainty.html) for how to build the
`thetaMat`/`nStud` simulation the band is meant for, and the [VPC
article](rxode2-vpc.html) for the diagnostic version of the same idea.
