---
title: "Parallel rxode2 solving with mirai"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
```

> This vignette builds on
> [Out-of-memory solving with rxode2](rxode2-oom-solve.html), which introduces
> chunked solving in detail.  Read that vignette first if you are new to the
> `file` / `chunkSize` arguments.

## Two parallelism layers in rxode2

rxode2 has two independent parallelism mechanisms:

| Layer | Mechanism | Granularity | When active |
|-------|-----------|-------------|-------------|
| 1 | OpenMP threads | Subject-level, within one `rxSolve()` call | Always; controlled by `rxControl(cores=)` |
| 2 | mirai daemons | Chunk-level, across separate processes | When `parallel > 0` |

**Layer 1** is always on and works well for most use cases.  **Layer
2** adds process-level parallelism on top: each chunk is an
independent `rxSolve()` call dispatched to a separate background
process (daemon).

## When mirai parallelism helps

mirai chunk-level parallelism is most beneficial when:

- Each chunk requires enough RAM that the solve is I/O-bound on a single core.

- You want to distribute chunks across compute nodes on a cluster or HPC.

If your population fits comfortably in RAM and OpenMP is already saturating your
cores, Layer 1 alone is usually sufficient.

## Model and event table

We reuse the same one-compartment oral PK model and 10-subject event table from
the [OOM vignette](rxode2-oom-solve.html).

```{r model}
mod <- rxode2({
  d/dt(depot)  <- -ka * depot
  d/dt(center) <- ka * depot - cl / v * center
  cp           <- center / v
})

ev <- et(amt = 100, ii = 24, until = 72) |>
  et(seq(0, 72, by = 4)) |>
  et(id = 1:10)

params <- data.frame(
  id = 1:10,
  ka = rlnorm(10, log(0.5), 0.3),
  cl = rlnorm(10, log(4),   0.2),
  v  = rlnorm(10, log(70),  0.1)
)
```

## Parallel chunked solving with `rxSolve()`

Add `parallel = n` alongside `file` and (optionally) `chunkSize`.
rxode2 calls `mirai::daemons(n)` internally, dispatches one chunk per daemon,
and shuts the daemons down when done.

```{r parallel-solve, eval = requireNamespace("mirai", quietly = TRUE)}
res <- rxSolve(mod, params, ev,
               file      = tempfile("rxPar"),
               chunkSize = 3,
               parallel  = 2)
res
```

```{r parallel-df, eval = requireNamespace("mirai", quietly = TRUE)}
head(as.data.frame(res))
```

`chunkSize = 3` gives four chunks of up to 3 subjects; `parallel
= 2` means two daemons solve chunks concurrently (chunks 1 and 2 run
together, then 3 and 4).

### Persistent daemons

When you call `rxSolve()` with `parallel > 0`, rxode2 creates **and
destroys** daemons around the solve.  If you run many consecutive chunked
solves, spinning daemons up and down for each one adds overhead.  Instead,
start them yourself before the first solve:

```{r persistent-daemons, eval = requireNamespace("mirai", quietly = TRUE)}
mirai::daemons(2)

res1 <- rxSolve(mod, params, ev,
                file      = tempfile("rxPar1"),
                chunkSize = 3,
                parallel  = 2)

res2 <- rxSolve(mod, params, ev,
                file      = tempfile("rxPar2"),
                chunkSize = 3,
                parallel  = 2)

mirai::daemons(0)  # shut down when done
```

When daemons are already running rxode2 reuses them rather than creating new
ones.

## Custom mirai workflows

Not every parallel use case fits the `file` pattern.  You can dispatch
arbitrary `rxSolve()` calls across mirai daemons for independent scenarios such
as sensitivity analyses or dose–response simulations.

```{r custom-mirai, eval = requireNamespace("mirai", quietly = TRUE)}
doses <- c(50, 100, 200)

mirai::daemons(3)

tasks <- lapply(doses, function(d) {
  mirai::mirai(
    {
      library(rxode2)
      ev_d <- et(amt = d, ii = 24, until = 72) |>
        et(seq(0, 72, by = 4))
      as.data.frame(rxSolve(mod, params, ev_d))
    },
    mod    = mod,
    params = params,
    d      = d
  )
})

results <- lapply(tasks, `[]`)   # block until each task finishes

mirai::daemons(0)

# Each element of results is a data frame for one dose level
sapply(results, nrow)
```

This pattern is useful for:

- **Sensitivity analysis** — vary one parameter at a time across daemons.
- **Dose–response** — simulate many dose levels simultaneously.
- **Bootstrap replicates** — each daemon processes an independent replicate
  data set.

## Reproducibility with seeds

rxode2's OOM path uses its own internal seed engine (`rxSetSeed()` /
`rxGetSeed()`), which is separate from R's `set.seed()`.  To make a chunked
parallel solve reproducible, call `rxSetSeed()` **before** `rxSolve()`:

```{r seeds, eval = requireNamespace("mirai", quietly = TRUE)}
mod_rand <- rxode2({
  eta.ka ~ 0.09
  d/dt(depot)  <- -ka * depot
  d/dt(center) <- ka * depot - cl / v * center
  cp           <- center / v
  ka           <- exp(log(0.5) + eta.ka)
})

omega <- lotri::lotri(eta.ka ~ 0.09)

rxSetSeed(42)
res_seed <- rxSolve(mod_rand, params[, c("id", "cl", "v")], ev,
                    omega        = omega,
                    file      = tempfile("rxSeed"),
                    chunkSize = 3,
                    parallel  = 2)

head(as.data.frame(res_seed)[, c("id", "time", "cp")])
```

When `parallel` is set, rxode2 pre-draws **all** subjects' etas in one pass
before splitting into chunks.  This means the eta draws match those of an
equivalent single-call `rxSolve()` at the same rxode2 seed — chunk boundaries
do not affect the random-number sequence.

For custom mirai workflows, seed each task independently:

```r
tasks <- lapply(seq_along(reps), function(i) {
  mirai::mirai(
    {
      library(rxode2)
      rxSetSeed(seed)
      as.data.frame(rxSolve(mod, params, ev))
    },
    mod = mod, params = params, ev = ev, seed = base_seed + i
  )
})
```

## Performance tips

| Guideline | Reasoning |
|-----------|-----------|
| Chunk size ≈ total_subjects / (2 × ndaemons) | Keeps daemons busy without under-utilising them |
| Start with `parallel = 4` for 4-core machines | Match physical cores; OpenMP within each chunk uses 1 thread by default in daemon processes |
| Avoid `chunkSize = 1` | Per-chunk mirai overhead outweighs any parallelism gain |
| Profile before tuning | `system.time(rxSolve(...))` is the fastest way to find the sweet spot |

For cluster / HPC use, see `mirai::daemons()` with the `url` and
`remote` arguments; the rxode2 `parallel` path works unchanged with
remote daemons.

## Session information

```{r session}
sessionInfo()
```
