---
title: "rxode2 model functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
```

## rxode2 model function object

### Creating model object and basic properties

A rxode2 model function is a way to specify a rxode2 model inside of an R function. This function has an optional `ini({})` block as well as a required `model({})` block that may or may not have a residual specification included. We will build up from the simplest model and add items to describe the structure of the `rxode2` model function (in the code called a `ui` function).

Starting with the model from the original `RxODE` tutorial in a `model({})` block we can setup this function:

```{r modelOnly}
m1 <-function() {
  model({
    C2          <- centr/V2
    C3          <- peri/V3
    d/dt(depot) <- -KA*depot
    f(depot)    <- fdepot
    dur(depot)  <- durDepot
    rate(depot) <- rateDepot
    d/dt(centr) <- KA*depot - CL*C2 - Q*C2 + Q*C3
    d/dt(peri)  <-                    Q*C2 - Q*C3
    d/dt(eff)   <- Kin - Kout*(1-C2/(EC50+C2))*eff
    eff(0)      <- 1
  })
}
```

As expected if you try to access any properties of this function without evaluating it in some way, it will fail. For example if we try to see the initial estimates data-frame an error occurs:

```{r badModelProperty}
try(m1$iniDf)
```

But we can see the information if we either evaluate the function by `m1()` or use `rxode2()`/`nlmixr2()` to evalute the function. If you are in interactive mode, the `rxode2()` /`nlmixr2()` will change comments in the initial estimates block to labels. If you evaluate the function, the labels are dropped.

Lets evaluate the model using `rxode2()` to find out a bit more about the function:

```{r evalModel}
m1 <- rxode2(m1)
m1$iniDf
```

You can see in the output above, there are no initial estimates in this model, the `$iniDf` from the model has no rows representing the model information (though it does list the columns that are present in the initialization data-frame (which we will discuss later).

The other thing you can see from the object is there is no end-points defined by looking at `m1$predDf` which will return `NULL` when there are no end-points.

```{r predDfNone}
m1$predDf
```

Lets change this a bit with model piping to include initial estimates and a residual error to see the difference

```{r pipeFull}
m2 <- m1 |>
  model({
    KA <- exp(tka+eta.ka)
    CL <- exp(tcl+eta.cl + covClWt*WT)
    V2 <- exp(tv2)
    Q <-exp(tq)
    V3 <- exp(tv3)
    fdepot <- expit(tfdepot)
    durDepot <- exp(tDurDepot)
    rateDepot <- exp(tRateDepot)
    },append=FALSE, cov="WT") |>
  model({
    C2 ~ pow(c2.pow.sd, c2.pow)
    eff ~ add(eff.sd)
  }, append=TRUE) |>
  ini({
    tka   <- log(2.94E-01)
    tcl   <- log(1.86E+01)
    tv2   <- log(4.02E+01)
    tq    <- log(1.05E+01)
    tv3   <- log(2.97E+02)
    Kin <- 1
    Kout <- 1
    EC50 <- 5
    tfdepot <- logit(0.99)
    tDurDepot <- log(8)
    tRateDepot <- log(1250)
    c2.pow.sd <- 0.1
    c2.pow <- c(0.1, 1, 10)
    eff.sd <- 0.1
    eta.ka ~ 0.01
    eta.cl ~ 0.01
  })

# Now you can see information in both $iniDf and $predDf
m2$iniDf

m2$predDf

```

The model structure itself is a compressed `rxode2` ui object; you can see it by the class:

```{r rxUiCompressCls}
class(m2)
```

This compressed object is actually a compressed R environment. If you decompress the environment you can interact with the environment like any typical R environment. This is done by `rxode2::rxUiDecompress():`

```{r rxUiDecompressCls}
m2 <- rxUiDecompress(m2)
class(m2)

ls(m2, all=TRUE)
```

You can see the two properties we queried are low level objects inside of the `roxde2` ui environment (or compressed environment). We will describe the other items in more detail later.

## `$iniDf`: Initialization data frame

The initialization data frame comes from the `lotri` package and converts the `ini({})` a data-frame that is easier to use in R code. This data frame has the following columns:

| Column          | Meaning                                                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
|-----------|-------------------------------------------------------------|
| `ntheta`        | (Integer) this represents the fixed/population effect parameter number (ie. `theta`) This should start with 1 and end at the largest value. It includes the error parameter estimates. When not part of the residual errors or population estimates, it should be `NA_integer_`                                                                                                                                                                                                                   |
| `neta1`         | (Real Number) this represents the row of the between subject variability matrix (ie `omega`). If not a between subject variability it should be a `NA_real_`                                                                                                                                                                                                                                                                                                                                      |
| `neta2`         | (Real Number) this represent the column of the between subject variability matrix (ie `oomega`). If not a between subject variability it should be a `NA_real_`                                                                                                                                                                                                                                                                                                                                   |
| `name`          | (Character) This is the parameter name of the parameter in the model                                                                                                                                                                                                                                                                                                                                                                                                                              |
| `lower`         | (Real Number) Lower bound of the parameter                                                                                                                                                                                                                                                                                                                                                                                                                                                        |
| `est`           | (Real Number) Value of the parameter                                                                                                                                                                                                                                                                                                                                                                                                                                                              |
| `upper`         | (Real Number) Upper bound of the parameter                                                                                                                                                                                                                                                                                                                                                                                                                                                        |
| `fix`           | (Logical) When `TRUE` the estimate is fixed, when `FALSE` the estimate is unfixed.                                                                                                                                                                                                                                                                                                                                                                                                                |
| `label`         | (Character) the parameter label. When not present, this should be `NA_character_`                                                                                                                                                                                                                                                                                                                                                                                                                 |
| `backTransform` | (Character) Represents the back-transformation that is performed for `nlmixr`'s `$parFixed`. This overrides the implied back-transformation from the `rxode2` model parsing                                                                                                                                                                                                                                                                                                                       |
| `condition`     | (Character) this represents the level for the between subject varaibilty `id` and eventually other levels like `occ` when between occasion variability is supported. For errors this represents the endpoint that is being applied. All other columns should be `NA_character_`                                                                                                                                                                                                                   |
| `err`           | (Character) this describes the error type for each parameter describing the residual endpoints or log-likelihood distributions. The `err` will say `add` for additive error, `lnorm` for lognormal error etc. For distributions with multiple parameters, like `pow`, it will display `pow` for the first parameter and `pow2` for the second parameter. This is true of distributions with even more parameters. All parameters that are not defined by a distribution should be `NA_character_` |

## `$predDf` endpoint/residual data.frame

Inside the parsed `rxode2` function, there is another element called `$predDf`. This describes the endpoints in the model defined by lines like `endPoint ~ add(add.sd)`. In this data frame the following columns are defined:

| Column         | Description                                                                                                                                                                                                                                                                                                                              |
|------------------------------------------|---------------------------------------------------------------------------------------------------------------------|
| `cond`         | (Character) This is the condition in the expression `var ~ add(add.sd) | cond`                                                                                                                                                                                                                                                           |
| `var`          | (Character) This is the left-hand sided variable in `var ~ add(add.sd) | cond`                                                                                                                                                                                                                                                           |
| `dvid`         | (Integer) This an integer that represents the `dvid` that needs to be used in multiple endpoint/residual models                                                                                                                                                                                                                          |
| `trLow`        | (Real Number) The low value for the current transformation error (like `logitNorm()`                                                                                                                                                                                                                                                     |
| `trHi`         | (Real Number) The high value for the current transformation error (like `logitNorm()`)                                                                                                                                                                                                                                                   |
| `transform`    | (Factor) This represents the type of transformation for this endpoint                                                                                                                                                                                                                                                                    |
| `errType`      | (Factor) This represents the common error types for pharmacometric models. This can be `add`, `prop`, `pow`, `add+prop`, `add+pow` or `none`                                                                                                                                                                                             |
| `errTypeF`     | (Factor) This specifies the error type for proportional and power error. It can be proportional to the `f` value (which is the default for `prop` and `pow`). In cases where you have a transformation, it can be the `tranformed` `f` value or the `untransformed` `f` value. It can also be `none` for no proportional/additive models |
| `addProp`      | Factor that specifies if the `add+prop` residual is type 1 (standard deviations add) or type 2 (variances add), or the default (which is determined by `option(rxode2.addProp="combined1")` or `option(rxode2.addProp="combined2")` but defaults to combined2 error                                                                      |
| `distribution` | Factor with 19 levels that name all the known residual/likelihood specifications                                                                                                                                                                                                                                                         |
| `line`         | (Integer) This is the line number of the original `model({})` block where the error structure occurs                                                                                                                                                                                                                                     |
| `a`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the first parameter                                                                                                                                                                         |
| `b`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the second parameter                                                                                                                                                                        |
| `c`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the third parameter                                                                                                                                                                         |
| `d`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the fourth parameter                                                                                                                                                                        |
| `e`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the fifth parameter                                                                                                                                                                         |
| `f`            | (Character) This is a parameter defined in the model that is applied to the error instead of an estimate from the ini block. Usually this is the sixth parameter                                                                                                                                                                         |
| `lambda`       | (Character) This is the lambda parameter in errors like Box-Cox transformation                                                                                                                                                                                                                                                           |
| `linCmt`       | (logical) This is the linear compartment model logical flag.                                                                                                                                                                                                                                                                             |
| `cmt`          | (Integer) This is the compartment specification for this particular endpoint.                                                                                                                                                                                                                                                            |

## Model (`$lstExpr`), and Model variables (`$mv0` / `$mvL`)

The most basic (ie stored) component of the model is the `ui$lstExpr` which is a list that has a language object representing each line of the model.

The `ui$mv0` is the `rxode2` model variables for the `model({})` block **with all error model lines excluded**.

To get the lower-level functions for linear compartmental models, a `rxode2` model is converted to lower-level functions `linCmtA` or `linCmtB` depending on if gradients are needed. When this occurs, `$mvL` will contain the lower-level model. **This still excludes all the error model lines.**

## Covariates

The parameters that are not defined in the model and are not defined as an input parameters are covariates. You can access them directly with `ui$covariates` or use the alias `ui$allCovs`.

There are a few data-frames that relate to the covariates alone (mu referencing will be discussed later)

### `$covLhsDf:` covariate relationship with left handed assignment

The data frame `$covLhsDf` lists the left handed assignments where a covariate is found. In the model `m2` it defines a covariate line `CL <- exp(tcl + eta.cl + covClWt * WT)`, you can see the relation with the two columns in this dataset `lhs` (for the lefthand sided value) and `cov` (for the covariate):

```{r showCovLhsDf}
m2$covLhsDf
```

The columns are:

| Column | Description |
|--------|-------------|
| `lhs`  | The left-hand side variable name that uses the covariate |
| `cov`  | The covariate name |

Two companion named vectors are also available: `$covLhs` maps covariate name to LHS variable name, and `$lhsCov` maps LHS variable name to covariate name:

```{r covLhsVectors}
# covariate -> LHS
m2$covLhs

# LHS -> covariate
m2$lhsCov
```

## Variable mapping tables

Because `rxode2` model functions express parameters indirectly (e.g. `CL <- exp(tcl + eta.cl + covClWt*WT)`), several named-vector look-ups are provided to translate between the raw parameter name and the LHS variable it defines.

### `$etaLhs` and `$lhsEta`: eta parameter mapping

`$etaLhs` gives a named vector where names are eta parameter names and values are the corresponding LHS variable. `$lhsEta` is the reverse.

```{r etaMaps}
# eta -> LHS
m2$etaLhs

# LHS -> eta
m2$lhsEta
```

### `$thetaLhs` and `$lhsTheta`: theta parameter mapping

`$thetaLhs` maps each population (theta) parameter to the LHS variable it defines. `$lhsTheta` is the reverse.

```{r thetaMaps}
# theta -> LHS
m2$thetaLhs

# LHS -> theta
m2$lhsTheta
```

### `$varLhs` and `$lhsVar`: combined variable mapping

`$varLhs` is a combined look-up across all etas, thetas, covariates, and level assignments: names are the raw parameter/covariate/level names and values are the LHS variables. `$lhsVar` is the reverse.

```{r varLhsMaps}
# all vars -> LHS
m2$varLhs

# LHS -> all vars
m2$lhsVar
```

## Mu-referencing (`$muRefTable`)

When population parameters are expressed in a mu-referenced form (e.g. `CL <- exp(tcl + eta.cl)`), `$muRefTable` returns a data frame summarising those relationships. It returns `NULL` when no mu-referencing is detected.

```{r muRefTable}
m2$muRefTable
```

The columns are:

| Column        | Description                                         |
|---------------|-----------------------------------------------------|
| `theta`       | The theta (population) parameter name               |
| `eta`         | The eta (random-effect) parameter name              |
| `lhs`         | The LHS variable that combines theta and eta        |
| `covariates`  | Any covariate terms added to the mu-reference expression (if present) |

## Model states

### `$state`: compartment/state names

`$state` returns a character vector of all ODE state (compartment) names in the order they appear in the model:

```{r stateVec}
m2$state
```

### `$stateDf`: compartment data frame

`$stateDf` returns a data frame of compartment numbers and names. For linear compartment models it also includes whether each compartment supports infusion (`Rate`) and turn-off (`Off`) events:

```{r stateDf}
m2$stateDf
```

### `$statePropDf`: compartment property data frame

`$statePropDf` lists which compartments have special dosing properties (`ini`, `f`, `alag`, `rate`, `dur`) explicitly defined in the model. A compartment only appears when at least one property is set:

```{r statePropDf}
m2$statePropDf
```

## Parameter vectors

### `$theta`: population parameter estimates

`$theta` returns a named numeric vector of all theta (population/fixed-effect and residual) initial estimates:

```{r thetaVec}
m2$theta
```

### `$thetaLower` and `$thetaUpper`: parameter bounds

Named numeric vectors of the lower and upper bounds for every theta:

```{r thetaBounds}
m2$thetaLower
m2$thetaUpper
```

### `$omega`: random-effects variance-covariance matrix

`$omega` returns the between-subject variability (BSV) matrix derived from the `ini({})` block, or `NULL` when no etas are specified:

```{r omegaMat}
m2$omega
```

## Model properties (`$props`)

`$props` is a comprehensive summary list of the model in one place:

```{r propsObj}
str(m2$props)
```

The list elements are:

| Element    | Description |
|------------|-------------|
| `pop`      | Character vector of population (theta) parameter names |
| `resid`    | Character vector of residual error parameter names |
| `group`    | Named list of eta vectors, one element per BSV condition (e.g. `id`) |
| `linCmt`   | Logical; `TRUE` when the model uses a linear compartment solution |
| `cmt`      | Character vector of all doseable compartment names |
| `output`   | List with `primary` (LHS vars defined by population params), `secondary` (other LHS vars), `endpoint` (endpoint variables), and `state` (ODE state names) |
| `cmtProp`  | Same as `$statePropDf` |

## Multiple endpoints (`$multipleEndpoint`)

When a model has more than one residual endpoint, `$multipleEndpoint` returns a data frame showing how to specify each endpoint in an event table (via `cmt=` or `dvid=`). It returns `NULL` for single-endpoint models.

```{r multipleEndpointNull}
# m2 has two endpoints: C2 and eff
m2$multipleEndpoint
```

The `cmt` column shows the compartment label or number to use, and the `dvid*` column shows the equivalent `dvid` specification.

## Model representations

### `$lstChr`: model as a character vector

`$lstChr` deparses every line of `$lstExpr` to a character string. This is convenient for searching or displaying model lines:

```{r lstChr}
m2$lstChr
```

### `$funTxt`: model as a single string

`$funTxt` is `$lstChr` collapsed with newlines — the entire model body as one character string:

```{r funTxt}
cat(m2$funTxt)
```

### `$modelDesc`: short description string

`$modelDesc` produces a human-readable one-line description of the model structure (free-form ODE, linear compartment, Pred, etc.):

```{r modelDesc}
m2$modelDesc
```

### `$iniFun`: quoted `ini()` block

`$iniFun` returns the `ini({})` block as a quoted R expression (a `lotri`-style call), which can be used to reconstruct or inspect the initialisation block programmatically:

```{r iniFun}
m2$iniFun
```

### `$modelFun`: quoted `model()` block

`$modelFun` returns the `model({})` block as a quoted R expression:

```{r modelFun}
m2$modelFun
```

### `$fun`: reconstructed model function

`$fun` reconstructs the full model as a standard R function (without arguments) whose body contains the meta-data assignments, then the `ini()` block, then the `model()` block:

```{r funRecon}
m2$fun
```

## Simulation models

When a model includes a residual error specification, `rxode2` can generate a simulation version of the model that incorporates the error draws directly.

### `$simulationSigma`: residual error (sigma) parameter values

`$simulationSigma` returns a named numeric vector of the residual (sigma) parameter initial estimates, extracted from `$iniDf`:

```{r simSigma}
m2$simulationSigma
```

### `$simulationModel`: simulation model object

`$simulationModel` returns an `rxode2` model object that appends the error-sampling code to the base ODE model, ready for stochastic simulation:

```{r simModel}
m2$simulationModel
```

### `$simulationIniModel`: simulation model with initial estimates prepended

`$simulationIniModel` is like `$simulationModel` but with the theta/sigma initial estimates prepended as parameter assignments, so the model is self-contained and can be run without supplying a parameter vector:

```{r simIniModel}
m2$simulationIniModel
```

## Model identity and hashing

Two hashes are computed over the normalised model structure (model expressions, `iniDf`, error lines, defined functions, and rxode2 version) and are suitable for caching or comparing models:

```{r modelHashes}
# MD5 hash
m2$md5

# SHA-1 hash
m2$sha1
```

These hashes change whenever the model equations, initial estimates, or error specification change, but are insensitive to label or comment differences.

## Extending `rxUi` with custom `$` properties

Any package (or user script) can add new `ui$myProperty` accessors by defining an S3 method for the `rxUiGet` generic. This is the same mechanism rxode2 uses internally for every property shown above.

### How the dispatch works

When you write `ui$myProperty`, the `$.rxUi` method:

1. Wraps `ui` and the `exact` flag into a small list and sets its class to `c("myProperty", "rxUiGet")`.
2. Calls `rxUiGet(x)`, which dispatches to `rxUiGet.myProperty(x)`.

So adding a new property is simply a matter of defining `rxUiGet.myProperty`.

### Method signature

```{r rxUiGetSignature, eval=FALSE}
rxUiGet.myProperty <- function(x, ...) {
  .ui <- x[[1]]   # the decompressed rxUi environment
  # x[[2]] is the `exact` logical — rarely needed
  # compute and return your value
}
```

`x[[1]]` is the fully decompressed `rxUi` environment, so all built-in properties (e.g. `.ui$iniDf`, `.ui$state`) are available inside your method.

### The `desc` attribute and autocomplete visibility

The `desc` attribute controls whether the property appears in `str(ui)` output and in RStudio's `$` autocompletion list. A non-empty string makes the property visible; `NULL` or an empty string hides it (useful for internal helpers):

```{r rxUiGetDesc, eval=FALSE}
attr(rxUiGet.myProperty, "desc") <- "A short description for autocomplete"
```

The `rstudio` attribute sets the placeholder value that RStudio shows during tab-completion (so it does not have to evaluate the property while you type):

```{r rxUiGetRstudio, eval=FALSE}
# Use a representative example value — character, numeric, NA (passthrough), etc.
attr(rxUiGet.myProperty, "rstudio") <- "example value"
# NA means: evaluate for real even during completion
attr(rxUiGet.myProperty, "rstudio") <- NA
```

### Discovering all registered properties

`rxUiDevelop(TRUE)` temporarily makes hidden methods (those with an empty `desc`) visible in `str()` output, which is useful when exploring the full set of registered properties:

```{r rxUiDevelop}
rxUiDevelop(TRUE)
str(m2)
rxUiDevelop(FALSE)
```

### Worked example: counting ODE states

The example below registers a new property `$nStates` that returns the number of ODE states in the model:

```{r rxUiGetExample}
rxUiGet.nStates <- function(x, ...) {
  length(x[[1]]$state)
}
attr(rxUiGet.nStates, "desc") <- "Number of ODE states in the model"
attr(rxUiGet.nStates, "rstudio") <- 0L  # integer placeholder

m2$nStates
```

### Storing persistent metadata in `ui$meta`

Each `rxUi` environment contains a child environment called `meta` for storing arbitrary key-value data that should travel with the model object (e.g. a model name or provenance tag). Values stored there are also accessible via the `$` operator through the `rxUiGet.default` fallback, and they are printed as assignment expressions when the model is displayed.

```{r rxUiMeta}
m2$meta$myTag <- "two-cmt PK/PD example"
m2$meta$myTag
# also accessible via the $ operator (falls through to meta via rxUiGet.default):
m2$myTag
```
