---
title: "Matrix Exponential Solving in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Background

A linear system of ODEs

```
dy/dt = A . y
```

has a closed-form solution given by the **matrix exponential**,

```
y(t) = exp(A * t) . y0
```

which rxode2 evaluates through an eigenvalue decomposition,

```
y(t) = V . diag(exp(lambda * t)) . V^-1 . y0
```

where `V` are the eigenvectors and `lambda` the eigenvalues of `A`. For a system
that really is linear this is exact -- there is no discretization error at any
step size, because there is no stepping.

Every standard compartmental PK model is of this form, which makes the matrix
exponential a natural fit: the rate matrix is built once from the model's rate
constants and the solution is propagated across each interval in one operation.

`rxSolve(..., method = "indLin")` selects this integrator.

## Three things sharing the `indLin` name

| Name | What it is | What it does |
|---|---|---|
| `indLin(model)` | An R function | Rewrites a `d/dt()` model into matrix-exponential (`matExp()`) form and returns the model code as text |
| `indLin(state) <- expr` | A model statement | Adds an inhomogeneous forcing term to one compartment of a `matExp()` model |
| `method = "indLin"` | An `rxSolve()` option | Selects the matrix-exponential integrator instead of a time-stepping solver |

## What `indLin()` does

`indLin()` is the symbolic half. Give it a model written with `d/dt()` and it
returns the same model written as a rate matrix:

```{r}
library(rxode2)

oralOde <- function() {
  ini({
    ka <- 1
    cl <- 0.1
    v <- 1
  })
  model({
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - (cl / v) * central
    cp <- central / v
  })
}

cat(indLin(oralOde))
```

which prints

```
matExp()
cmt(depot)
cmt(central)
k_depot_central = ka
k_central_output = cl/v
cp=central/v
```

`indLin()` differentiated each right-hand side with respect to each state to get

```
A = [ -ka       0    ]
    [  ka   -cl/v    ]
```

and then reported `A` through its *micro-constants* rather than as a matrix: the
off-diagonal `A[central, depot] = ka` becomes `k_depot_central`, and the column
sum lost to the outside world, `-(A[depot, depot] + A[central, depot])`, becomes
`k_central_output`. The diagonal is implied, so it never has to be written down.
Nothing else changes -- the output line `cp = central/v` is carried through
untouched.

A nonlinear model converts too. The linear part goes into the rate matrix and
the nonlinear remainder becomes an `indLin()` forcing:

```{r}
mmOde <- function() {
  ini({
    ka <- 1
    km <- 0.5
    vmax <- 0.2
    v <- 1
  })
  model({
    d/dt(depot) <- -ka * depot
    cp <- central / v
    d/dt(central) <- ka * depot - vmax * cp / (km + cp)
  })
}

cat(indLin(mmOde))
```

```
matExp()
cmt(depot)
cmt(central)
k_depot_central = ka
indLin(central) <- -vmax*central/(v*(km + central/v))
cp=central/v
```

Absorption stays in the rate matrix, because `ka` is a genuine constant.
Michaelis-Menten elimination does not: its rate depends on how much drug is in
`central`. A rate constant that reads a compartment would make the rate matrix
change with the solution, which is exactly what the matrix exponential assumes
does not happen -- so rxode2 rejects one:

```r
rxode2("
matExp()
cmt(central)
vmax <- 10
km <- 5
k_central_output = vmax/(km + central)
")
#> Error: matrix exponential rate constant 'k_central_output' depends on the
#> compartment 'central'; rate constants must be constant in the states --
#> put the state-dependent part in 'indLin(central) <- ...' instead
```

The forcing is where a state-dependent term belongs, and it is what the solver
iterates on.

You rarely have to call `indLin()` yourself: `rxSolve(..., method = "indLin")` on
an ODE model runs this conversion internally. Calling it directly is how you
inspect the factorization, or obtain a `matExp()` model you want to edit and
keep.

## How rxode2 solves it

Over each substep the rate matrix is evaluated once at the current state, the
matrix exponential advances the states exactly for that frozen matrix, and the
next substep starts from the updated state.

For a **linear** model this is exact and the substep length is irrelevant:

```{r}
oralMe <- function() {
  ini({
    ka <- 1
    cl <- 0.1
    v <- 1
  })
  model({
    matExp()
    cmt(depot)
    cmt(central)
    k_depot_central <- ka
    k_central_output <- cl / v
    cp <- central / v
  })
}

ev <- et(amt = 100, ii = 12, until = 48) |> et(seq(0, 72, by = 0.5))

meSol <- rxSolve(oralMe, ev, method = "indLin")
odeSol <- rxSolve(oralOde, ev, method = "liblsoda")

## differences are at machine precision (~3e-13)
max(abs(meSol$cp - odeSol$cp))
```

For a model with a state-dependent `indLin()` forcing, the rate matrix is still
constant but the forcing is not, so the substep is only an approximation. The
solver therefore iterates the forcing to a fixed point within each substep and
chooses the substep itself from a local error estimate -- so `atol` and `rtol`
control the accuracy, the same way they do for every other adaptive method:

```{r}
solveMm <- function(...) {
  rxSolve(mmOde, et(amt = 3) |> et(c(0.1, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12, 16, 24, 30)), ...)
}

refMm <- solveMm(method = "liblsoda", atol = 1e-12, rtol = 1e-12)

maxRelErr <- function(sol) max(abs(sol$cp - refMm$cp) / refMm$cp)

maxRelErr(solveMm(method = "indLin"))
maxRelErr(solveMm(method = "indLin", atol = 1e-10, rtol = 1e-10))
```

| `atol`/`rtol` | Maximum relative error vs LSODA | Steps (`$counts$slvr`) |
|---|---|---|
| `1e-4` | 4.0e-03 | 148 |
| `1e-6` | 7.3e-05 | 1307 |
| default (`1e-8`/`1e-6`) | 7.6e-06 | 3212 |
| `1e-8` | 6.8e-07 | 12865 |
| `1e-10` | 6.4e-09 | 128442 |

The step is sized from a first-order error estimate, but each step advances on
the average of two answers whose leading errors are equal and opposite -- the
one built at the step's starting state and the converged one -- so what is
propagated is second order. The practical consequence is the scaling: the error
falls roughly in proportion to the tolerance, not to its square root, while the
step count still grows only as its square root. So the work needed for a given
accuracy grows like `1/sqrt(error)` rather than `1/error`, which is worth a
factor of tens to hundreds of steps at the accuracies people usually want.

For a tight tolerance, `indLinRichardson = TRUE` is worth trying. It runs each
step once whole and twice at half length and extrapolates, which raises the step
from second to third order at three times the cost per step. That is a losing
trade at loose tolerances and a winning one at tight ones -- on this model it is
a wash at a relative error of `1e-4`, about three times faster at `1e-6`, and
about six times faster at `1e-8`:

```{r}
maxRelErr(solveMm(method = "indLin", atol = 1e-8, rtol = 1e-8))
maxRelErr(solveMm(method = "indLin", atol = 1e-8, rtol = 1e-8,
                  indLinRichardson = TRUE))
```

`hmax` is now only an upper bound on the substep. Lowering it does not
meaningfully improve the answer, because the solver is already choosing a
shorter step than `hmax` allows:

```{r}
maxRelErr(solveMm(method = "indLin", hmax = 1))    # 8.4e-06
maxRelErr(solveMm(method = "indLin", hmax = 0.01)) # 1.9e-06
```

> **This changed in rxode2 5.1.7.** Previously the nonlinearity was folded into
> a rate constant, there was no iteration and no error control, and accuracy was
> governed entirely by `hmax` -- the default gave a 71% error on the sampling
> grid above. If you set an explicit `hmax` to work around that, you can drop
> it and set `atol`/`rtol` instead.

## Matrix exponential algorithms

Three algorithms are available via `indLinMatExpType`:

| Value | Algorithm | Notes |
|---|---|---|
| `"expokit"` | Sidje (1998) Krylov/Pade | **Default.** Good for sparse or large matrices; controlled by `indLinPhiTol` and `indLinPhiM` |
| `"Al-Mohy"` | Al-Mohy and Higham (2009) scaling and squaring | Accurate; order controlled by `indLinMatExpOrder` |
| `"arma"` | RcppArmadillo `expmat` | Fastest for small dense matrices |

```{r}
almohySol <- rxSolve(mmOde, ev, method = "indLin", hmax = 0.01,
                     indLinMatExpType = "Al-Mohy", indLinMatExpOrder = 8L)

armaSol <- rxSolve(mmOde, ev, method = "indLin", hmax = 0.01,
                   indLinMatExpType = "arma")
```

For the expokit method, two additional controls are available:

- `indLinPhiTol` (default `1e-7`): tolerance for the exponential-matrix action
  computation.
- `indLinPhiM` (default `0L`, meaning automatic): maximum Krylov basis size.

## Multi-state terms

When an ODE term is the product of two or more states (`y * z`), no way of
factoring it produces a legal rate constant: divide out `y` and the coefficient
still contains `z`, and vice versa. Such a term therefore goes to the forcing
whole. The van der Pol oscillator, whose `mu * (1 - y^2) * dy` term mixes both
states, shows what that looks like.

```{r}
vanPol <- function() {
  ini({
    mu <- 1
  })
  model({
    y(0) <- 2
    dy(0) <- 0
    d/dt(y) <- dy
    d/dt(dy) <- mu * (1 - y^2) * dy - y
  })
}

cat(indLin(vanPol))
```

```
matExp()
cmt(y)
cmt(dy)
k_y_dy = -1
k_y_output = 1
k_dy_y = 1
k_dy_output = -(1 + mu)
indLin(dy) <- -mu*dy*Rx_pow_di(y, 2)
y(0)=2
dy(0)=0
```

Expanding `mu * (1 - y^2) * dy` gives `mu*dy - mu*y^2*dy`. The first part is
linear in `dy` and lands in the rate matrix; only the second, which mixes `y`
and `dy`, becomes a forcing. The rate matrix stays constant in the states, which
is what makes the matrix exponential valid, and the solver iterates the forcing.

`rxIndLinStrategy()` and `rxIndLinState()` used to choose which state a
multi-state product was factored onto. They no longer affect the conversion --
no choice of state yields a state-free coefficient -- and are kept only so
existing code keeps running.

## Writing the rate matrix directly with `matExp()`

Everything above went through `symengine` to factor the ODEs symbolically. For
complex expressions that step can fail or produce an unexpected factorization.
`matExp()` skips it entirely: you write the rate constants, and rxode2 builds the
rate matrix from them at compile time. This is:

- **The most stable path** -- no CAS factorization, no symengine dependency, no
  expression simplification.
- **Familiar to NONMEM users** -- it mirrors ADVAN5/7 `K(i,j)` notation with the
  same source-first index convention.
- **Self-documenting** -- every rate constant name says what it connects.

### `matExp()` model syntax

| Statement | Meaning |
|---|---|
| `matExp()` | Declare this model uses matrix exponential integration. **Required.** Cannot be combined with `d/dt()`. |
| `k_from_to <- expr` | First-order rate from compartment `from` to compartment `to`. |
| `k.from.to <- expr` | Dot-separator alternative (identical meaning). |
| `k_cmt_output <- expr` | Elimination from `cmt` to outside the system (like NONMEM's `K(i,0)`). |
| `k_from_to_nd <- expr` | Non-depleting transfer: only `to` gains; `from` is unchanged. |
| `indLin(state) <- expr` | Add an inhomogeneous forcing term to `state`. |
| `cmt(name)` | Optionally declare a compartment explicitly. |
| `other <- expr` | Standard derived-variable assignment (output calculation). |

Compartments are **auto-detected** from the `k_from_to` names, and the diagonal
entries of the rate matrix are computed automatically from the outflow terms.

### Declaring compartment order with `cmt()`

Without `cmt()` declarations, compartment order in the rate matrix -- and
therefore the CMT numbers used in the event table -- follows the order in which
names first appear in the `k_from_to` assignments. Use `cmt()` before the rate
constants to pin the numbering, for example to match a NONMEM control stream
where the depot is CMT=1:

```{r}
mmComp <- function() {
  ini({
    ka <- 1
    km <- 0.5
    vmax <- 0.2
    v <- 1
  })
  model({
    matExp()
    cmt(depot)   ## CMT = 1
    cmt(central) ## CMT = 2
    cp <- central / v
    k_depot_central <- ka ## absorption: depot -> central
    ## MM elimination is state dependent, so it is a forcing, not a rate
    ## constant -- writing it as k_central_output would be an error.
    indLin(central) <- -vmax * cp / (km + cp)
  })
}
```

Compartments that appear in rate constants but have no `cmt()` declaration are
appended after the declared ones in first-appearance order, so you can fix the
positions that matter and let the rest be auto-ordered. `cmt()` declarations are
purely for ordering; they do not change the model mathematics.

### Rules and restrictions

```{r}
## ERROR: matExp() cannot be combined with d/dt()
rxode2({
  matExp()
  d/dt(depot) <- -ka * depot
})
#> :ERR: Matrix exponential models cannot be used with any ODEs

## ERROR: self-transfer is not allowed
rxode2({
  matExp()
  k_depot_depot <- 0.1
})
#> :ERR: transfer from a compartment to itself (e.g., k_cmt1_cmt1) is not allowed

## ERROR: indLin() without matExp()
rxode2({
  cmt(depot)
  indLin(depot) <- 0.5
})
#> :ERR: indLin() cannot be used without matExp() defined in the model

## ERROR: a rate constant that reads a compartment
rxode2({
  matExp()
  cmt(central)
  k_central_output <- vmax / (km + central)
})
#> :ERR: matrix exponential rate constant 'k_central_output' depends on the
#>       compartment 'central'; rate constants must be constant in the states --
#>       put the state-dependent part in 'indLin(central) <- ...' instead
```

### State-dependent rate constants

Rate expressions can reference derived variables that depend on the current
state, as in the Michaelis-Menten example above. The rate matrix is then no
longer constant, and the substep argument from the accuracy section applies --
set `hmax`:

```{r}
library(ggplot2)

ev <- et(amt = 3) |> et(seq(0, 30, by = 0.25))

mmSol <- rxSolve(mmComp, ev, method = "indLin", hmax = 0.01)

ggplot(as.data.frame(mmSol), aes(time, cp)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Michaelis-Menten PK solved with matExp()")
```

rxode2 infers `depot` and `central` as compartments from the rate constant
names. The `output` label is reserved for the elimination sink, analogous to
compartment 0 in NONMEM.

### Two-compartment PK with a peripheral compartment

```{r}
twoCmt <- function() {
  ini({
    ka <- 1
    cl <- 2
    vc <- 10
    q <- 1
    vp <- 20
  })
  model({
    matExp()
    k_depot_central <- ka   ## absorption
    k_central_output <- cl / vc ## central elimination
    k_central_periph <- q / vc  ## distribution to peripheral
    k_periph_central <- q / vp  ## redistribution from peripheral
    cp <- central / vc
  })
}

twoCmtSol <- rxSolve(twoCmt, et(amt = 100) |> et(seq(0, 48, by = 0.25)),
                     method = "indLin")

ggplot(as.data.frame(twoCmtSol), aes(time, cp)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)")
```

### Adding a forcing term with `indLin()`

`indLin(state) <- expr` supplies an inhomogeneous term: one that enters a
compartment without being proportional to any state. This covers endogenous
production, constant inputs that are not in the dosing event table, and
steady-state baselines. A turnover response model is the smallest useful example
-- production `kin` is a forcing term, loss `kout * resp` is a rate constant:

```{r}
turnover <- function() {
  ini({
    kin <- 5
    kout <- 0.1
  })
  model({
    matExp()
    cmt(resp)
    k_resp_output <- kout   ## first-order loss
    indLin(resp) <- kin     ## zero-order production
  })
}

turnoverSol <- rxSolve(turnover, et(seq(0, 100, by = 1)), method = "indLin")

## baseline approaches kin/kout = 50
ggplot(as.data.frame(turnoverSol), aes(time, resp)) +
  geom_line() +
  geom_hline(yintercept = 5 / 0.1, linetype = 2) +
  labs(x = "Time (h)", y = "Response")
```

The forcing is treated as constant across a substep, which makes it exact for a
genuinely constant term like `kin` above.

### Comparison with NONMEM ADVAN5/7

NONMEM's general linear solver (ADVAN5 for non-steady-state, ADVAN7 for
steady-state) specifies rate constants as `K(i,j)` scalars in `$PK`, where
**`K(i,j)` is the first-order transfer rate from compartment `i` to compartment
`j`** (source first, destination second). rxode2's `matExp()` uses the **same
source-first convention**: `k_from_to`.

| NONMEM | rxode2 `matExp()` | Meaning |
|---|---|---|
| `K12 = ka` | `k_cmt1_cmt2 <- ka` | depot (1) to central (2) |
| `K20 = ke` | `k_cmt2_output <- ke` | central (2) elimination |
| `K23 = Q/V2` | `k_cmt2_cmt3 <- q / v2` | central to peripheral |
| `K32 = Q/V3` | `k_cmt3_cmt2 <- q / v3` | peripheral to central |

Differences from NONMEM:

- **Compartment naming**: NONMEM numbers compartments (1, 2, 3, ...); rxode2 uses
  descriptive names (`depot`, `central`, `periph`).
- **Compartment declarations**: NONMEM requires explicit `COMP` declarations;
  rxode2 auto-detects from rate constant names.
- **Elimination sink**: NONMEM uses `K(i,0)` (compartment 0 = outside); rxode2
  uses `k_cmt_output` (`output` is the reserved sink label).
- **Diagonal**: both compute the diagonal (self-decay) entries automatically from
  the listed outflow rates.
- **Non-depleting inputs**: rxode2 supports `k_from_to_nd` for zero-depletion
  transfers; NONMEM has no direct equivalent.
- **Forcing terms**: rxode2's `indLin(state) <- expr` adds an inhomogeneous term;
  in NONMEM this requires explicit ODEs (ADVAN6+).

### When to use `matExp()` vs `d/dt()`

| Situation | Recommendation |
|---|---|
| Linear compartmental PK/PD with known rate constants | `matExp()` -- most stable, no symengine |
| Model migrated from NONMEM ADVAN5/7 | `matExp()` -- direct `K(i,j)` translation |
| Exploratory model with changing structure | `d/dt()` -- faster iteration |
| Nonlinear model | `d/dt()` with a time-stepping solver, unless you have a specific reason not to |

## Use in nlmixr2

`matExp()` and `indLin()` models estimate in nlmixr2 across the focei, nlm and
SAEM families and match the equivalent `d/dt()` model. Be aware of how that
currently works: nlmixr2est materializes the rate constants back into ordinary
ODEs before building each estimation model, so a fit solves the ODE form rather
than propagating a matrix exponential. The syntax is therefore a modeling
convenience there, not a change of integrator.

Full matrix exponential support during estimation is upcoming.

## When to use `method = "indLin"`

| Scenario | Recommendation |
|---|---|
| Purely linear system | `method = "indLin"` gives an exact, closed-form solution |
| Multi-subject population simulation, linear PK | `method = "liblsoda"` (default) is typically fastest overall |
| State-dependent rate matrix | `method = "indLin"` only with an explicit, verified `hmax` |
| Heavily nonlinear or chaotic ODE | Time-stepping solvers (`liblsoda`, `dop853`) |
| Stiff system | `liblsoda` or `lsoda`; the eigenvalue decomposition can struggle when eigenvalues span many orders of magnitude |

### Advantages

1. **Exact for linear systems**: no discretization error at all, at any step
   size.
2. **Interval-based**: the system is propagated across an interval rather than
   stepped through it, so no local error control is needed inside the interval.

### Limitations

1. The eigenvalue decomposition does not exist for all square matrices
   (defective matrices).
2. For large systems the matrix operations become expensive; the decomposition
   scales as O(n^3).
3. Accuracy on a state-dependent model depends entirely on `hmax`, which has no
   automatic error control behind it.
4. The implementation is under active development; results may change between
   versions.

## References

- Moler C, Van Loan C (2003). Nineteen dubious ways to compute the exponential of
  a matrix, twenty-five years later. *SIAM Rev* 45(1):3--49.
  <https://doi.org/10.1137/S00361445024180>

- Al-Mohy AH, Higham NJ (2009). A new scaling and squaring algorithm for the
  matrix exponential. *SIAM J Matrix Anal Appl* 31(3):970--989.

- Sidje RB (1998). Expokit: a software package for computing matrix
  exponentials. *ACM Trans Math Softw* 24(1):130--156.

- Beal S, Sheiner LB, Boeckmann A, Bauer RJ (2009). NONMEM Users Guides
  (1989--2009). Icon Development Solutions, Ellicott City, MD. (ADVAN5/7
  documentation for K(i,j) rate-constant specification.)
