---
title: "Clinical Trial Simulation with rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxode2)
library(ggplot2)
library(data.table)
```

## Overview

Clinical trial simulation (CTS) uses a pharmacometric model to
predict trial outcomes *in silico* before running an expensive study.
Common applications include:

- **Dose regimen selection** — which dose and frequency optimises
  exposure relative to a PK/PD target?
- **Target attainment analysis (TAA)** — what fraction of patients
  achieves the desired exposure target?
- **Sample-size estimation by simulation** — how many subjects are
  needed to detect a treatment difference with adequate power?
- **Special population predictions** — paediatric, renal impairment,
  or other covariate subgroups.

## Model setup

We use a one-compartment oral PK model linked to a simple Emax PD
model to illustrate several CTS tasks.

```{r model}
pkpdModel <- function() {
  ini({
    tka   <- log(1.0);  label("log Ka (1/h)")
    tcl   <- log(4.0);  label("log CL (L/h)")
    tv    <- log(35.0); label("log V (L)")
    temax <- log(80);   label("log Emax (%)")
    tec50 <- log(2.0);  label("log EC50 (mg/L)")
    eta.cl ~ 0.09
    eta.v  ~ 0.09
    prop.sd <- 0.15
  })
  model({
    ka   <- exp(tka)
    cl   <- exp(tcl + eta.cl)
    v    <- exp(tv  + eta.v)
    emax <- exp(temax)
    ec50 <- exp(tec50)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp   <- center / v
    ## Emax PD model
    effect <- emax * cp / (ec50 + cp)
    cp ~ prop(prop.sd)
  })
}
```

## Dose regimen comparison

Compare three once-daily oral doses (50, 100, 200 mg) in a
1000-subject population.

```{r dosingScenarios, message=FALSE}
nSub  <- 1000
doses <- c(50, 100, 200)

# Build a list of event tables — one per dose
evList <- lapply(doses, function(d) {
  et(amt = d, ii = 24, addl = 6) |>   # 7-day QD
    et(seq(0, 168, by = 1))            # hourly observations
})

# Simulate all doses in one call using a named list
simList <- lapply(seq_along(doses), function(i) {
  res <- rxSolve(pkpdModel, evList[[i]], nSub = nSub,
                 returnType = "data.frame")
  res$dose <- doses[i]
  res
})

simAll <- rbindlist(simList)
```

### Concentration-time profiles by dose

```{r ctProfiles, fig.width=7, fig.height=4}
dt <- as.data.table(simAll)

medCI <- dt[, .(
  p05 = quantile(cp, 0.05),
  p50 = quantile(cp, 0.50),
  p95 = quantile(cp, 0.95)
), by = .(time, dose)]

ggplot(medCI, aes(x = time, group = factor(dose), colour = factor(dose),
                  fill = factor(dose))) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line(aes(y = p50), linewidth = 1) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       colour = "Dose (mg)", fill = "Dose (mg)",
       title = "Dose comparison — steady-state concentration profiles") +
  theme_bw()
```

## Target attainment analysis (TAA)

Define a PK target as trough concentration ≥ 1 mg/L at steady state
(day 7, trough = 168 h).

```{r taa, fig.width=6, fig.height=4}
trough <- dt[time == 168, .(
  pctAbove = mean(cp >= 1.0) * 100
), by = dose]

ggplot(trough, aes(x = factor(dose), y = pctAbove)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "red") +
  labs(x = "Dose (mg)", y = "% subjects with trough ≥ 1 mg/L",
       title = "Target attainment at steady-state trough",
       subtitle = "Dashed line: 90% target attainment criterion") +
  theme_bw()
```

## PK/PD target: % of population with effect ≥ 50%

Using the Emax model, what fraction achieves ≥ 50% of maximum effect
at the trough?

```{r pdTarget, fig.width=6, fig.height=4}
pdTarget <- dt[time == 168, .(
  pctTarget = mean(effect >= 50) * 100
), by = dose]

ggplot(pdTarget, aes(x = factor(dose), y = pctTarget)) +
  geom_col(fill = "coral", width = 0.5) +
  geom_hline(yintercept = 80, linetype = "dashed", colour = "red") +
  labs(x = "Dose (mg)", y = "% subjects with effect ≥ 50%",
       title = "PD target attainment at steady-state trough",
       subtitle = "Dashed line: 80% target attainment criterion") +
  theme_bw()
```

## Exposure metric distributions

Compute AUC over one dosing interval at steady state (day 7,
hours 144–168) for each dose.

```{r aucDist, fig.width=7, fig.height=4}
ss <- dt[time >= 144 & time <= 168]

aucSS <- ss[, .(
  AUCtau = sum(diff(time) * (cp[-length(cp)] + cp[-1]) / 2)
), by = .(id, dose)]

ggplot(aucSS, aes(x = AUCtau, fill = factor(dose))) +
  geom_density(alpha = 0.4) +
  labs(x = "AUC_tau (mg·h/L)", y = "Density",
       fill = "Dose (mg)",
       title = "Steady-state AUC_tau distribution by dose") +
  theme_bw()
```

## Sample-size estimation by simulation

Estimate the power to detect a difference in mean trough concentration
between 100 mg and 50 mg using a two-sample t-test, as a function of
sample size.

```{r powerSim, message=FALSE}
set.seed(99)
nSizes  <- c(20, 40, 60, 80, 100)
nTrials <- 200   # increase to >= 1000 for stable power estimates

powerRes <- rbindlist(lapply(nSizes, function(n) {
  pvals <- vapply(seq_len(nTrials), function(trial) {
    # Simulate n subjects per arm
    sim50  <- rxSolve(pkpdModel,
                      et(amt = 50,  ii = 24, addl = 6) |> et(168),
                      nSub = n, returnType = "data.frame")
    sim100 <- rxSolve(pkpdModel,
                      et(amt = 100, ii = 24, addl = 6) |> et(168),
                      nSub = n, returnType = "data.frame")
    t.test(sim100$cp, sim50$cp)$p.value
  }, numeric(1))
  data.table(n = n, power = mean(pvals < 0.05))
}))
```

```{r powerPlot, fig.width=6, fig.height=4}
ggplot(powerRes, aes(x = n, y = power)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "red") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Subjects per arm", y = "Power",
       title = "Power by simulation — 100 mg vs 50 mg trough comparison",
       subtitle = "Dashed line: 80% power target") +
  theme_bw()
```

## Special population simulation

Simulate the 100 mg dose in a paediatric population with lower body
weight driving CL allometrically:

```{r paeds, message=FALSE, fig.width=7, fig.height=4}
# Covariate table for a paediatric cohort (n=200, WT 10-40 kg)
set.seed(7)
pedCovs <- data.frame(
  id = 1:200,
  WT = runif(200, 10, 40)
)

# Model with allometric CL scaling
pedModel <- pkpdModel |>
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75
    v  <- exp(tv  + eta.v)  * (WT / 70)
  }, append = NA)

pedSim <- rxSolve(
  pedModel,
  et(amt = 100, ii = 24, addl = 6) |> et(seq(0, 168, by = 2)),
  iCov = pedCovs,
  nSub = nrow(pedCovs),
  returnType = "data.frame"
)

pedDT <- as.data.table(pedSim)
pedCI <- pedDT[, .(p05 = quantile(cp, 0.05),
                   p50 = quantile(cp, 0.50),
                   p95 = quantile(cp, 0.95)), by = time]

ggplot(pedCI, aes(x = time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "coral", alpha = 0.3) +
  geom_line(aes(y = p50), colour = "coral", linewidth = 1) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Paediatric simulation — 100 mg QD with allometric scaling",
       subtitle = "WT range 10–40 kg") +
  theme_bw()
```

## Tips

- Use `rxSetSeed()` / `rxWithSeed()` for reproducible simulations.
- Increase `nSub` and `nTrials` (for power simulations) for production
  analyses; the values used here are kept small for build speed.
- Combine parameter uncertainty (see the
  [parameter uncertainty vignette](rxode2-parameter-uncertainty.html))
  with BSV for the most complete uncertainty propagation.
- The `nlmixr2lib` package provides published models ready for use in
  clinical trial simulations.
