---
title: "Matrix Exponentials and Inductive Linearization in rxode2"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Background

Nonlinear ordinary differential equations (ODEs) arise naturally in
pharmacokinetic--pharmacodynamic (PKPD) models.  Consider a drug that
undergoes Michaelis--Menten (saturable) elimination: the clearance
depends on the current concentration, making the ODE nonlinear.
Standard time-stepping solvers (LSODA, Runge--Kutta) handle this by
advancing one small step at a time, using only local information about
the solution.

**Inductive Linearization** (IndLin), introduced by Duffull and Hegarty
(2014) and further explored by Sharif, Hasegawa and Duffull (2022),
offers an alternative.  Rather than stepping forward in time, it
iteratively converts the nonlinear ODE into a linear time-varying (LTV)
system.  The LTV system is of the form

```
dy/dt = f(t) + A(t, y^[n-1]) * y^[n]
```

where the matrix `A` is formed using the solution from the previous
iteration `y^[n-1]`.  Starting from `y^[0] = 0`, each iteration
produces a more accurate approximation of the true solution.  Crucially,
because the system is now linear, it can be solved exactly over the
whole time span using the **matrix exponential** (ME), rather than
approximated with finite steps.

rxode2 pairs IndLin with the matrix-exponential integrator using
eigenvalue decomposition (EVD):

```
y^[n](t) = V * diag(exp(lambda * t)) * V^{-1} * y0
```

where `V` are the eigenvectors and `lambda` the eigenvalues of the rate
matrix `K`.

### Key properties

- The solution is available at **all** times in `[0, T]`
  simultaneously, not just at local steps.  Perturbations such as dose
  events at any time are handled globally.
- For purely **linear** ODE systems the iteration converges in one
  step, giving an exact solution via ME alone.
- For **nonlinear** systems the method is approximate but converges
  with a user-controlled tolerance.

The reference comparing IndLin to MATLAB's `ode45` (Sharif et al.
2022) showed that IndLin with an adaptive step size and a convergence
stopping rule was six times faster than `ode45` for a
Michaelis--Menten PK example, while recovering parameter estimates of
comparable accuracy.

---

## Quick start

### Step 1 -- build the model with `indLin = TRUE`

Activate inductive linearization by passing `indLin = TRUE` to
`rxode2()`:

```{r}
library(rxode2)

mmPK <- rxode2(
  {
    d/dt(depot)  <- -ka * depot
    Cp           <- center / V
    d/dt(center) <- ka * depot - Vmax * Cp / (Km + Cp)
  },
  indLin = TRUE
)
```

This step uses the `symengine` computer-algebra system to factor each
ODE right-hand side into a rate matrix `A` (the part linear in the
state variables) and an inhomogeneous forcing vector `f` (the part that
depends only on time and parameters).  A summary of the factorization
is printed when you inspect the model:

```{r}
summary(mmPK)
```

### Step 2 -- solve with `method = "indLin"`

```{r}
library(rxode2)

params <- c(ka = 1, Km = 0.5, Vmax = 0.2, V = 1)

ev <- eventTable()
ev$add.dosing(dose = 3, start.time = 0)
ev$add.sampling(c(0.1, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12, 16, 24, 30))

sol_indlin <- rxSolve(mmPK, params, ev, method = "indLin")
sol_lsoda  <- rxSolve(mmPK, params, ev, method = "liblsoda")

plot(sol_indlin)
```

For a purely linear model the two solutions are numerically identical;
for a Michaelis--Menten model they are close, with differences
controlled by the convergence tolerance.

### Comparing against LSODA

```{r}
library(ggplot2)

df <- merge(
  as.data.frame(sol_lsoda)[, c("time", "Cp")],
  as.data.frame(sol_indlin)[, c("time", "Cp")],
  by = "time", suffixes = c(".lsoda", ".indlin")
)

ggplot(df, aes(time)) +
  geom_line(aes(y = Cp.lsoda,  colour = "LSODA")) +
  geom_line(aes(y = Cp.indlin, colour = "IndLin"), linetype = 2) +
  labs(x = "Time (h)", y = "Concentration (mg/L)", colour = "Solver")
```

---

## Purely linear systems: matrix exponential alone

When every ODE is linear in the state variables, the inductive
linearization loop converges immediately and the solution is exact.
This includes all standard compartmental PK models.

```{r}
simplePK <- rxode2(
  {
    d/dt(depot)  <- -ka * depot
    d/dt(center) <- ka * depot - (CL / V) * center
    Cp           <- center / V
  },
  indLin = TRUE
)

params2 <- c(ka = 1, CL = 0.1, V = 1)

ev2 <- eventTable()
ev2$add.dosing(dose = 100, start.time = 0, nbr.doses = 5, dosing.interval = 12)
ev2$add.sampling(seq(0, 72, by = 0.5))

pk_me    <- rxSolve(simplePK, params2, ev2, method = "indLin")
pk_lsoda <- rxSolve(simplePK, params2, ev2, method = "liblsoda")

## Differences should be at machine precision
max(abs(pk_me$Cp - pk_lsoda$Cp))
```

---

## Matrix exponential algorithms

Three ME algorithms are available via `indLinMatExpType`:

| Value | Algorithm | Notes |
|---|---|---|
| `"Al-Mohy"` | Al-Mohy & Higham (2009) scaling and squaring | Accurate; order controlled by `indLinMatExpOrder` |
| `"expokit"` | Sidje (1998) Krylov/Pade | Good for sparse/large matrices; controlled by `indLinPhiTol` and `indLinPhiM` |
| `"arma"` | RcppArmadillo `expmat` | Fastest for small dense matrices |

The default is `"expokit"`.

```{r}
## Al-Mohy / Higham algorithm, Pade order 8
sol_almohy <- rxSolve(mmPK, params, ev,
                      method          = "indLin",
                      indLinMatExpType  = "Al-Mohy",
                      indLinMatExpOrder = 8L)

## RcppArmadillo expmat
sol_arma <- rxSolve(mmPK, params, ev,
                    method         = "indLin",
                    indLinMatExpType = "arma")
```

For the expokit method, two additional controls are available:

- `indLinPhiTol` (default `1e-7`): tolerance for the exponential-matrix
  action computation.
- `indLinPhiM` (default `0L`, meaning automatic): maximum Krylov basis
  size.

---

## Controlling the linearization strategy

### `rxIndLinStrategy()`

When an ODE term contains the product of two or more state variables
(e.g., `y * z`), the solver must decide which state to use as the
linear factor.  `rxIndLinStrategy()` selects the algorithm:

- `"curState"` (default): prefer the state whose derivative is being
  computed (`d/dt(y) <- ... y*z ...` factors by `y`).  When the
  current state is not present, fall back to the first state found.
- `"split"`: divide the term equally among all states involved.

```{r}
## Use split strategy for the Van der Pol equation
rxIndLinStrategy("split")

vanPol <- rxode2(
  {
    y(0)  <- 2
    dy(0) <- 0
    d/dt(y)  <- dy
    d/dt(dy) <- mu * (1 - y^2) * dy - y
  },
  indLin = TRUE
)

rxIndLinStrategy("curState") # restore default
```

### `rxIndLinState()`

For fine-grained control you can specify, per state, which state
variable should appear as the linear factor for each ODE.

```{r}
## For the Van der Pol equation: factor d/dt(y) by dy, and d/dt(dy) by y
rxIndLinState(list(y = "dy", dy = "y"))

vanPol2 <- rxode2(
  {
    y(0)  <- 2
    dy(0) <- 0
    d/dt(y)  <- dy
    d/dt(dy) <- mu * (1 - y^2) * dy - y
  },
  indLin = TRUE
)

rxIndLinState(NULL)  # clear preferences
```

---

## A full example: Michaelis--Menten PK

This example reproduces the scenario from Sharif et al. (2022): oral
absorption with Michaelis--Menten elimination.

```{r}
library(rxode2)
library(ggplot2)

mmPK <- rxode2(
  {
    d/dt(depot)  <- -ka * depot
    Cp           <- center / V
    d/dt(center) <- ka * depot - Vmax * Cp / (Km + Cp)
  },
  indLin = TRUE
)

params <- c(ka = 1, Km = 0.5, Vmax = 0.2, V = 1)

ev <- eventTable()
ev$add.dosing(dose = 3, start.time = 0)
ev$add.sampling(c(0.1, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12, 16, 24, 30))

## Three solvers
sol_ll  <- rxSolve(mmPK, params, ev, method = "liblsoda")
sol_il  <- rxSolve(mmPK, params, ev, method = "indLin")
sol_ila <- rxSolve(mmPK, params, ev,
                   method = "indLin", indLinMatExpType = "Al-Mohy")

df <- rbind(
  cbind(as.data.frame(sol_ll),  solver = "liblsoda"),
  cbind(as.data.frame(sol_il),  solver = "IndLin/expokit"),
  cbind(as.data.frame(sol_ila), solver = "IndLin/Al-Mohy")
)

ggplot(df, aes(time, Cp, colour = solver, linetype = solver)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Michaelis-Menten PK: solver comparison")
```

---

## Direct coefficient specification with `matExp()`: the most stable approach

The auto-factorization path (`indLin = TRUE`) uses the `symengine`
computer-algebra system to symbolically factor each ODE into a rate
matrix and a forcing vector.  For complex expressions this symbolic
step can fail or produce unexpected factorizations.

The `matExp()` model declaration bypasses symbolic manipulation
entirely.  You write rate constants in the form `k_from_to` (or
`k.from.to`), and rxode2 constructs the rate matrix directly from
those coefficients at compile time.  This is:

- **The most stable approach** -- no CAS factorization, no symengine
  dependency, no expression simplification.
- **Familiar to NONMEM users** -- mirrors NONMEM's ADVAN5/7 `K(i,j)`
  notation with the same source-first index convention.
- **Self-documenting** -- every rate constant name encodes what it
  connects.

### `matExp()` model syntax

A `matExp()` model has three kinds of statements:

| Statement | Meaning |
|---|---|
| `matExp()` | Declare this model uses matrix exponential integration. **Required.** Cannot be combined with `d/dt()`. |
| `k_from_to = expr` | First-order rate from compartment `from` to compartment `to`. |
| `k.from.to = expr` | Dot-separator alternative (identical meaning). |
| `k_cmt_output = expr` | Elimination from `cmt` to outside the system (like NONMEM's `K(i,0)`). |
| `k_from_to_nd = expr` | Non-depleting transfer: only `to` gains; `from` is unchanged. |
| `indLin(state) <- expr` | Add a time-varying forcing term to `state` (inhomogeneous part). |
| `cmt(name)` | Optionally declare a compartment explicitly. Usually not needed. |
| `other = expr` | Standard derived-variable assignment (output calculation). |

Compartments are **auto-detected** from the `k_from_to` names -- no
explicit `cmt()` declarations are required.  The diagonal entries of
the rate matrix (self-decay) are computed automatically from the
outflow terms.

### Optional: declaring compartment order with `cmt()`

By default the compartment ordering in the rate matrix (and therefore
the CMT numbers used in the event table) follows the order in which
compartment names first appear in the `k_from_to` assignments.  If you
need a specific CMT numbering -- for example, to match a NONMEM control
stream where the depot is CMT=1 and central is CMT=2 -- you can pin the
order with explicit `cmt()` declarations placed before the rate
constants:

```{r}
mmOrdered <- rxode2({
  matExp()
  cmt(depot)      ## CMT = 1
  cmt(central)    ## CMT = 2
  Cp = central / V
  k_depot_central  = ka                ## absorption: depot -> central
  k_central_output = Vmax / (Km + Cp) ## MM elimination: state-dependent rate
})
```

Any compartment listed in a `cmt()` call is assigned CMT numbers in
declaration order.  Compartments that appear in rate constants but have
no `cmt()` declaration are appended after the declared ones in their
first-appearance order.  This lets you fix the positions that matter
(e.g., the dosing compartment) while letting the rest be auto-ordered.

`cmt()` declarations are purely for ordering; they do not change the
model mathematics.

### Rules and restrictions

```{r}
library(rxode2)

## ERROR: matExp() cannot be combined with d/dt()
rxode2({
  matExp()
  d/dt(depot) <- -ka * depot   # error
})

## ERROR: self-transfer is not allowed
rxode2({
  matExp()
  k_depot_depot = 0.1          # error
})

## ERROR: indLin() without matExp()
rxode2({
  cmt(depot)
  indLin(depot) <- 0.5         # error
})
```

### Michaelis-Menten elimination

Rate expressions in `matExp()` can reference derived variables that
depend on the current state.  When such a rate is state-dependent the
system is no longer strictly linear, but `method = "indLin"` handles
this via iteration: at each IndLin pass the rate matrix is recomputed
using the current state estimate, and the matrix exponential is then
solved exactly for that linearised system.  Michaelis-Menten
elimination is the canonical example.

```{r}
library(rxode2)

mmComp <- rxode2({
  matExp()
  Cp = central / V                   ## derived variable (used in rate expression)
  k_depot_central  = ka              ## absorption: depot -> central
  k_central_output = Vmax / (Km + Cp) ## MM elimination expressed as effective clearance
})

params <- c(ka = 1, Km = 0.5, Vmax = 0.2, V = 1)

ev <- eventTable()
ev$add.dosing(dose = 3, start.time = 0)
ev$add.sampling(c(0.1, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12, 16, 24, 30))

## method = "indLin" iterates to account for the state-dependent rate
sol <- rxSolve(mmComp, params, ev, method = "indLin")
plot(sol, Cp)
```

rxode2 infers `depot` and `central` as state compartments from the rate
constant names.  The `output` label is reserved for the elimination
sink, analogous to compartment 0 in NONMEM.

### Three-compartment PK with peripheral

```{r}
threeComp <- rxode2({
  matExp()
  k_depot_central   = ka            ## absorption
  k_central_output  = CL / Vc       ## central elimination
  k_central_periph  = Q  / Vc       ## distribution to peripheral
  k_periph_central  = Q  / Vp       ## redistribution from peripheral
  Cp = central / Vc
})

params3 <- c(ka = 1, CL = 2, Vc = 10, Q = 1, Vp = 20)
sol3 <- rxSolve(threeComp, params3, ev, method = "indLin")
```

### Adding a forcing term with `indLin()`

The `indLin(state)` statement adds a time-varying (inhomogeneous) term
to a compartment.  This is useful for constant infusions that are not
handled by the dosing event table, or for steady-state baseline inputs:

```{r}
glucoseModel <- rxode2({
  matExp()
  k_Gc_output = Clg / Vg       ## glucose clearance
  k_Gp_Gc    = Q   / Vp        ## peripheral to central
  k_Gc_Gp    = Q   / Vg        ## central to peripheral
  ## Constant endogenous glucose production enters Gc
  indLin(Gc) <- Gprod
})
```

### Comparison with NONMEM ADVAN5/7

NONMEM's general linear solver (ADVAN5 for non-steady-state, ADVAN7
for steady-state) specifies rate constants as `K(i,j)` scalars in
`$PK`, where **`K(i,j)` is the first-order transfer rate from
compartment `i` to compartment `j`** (source first, destination
second).

rxode2's `matExp()` uses the **same source-first convention**:
`k_from_to`.

| NONMEM | rxode2 `matExp()` | Meaning |
|---|---|---|
| `K12 = ka` | `k_cmt1_cmt2 = ka` | depot (1) to central (2) |
| `K20 = ke` | `k_cmt2_output = ke` | central (2) elimination |
| `K23 = Q/V2` | `k_cmt2_cmt3 = Q/V2` | central to peripheral |
| `K32 = Q/V3` | `k_cmt3_cmt2 = Q/V3` | peripheral to central |

Differences from NONMEM:

- **Compartment naming**: NONMEM numbers compartments (1, 2, 3, ...);
  rxode2 uses descriptive names (`depot`, `central`, `periph`).
- **Compartment declarations**: NONMEM requires explicit `COMP`
  declarations; rxode2 auto-detects from rate constant names.
- **Elimination sink**: NONMEM uses `K(i,0)` (compartment 0 = outside);
  rxode2 uses `k_cmt_output` (`output` is the reserved sink label).
- **Diagonal**: Both tools compute the diagonal (self-decay) entries
  automatically from the listed outflow rates.
- **Non-depleting inputs**: rxode2 supports `k_from_to_nd` for
  zero-depletion transfers; NONMEM has no direct equivalent.
- **Forcing terms**: rxode2's `indLin(state) <- expr` adds an
  inhomogeneous term; in NONMEM this requires explicit ODE (ADVAN6+).

### Why `matExp()` is the most stable path

When using `indLin = TRUE` (ODE auto-factorization), rxode2 must:

1. Parse the ODE right-hand sides symbolically via `symengine`
2. Factor each expression into linear-in-state terms vs. remaining
   (forcing) terms
3. Handle multi-state products using heuristics
   (`rxIndLinStrategy`, `rxIndLinState`)

Any of these steps can fail or produce a suboptimal factorization for
complex models.  With `matExp()`, none of this happens: you specify
the rate matrix entries directly, and rxode2 generates the matrix
exponential code without any symbolic processing.

### When to use `matExp()` vs `indLin = TRUE`

| Situation | Recommendation |
|---|---|
| Linear compartmental PK/PD (known rate constants) | `matExp()` -- most stable, no symengine |
| Model migrated from NONMEM ADVAN5/7 | `matExp()` -- direct `K(i,j)` translation |
| Nonlinear ODE (Michaelis--Menten, etc.) | `indLin = TRUE` -- auto-factorization required |
| Exploratory model with changing ODE structure | `indLin = TRUE` -- faster iteration |
| Complex nonlinear ODE where auto-factorization fails | Consider `matExp()` if you can state the rate matrix analytically |

---

## When to use inductive linearization

| Scenario | Recommendation |
|---|---|
| Multi-subject population simulation, linear PK | `method = "liblsoda"` (default) is typically fastest overall |
| Purely linear system, single subject | `method = "indLin"` gives an exact, closed-form solution |
| Mildly nonlinear system (Michaelis--Menten saturation) | `method = "indLin"` can be faster than LSODA when parameters are evaluated many times (e.g., in estimation) |
| Heavily nonlinear or chaotic ODE | Time-stepping solvers (`liblsoda`, `dop853`) are safer; IndLin convergence is not guaranteed |
| Stiff system | `liblsoda` or `lsoda` with Jacobian specification; IndLin with EVD can struggle when eigenvalues span many orders of magnitude |

### Advantages over time-stepping solvers

1. **All-time domain access**: because the linearization is performed
   over the entire interval, information from all time points is
   available.  Dose events at any time are incorporated without the
   solver needing an explicit hint about discontinuities.
2. **Exact solution for linear systems**: no discretization error.
3. **Estimation warm-start**: during iterative parameter estimation, the
   previous parameter vector's solution can seed the next IndLin
   iteration (the "smart update" described in Sharif et al. 2022),
   reducing the number of linearization iterations needed.

### Limitations

1. EVD does not exist for all square matrices (defective matrices).
2. For large systems (many ODEs) the matrix operations become
   expensive; EVD scales as O(n^3).
3. Convergence for highly nonlinear systems may be slow or fail.
4. The current implementation is under active development; results may
   change between versions.

---

## References

- Sharif, Hasegawa, Duffull (2022). Exploring *Inductive Linearization*
  for simulation and estimation with an application to the
  Michaelis--Menten model. *Journal of Pharmacokinetics and
  Pharmacodynamics* 49:445--453. <https://doi.org/10.1007/s10928-022-09813-z>

- Duffull SB, Hegarty G (2014). An inductive approximation to the
  solution of systems of nonlinear ordinary differential equations in
  pharmacokinetics-pharmacodynamics. *J Theor Comput Sci* 1(4):1000119.

- Hasegawa C, Duffull SB (2018). Exploring inductive linearization for
  pharmacokinetic-pharmacodynamic systems of nonlinear ordinary
  differential equations. *J Pharmacokinet Pharmacodyn* 20(1):35--47.

- Al-Mohy AH, Higham NJ (2009). A new scaling and squaring algorithm
  for the matrix exponential. *SIAM J Matrix Anal Appl* 31(3):970--989.

- Sidje RB (1998). Expokit: a software package for computing matrix
  exponentials. *ACM Trans Math Softw* 24(1):130--156.

- Beal S, Sheiner LB, Boeckmann A, Bauer RJ (2009). NONMEM Users
  Guides (1989--2009). Icon Development Solutions, Ellicott City, MD.
  (ADVAN5/7 documentation for K(i,j) rate-constant specification.)
