---
title: "rxode2 advanced features"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Explore advanced rxode2 features like solved compartment models (linCmt),
  transit compartments, and mixing solved systems with ODEs.
---

```{r setup, include=FALSE}
library(learnr)
library(rxode2)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

## Review: Putting it all Together

We've come a long way! From basic syntax to complex population simulations with covariates. Let's do one final review exercise that combines several concepts.

### Exercise: A Complete Workflow

1.  Define a one-compartment model function `final_mod` with:
    *   `tka = 0.5`, `tcl = 1`, `tv = 10`.
    *   `eta.cl ~ 0.1`.
    *   Additive error `add.sd = 0.1`.
    *   Include a weight effect on `cl`: `cl <- exp(tcl + eta.cl) * (WT/70)^0.75`.
2.  Create an event table for 100 mg dose at time 0, sampling 0-24h.
3.  Simulate 10 subjects with weights ranging from 50 to 100 kg.

```{r final_review, exercise=TRUE}
# 1. Define final_mod
final_mod <- function() {
  ini({
    tka <- 0.5; tcl <- 1; tv <- 10
    eta.cl ~ 0.1
    add.sd <- 0.1
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75
    v  <- exp(tv)
    d/dt(center) <- - cl/v * center
    cp <- center/v
    cp ~ add(add.sd)
  })
}

# 2. Create event table
ev <- et(amt = 100) |> et(0:24)

# 3. Define iCov and solve
my_covs <- data.frame(id = 1:10, WT = seq(50, 100, length.out = 10))
res <- rxSolve(final_mod, ev, iCov = my_covs)

plot(res, cp)
```

## Introduction

While `rxode2` is primarily an ODE solver, it also supports:
- **Prediction only models**: Models without differential equations.
- **Solved compartment models**: 1, 2, or 3 compartment models that use analytical solutions instead of numerical integration.
- **Mixed models**: Combining solved systems with ODEs.

## Solved Compartment Models (`linCmt()`)

Instead of writing out the differential equations for standard PK models, you can use the `linCmt()` function. `rxode2` will automatically determine the model (1, 2, or 3 compartments) based on the parameters you provide.

### Parameterization

`linCmt()` recognizes many common parameterizations:
- **Clearance/Volume**: `CL`, `V`, `Q`, `V2`, etc.
- **Micro-constants**: `ke`, `k12`, `k21`, etc.

```r
mod <- function() {
  ini({
    cl <- 1; v <- 10
  })
  model({
    cp <- linCmt()
  })
}
```

### Exercise: Use `linCmt()`

Define a 2-compartment model using `linCmt()` with parameters `cl`, `v`, `q`, and `v2`.

```{r lincmt_ex, exercise=TRUE}
mod <- function() {
  ini({
    tcl <- log(1); tv <- log(10)
    tq <- log(0.5); tv2 <- log(20)
  })
  model({
    cl <- exp(tcl); v <- exp(tv)
    q <- exp(tq); v2 <- exp(tv2)
    
    # Use linCmt() here
    cp <- linCmt()
  })
}

ev <- et(amt = 100) |> et(0:24)
res <- rxSolve(mod, ev)
plot(res, cp)
```

## Mixing Solved Systems and ODEs

You can use `linCmt()` to handle the PK part of a model and use ODEs for the PD part.

```r
mod_mixed <- function() {
  ini({
    cl <- 1; v <- 10; kin <- 1; kout <- 1; ec50 <- 10
  })
  model({
    cp <- linCmt(cl, v)
    d/dt(effect) <- kin - kout * (1 - cp/(ec50 + cp)) * effect
  })
}
```

**Note**: When mixing, the `linCmt()` compartments are usually numbered after the ODE compartments.

## Transit Compartments

`rxode2` has built-in support for transit compartment models, which are often used to describe delayed absorption.

### Exercise: Transit Compartment

Use the `transit()` function to define a model with a transit delay.

```{r transit_ex, exercise=TRUE}
mod_transit <- function() {
  ini({
    tka <- log(0.5); tcl <- log(1); tv <- log(10)
    ktr <- log(1); n <- log(3)
  })
  model({
    ka <- exp(tka); cl <- exp(tcl); v <- exp(tv)
    ktr <- exp(ktr); n <- exp(n)
    
    # transit(n, ktr, ka) returns the amount entering the depot
    d/dt(depot) <- transit(n, ktr, ka) - ka * depot
    d/dt(center) <- ka * depot - cl/v * center
    cp <- center/v
  })
}

ev <- et(amt = 100) |> et(0:24)
res <- rxSolve(mod_transit, ev)
plot(res, cp)
```

## Summary

- `linCmt()` provides fast, analytical solutions for standard PK models.
- You can mix analytical PK with numerical PD ODEs.
- `transit()` simplifies the implementation of transit compartment models.

## Congratulations!

You have completed the `rxode2` tutorial series! You now have the skills to:
- Write and modify `rxode2` models.
- Create complex dosing schedules and sampling plans with event tables.
- Perform individual and population simulations.
- Incorporate covariates and handle parameter uncertainty.
- Use advanced features like `linCmt()` and transit compartments.

For more information and advanced topics, visit the [rxode2 documentation](https://nlmixr2.github.io/rxode2/).
